### 2020-02-18 data-raw/ep1162_prelimDataScreening.Rmd
Data cleaning

---
title: "EP1162_Prelim_Data_Screening"
author: "Hardy Griesbauer"
date: "18/02/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
library(tidyverse)

# Load data
load("../data/ep1162_Data.RData")

```

## Preliminary screening
I did some preliminary screening, to confirm data entry and detect any potential outliers.  Some findings below.

## Outliers
Quick boxplot to confirm DBH and height measurements

#### DBH

```{r echo=FALSE}

dat %>% 
  ggplot()+
  aes(group=Species,y=DBH,fill=Species) +
  geom_boxplot()

```

DBH measurements look reasonable, and no major outliers.  Note that there is a tree with species recorded as "l".

#### Height

```{r echo=FALSE}

dat %>% 
  ggplot()+
  aes(group=Species,y=Height,fill=Species) +
  geom_boxplot()

```
Again, measurements look reasonable here.

### Data entry

One tree has species recorded as "l".  I assume that it's SAF, but might be worthwhile to verify with the field cards?

```{r echo=FALSE}

dat %>% 
  filter(Species=="l") %>% 
  knitr::kable()

```

There are also two trees recorded with id=631:

```{r echo=FALSE}
# Look at entries where tree number is duplicated within plot
dat %>% 
  add_count(Plot,TreeID,Year) %>% # get a number by this grouping
  filter(n>1) %>% 
  drop_na(TreeID) %>% 
  filter(TreeID!="New") %>% 
    knitr::kable()

```

Looks like it is the same tree (exact same DBH and Height measured in 2019), so assume it's safe to delete the duplicated entry?

There are also quite a few trees with their ID recorded as "New".  Not an issue from an analysis point of view, however, I seem to recall that we were assigning new tags/IDs for all trees.  But I also remember a discussion about tagging trees once they get to a larger size!

### Decreasing DBH 
I also screened trees that had a smaller DBH measurements than a previous measurement at some point.  Below are some examples:

```{r echo=FALSE}
# Look at entries where recorded DBH is less from one measurement to the other

dbhDecrease<-
  dat %>% 
  filter(TreeID!="New") %>% 
  filter(TreeID!=631) %>% # this is a duplicated tree
  pivot_wider(id_cols=c(Plot,TreeID),names_from="Year",values_from="DBH") %>% 
  mutate(diff.2019=`2019`-`2009`) %>% 
  mutate(diff.2009=`2009`-`1997`) %>% 
  mutate(diff.1997=`1997`-`1994`) %>% 
  mutate(diff.1994=`1994`-`1992`) %>% 
  filter(diff.2019<0 | diff.2009<0 | diff.1997<0 | diff.1994<0) %>% 
  dplyr::select(Plot,TreeID) %>% # select these trees with decreasing DBH measurements
  left_join(dat,by=c("Plot","TreeID")) %>% 
  mutate_at(vars(contains("DBH")),round,1) %>% # round to one decimal place
  dplyr::select(Plot:DBH)

  # Show example rows
  dbhDecrease %>% 
  slice(1:20) %>% 
  knitr::kable()

```

In most cases, the differences are very small, and likely can be atrributed to slight measurement differences between years.  We can likely ignore this error and simply describe it if necessary in the results section?

In total, there are `r dbhDecrease %>% count(TreeID) %>% nrow()` trees with this situation.  I have the treeID saved and can provide them if we want to double check these trees.

*****
## Actions required
- Confirm species for tree with Species recorded as "l" (likely SAF)
- COnfirm that ok to delete duplicate record for Tree #631 (Plot 10)
- Determine what to do about trees where DBH measurements aren't monotonically increasing.

### 2020-02-18 data-raw/dataClean1.R

@@ -104,15 +104,19 @@ boxplot(dat$Height) # all good there, nothing seems out of the ordinary
 DBH_decrease<-
   dat %>% 
   filter(TreeID!="New") %>% 
-  filter(TreeID!=631) %>% 
-  mutate(ID=paste(Plot,TreeID,sep=".")) %>% 
-  # distinct(.,ID,.keep_all = T) %>% # drop duplicated ID
-  pivot_wider(id_cols=ID,names_from="Year",values_from="DBH") %>% 
+  filter(TreeID!=631) %>% # this is a duplicated tree
+  pivot_wider(id_cols=c(Plot,TreeID),names_from="Year",values_from="DBH") %>% 
   mutate(diff.2019=`2019`-`2009`) %>% 
   mutate(diff.2009=`2009`-`1997`) %>% 
   mutate(diff.1997=`1997`-`1994`) %>% 
   mutate(diff.1994=`1994`-`1992`) %>% 
-  filter(diff.2019<0 | diff.2009<0 | diff.1997<0 | diff.1994<0)
-  
+  filter(diff.2019<0 | diff.2009<0 | diff.1997<0 | diff.1994<0) %>% 
+  dplyr::select(Plot,TreeID) %>% # select these trees with decreasing DBH measurements
+  left_join(dat,by=c("Plot","TreeID")) %>%  # left join to filter for trees with decreasing DBH
+  mutate_at(vars(contains("DBH")),round,1)
+
+# Export this table
+write.csv(DBH_decrease,file="data-raw/EP1162_trees_DBHDecreasing.csv",row.names = FALSE)
+
   
 

### 2020-02-28 data-raw/ep1162_prelimDataScreening.Rmd
DBH interpolation

@@
 
+*ALRF staff confirmed that the species is "Bl"*, so we'll change the species code accordingly in the next line of code:
+
+```{r}
+dat<-
+  dat %>% 
+  mutate(Species=replace(Species,Species=="l","Bl"))
+
+```
+
 There are also two trees recorded with id=631:
@@
 
-There are also quite a few trees with their ID recorded as "New".
-Not an issue from an analysis point of view, however, I seem to recall that we were assigning new tags/IDs for all trees.
-But I also remember a discussion about tagging trees once they get to a larger size!
+*Confirmed that this tree was duplicated by mistake.  We will remove one entry.*
+
+```{r}
+dat<-
+  dat %>% 
+  filter(!(Plot==10 & TreeID==631 & Sector=="NA")) # remove entry
+
+```
+
+There are also quite a few trees with their ID recorded as "New".  Not an issue from an analysis point of view,
 
@@
 In most cases, the differences are very small, and likely can be atrributed to slight measurement differences between years.
-We can likely ignore this error and simply describe it if necessary in the results section?
 
@@
 *****
-## Actions required
-- Confirm species for tree with Species recorded as "l" (likely SAF)
-- COnfirm that ok to delete duplicate record for Tree #631 (Plot 10)
-- Determine what to do about trees where DBH measurements aren't monotonically increasing.
+## Trees without monotonically increasing DBH measurements
+After discussion, we will handle trees without monotonically increasing DBH measurements as follows:
+
+- Where DBH decreases occur in dead trees or during periods where trees died, we will treat these decreases as likely due to wood shrinkage/rot associated with mortality.
+- Where DBH decreases occur during periods where trees were live, we will treat these decreases as a measurement error.  We will use linear interpolation to correct these errors.
+
+Code below.
+
+### Trees with decreasing DBH over dead periods
+
+```{r}
+# Step one: identify periods where DBH decrease occured during a period when tree waslive
+
+dbhDecrease %>% 
+  filter(Status=="Dead") %>% # filter trees with decreasing DBH that had dead status at any point
+  left_join(dat,by=c("Plot","TreeID")) %>%  # join with dataset
+  dplyr::select(Plot,TreeID,Year=Year.y,Status=Status.y,DBH=DBH.y)
+
+```
+
+A visual inspection reveals that the following trees had DBH decreases during periods where the tree was dead:
+
+- Plot 3, Tree 457
+- Plot 10, Tree 409
+
+These trees should be filtered from further adjustment.
+
+```{r}
+# Filter out trees with DBH decrease with dead periods
+dbhDecrease<-
+  dbhDecrease %>% 
+  filter(!(Plot==3 & TreeID==457)) %>% 
+  filter(!(Plot==10 & TreeID==409))
+
+```
+
+### Trees with decreasing DBH over live periods
+We will make adjustments on a tree by tree basis.
+
+First, reshape the dbhDecrease dataframe to make it easier to see DBH by year, then save it to a .csv file for manual editing.
+
+*I set the following chunk not to run, because we only want to do this once*
+
+```{r eval=FALSE}
+dbhEdit1<-
+  dbhDecrease%>% 
+  dplyr::select(-Status) %>% # Don't need this column anymore
+  pivot_wider(names_from="Year",values_from="DBH") %>% 
+  write.csv(file="dbh_for_edit.csv",row.names=FALSE)
+
+# Now edit .csv file manually
+
+```
+
+One thing to note: the following trees had fallen as of 2019, which means we should not interpolate their 2019 values:
+```{r}
+dbhDecrease %>% 
+  filter(Year==2019 & is.na(DBH))
+
+```
+
+Also, there are trees that were <7.5cm DBH during the 1992 or 1994 measurements.  DBH values for those trees/periods should not be interpolated either.
+
+For values that shouldn't be interpolated, I entered 'DoNotReplace' in the corresponding cell.
+
+Values to be replaced by interpolation had 'Interpolate' entered into their cell.
+
+```{r}
+dbhEdit2<-
+  readr::read_csv(here::here("data-raw","dbh_for_edit.csv"))
+
+# save this file so we can go back to it
+save(dbhEdit2,file=paste("editedDBH",Sys.Date(),"RData",sep="."))
+
+# Show data with replacements
+dbhEdit2
+
+```
+
+Now we can interpolate the missing values
+
+```{r}
+library(imputeTS)
+library(padr)
+library(lubridate)
+library(Hmisc)
+
+# Create a vector of years that are not to be interpolated
+DoNotReplace<-  
+  dbhEdit2 %>% 
+  pivot_longer(cols=`1992`:`2019`,names_to="Year",values_to="DBH") %>% 
+  filter(DBH=="DoNotReplace")
+
+# Interpolation
+dbhEdit3<-
+  dbhEdit2 %>% 
+  pivot_longer(cols=`1992`:`2019`,names_to="Year",values_to="DBH") %>% 
+  mutate(DBH=as.numeric(DBH)) %>% 
+  mutate(Year=paste("01-01",Year,sep="-")) %>% 
+  mutate(Year=dmy(Year)) %>% 
+  group_by(TreeID) %>% 
+  pad() %>% 
+  ungroup() %>% 
+  mutate(Year=year(Year)) %>% 
+  as.data.frame()
+
+# Define function to predict DBH values
+dbhInterp<-function() {
+
+  YearList=data.frame(Year=c(1992,1994,1997,2009,2019))
+
+ x<-
+   dbhEdit3 %>% 
+   lm(DBH~Year,data=.) %>% 
+   predict(.,newdata=YearList)
+
+ return(data.frame(Year=YearList$Year,DBH.pred=x))
+
+}
+
+# Run function by TreeID
+dbhPred<-by(dbhEdit3, dbhEdit3[,"TreeID"],
+   function(x) predict(lm(DBH ~ Year, data = x),new=data.frame(Year=c(1992,1994,1997,2009,2019))),simplify=F)
+
+do
+
+```

### 2020-02-28 data-raw/ep1162_prelimDataScreening.Rmd
More data screening

@@
 date: "18/02/2020"
-output: html_document
+output:
+  word_document: default
+  html_document: default
 ---
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
+```{r setup, warning=FALSE, message=FALSE}
+knitr::opts_chunk$set(message=FALSE, warning = FALSE)
 library(tidyverse)
@@
 
-# Now edit .csv file manually
-
 ```
@@
   filter(Year==2019 & is.na(DBH))
+
+dbhEdit2<-
+  readr::read_csv(here::here("data-raw","dbh_for_edit.csv"))
+
+# save this file so we can go back to it
+save(dbhEdit2,file=paste("editedDBH",Sys.Date(),"RData",sep="."))
 
@@
 ```{r}
-dbhEdit2<-
-  readr::read_csv(here::here("data-raw","dbh_for_edit.csv"))
-
-# save this file so we can go back to it
-save(dbhEdit2,file=paste("editedDBH",Sys.Date(),"RData",sep="."))
+# Load tree data with values to replace
+load(here::here("data-raw","editedDBH.2020-02-27.RData"))
 
@@
   pivot_longer(cols=`1992`:`2019`,names_to="Year",values_to="DBH") %>% 
+  mutate(TreeID=factor(TreeID)) %>% 
+  mutate(Year=as.numeric(Year)) %>% 
   filter(DBH=="DoNotReplace")
@@
   mutate(Year=dmy(Year)) %>% 
+  mutate(TreeID=factor(TreeID)) %>% 
   group_by(TreeID) %>% 
@@
 # Run function by TreeID
-dbhPred<-by(dbhEdit3, dbhEdit3[,"TreeID"],
+dbhPred1<-by(dbhEdit3, dbhEdit3[,"TreeID"],
    function(x) predict(lm(DBH ~ Year, data = x),new=data.frame(Year=c(1992,1994,1997,2009,2019))),simplify=F)
 
-do
+# Bring datasets together 
+dbhPred<-data.frame(TreeID=names(dbhPred1),do.call(rbind.data.frame,dbhPred1)) %>% # unpack output from by() function
+  setNames(c("TreeID",c(1992,1994,1997,2009,2019))) %>% # rename columns
+  pivot_longer(cols=`1992`:`2019`,names_to="Year",values_to = "DBH.Pred") %>% # pivot longer 
+  mutate(Year=as.numeric(Year)) %>%
+
+  # Join with dbhEdit3
+  left_join(dbhEdit3,by=c("TreeID","Year")) %>%
+
+  # Join with doNot Replace
+  left_join(DoNotReplace,by=c("TreeID","Year")) %>%
+
+  # Formatting
+  dplyr::select(Plot=Plot.x,TreeID,Year,DBH=DBH.x,DBH.Pred,DoNotRep=DBH.y) %>% 
+  mutate_at(vars(DBH.Pred),round,1) %>% # round DBH.Pred
+
+  # Find and replace
+  mutate(DBH.New=DBH) %>% 
+  mutate(DBH.New=coalesce(DBH.New,DBH.Pred)) %>% 
+  mutate(DBH.New=replace(DBH.New,!is.na(DoNotRep),"NA")) %>%
+
+  # Set new DBH column
+  dplyr::select(Plot:Year,DBH=DBH.New) %>% 
+  mutate(DBH.Pred=as.numeric(DBH)) %>% 
+  dplyr::select(-DBH) %>% 
+  mutate(Plot=factor(Plot)) %>% 
+  mutate(Year=factor(Year)) %>% 
+  mutate(TreeID=factor(TreeID))
 
 ```
+
+Now let's test to make sure that all DBH are now monotonic.
+
+```{r}
+library(MonoInc)
+
+x<-
+  dbhPred %>% 
+  drop_na(DBH)
+
+ mono.x<- by(x$DBH,x$TreeID,monotonic)
+
+ # One tree still has decreasing DBH
+ dbhPred %>% 
+   filter(TreeID==as.numeric(names(which(mono.x==FALSE))))
+
+```
+
+One tree still has decreasing DBH.
+1994 value for this tree is less than 1992.
+This is because the linear function predicts values below the 1992 value.
+Let's manually replace the 1994 value with interpolation between 1992 and 1997 values.
+
+```{r}
+
+  dbhPred[dbhPred$TreeID==31 & dbhPred$Year==1994,"DBH"]=22.5
+
+```
+
+*** 
+## Now merge DBH values back into the dataset
+```{r}
+dat<-
+  dat %>% 
+  left_join(dbhPred,by=c("Plot","TreeID","Year")) %>% 
+  arrange(Plot,TreeID,Year) %>% 
+  mutate(DBH=coalesce(DBH.Pred,DBH)) %>% 
+  mutate_at(vars(DBH),round,1) %>% 
+  dplyr::select(-DBH.Pred)
+
+```
+
+# Heights
+Now let's look at trees where height measurements aren't monotonically increasing.
+
+*The chunk below isn't set to run, because we only want to do this once*
+
+```{r eval=FALSE}
+x<-
+  dat %>% 
+  drop_na(Height) %>% # drop rows with missing Height values 
+  arrange(Plot,TreeID,Year) %>% 
+  mutate(id=paste(Plot,TreeID,sep="."))
+
+ mono.x<- by(x$Height,x$id,monotonic)
+
+  # Trees with decreasing height measurements
+ # write to .csv file
+  x %>% 
+   group_by(id) %>% 
+   summarise(n=n()) %>% 
+   filter(n>1) %>%  # filter tree records with more than one height measurement
+   filter(id%in%names(which(mono.x==FALSE))) %>% 
+   left_join(x,by="id") %>% 
+   arrange(id) %>% 
+   dplyr::select(Plot,TreeID,Species,Year,Height) %>% 
+   pivot_wider(names_from="Year",values_from = "Height")
+
+```
+
+42 trees have decreasing DBH.  A visual inspection of the plot cards revealed several trees with broken tops (e.g., Plot 12 Tree 22, and Plot 18 Tree 346).
+
+In almost all other cases, there is too much fluctuation to confidently delete height measurements for interpolation.  I recommend we leave the height measurements as is.

### 2020-02-28 data-raw/ep1162_prelimDataScreening.Rmd
Data cleaning

@@
 output:
+  html_document: default
   word_document: default
-  html_document: default
 ---
@@
 
-In total, there are `r dbhDecrease %>% count(TreeID) %>% nrow()` trees with this situation.  I have the treeID saved and can provide them if we want to double check these trees.
+In total, there are `r dbhDecrease %>% count(TreeID) %>% nrow()` trees with this situation.
 
@@
 One thing to note: the following trees had fallen as of 2019, which means we should not interpolate their 2019 values:
+
 ```{r}
@@
   filter(Year==2019 & is.na(DBH))
+```
 
+*The following chunk won't run, because we only want to do this once*
+```{r eval=FALSE}
 dbhEdit2<-
@@
   group_by(TreeID) %>% 
-  pad() %>% 
+  pad() %>% # pad out time series so we can interpolate
   ungroup() %>% 
@@
   dbhPred %>% 
-  drop_na(DBH)
+  drop_na(DBH.Pred)
 
- mono.x<- by(x$DBH,x$TreeID,monotonic)
+ mono.x<- by(x$DBH.Pred,x$TreeID,monotonic)
 
@@
 
-  dbhPred[dbhPred$TreeID==31 & dbhPred$Year==1994,"DBH"]=22.5
+  dbhPred[dbhPred$TreeID==31 & dbhPred$Year==1994,"DBH.Pred"]=22.5
 
@@
 
-*The chunk below isn't set to run, because we only want to do this once*
-
-```{r eval=FALSE}
+```{r}
 x<-

### 2020-03-02 data-raw/ep1162_prelimDataScreening.Rmd
final data checking

@@
   # Trees with decreasing height measurements
- # write to .csv file
+
   x %>% 
@@
 In almost all other cases, there is too much fluctuation to confidently delete height measurements for interpolation.  I recommend we leave the height measurements as is.
+
+*** 
+# FINAL SAVING
+We'll save the final dataset and export it as well to .csv for final checks.
+
+```{r}
+dat %>% 
+  write.csv(file=here::here("data-raw","ep1162_finalDataCheck.csv"))
+
+dat %>% 
+  save(file=here::here("data","ep1162_Data.RData")
+
+```

### 2020-03-02 data-raw/ep1162_prelimDataScreening.Rmd
Data cleaning

@@
 
+### Remove trees with DBH<7.5cm DBH for that year
+We need to remove trees from the record that have less than 7.5cm DBH recorded.
+
+There are `dat %>% filter(DBH<7.5) %>% nrow()` records of trees with less than 7.5cm DBH that will be removed using the following chunk.
+
+```{r}
+
+dat<- # re-assign to dat
+  dat %>% 
+  filter(DBH>=7.5) # keep trees >= 7.5cm DBH
+
+```
+
 *** 
@@
 dat %>% 
-  write.csv(file=here::here("data-raw","ep1162_finalDataCheck.csv"))
+  write.csv(file=here::here("data-raw","ep1162_finalData.csv"),
+            row.names=F,
+            na="")
 
 dat %>% 
-  save(file=here::here("data","ep1162_Data.RData")
+  save(file=here::here("data","ep1162_Data.RData"))
 

### 2020-03-09 data-raw/dataClean1.R
datacleaning

@@ -1,9 +1,10 @@
 # Script to clean and format EP1162 raw data
 
 library(tidyverse)
+library(here)
 
 # Read Excel file
-X<-readxl::read_excel("data-raw/4.0 - FINAL DATASET FOR ANALYSIS - EP 1162 -  1992-2019 FIELD DATA (Feb 4th 2019) MJ.xlsx",
+X<-readxl::read_excel(here("data-raw","4.0 - FINAL DATASET FOR ANALYSIS - EP 1162 -  1992-2019 FIELD DATA (Feb 4th 2019) MJ.xlsx"),
                       skip=10)
 
 

### 2020-03-09 analysis/fdiBasalAreaAnalysis.Rmd
data cleaning

---
title: "Fdi basal area analysis"
author: "Hardy Griesbauer"
date: "06/03/2020"
output: html_document
---

## The mystery of the disappearing Fdi basal area
The figure below suggests that Douglas-fir basal area declined noticeably between 1992 and 1994 in the control and 20m2 residual treatment units.  Below is some script to analyze this phenomenon further.

```{r setup,include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width=8)

# Load libraries
library(tidyverse)

# Load dataset
load(here::here("data","ep1162_Data.RData"))
load(here::here("data","ep1162_plotData.RData"))

# set up diameter classes
diamClass=seq(from=7.5,to=95,by=5)

dat %>% 
    filter(Species!="Ep") %>% # remove birch 
  filter(Status=="Live") %>% # filter for live only trees
   left_join(plotDat,by="Plot") %>% # join with plot data
   mutate(DBH.class=cut(DBH,diamClass,ordered_result = TRUE,right=F)) %>% # assign DBH classes
  mutate(BA=(DBH/100/2)^2*pi) %>% # convert tree DBH to basal area
  group_by(BA.Target,Plot,Year,Species) %>% 
  summarise(BA=sum(BA)*20) %>% # multiply by 20 for BA/ha
  ungroup() %>%

  # create mean and SD
  group_by(BA.Target,Year,Species) %>% 
  summarise(BA.Mean=mean(BA),
            BA.se=sd(BA)/sqrt(length(BA))) %>% # calculate standard error of the mean

    # Add labeller column
  mutate(label=factor(paste("Trt:",BA.Target,sep=""))) %>%

  # # pad out the year column
  # right_join(data.frame(Year=as.factor(1992:2019)),by="Year") %>%

  # Plotting
  ggplot()+
  aes(x=Year,y=BA.Mean,group=Species,fill=Species)+
  geom_bar(stat="Identity")+
  facet_wrap(~label)+
  ylab("Mean basal area per plot (m2/ha)")+
  xlab("Measurement year")

```

First step, let's isolate plots with Fdi in the two treatment units:
```{r}
fdiDat<-
  dat %>% 
  filter(Species=="Fd") %>% # filter for Douglas-fir entries
  left_join(plotDat,by="Plot") %>% 
  mutate(DBH.class=cut(DBH,diamClass,ordered_result = TRUE,right=F)) # assign DBH classes

```

Looks like there are two trees that are missing measurements from 1994 and 1997.

```{r}
fdiDat %>% 
  filter(TreeID %in% c(400,188))

```

Are there any other trees (non-Fd) that are missing mid-point measurements?

```{r}

  dat %>% 
  filter(TreeID!="New") %>% # have to remove these from dataset
  dplyr::select(Plot,TreeID,Year,DBH) %>% # select columns to run analysis on
  pivot_wider(names_from=Year,values_from=DBH) %>% 
  filter(!is.na(`1992`) & !is.na(`2019`)) %>% # filter for trees with measurements at 1992 and 2019
  pivot_longer(cols=3:7,names_to="Year",values_to="DBH") %>% 
  drop_na(DBH) %>% 
  group_by(Plot,TreeID) %>% 
  summarise(num=n()) %>% 
  filter(num<5)

```

I suggest we interpolate these using linear regression:

```{r}

### 2020-03-09 analysis/prelimAnalysis.Rmd
data cleaning

---
title: "EP1162 Preliminary Analysis"
author: "Hardy Griesbauer"
date: "04/03/2020"
output: html_document
---

```{r setup, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width=8)

# Load libraries
library(tidyverse)

# Load dataset
load(here::here("data","ep1162_Data.RData"))
load(here::here("data","ep1162_plotData.RData"))

```

## Prelim analysis
Below are some preliminary figures of the EP1162 data.  I tried different types of figures just to give us a few options to discuss when putting the final report together.  I was excited to start taking a look at the data!

NB: this code is also posted in the github repo if you want to run it for your self.  Just clone the entire repo to your local drive and take it from there.

https://github.com/hgriesbauer/EP1162

*Also - there may still be some adjustments made to two height measurements in the dataset.  It will be easy to re-run this analysis with an updated dataset if need  be!*

### Number of trees by diameter class
```{r}

# set up diameter classes
diamClass=seq(from=7.5,to=95,by=5)

dat %>% 
  left_join(plotDat,by="Plot") %>% 
  mutate(DBH.class=cut(DBH,diamClass,ordered_result = TRUE,right=F)) %>% # assign DBH classes

  # Group and summarise
  group_by(BA.Target,Plot,Year,DBH.class) %>% 
  summarise(Num.Trees=n()) %>%
  mutate(Num.Trees=Num.Trees*20) %>% # multiply to get trees per hectare

  # Generate means and SD
  ungroup() %>% 
  group_by(BA.Target,Year,DBH.class) %>% 
  summarise(Tree.Count.mean=mean(Num.Trees),
            Tree.Count.SD=sd(Num.Trees)) %>%

  # Add labeller column
  mutate(label=factor(paste("Trt:",BA.Target,sep=""))) %>%

  # Plotting
  ggplot()+
  aes(x=DBH.class,y=Tree.Count.mean)+
  geom_bar(stat="Identity")+
  facet_wrap(~label~Year,ncol=5)+
  theme(axis.text.x = element_text(angle = 90, hjust = 1,size=6,vjust=0.25))+
  ylab("Trees per hectare")+
  xlab("Diameter classes (cm)")

```

In this figure, Trt:10 refers to the treatment units with a target of 10m2/hectare post-harvest.

I didn't place standard error bars on the graph.  They are quite wide (high variation within treatment units).

Interesting to note that in all three treatment units, the number of trees in the smaller diameter classes actually decrease between 1992 and 1994 - will look at the plots to see where this ocurred, but could also be related to mortality in the understory?

A few other observations from this figure:

- By 2019, the 3 treatment units are similar in termms of trees per hectare in the diameter classes above 17.5cm DBH.  The control has more stems in the smaller diameter classes.

*Could compare Q values between treatment units.*

### Trees per hectare by species in 2019
Let's look at species composition by treatment unit in 2019:

```{r}

dat %>% 
  left_join(plotDat,by="Plot") %>% # join with plot data
   mutate(DBH.class=cut(DBH,diamClass,ordered_result = TRUE,right=F)) %>% # assign DBH classes
  filter(Year==2019) %>% # filter for 2019
  filter(Species%in%c("Sx","Bl")) %>% #filter for Sx and Bl
  group_by(BA.Target,Plot,Species,DBH.class) %>% 
  summarise(Spp.Plot=n()) %>% # summarise number of trees per plot by species
  mutate(Spp.Plot=Spp.Plot*20) %>%# plot multiplier to get sph

  # create means
  ungroup() %>% 
  group_by(BA.Target,Species,DBH.class) %>%
  summarise(SPH=mean(Spp.Plot)) %>%

   # Add labeller column
  mutate(label=factor(paste("Trt:",BA.Target,sep=""))) %>%

  # Plotting
  ggplot()+
  aes(x=DBH.class,y=SPH,fill=Species)+
  geom_bar(stat="Identity")+
  facet_wrap(~label~Species,ncol=2)+
  theme(axis.text.x = element_text(angle = 90, hjust = 1,size=6,vjust=0.25))+
  ylab("Mean trees per hectare per plot")+
  xlab("Diameter classes (cm)")

```

A few observations:

- There are actually more spruce trees per hectare in the high-removal treatment than the other two.  
- Higher Q value for Control treatment SAF compared to harvested units.
-

### Basal area by year and treatment

```{r}
dat %>% 
   left_join(plotDat,by="Plot") %>% # join with plot data
   mutate(DBH.class=cut(DBH,diamClass,ordered_result = TRUE,right=F)) %>% # assign DBH classes
  mutate(BA=(DBH/100/2)^2*pi) %>% # convert tree DBH to basal area
  group_by(BA.Target,Plot,Year) %>% 
  summarise(BA=sum(BA)*20) %>% # multiply by 20 for BA/ha
  ungroup() %>%

  # create mean and SD
  group_by(BA.Target,Year) %>% 
  summarise(BA.Mean=mean(BA),
            BA.se=sd(BA)/sqrt(length(BA))) %>% # calculate standard error of the mean

    # Add labeller column
  mutate(label=factor(paste("Trt:",BA.Target,sep=""))) %>%

  # # padd out the year column
  # right_join(data.frame(Year=as.factor(1992:2019)),by="Year") %>%

  # Plotting
  ggplot()+
  aes(x=Year,y=BA.Mean,color=label,group=label)+
  geom_point(size=3)+
  geom_line()+
  geom_errorbar(aes(ymin=BA.Mean-BA.se,ymax=BA.Mean+BA.se),width=0.1)+
  xlab("Measurement year")+
  ylab("Mean basal area per plot (m2/ha)")

```

I included error bars in this code.

Observations:

- Basal area actually decreased between 1992 and 1994 for control and Trt20.  I will look into this a bit more.
- Both 10 and 20 Treatment Units have converged in terms of basal area per hectare by 2019, and are >10m2/ha less than the control.

### Basal area contribution by species

```{r}

dat %>% 
    filter(Species!="Ep") %>% # remove birch 
  filter(Status=="Live") %>% # filter for live only trees
   left_join(plotDat,by="Plot") %>% # join with plot data
   mutate(DBH.class=cut(DBH,diamClass,ordered_result = TRUE,right=F)) %>% # assign DBH classes
  mutate(BA=(DBH/100/2)^2*pi) %>% # convert tree DBH to basal area
  group_by(BA.Target,Plot,Year,Species) %>% 
  summarise(BA=sum(BA)*20) %>% # multiply by 20 for BA/ha
  ungroup() %>%

  # create mean and SD
  group_by(BA.Target,Year,Species) %>% 
  summarise(BA.Mean=mean(BA),
            BA.se=sd(BA)/sqrt(length(BA))) %>% # calculate standard error of the mean

    # Add labeller column
  mutate(label=factor(paste("Trt:",BA.Target,sep=""))) %>%

  # # pad out the year column
  # right_join(data.frame(Year=as.factor(1992:2019)),by="Year") %>%

  # Plotting
  ggplot()+
  aes(x=Year,y=BA.Mean,group=Species,fill=Species)+
  geom_bar(stat="Identity")+
  facet_wrap(~label)+
  ylab("Mean basal area per plot (m2/ha)")+
  xlab("Measurement year")

```

*Note: I removed birch from the dataset for this figure*

A few observations:

- Spruce basal area increased relatively faster in the high-removal treatment unit compared to the other two treatment units.  Interesting!
- Douglas-fir basal area decreased between 1992 and 1994.  This likely explains the overall basal area decreases between 92 and 94 in previous figure.  Will look into this more  *Note - I removed dead trees for this analysis, just to see if this explained the decrease in Fdi BA - no effect*
- Close to 30 years afterwards, both harvest treatments are similar in terms of total basal area, and SAF area.

### 2020-03-10 analysis/prelimAnalysis.Rmd
updated prelim analysis

@@
 Below are some preliminary figures of the EP1162 data.
-I tried different types of figures just to give us a few options to discuss when putting the final report together.
-I was excited to start taking a look at the data!
 
@@
 
-*Also - there may still be some adjustments made to two height measurements in the dataset.  It will be easy to re-run this analysis with an updated dataset if need  be!*
+NB2: Plots 16,17, 18 and 19 were omitted from the dataset for these analyses because they were not measured in 1994 or 1997.
+
+```{r}
+dat<-
+  dat %>% 
+  filter(!Plot %in% c(16,17,18,19)) # remove plots 16-19 from data for further analysis
+
+```
+
+This means that the following analyses are based on `r length(unique(dat$Plot))` plots.
 
@@
   left_join(plotDat,by="Plot") %>% 
+   filter(Status=="Live") %>% # filter for live trees only
   mutate(DBH.class=cut(DBH,diamClass,ordered_result = TRUE,right=F)) %>% # assign DBH classes
@@
 
-*Could compare Q values between treatment units.*
-
-### Trees per hectare by species in 2019
+### Trees per hectare by species and diameter class in 2019
 Let's look at species composition by treatment unit in 2019:
@@
   left_join(plotDat,by="Plot") %>% # join with plot data
+   filter(Status=="Live") %>% # filter for live trees only
    mutate(DBH.class=cut(DBH,diamClass,ordered_result = TRUE,right=F)) %>% # assign DBH classes
@@
 - Higher Q value for Control treatment SAF compared to harvested units.
--
+
+### Basal area per hectare by species and diameter class in 2019
+Let's look at basal area by species, diameter class and treatment unit in 2019:
+
+```{r}
+dat %>% 
+  left_join(plotDat,by="Plot") %>% # join with plot data
+   filter(Status=="Live") %>% # filter for live trees only
+   mutate(DBH.class=cut(DBH,diamClass,ordered_result = TRUE,right=F)) %>% # assign DBH classes
+  filter(Year==2019) %>% # filter for 2019
+  filter(Species%in%c("Sx","Bl")) %>% #filter for Sx and Bl
+  mutate(BA=(DBH/2/100)^2*pi) %>% # convert DBH to basal area
+  group_by(BA.Target,Plot,Species,DBH.class) %>% 
+  summarise(BA.Plot=sum(BA)*20) %>%   # sum basal area per diam class and multiply to get per ha
+
+  # create means
+  ungroup() %>% 
+  group_by(BA.Target,Species,DBH.class) %>%
+  summarise(BA=mean(BA.Plot)) %>%
+
+   # Add labeller column
+  mutate(label=factor(paste("Trt:",BA.Target,sep=""))) %>%
+
+  # Plotting
+  ggplot()+
+  aes(x=DBH.class,y=BA,fill=Species)+
+  geom_bar(stat="Identity")+
+  facet_wrap(~label~Species,ncol=2)+
+  theme(axis.text.x = element_text(angle = 90, hjust = 1,size=6,vjust=0.25))+
+  ylab("Mean basal area per hectare per plot (m2)")+
+  xlab("Diameter classes (cm)")
+
+```
+
+A few observations:
+
+- Sx basal area is concentrated in the larger diameter classes.
+- Bl basal area is more evenly distributed across diameter classes.
 
@@
    left_join(plotDat,by="Plot") %>% # join with plot data
+  filter(Status=="Live") %>% 
    mutate(DBH.class=cut(DBH,diamClass,ordered_result = TRUE,right=F)) %>% # assign DBH classes
@@
     # Add labeller column
-  mutate(label=factor(paste("Trt:",BA.Target,sep=""))) %>%
+  mutate(Trt.Unit=factor(paste("Trt:",BA.Target,sep=""))) %>%
 
@@
   ggplot()+
-  aes(x=Year,y=BA.Mean,color=label,group=label)+
+  aes(x=Year,y=BA.Mean,color=Trt.Unit,group=Trt.Unit)+
   geom_point(size=3)+
@@
 
-- Basal area actually decreased between 1992 and 1994 for control and Trt20.  I will look into this a bit more.
 - Both 10 and 20 Treatment Units have converged in terms of basal area per hectare by 2019, and are >10m2/ha less than the control.
+- Treatment Unit 10 has a steeper temporal trend than Treatment Unit 20.  Growing faster?
 
@@
   ggplot()+
-  aes(x=Year,y=BA.Mean,group=Species,fill=Species)+
+  aes(x=Year,y=BA.Mean,fill=Species)+
   geom_bar(stat="Identity")+
@@
 - Spruce basal area increased relatively faster in the high-removal treatment unit compared to the other two treatment units.  Interesting!
-- Douglas-fir basal area decreased between 1992 and 1994.
-This likely explains the overall basal area decreases between 92 and 94 in previous figure.
-Will look into this more  *Note - I removed dead trees for this analysis, just to see if this explained the decrease in Fdi BA - no effect*
 - Close to 30 years afterwards, both harvest treatments are similar in terms of total basal area, and SAF area.
+- 2019 BA in both treatment units are similar to control in 2009.  Treatment units are about 10 years behind at this point?
+
+### PBAI
+Below is code to recreate Mike's PBAI graph by initial BA:
+
+```{r}
+
+dat %>% 
+  left_join(plotDat,by="Plot") %>% # join with plot data
+   filter(Status=="Live") %>% # filter for live trees only
+  mutate(BA=(DBH/2/100)^2*pi*20) %>% # convert DBH to basal area and multiply for per hectare
+  dplyr::select(BA.Target,Plot,TreeID,Species,Year,BA) %>% 
+  group_by(BA.Target,Plot,Year) %>% 
+  summarise(BA.Sum=sum(BA,na.rm=T)) %>% 
+  pivot_wider(names_from=Year,values_from=BA.Sum) %>% 
+  mutate(`1992-1994`=`1994`-`1992`) %>% 
+  mutate(`1994-1997`=`1997`-`1994`) %>% 
+  mutate(`1997-2009`=`2009`-`1997`) %>% 
+  mutate(`2009-2019`=`2019`-`2009`) %>%
+
+  # pivot back
+  dplyr::select(BA.Target,Plot,`1992-1994`:`2009-2019`) %>% 
+  pivot_longer(cols=`1992-1994`:`2009-2019`,names_to="Period",values_to="PBAI") %>% 
+  mutate(Period=factor(Period)) %>%
+
+  # summarise PBAI by BA.Target
+  group_by(BA.Target,Period) %>% 
+  summarise(PBAI.mean=mean(PBAI,na.rm=T),
+            PBAI.se=sd(PBAI,na.rm=T)/sqrt(length(PBAI))) %>%
+
+    # Add labeller column
+  mutate(Trt.Unit=factor(paste("Trt:",BA.Target,sep=""))) %>%
+
+  # Plotting
+  ggplot()+
+  aes(x=Period,y=PBAI.mean,fill=Trt.Unit,color=Trt.Unit,group=Trt.Unit)+
+  geom_point(size=3)+
+  geom_line()+
+  geom_errorbar(aes(ymin=PBAI.mean-PBAI.se,ymax=PBAI.mean+PBAI.se),width=0.1)+
+  xlab("Period")+
+  ylab("Mean basal area per plot (m2/ha)")
+
+```
+
+Graph looks similar to Mike's original figure.
+
+Some very interesting trends here:
+
+- Treatment 10 plots have higher increment than other treatment units.
+- Control treatment plots increment is starting to stabilize now?  Does this suggest that in the next 10-20 years, the treatment units may converge with control in terms of BAI?
+
+### Effect of initial BA on PBAI
+Below is R code to re-create Mike's figure of initial BA on PBAI by period
+
+```{r}
+
+pbai<- # assign output so we can plot several periods
+  dat %>% 
+  left_join(plotDat,by="Plot") %>% # join with plot data
+   filter(Status=="Live") %>% # filter for live trees only
+  mutate(BA=(DBH/2/100)^2*pi*20) %>% # convert DBH to basal area and multiply for per hectare
+  dplyr::select(BA.Target,Plot,TreeID,Species,Year,BA) %>% 
+  group_by(BA.Target,Plot,Year) %>% 
+  summarise(BA.Sum=sum(BA,na.rm=T)) %>% 
+  pivot_wider(names_from=Year,values_from=BA.Sum) %>% 
+  mutate(`1992-1994`=`1994`-`1992`) %>% 
+  mutate(`1994-1997`=`1997`-`1994`) %>% 
+  mutate(`1997-2009`=`2009`-`1997`) %>% 
+  mutate(`2009-2019`=`2019`-`2009`)
+
+  # Plotting function
+pbai.Plot<-function(initYear,baPeriod,data=pbai) {
+
+  p1<-
+    data.frame(data[initYear],data[baPeriod]) %>% 
+    setNames(c("initialBA","PBAI")) %>% 
+     # filter(PBAI>=0) %>% 
+      ggplot()+
+    aes(x=initialBA,y=PBAI)+
+    geom_point(size=3)+
+    xlab(paste(initYear," plot basal area per hectare",sep=""))+
+    ylab(paste("BAI per hectare for ",baPeriod, "Period"))
+
+  return(p1)
+
+}
+
+pbai.Plot(initYear="2009",baPeriod="2009-2019")
+
+```
+
+We see a negative relationship between plot basal area and 2009-2019 PBAI.
+Note the outlier - this plot is a control plot that had a large Sx tree die between 2009-2019, resulting in negative PBAI for that period.
+
+The relationship is consistent using initial BA for earlier periods (e.g., 1994):
+
+```{r}
+pbai.Plot(initYear="1994",baPeriod="2009-2019")
+```
+
+*For discussion*  Although the relationship seems nonlinear negative, I think that's mostly due to the plot (Treatment: Control Plot 3) with negative PBAI.
+I don't think we would want to model plots with negative PBAI, as these are due to stand dynamics/forest health factors, and we are primarily interested in effects of treatment on growth.
+
+With plot 3 removed, the relationship looks more linear:
+
+```{r}
+pbai.2<-
+  pbai %>% 
+  filter(Plot!=3)
+
+pbai.Plot(initYear="1994",baPeriod="2009-2019",data=pbai.2)
+
+```
+
+***
+This linear relationship doesn't hold for periods earlier than the 2009-2019 period:
+
+```{r}
+
+pbai.Plot(initYear="1994",baPeriod="1994-1997")
+```

### 2020-03-10 analysis/prelimAnalysis.Rmd
A few adjustments to analysis.

@@
   xlab("Period")+
-  ylab("Mean basal area per plot (m2/ha)")
+  ylab("Mean BAI per plot (m2/ha/period)")
 

### 2020-03-20 analysis/heightModel.Rmd
add more files and heigh:dbh differential

---
title: "Height estimates from DBH:Height relationship"
author: "Hardy Griesbauer"
date: "20/03/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Libraries
library(tidyverse)
library(here)

# Load data
load(here("data","ep1162_Data.RData"))

```

## Estimating tree heights based on dbh:height relationships
In order to estimate stand-level volumes for each treatment, we need to first estimate heights for trees that do not have height measurements.  We will do the following:

1. Model DBH:Height relationships separately for each measurement period, and for Bl:Sx separately (at first).
2. Compare Bl and Sx relationships, compare estimates between years, and determine if we can pool species and/or years to estimate Height.

### Step 1: Explore height:dbh relationships with scatterplots
First, let's explore scatterplots for each species and measurement year:

```{r dbh:height scatterplots,fig.height=12, fig.width=10}

  dat %>% 
    filter(Species %in% c("Bl","Sx")) %>% # filter for Bl and Sx for now
    filter(Status=="Live") %>% 
    drop_na(Height) %>% # remove trees where height was not measured

    # Plotting
  ggplot()+
  aes(x=DBH,y=Height,fill=Species)+
  geom_point()+
  facet_wrap(Year~Species,ncol=2)+
  theme(legend.position = "none")+
  ylab("Height (m)")+
  xlab("DBH (cm)")+
  geom_smooth(formula = y~log(x),method="lm",se=FALSE)

```

It looks like there are differences between measurement years.  There may be subtle differences between species within a measurement year.  Let's explore that now by superposing Bl and Sx graphs within same year:

```{r dbh:height scatterplots species superposed,fig.height=12, fig.width=10}

  dat %>% 
    filter(Species %in% c("Bl","Sx")) %>% # filter for Bl and Sx for now
    filter(Status=="Live") %>% 
    drop_na(Height) %>% # remove trees where height was not measured

    # Plotting
  ggplot()+
  aes(x=DBH,y=Height,color=Species)+
  geom_point()+
  facet_wrap(~Year)+
  theme(legend.position = "bottom")+
  ylab("Height (m)")+
  xlab("DBH (cm)")+
  geom_smooth(formula = y~log(x),method="lm",se=FALSE)

```

We can see subtle differences between Bl and Sx DBH:Height relationships in first three measurement periods.  By 2019, it looks like DBH:height relationships converged between the species.

***

Let's look at differences between years by superposing different year graphs for species separately:

```{r dbh:height scatterplots years superposed,fig.height=12, fig.width=10}

  dat %>% 
    filter(Species %in% c("Bl","Sx")) %>% # filter for Bl and Sx for now
    filter(Status=="Live") %>% 
    drop_na(Height) %>% # remove trees where height was not measured

    # Plotting
  ggplot()+
  aes(x=DBH,y=Height,color=Year)+
  geom_point()+
  facet_wrap(~Species,ncol=2)+
  theme(legend.position = "bottom")+
  ylab("Height (m)")+
  xlab("DBH (cm)")+
  geom_smooth(formula = y~log(x),method="lm",se=FALSE)

```

Looks like Bl maintained the same DBH:Height relationship over all measurement periods.  Sx 2019 relationship differs from other time periods.

****
### Can we pool species and measurement years?

I believe we can test this statistically:

*Note: I will get Peter Ott to confirm this is the correct way to test regression models between groups*

```{r test for species and year effect}

# Assign data
x<-
   dat %>% 
    filter(Species %in% c("Bl","Sx")) %>% # filter for Bl and Sx for now
    filter(Status=="Live") %>% 
    drop_na(Height) # remove trees where height was not measured

# fit a model with all terms
mod.full<-lm(Height~log(DBH)+log(DBH):Species+log(DBH):Year+log(DBH):Species:Year,data=x)

# fit model with only DBH and species terms
mod.species<-lm(Height~log(DBH)+log(DBH):Species,data=x)

# fit model with only DBH and year terms
mod.year<-lm(Height~log(DBH)+log(DBH):Year,data=x)

# fit model with only DBH terms
mod.dbh<-lm(Height~log(DBH),data=x)

# Run through these iteratively
anova(mod.full,mod.dbh)

```

Based on this ANOVA table, there is a significant reduction in residual sum of squares between the full model and the DBH-only model.  The species+DBH and year+DBH models don't seem to add much (not shown).

***
### Compare R^2^ between models
How much variance is explained by each model?  Summary below:

```{r extract r2 and other stats for each model}
rbind(data.frame(Model="full",broom::glance(mod.full)),
      data.frame(Model="dbh-only",broom::glance(mod.dbh)),
      data.frame(Model="dbh+species",broom::glance(mod.species)),
      data.frame(Model="dbh+year",broom::glance(mod.year))

      ) %>% 
  dplyr::select(1:7)

```

Based on this, and the extremely modest increase in R^2^ from DBH-only to full model, I'm tempted to say that we can pool all species and years into the most parsimonious (ie., DBH:Height) model.  But definitely up for discussion!

### DBH:Height equation for pooled data
If we decide to use pooled data for the DBH:Height equation:

```{r dbh:height equation}

paste("Predicted.Height = ",coef(mod.dbh)[2],
      " x ln(DBH) ",
      coef(mod.dbh)[1])
```

Here are the coefficients from Mike's analysis:

*Subalpine fir*
HT = 11.188 x ln(DBH)  - 17.609

*Hybrid white spruce*
HT = 11.79 x ln(DBH)  - 19.637

The coefficients using the pooled species/years are pretty similar to Mike's formula for spruce.

### 2020-03-20 analysis/prelimAnalysis.Rmd
updating figures

@@
   # Plotting function
-pbai.Plot<-function(initYear,baPeriod,data=pbai) {
+pbai.Plot<-function(initYear,baPeriod,data=pbai,x=15,y=5) {
+
+  # Linear model to extract coefficients and R2
+  mod1<-
+      data.frame(data[initYear],data[baPeriod]) %>% 
+      setNames(c("initialBA","PBAI")) %>% 
+      lm(PBAI~initialBA,data=.)
+
+  text1=paste(round(broom::tidy(mod1)[1,"estimate"],3),
+              " + (",
+              round(broom::tidy(mod1)[2,"estimate"],3),
+              "X)\n",
+              "R² = ",
+              round(broom::glance(mod1)[1],3),
+              "\np = ",
+              round(broom::glance(mod1)["p.value"],3),
+              sep="")
 
@@
     geom_point(size=3)+
+    geom_text(label=text1,x=x,y=y,check_overlap=TRUE,hjust=0)+
+    geom_smooth(method="lm",se=FALSE)+ # add regression line
     xlab(paste(initYear," plot basal area per hectare",sep=""))+
@@
 
-pbai.Plot(initYear="2009",baPeriod="2009-2019")
+# iteratively adjust x and y to line up text properly
+pbai.Plot(initYear="2009",baPeriod="2009-2019",x=10,y=2)
 
@@
 Note the outlier - this plot is a control plot that had a large Sx tree die between 2009-2019, resulting in negative PBAI for that period.
-
-The relationship is consistent using initial BA for earlier periods (e.g., 1994):
-
-```{r}
-pbai.Plot(initYear="1994",baPeriod="2009-2019")
-```
 
@@
 
-pbai.Plot(initYear="1994",baPeriod="2009-2019",data=pbai.2)
+pbai.Plot(initYear="1994",baPeriod="2009-2019",data=pbai.2,x=5,y=3)
 
@@
 
+1994 period:
+
 ```{r}
 
-pbai.Plot(initYear="1994",baPeriod="1994-1997")
+pbai.Plot(initYear="1994",baPeriod="1994-1997",y=3.5,x=25)
 ```
+
+1997 period:
+
+```{r}
+
+pbai.Plot(initYear="1997",baPeriod="1997-2009",x=5,y=1)
+```

### 2020-03-21 analysis/heightModel.Rmd
more updates

@@
 
+As per previous discussion, we shouldn't pool measurements across measurement years because this would violate assumption of independence among samples.
+
 ****
-### Can we pool species and measurement years?
+### Can we pool species?
 
@@
 
-```{r test for species and year effect}
+```{r test for species effect}
+
+# function to compare three models
+# separate models will be fit for each year
+
+speciesMod1<-function(year) {
 
@@
     filter(Status=="Live") %>% 
+  filter(Year==year) %>% 
     drop_na(Height) # remove trees where height was not measured
@@
 # fit a model with all terms
-mod.full<-lm(Height~log(DBH)+log(DBH):Species+log(DBH):Year+log(DBH):Species:Year,data=x)
-
-# fit model with only DBH and species terms
-mod.species<-lm(Height~log(DBH)+log(DBH):Species,data=x)
-
-# fit model with only DBH and year terms
-mod.year<-lm(Height~log(DBH)+log(DBH):Year,data=x)
+mod.full<-lm(Height~log(DBH)+log(DBH):Species,data=x)
 
@@
 
-# Run through these iteratively
-anova(mod.full,mod.dbh)
+# extract coef and other values
+out<-data.frame(year=year,intercept=coef(mod.dbh)[1],
+                slope=coef(mod.dbh)[2],
+                p.value=broom::glance(mod.dbh)$p.value,
+                r2=broom::glance(mod.dbh)[1])
+rownames(out)=year
+
+# Return
+return(list(anov=anova(mod.full,mod.dbh),out=out))
+
+}
 
@@
 
-Based on this ANOVA table, there is a significant reduction in residual sum of squares between the full model and the DBH-only model.
-The species+DBH and year+DBH models don't seem to add much (not shown).
+Now let's run the model for the four measurement periods:
+
+1992:
+```{r}
+speciesMod1(1992)$anov
+```
 
 ***
-### Compare R^2^ between models
-How much variance is explained by each model?  Summary below:
+1994:
 
-```{r extract r2 and other stats for each model}
-rbind(data.frame(Model="full",broom::glance(mod.full)),
-      data.frame(Model="dbh-only",broom::glance(mod.dbh)),
-      data.frame(Model="dbh+species",broom::glance(mod.species)),
-      data.frame(Model="dbh+year",broom::glance(mod.year))
-
-      ) %>% 
-  dplyr::select(1:7)
-
+```{r}
+speciesMod1(1994)$anov
 ```
 
-Based on this, and the extremely modest increase in R^2^ from DBH-only to full model, I'm tempted to say that we can pool all species and years into the most parsimonious (ie., DBH:Height) model.
-But definitely up for discussion!
+***
+1997:
 
-### DBH:Height equation for pooled data
-If we decide to use pooled data for the DBH:Height equation:
+```{r}
+speciesMod1(1997)$anov
+```
+
+***
+2019:
+
+```{r}
+speciesMod1(2019)$anov
+```
+
+For each of the four measurement periods, there is no significant difference between models that include DBH:species interactions and just use DBH alone.
+Therefore, we can develop a single height model for each measurement period.
+
+***
+
+### DBH:Height equation for pooled species data
 
@@
 
-paste("Predicted.Height = ",coef(mod.dbh)[2],
-      " x ln(DBH) ",
-      coef(mod.dbh)[1])
+rbind(speciesMod1(1992)$out,
+      speciesMod1(1994)$out,
+      speciesMod1(1997)$out,
+      speciesMod1(2019)$out) %>% 
+  as_tibble()
+
 ```
@@
 HT = 11.79 x ln(DBH)  - 19.637
-
-The coefficients using the pooled species/years are pretty similar to Mike's formula for spruce.

### 2020-03-21 analysis/prelimAnalysis.Rmd
more updates

@@
 knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
-                      fig.width=8)
+                      fig.width=10)
 
@@
   mutate(`1997-2009`=`2009`-`1997`) %>% 
-  mutate(`2009-2019`=`2019`-`2009`)
+  mutate(`2009-2019`=`2019`-`2009`) %>% 
+  mutate(`1994-2019`=`2019`-`1994`)
 
@@
 
-pbai.Plot(initYear="1994",baPeriod="2009-2019",data=pbai.2,x=5,y=3)
+pbai.Plot(initYear="2009",baPeriod="2009-2019",data=pbai.2,x=10,y=3)
 
@@
 ```
+
+1994-2019 period:
+
+```{r}
+
+pbai.Plot(initYear="1994",baPeriod="1994-2019",x=5,y=1)
+```
+
+Plot 3 is an outlier because it has a negative BAI over the 1994-2019 period.  With this outlier removed, the slope would likely be p<0.05.

### 2020-03-23 analysis/March23_updates.Rmd
March23 updates

---
title: "March 23 Updates to report"
author: "Hardy Griesbauer"
date: "23/03/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE,message = FALSE)

# libraries
library(tidyverse)

# Load dataset
load(here::here("data","ep1162_Data.RData"))
load(here::here("data","ep1162_plotData.RData"))
```

### Updates for technical report
Below is code and output for the technical report, as per Mike's email March 23,2020.

NB: this code is also posted in the github repo if you want to run it for your self.  Just clone the entire repo to your local drive and take it from there.

https://github.com/hgriesbauer/EP1162

NB2: Plots 16,17, 18 and 19 were omitted from the dataset for these analyses because they were not measured in 1994 or 1997.

```{r initial setup}
dat<-
  dat %>% 
  filter(!Plot %in% c(16,17,18,19)) # remove plots 16-19 from data for further analysis

# set up diameter classes
diamClass=seq(from=7.5,to=95,by=5)

```

This means that the following analyses are based on `r length(unique(dat$Plot))` plots.

***
From Mike's email:

**Figure 4a and 5 (graphs of BA over time, and BAI over time, by treatment)**

These graphs are great, and we can manage with them as is, if need be. The one tweak I would ask for if possible, is a modification to the X-axis so that the timescales are linear (i.e. - proportional scale to the time elapsed between remeasurements).

```{r Figure 4a}

dat %>% 
   left_join(plotDat,by="Plot") %>% # join with plot data
  filter(Species!="Ep") %>% # remove birch
  filter(Status=="Live") %>% 
  mutate(Year=as.numeric(as.character(Year))) %>% 
   mutate(DBH.class=cut(DBH,diamClass,ordered_result = TRUE,right=F)) %>% # assign DBH classes
  mutate(BA=(DBH/100/2)^2*pi) %>% # convert tree DBH to basal area
  group_by(BA.Target,Plot,Year) %>% 
  summarise(BA=sum(BA)*20) %>% # multiply by 20 for BA/ha
  ungroup() %>%

  # create mean and SD
  group_by(BA.Target,Year) %>% 
  summarise(BA.Mean=mean(BA),
            BA.se=sd(BA)/sqrt(length(BA))) %>% # calculate standard error of the mean

    # Add labeller column
  mutate(Trt.Unit=factor(paste("Trt:",BA.Target,sep=""))) %>%

  # # padd out the year column
  # right_join(data.frame(Year=as.factor(1992:2019)),by="Year") %>%

  # Plotting
  ggplot()+
  aes(x=Year,y=BA.Mean,color=Trt.Unit,group=Trt.Unit)+
  geom_point(size=3)+
  geom_line()+
  geom_errorbar(aes(ymin=BA.Mean-BA.se,ymax=BA.Mean+BA.se),width=0.1)+
  xlab("Measurement year")+
  ylab("Mean basal area per plot (m²/ha)")+
  scale_x_continuous(limits=c(1992,2019),breaks=c(1992,1994,1997,2009,2019),
                     minor_breaks = c(1992,1994,1997,2009,2019),
                     name="Measurement year") # stretch out x-axis

```

**Figure 5**

```{r Fig 5}
dat %>% 
  left_join(plotDat,by="Plot") %>% # join with plot data
   filter(Status=="Live") %>% # filter for live trees only
  mutate(BA=(DBH/2/100)^2*pi*20) %>% # convert DBH to basal area and multiply for per hectare
  dplyr::select(BA.Target,Plot,TreeID,Species,Year,BA) %>% 
  group_by(BA.Target,Plot,Year) %>% 
  summarise(BA.Sum=sum(BA,na.rm=T)) %>% 
  pivot_wider(names_from=Year,values_from=BA.Sum) %>% 
  mutate(`1992-1994`=`1994`-`1992`) %>% 
  mutate(`1994-1997`=`1997`-`1994`) %>% 
  mutate(`1997-2009`=`2009`-`1997`) %>% 
  mutate(`2009-2019`=`2019`-`2009`) %>%

  # pivot back
  dplyr::select(BA.Target,Plot,`1992-1994`:`2009-2019`) %>% 
  pivot_longer(cols=`1992-1994`:`2009-2019`,names_to="Period",values_to="PBAI") %>% 
  mutate(Period=factor(Period)) %>%

  # summarise PBAI by BA.Target
  group_by(BA.Target,Period) %>% 
  summarise(PBAI.mean=mean(PBAI,na.rm=T),
            PBAI.se=sd(PBAI,na.rm=T)/sqrt(length(PBAI))) %>%

    # Add labeller column
  mutate(Trt.Unit=factor(paste("Trt:",BA.Target,sep=""))) %>%

  # Add new column for plotting
  mutate(Meas.Year=str_split_fixed(Period,"-",n=2)[,2]) %>% # extract measurement year
  mutate(Meas.Year=as.numeric(Meas.Year)) %>%

  # Plotting
  ggplot()+
  aes(x=Meas.Year,y=PBAI.mean,fill=Trt.Unit,color=Trt.Unit,group=Trt.Unit)+
  geom_point(size=3)+
  geom_line()+
  geom_errorbar(aes(ymin=PBAI.mean-PBAI.se,ymax=PBAI.mean+PBAI.se),width=0.1)+
  xlab("Period")+
  ylab("Mean BAI per plot (m²/ha/period)")+
  scale_x_continuous(limits=c(1994,2019),breaks=c(1994,1997,2009,2019),
                     labels = c("1992-1994","1994-1997","1997-2009","2009-2019"),
                     name="Period",
                     minor_breaks = c(1994,1997,2009,2019))+ # stretch out x-axis
  theme(axis.text.x = element_text(angle = 50, hjust = 1))

```

**Figure 4b (NEW) - graph of live BA over time for each of the PSP's**

Could you please generate a graph of live BA over time for each of the PSP's, so their collective trajectories and similarities / differences can be compared by the reader? The behaviour of the BA of individual plots varies somewhat from the treatment means, esp in the first few years, when some plots experience mortality, and others not so much.

```{r Figure 4b graph live BA over time}

dat %>% 
   left_join(plotDat,by="Plot") %>% # join with plot data
  filter(Status=="Live") %>% 
  mutate(Plot=paste("Plot:",Plot,sep="")) %>% 
  mutate(Year=as.numeric(as.character(Year))) %>% 
   mutate(DBH.class=cut(DBH,diamClass,ordered_result = TRUE,right=F)) %>% # assign DBH classes
  mutate(BA=(DBH/100/2)^2*pi) %>% # convert tree DBH to basal area
 mutate(Trt.Unit=factor(paste("Trt:",BA.Target,sep=""))) %>% 
  group_by(Trt.Unit,Plot,Year) %>% 
  summarise(BA=sum(BA)*20) %>% # multiply by 20 for BA/ha

  # Plotting
  ggplot()+
  aes(x=Year,y=BA,color=Trt.Unit,group=Trt.Unit)+
  geom_point(size=3)+
  geom_line()+
  ylab("Mean basal area per plot (m²/ha)")+
  scale_x_continuous(limits=c(1992,2019),breaks=c(1992,1994,1997,2009,2019),
                     minor_breaks = c(1992,1994,1997,2009,2019),
                     name="Measurement year")+ # stretch out x-axis
  facet_wrap(Trt.Unit~Plot,ncol=5)+
  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5),
        legend.position = "bottom")

```

**Table 4 (NEW)  - table(s) comparing Bl, Sx, and Fd basal area by treatment, for the years 1994 and 2019**

I would like to reference this data, to highlight quantitative changes in Bl / Sx abundance over time, by treatment, for two reference dates.

We discuss this qualitatively in reference to the graphs, but it would be good to cite specific quantitative data too.

```{r Table 4}

table4<-
  dat %>% 
   left_join(plotDat,by="Plot") %>% # join with plot data
  filter(Status=="Live") %>% 
  filter(Species!="Ep") %>% # remove birch from dataset
  mutate(Year=as.numeric(as.character(Year))) %>% 
  filter(Year %in% c(1994,2019)) %>% 
    mutate(BA=(DBH/100/2)^2*pi) %>% # convert tree DBH to basal area
 mutate(Trt.Unit=factor(paste("Trt:",BA.Target,sep=""))) %>% 
  group_by(Trt.Unit,Species,Year) %>% 
  summarise(BA=mean(BA,na.rm=TRUE)*20) %>% 
  mutate_at(vars("BA"),round,2) %>% 
  pivot_wider(names_from="Year",values_from="BA")

# Print to screen
  knitr::kable(table4,format="markdown")

# write to .csv
  write.csv(table4,file="table4.csv")

```

Note: this code also writes Table 4 to a .csv, so you can adjust formatting to fit the paper.

***
**Figure 8 (QUESTIONS)  - changes over time in trees species comp and distribution by BA**

The Trt20 basal areas for 2019 seem higher than they do in Figure 4a. In Figure 4a, Trt10 and Trt20 BA values converge and are almost identical.  Are we comparing apples to apples (i.e. live basal area) for both Figures 4a and 8?

*From Hardy*
Good catch!  There was a mathematical issue with how the mean BA per plot per species was being calculated.  I've fixed it in the code below.

```{r}

dat %>% 
    filter(Species!="Ep") %>% # remove birch 
    filter(Status=="Live") %>% # filter for live only trees
   left_join(plotDat,by="Plot") %>% # join with plot data
   mutate(DBH.class=cut(DBH,diamClass,ordered_result = TRUE,right=F)) %>% # assign DBH classes
  mutate(BA=(DBH/100/2)^2*pi) %>% # convert tree DBH to basal area
  group_by(BA.Target,Plot,Year,Species) %>% 
  summarise(BA=sum(BA)*20) %>% # multiply by 20 for BA/ha
  ungroup() %>%

  # have to assign plots missing species with BA=0
  pivot_wider(names_from="Species",values_from="BA") %>% 
  mutate_at(vars(c("Sx","Fd")),replace_na,0) %>% 
  pivot_longer(cols=Bl:Fd,names_to="Species",values_to="BA") %>%

  # create mean and SD
  group_by(BA.Target,Year,Species) %>% 
  summarise(BA.Mean=mean(BA),
            BA.se=sd(BA)/sqrt(length(BA))) %>% # calculate standard error of the mean

    # Add labeller column
  mutate(label=factor(paste("Trt:",BA.Target,sep=""))) %>%

  # # pad out the year column
  # right_join(data.frame(Year=as.factor(1992:2019)),by="Year") %>%

  # Plotting
  ggplot()+
  aes(x=Year,y=BA.Mean,fill=Species)+
  geom_bar(stat="Identity")+
  facet_wrap(~label)+
  ylab("Mean basal area per plot (m2/ha)")+
  xlab("Measurement year")

```

**NB**
However, this figure (and subsequent digging) has revealed another issue with the data - some plots were not measured in 1992, which that mean basal area contribution by species will vary between 1992 and 1994.  Note the decrease in Fd basal area per plot between 1992 and 1994.  This is due to different number of plots between 1992 and 1994, not because of Fd mortality (as this figure might imply).  I see a couple of options:

1. Explain this in the report; or
2. Limit analysis for these types of figures to 1994-2019 period.

Might be other options too.

***
**Which plots have live Fd stems in them, and how many stems are they?**

I am guessing that the significant Fd basal area in Figure 8 is due to few stems.

```{r extract all Douglas-fir trees in dataset}
dat %>% filter(Species=="Fd")

```

There is only one Fd tree in the entire dataset.

**Figure 7 (QUESTION)  - total sph by dbh class - diameter distribution over time**

You mentioned that we see a drop in sph of the smaller diameter classes between 1992 and 1994, and we attribute this to mortality processes. This is vaguely apparent in the graphs, but at a small scale. Could you please generate a table of sph numbers by dbh class for 1992 and 1994 for each treatment, that would allow us to address our comments more quantitatively and less qualitatively?

```{r}

sphCount<-
dat %>% 
    filter(Species!="Ep") %>% # remove birch 
  filter(Year %in% c(1992,1994)) %>% 
    filter(Status=="Live") %>% # filter for live only trees
   left_join(plotDat,by="Plot") %>% # join with plot data
   mutate(DBH.class=cut(DBH,diamClass,ordered_result = TRUE,right=F)) %>% # assign DBH classes
  group_by(BA.Target,Plot,Year,Species,DBH.class) %>% 
  summarise(Num.Trees=n()) %>% 
  ungroup() %>% 
  group_by(BA.Target,Year,Species,DBH.class) %>% 
    summarise(Num.Trees=mean(Num.Trees)) %>% 
  pivot_wider(names_from="Year",values_from = "Num.Trees") %>% 
arrange(BA.Target,Species,DBH.class)

# Print table to screen
knitr::kable(sphCount,digits=2)

# write to csv
write.csv(sphCount,file="tableForFigure7.csv")

```

*This plot shows the mean number of stems per plot per Treatment Unit/Species/DBH Class/Year*

Would this sort of table work?  I think a lot of the differences between 1992 and 1994 might be in the difference in plots measured between 1992 and 1994.

Note the funny formatting in the DBH class actually has a purpose: the '[' denotes the lower value that is contained within the DBH class, and the ')' denotes the upper value that forms the lower bound of the next higher class.  In other words, in the [7.5-12.5) class, a tree that measured 12.5 would be included in the next higher class (i.e., [12.5-17.5)).

### 2020-03-25 analysis/March23_updates.Rmd
some fixes for the technical report

@@
 
-```{r Figure 4b graph live BA over time}
+```{r Figure 4b graph live BA over time with each PSP separate window}
 
@@
   geom_line()+
-  ylab("Mean basal area per plot (m²/ha)")+
+  ylab("Total basal area per hectare (m²/ha)")+
   scale_x_continuous(limits=c(1992,2019),breaks=c(1992,1994,1997,2009,2019),
@@
   facet_wrap(Trt.Unit~Plot,ncol=5)+
+  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5),
+        legend.position = "bottom")
+
+```
+
+**Updated Figure 4b - All PSPs on one graph**
+
+```{r Figure 4b graph live BA over time with each PSP on same graph}
+
+dat %>% 
+   left_join(plotDat,by="Plot") %>% # join with plot data
+  filter(Status=="Live") %>% 
+  mutate(Plot=paste("Plot:",Plot,sep="")) %>% 
+  mutate(Year=as.numeric(as.character(Year))) %>% 
+   mutate(DBH.class=cut(DBH,diamClass,ordered_result = TRUE,right=F)) %>% # assign DBH classes
+  mutate(BA=(DBH/100/2)^2*pi) %>% # convert tree DBH to basal area
+ mutate(Trt.Unit=factor(paste("Trt:",BA.Target,sep=""))) %>% 
+  group_by(Trt.Unit,Plot,Year) %>% 
+  summarise(BA=sum(BA)*20) %>% # multiply by 20 for BA/ha
+
+  # Plotting
+  ggplot()+
+  aes(x=Year,y=BA,color=Trt.Unit,group=Plot)+
+  geom_point(size=3)+
+  geom_line()+
+  ylab("Total basal area per hectare (m²/ha)")+
+  scale_x_continuous(limits=c(1992,2019),breaks=c(1992,1994,1997,2009,2019),
+                     minor_breaks = c(1992,1994,1997,2009,2019),
+                     name="Measurement year")+ # stretch out x-axis
   theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5),
@@
 
+# Create a matrix with number of plots (for calculating average BA)
+table4.Numplots<-
+    dat %>% 
+   left_join(plotDat,by="Plot") %>% # join with plot data
+  filter(Status=="Live") %>% 
+  filter(Species!="Ep") %>% # remove birch from dataset
+  mutate(Year=as.numeric(as.character(Year))) %>% 
+  filter(Year %in% c(1994,2019)) %>% 
+  mutate(Trt.Unit=factor(paste("Trt:",BA.Target,sep=""))) %>% 
+  group_by(Trt.Unit,Year,Plot) %>% 
+  summarise(BA=mean(DBH)) %>% 
+  group_by(Trt.Unit,Year) %>% 
+  summarise(num.plots=n())
+
 table4<-
@@
   group_by(Trt.Unit,Species,Year) %>% 
-  summarise(BA=mean(BA,na.rm=TRUE)*20) %>% 
-  mutate_at(vars("BA"),round,2) %>% 
-  pivot_wider(names_from="Year",values_from="BA")
+  summarise(BA=sum(BA,na.rm=TRUE)) %>% # summarise BA for each treatment unit/year combination
+  mutate(BA=BA*20) %>% # multiply by 20 to get per hectare
+
+  # join with number of plots
+  left_join(table4.Numplots,by=c("Trt.Unit","Year")) %>% 
+  mutate(BA.ha=BA/num.plots) %>% # divide by number of plots to get mean BA by plot
+
+  # remove columns for reshaping
+  dplyr::select(-BA,-num.plots) %>%
+
+  # pivot wider 
+  pivot_wider(names_from="Year",values_from="BA.ha")
 
 # Print to screen
-  knitr::kable(table4,format="markdown")
+  knitr::kable(table4,format="markdown",digits=2)
 
@@
 ```
+
+Table shows mean BA per plot, multiplied by 20 to get BA/ha.
 
@@
   facet_wrap(~label)+
-  ylab("Mean basal area per plot (m2/ha)")+
+  ylab("Mean basal area per plot (m²/ha)")+
   xlab("Measurement year")
@@
 **NB**
-However, this figure (and subsequent digging) has revealed another issue with the data - some plots were not measured in 1992, which that mean basal area contribution by species will vary between 1992 and 1994.
 Note the decrease in Fd basal area per plot between 1992 and 1994.
-This is due to different number of plots between 1992 and 1994, not because of Fd mortality (as this figure might imply).
+This is due to how I calculated these graphs (mean BA per plot), and the different number of plots between 1992 and 1994.
+This is not because of Fd mortality (as this figure might imply).
 I see a couple of options:
@@
 1. Explain this in the report; or
-2. Limit analysis for these types of figures to 1994-2019 period.
+2. Limit analysis for these types of figures to 1994-2019 period; or
+3. Come up with a different method to calculate BA by period (open to suggestions!).
 
@@
 
-There is only one Fd tree in the entire dataset.
+There is only one (albeit large) Fd tree in the entire dataset (note that there may be Fd in plots 16-19, which are not included in this analysis).
 
@@
   group_by(BA.Target,Year,Species,DBH.class) %>% 
-    summarise(Num.Trees=mean(Num.Trees)) %>% 
+    summarise(Num.Trees=mean(Num.Trees)*20) %>% # multiply by 20 to get per hectare
   pivot_wider(names_from="Year",values_from = "Num.Trees") %>% 
@@
 
-*This plot shows the mean number of stems per plot per Treatment Unit/Species/DBH Class/Year*
+*This plot shows the mean number of stems per hectare plot per Treatment Unit/Species/DBH Class/Year*
 

### 2020-03-27 analysis/heightModel.Rmd
volume estimates

@@
 ---
-title: "Height estimates from DBH:Height relationship"
+title: "DBH:Height models for EP1162"
 author: "Hardy Griesbauer"
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = TRUE,warning = FALSE)
 
@@
 # Load data
-load(here("data","ep1162_Data.RData"))
+load(here::here("data","ep1162_Data.RData"))
 
@@
 1. Model DBH:Height relationships separately for each measurement period, and for Bl:Sx separately (at first).
-2. Compare Bl and Sx relationships, compare estimates between years, and determine if we can pool species and/or years to estimate Height.
+2. Compare Bl and Sx relationships, and determine if we can pool species to estimate Height.
+
+We will generate different dbh:height models for each measurement year, as pooling years would violate assumptions of independence.
+
+### Data preparation: remove trees with broken tops
+Trees with broken or dead tops that have height measurements should not be included in DBH:height models.  We can summarise and remove these:
+
+```{r}
+dat %>% 
+  filter(!is.na(DBT)) %>% # filter for trees with recorded broken top
+  filter(!is.na(Height)) %>% # filter for trees with height measurements
+  group_by(Species,Year) %>% # summarise trees with broken tops at the species and year level
+  summarise(Num.DBT=n()) %>% 
+  pivot_wider(names_from="Species",values_from="Num.DBT")
+
+```
 
@@
     filter(Species %in% c("Bl","Sx")) %>% # filter for Bl and Sx for now
-    filter(Status=="Live") %>% 
+    filter(Status=="Live") %>% # filter for live trees
+    filter(is.na(DBT)) %>% # filter for trees with no broken tops
     drop_na(Height) %>% # remove trees where height was not measured
@@
     filter(Status=="Live") %>% 
+   filter(is.na(DBT)) %>% # filter for trees with no broken tops
     drop_na(Height) %>% # remove trees where height was not measured
@@
     filter(Status=="Live") %>% 
+   filter(is.na(DBT)) %>% # filter for trees with no broken tops
   filter(Year==year) %>% 
@@
 
+heightModels<- # assign 
   rbind(speciesMod1(1992)$out,
@@
 
+print(heightModels)
+
 ```
 
-Here are the coefficients from Mike's analysis:
+#### Height model for 2009 measurement period
+Height was not recorded during the 2009 measurements, however, we should be able to estimate heights using the 2009 DBH data.  I propose the following model to estimate 2009 heights:
 
-*Subalpine fir*
-HT = 11.188 x ln(DBH)  - 17.609
+### Graph out intercept and slope for four measurement period
+I noticed that the intercept and slope vary monotonically with years, which suggests that we might be able to predict intercept/slope for a given year (in this case 2009).
+Let's examine this with a scatterplot:
 
-*Hybrid white spruce*
-HT = 11.79 x ln(DBH)  - 19.637
+```{r}
+
+heightModels %>% 
+  dplyr::select(1:3) %>% #select first three columns
+  pivot_longer(-year,values_to="coefficient",names_to="component") %>% # pivot longer for facet plot
+  ggplot()+
+  aes(x=year,y=coefficient)+
+  geom_point()+
+  geom_line(color="red")+
+  geom_vline(xintercept=2009,linetype="dashed")+
+  facet_wrap(~component,nrow=2,scales="free")
+
+```
+
+Based on these relationships, it seems we can reasonably estimate the slope and intercept for a dbh:height model at year 2009.  Thoughts?
+
+### Calculate 2009 intercept and slope coefficients for dbh:height model
+
+```{r}
+
+# Calculate slope to estimate intercept:
+int.slope<-
+  with(heightModels,heightModels[year==2019,"intercept"]-heightModels[year==1997,"intercept"])/(2019-1997)
+  int.slope$run=diff(c(1997,2009))
+  int.2009<-  as.numeric(with(int.slope,run*intercept)+with(heightModels,heightModels[year==1997,"intercept"]))
+
+####
+slope<-
+  with(heightModels,heightModels[year==2019,"slope"]-heightModels[year==1997,"slope"])/(2019-1997)
+  slope$run=diff(c(1997,2009))
+  slope.2009<-  as.numeric(with(slope,run*slope)+with(heightModels,heightModels[year==1997,"slope"]))
+
+### Add these values into heightModels table
+  heightModels<-
+    heightModels %>% 
+    add_row(year=2009,intercept=int.2009,slope=slope.2009) %>% # add 2009 coefficients to table
+    arrange(year) # order by year
+
+# print to screen
+  heightModels
+
+```
+
+### Save file for volume estimates
+```{r}
+# save coefficients for volume estimates
+save(heightModels,file=here("data","heightModels.RData"))
+```

### 2020-03-27 analysis/volumeEstimates.Rmd
volume estimates

---
title: "EP1162 volume estimates"
author: "Hardy Griesbauer"
date: "26/03/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE)

# Libraries
library(tidyverse)

# load data
# Load data

# Measurements
load(here::here("data","ep1162_Data.RData"))
load(here::here("data","ep1162_plotData.RData"))

# Height model coefficients
load(here::here("data","heightModels.RData"))  # load heightModels data frame

# initial setup
dat<-
  dat %>% 
  filter(!Plot %in% c(16,17,18,19)) # remove plots 16-19 from data for further analysis

```

## Volume estimates for EP1162

We will estimate volume estimates using the following approach:

1. Estimate heights for trees without height measurements, using DBH:height models developed earlier using EP1162 data
2. Estimate volume using equations from Nigh (2016) for specific species and BGC units

Source for Nigh (2016): https://www.for.gov.bc.ca/hfd/pubs/docs/TR/TR106.htm

**NB Plots 16-19 omitted from this analysis**

#### Estimate Heights
From earlier analyses, we calculated intercept and slope coefficients to estimate spruce and subalpine fir heights using DBH.  The formula is: Height = b1*ln(DBH)+b0

Coefficients were developed separately for each measurement period.  Analysis showed that the regression did not vary significantly between spruce and subalpine fir, therefore, we developed the coefficients by pooling data for both species within a measurement year.

```{r show height models,}
 hm<-
  heightModels %>% 
  dplyr::rename(b0=intercept,b1=slope) %>%   # rename columns
  mutate(year=factor(year))

# Print to screen 
hm

```
Table shows coefficients for height models for each measurement period.  P value and R^2^ values are from linear regression.

Note missing p values and r square for 2009 year - these coefficients were estimated from 1997 and 2019 data.

***
#### Nigh equations
Hardcode Nigh's 2016 total (Table 9) and merchantable (Table 11) volume equations for species and SBS BGC Zone into a table:

```{r}
volCoef<-

  ## TOTAL VOLUME
  #  Subalpine fir 
  data.frame(Species="Bl",VolType="total",b0=-9.79,b1=1.813,b2=1.033) %>%

  # Douglas-fir
  add_row(Species="Fd",VolType="total",b0=-10.19,b1=1.73,b2=1.206) %>%

  # Add spruce
  add_row(Species="Sx",VolType="total",b0=-10.024,b1= 1.801,b2= 1.107) %>%

  ## MERCHANTABLE VOLUME
  # Subalpine fir
  add_row(Species="Bl",VolType="merch",b0=-9.980,b1= 1.792,b2= 1.102) %>%

  # Douglas-fir
  add_row(Species="Fd",VolType="merch",b0=-10.503,b1= 1.724,b2= 1.296) %>%

  # Spruce
  add_row(Species="Sx",VolType="merch",b0=-10.280,b1= 1.769,b2= 1.206 ) %>%

  mutate(BGC.Zone="SBS") %>%

  # convert species to factor
  mutate(Species=factor(Species)) %>% 
  dplyr::select(Species,BGC.Zone,everything())

# print to screen
volCoef

```

This table shows the coefficients used in the volume equations.

Equation: vol = e^b~0~^ x dbh^b~1~^ x height^b~2~^

#### Estimate heights and volumes in dataset
We can now apply the coefficients above to estimate heights for trees that do not have height measurements:

```{r estimate heights and volumes, warning=FALSE}
dat<- # we will update dat dataset
  dat %>%

  # Height estimates
  left_join(hm[,1:3],by=c("Year"="year")) %>% # add coefficients to table
  mutate(heightVol=coalesce(Height,b0+b1*log(DBH))) %>%   # populate new height column height estimates
  dplyr::select(-b0,-b1) %>%  # remove coefficients

  # Total volume estimates
  left_join(filter(volCoef,VolType=="total"),by=c("Species")) %>% #add total volume coefficients to table
  mutate(totalVol=exp(1)^b0*DBH^b1*heightVol^b2) %>% # estimate volume
  dplyr::select(Plot:heightVol,totalVol) %>% # remove coefficients and other extraneous columns

  # Merchantable volume estimates
    left_join(filter(volCoef,VolType=="merch"),by=c("Species")) %>% #add merch volume coefficients to table
  mutate(merchVol=exp(1)^b0*DBH^b1*heightVol^b2) %>% # estimate volume
  dplyr::select(Plot:totalVol,merchVol)

# save output
save(dat,file=here::here("data","ep1162_Data_Volumes.RData"))

# write to .csv
write.csv(dat,file=here::here("data","ep1162_Data_Volumes.csv"))

```

**As per Mike's suggestion, I double checked that the estimates are reasonable by comparing estimates to the volume formula for a cone (Height*BA/3) - estimates appear correct**

## Preliminary volume analysis
Produce some preliminar graphs based on total volume estimates:

### Histogram of total volume estimates
```{r histogram of volume values, message=F}
dat %>% 
  left_join(plotDat,by="Plot") %>% # join with plot data
  filter(Status=="Live") %>% # filter for live trees only
  filter(Species %in% c("Bl","Sx")) %>% # only for Sx and Bl

  ggplot()+
  aes(x=totalVol,color=Species)+
  geom_histogram(fill="white", alpha=0.5, position="identity")+
  xlab("Volume per tree")

```

Data are right skewed, as expected in an uneven-aged stand.

### Mean volume per plot per year and treatment

```{r}
dat %>% 
  left_join(plotDat,by="Plot") %>% # join with plot data
  filter(Status=="Live") %>% # filter for live trees only

  # plot multiplier
  mutate(merchVol=merchVol*20) %>% 
  mutate(totalVol=totalVol*20) %>%

  # Summarise mean plot volume at the treatment unit and year levels
  group_by(BA.Target,Plot,Year) %>% 
  summarise(merchVol=sum(merchVol,na.rm=TRUE),
            totalVol=sum(totalVol,na.rm=TRUE)) %>% 
  ungroup() %>% 
    group_by(BA.Target,Year) %>% 
    summarise(merch.Vol=mean(merchVol,na.rm=TRUE),
            total.Vol=mean(totalVol,na.rm=TRUE),
            total.se=sd(totalVol)/sqrt(length(totalVol)),
            merch.se=sd(merchVol)/sqrt(length(merchVol)))  %>%

    # Add labeller column
  mutate(Trt.Unit=factor(paste("Trt:",BA.Target,sep=""))) %>%

  # Need to transform year format
    mutate(Year=as.numeric(as.character(Year))) %>%

ggplot()+
  aes(x=Year,y=total.Vol,color=Trt.Unit,group=Trt.Unit)+  
  geom_point(size=3)+
  geom_line()+
  geom_errorbar(aes(ymin=total.Vol-total.se,ymax=total.Vol+total.se),width=0.1)+
  xlab("Measurement year")+
  ylab("Mean total volume per plot (m³/ha)")+
  theme(legend.position = "bottom")+
    scale_x_continuous(limits=c(1992,2019),breaks=c(1992,1994,1997,2009,2019),
                     minor_breaks = c(1992,1994,1997,2009,2019),
                     name="Measurement year")

```

### Volume for all plots 
Plot out volume by year for all plots:

```{r plot volume by year,message=FALSE,warning=FALSE}

dat %>% 
   left_join(plotDat,by="Plot") %>% # join with plot data
  filter(Status=="Live") %>% 
  mutate(Plot=paste("Plot:",Plot,sep="")) %>% 
  mutate(Year=as.numeric(as.character(Year))) %>% 
  mutate(Trt.Unit=factor(paste("Trt:",BA.Target,sep=""))) %>% 
  group_by(Trt.Unit,Plot,Year) %>% 
  summarise(vol=sum(totalVol)*20) %>% # multiply by 20 for volume/ha

  # Plotting
  ggplot()+
  aes(x=Year,y=vol,color=Trt.Unit,group=Plot)+
  geom_point(size=3)+
  geom_line()+
  ylab("Plot volume per hectare (m³/ha)")+
  scale_x_continuous(limits=c(1992,2019),breaks=c(1992,1994,1997,2009,2019),
                     minor_breaks = c(1992,1994,1997,2009,2019),
                     name="Measurement year")+ # stretch out x-axis
  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5),
        legend.position = "bottom")
```

### Periodic volume increment
```{r periodic volume increment,message=FALSE}
dat %>% 
  left_join(plotDat,by="Plot") %>% # join with plot data
   filter(Status=="Live") %>% # filter for live trees only
  dplyr::select(BA.Target,Plot,TreeID,Species,Year,totalVol) %>% 
  group_by(BA.Target,Plot,Year) %>% 
  summarise(vol.Sum=sum(totalVol,na.rm=T)*20) %>% # plot multiplier
  pivot_wider(names_from=Year,values_from=vol.Sum) %>% 
  mutate(`1992-1994`=`1994`-`1992`) %>% 
  mutate(`1994-1997`=`1997`-`1994`) %>% 
  mutate(`1997-2009`=`2009`-`1997`) %>% 
  mutate(`2009-2019`=`2019`-`2009`) %>%

  # pivot back
  dplyr::select(BA.Target,Plot,`1992-1994`:`2009-2019`) %>% 
  pivot_longer(cols=`1992-1994`:`2009-2019`,names_to="Period",values_to="PVI") %>% 
  mutate(Period=factor(Period)) %>%

  # summarise PBAI by BA.Target
  group_by(BA.Target,Period) %>% 
  summarise(PVI.mean=mean(PVI,na.rm=T),
            PVI.se=sd(PVI,na.rm=T)/sqrt(length(PVI))) %>%

    # Add labeller column
  mutate(Trt.Unit=factor(paste("Trt:",BA.Target,sep=""))) %>%

  # Add new column for plotting
  mutate(Meas.Year=str_split_fixed(Period,"-",n=2)[,2]) %>% # extract measurement year
  mutate(Meas.Year=as.numeric(Meas.Year)) %>%

  # Plotting
  ggplot()+
  aes(x=Meas.Year,y=PVI.mean,fill=Trt.Unit,color=Trt.Unit,group=Trt.Unit)+
  geom_point(size=3)+
  geom_line()+
  geom_errorbar(aes(ymin=PVI.mean-PVI.se,ymax=PVI.mean+PVI.se),width=0.1)+
  xlab("Period")+
  ylab("Mean VI per plot (m³/ha/period)")+
  scale_x_continuous(limits=c(1994,2019),breaks=c(1994,1997,2009,2019),
                     labels = c("1992-1994","1994-1997","1997-2009","2009-2019"),
                     name="Period",
                     minor_breaks = c(1994,1997,2009,2019))+ # stretch out x-axis
  theme(axis.text.x = element_text(angle = 50, hjust = 1))

```

### 2020-04-22 analysis/treeChanges.Rmd
tree level changes

---
title: "Tree Level Changes"
author: "Hardy Griesbauer"
date: "21/04/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load libraries
library(tidyverse)
library(here)
library(lme4)
library(lmerTest)
library(broom)
library(MuMIn)
library(ggeffects)
library(car)

# Load data
load(here::here("data","ep1162_Data.RData"))
load(here::here("data","ep1162_plotData.RData"))

```
Below is code and output looking at tree-level changes over time

Also, we can include plots 16-19 even though they weren't measured across all periods, because we are looking at tree-level responses.

### PBAI at the tree level
We can look at PBAi at the tree-level, by tree class.  This will tell us where PBAI is concentrated.  Hypothesis is that smaller trees respond with higher PBAI and responses are stronger with in plotss with higher basal area removal.

```{r treeList and other starting variables}

# Let's filter only for trees that have five measurements 
treeList<- # assign variable
  dat %>% 
  filter(TreeID!="New") %>% #remove new trees
  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
  filter(Species %in% c("Sx","Bl")) %>% # filter birch out of dataset
  filter(Status=="Live") %>% # filter for live trees
  group_by(TreeID) %>% 
  summarise(N=n()) %>% 
  filter(N==5) %>% #filter for trees with 5 measurements
  pull(TreeID)

plotBA<-
  dat %>% 
  filter(Status=="Live") %>% # filter for live trees
   mutate(BA=(DBH/2/100)^2*pi) %>% # convert DBH to basal area 
  group_by(Plot,Year) %>% # grouping
  summarise(plotBA=sum(BA))

# derive PBAI
pbaiDat<-
  dat %>% 
  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
  filter(TreeID %in% treeList) %>% 
  mutate(BA=(DBH/2/100)^2*pi) %>% # convert DBH to basal area
  dplyr::select(Plot,TreeID,Species,Year,BA) %>% 
  pivot_wider(names_from=Year,values_from=BA) %>% 
  mutate(PBAI=`2019`-`1992`) %>% 
  dplyr::select(Plot,TreeID,Species,PBAI,baInit=`1992`)

```

LMER to predict basal area slope per tree after treatment
```{r}
x<- # assign output 
  pbaiDat %>% 
  left_join(plotDat,by="Plot")

```

```{r lmer by PBAI}
fit1<-
  x %>% 
  lmer(log(PBAI)~BA.Target+Species+log(baInit)+(1|Plot),data=.)

Anova(fit1)

plot(fit1)

glance(fit1)

```

```{r plot marginal means}

mydf<-  ggpredict(fit1,terms=c("baInit [exp]","Species","BA.Target"))

ggplot(mydf, aes(x, predicted, colour = group)) + 
  geom_line() +
  facet_wrap(~facet)+
  ylab("Predicted BAI (m2) per tree from 1992 to 2019")+
  xlab("Tree basal area (m2) in 1992")

ggsave("predictedPBAI.png")
```

```{r r squared values from model}

r.squaredGLMM(fit1)

```

## Let's look at basal area composition by tree size per treatment per year
```{r ba by tree size, year and treatment}

# set up diameter classes
diamClass=seq(from=7.5,to=95,by=5)

dat %>% 
   filter(!Plot %in% c(16,17,18,19)) %>%  # remove plots 16-19 from data for further analysis
  left_join(plotDat,by="Plot") %>% # join with plot data
   filter(Status=="Live") %>% # filter for live trees only
   mutate(DBH.class=cut(DBH,diamClass,ordered_result = TRUE,right=F)) %>% # assign DBH classes
  mutate(BA=(DBH/2/100)^2*pi) %>% # convert DBH to basal area
  group_by(BA.Target,Plot,Year,DBH.class) %>% 
  summarise(BA.Plot=sum(BA)*20) %>%   # sum basal area per diam class and multiply to get per ha

  # create means
  ungroup() %>% 
  group_by(BA.Target,Year,DBH.class) %>%
  summarise(BA=mean(BA.Plot)) %>%

   # Add labeller column
  mutate(label=factor(paste("Trt:",BA.Target,sep=""))) %>%

  # Plotting
  ggplot()+
  aes(x=DBH.class,y=BA,fill=label)+
  geom_bar(stat="Identity")+
  facet_grid(Year~label)+
  theme(axis.text.x = element_text(angle = 90, hjust = 1,size=6,vjust=0.25))+
  ylab("Mean basal area per hectare per plot (m2)")+
  xlab("Diameter classes (cm)")

ggsave(filename="baDiamClass.png")

```

### 2020-04-30 manuscript/EP1162_manuscript_draft.Rmd
starting on manuscript

--- 2020-04-22 analysis/treeChanges.Rmd
@@
 ---
-title: "Tree Level Changes"
-author: "Hardy Griesbauer"
-date: "21/04/2020"
-output: html_document
+title: Analyses of 27 years of stand structure development and basal area growth response to a range of initial basal area density levels, 1992-2019 
+author: "Mike Jull and Hardy Griesbauer"
+date: "29/03/2020"
+output:
+  word_document:
+    reference_docx: wordTemplate2.docx
+  html_document:
+    df_print: paged
+csl: cjfr.csl
+bibliography: ep1162Bib.bib
 ---
@@
 ```
-Below is code and output looking at tree-level changes over time
-
-Also, we can include plots 16-19 even though they weren't measured across all periods, because we are looking at tree-level responses.
-
-### PBAI at the tree level
-We can look at PBAi at the tree-level, by tree class.
-This will tell us where PBAI is concentrated.
-Hypothesis is that smaller trees respond with higher PBAI and responses are stronger with in plotss with higher basal area removal.
-
-```{r treeList and other starting variables}
-
-# Let's filter only for trees that have five measurements 
-treeList<- # assign variable
-  dat %>% 
-  filter(TreeID!="New") %>% #remove new trees
-  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
-  filter(Species %in% c("Sx","Bl")) %>% # filter birch out of dataset
-  filter(Status=="Live") %>% # filter for live trees
-  group_by(TreeID) %>% 
-  summarise(N=n()) %>% 
-  filter(N==5) %>% #filter for trees with 5 measurements
-  pull(TreeID)
-
-plotBA<-
-  dat %>% 
-  filter(Status=="Live") %>% # filter for live trees
-   mutate(BA=(DBH/2/100)^2*pi) %>% # convert DBH to basal area 
-  group_by(Plot,Year) %>% # grouping
-  summarise(plotBA=sum(BA))
-
-# derive PBAI
-pbaiDat<-
-  dat %>% 
-  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
-  filter(TreeID %in% treeList) %>% 
-  mutate(BA=(DBH/2/100)^2*pi) %>% # convert DBH to basal area
-  dplyr::select(Plot,TreeID,Species,Year,BA) %>% 
-  pivot_wider(names_from=Year,values_from=BA) %>% 
-  mutate(PBAI=`2019`-`1992`) %>% 
-  dplyr::select(Plot,TreeID,Species,PBAI,baInit=`1992`)
-
-```
-
-LMER to predict basal area slope per tree after treatment
-```{r}
-x<- # assign output 
-  pbaiDat %>% 
-  left_join(plotDat,by="Plot")
-
-```
-
-```{r lmer by PBAI}
-fit1<-
-  x %>% 
-  lmer(log(PBAI)~BA.Target+Species+log(baInit)+(1|Plot),data=.)
-
-Anova(fit1)
-
-plot(fit1)
-
-glance(fit1)
-
-```
-
-```{r plot marginal means}
-
-mydf<-  ggpredict(fit1,terms=c("baInit [exp]","Species","BA.Target"))
-
-ggplot(mydf, aes(x, predicted, colour = group)) + 
-  geom_line() +
-  facet_wrap(~facet)+
-  ylab("Predicted BAI (m2) per tree from 1992 to 2019")+
-  xlab("Tree basal area (m2) in 1992")
-
-ggsave("predictedPBAI.png")
-```
-
-```{r r squared values from model}
-
-r.squaredGLMM(fit1)
-
-```
-
-## Let's look at basal area composition by tree size per treatment per year
-```{r ba by tree size, year and treatment}
-
-# set up diameter classes
-diamClass=seq(from=7.5,to=95,by=5)
-
-dat %>% 
-   filter(!Plot %in% c(16,17,18,19)) %>%  # remove plots 16-19 from data for further analysis
-  left_join(plotDat,by="Plot") %>% # join with plot data
-   filter(Status=="Live") %>% # filter for live trees only
-   mutate(DBH.class=cut(DBH,diamClass,ordered_result = TRUE,right=F)) %>% # assign DBH classes
-  mutate(BA=(DBH/2/100)^2*pi) %>% # convert DBH to basal area
-  group_by(BA.Target,Plot,Year,DBH.class) %>% 
-  summarise(BA.Plot=sum(BA)*20) %>%   # sum basal area per diam class and multiply to get per ha
-
-  # create means
-  ungroup() %>% 
-  group_by(BA.Target,Year,DBH.class) %>%
-  summarise(BA=mean(BA.Plot)) %>%
-
-   # Add labeller column
-  mutate(label=factor(paste("Trt:",BA.Target,sep=""))) %>%
-
-  # Plotting
-  ggplot()+
-  aes(x=DBH.class,y=BA,fill=label)+
-  geom_bar(stat="Identity")+
-  facet_grid(Year~label)+
-  theme(axis.text.x = element_text(angle = 90, hjust = 1,size=6,vjust=0.25))+
-  ylab("Mean basal area per hectare per plot (m2)")+
-  xlab("Diameter classes (cm)")
-
-ggsave(filename="baDiamClass.png")
-
-```

### 2020-05-01 manuscript/EP1162_manuscript_draft.Rmd
Results and discussion work

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = FALSE,message=FALSE,warning=FALSE)
 
@@
 library(car)
+library(flextable)
+library(captioner)
 
@@
 
+# set up captions
+
+# Code to create captions for tables
+tableNums <- captioner(prefix = "Table")
+
+# Code to create captions for figures
+figNums <- captioner(prefix="Figure")
+
+# Code to create captions for appendices
+appNums <- captioner(prefix="Appendix")
+
 ```
+
+# Methods
+
+## Analysis
+
+### Tree-level growth responses
+To test the hypothesis that tree-level growth response over the 1992-2019 period varied with interactions between tree basal area at the time of treatment (1992) and (i) residual basal area; and (ii) species, we fit linear mixed effects models with log-transformed 1992-2019 basal area increment as the response variable, and treatment, species and log-transformed 1992 basal area as interacting fixed effects, with plot as a random intercept effect.
+We also fit a similar model with the 1992-2019 basal area increment expressed as a percentage of 1992 basal.
+Plots of residuals and fit lines with orginal data were examined to evaluate model goodness of fit.
+Variance explained for each model are reported using the approach described in Nakagaw et al.
+[-@Nakagawa2013], as implemented in the MuMIn package for R [@add later].
+
+We also examined how basal area increment at the tree-level varied over the four measurement periods, by fitting a similar model as described above, with an additional fixed effect of measurement period.
+This allowed us to examine potential lag responses in growth.
+
+## Stand structure and growth dynamics
+
+### Changes in tree density over time
+We qualitatively assessed changes in tree density over time through density diagrams that displayed tree density (stems per hectare) by diameter class, faceted by measurement period, species and treatment unit.
+The effects of tree density changes on basal area was also qualitatively evaluated by summarizing basal area per hectare by diameter class, species and treatment unit.
+
+### Stand-level basal area and volume growth
+
+# Results
+
+### Tree-level growth responses
+```{r tree-level growth responses}
+
+# Let's filter only for trees that have measurements in 1992 and 2019 
+treeList<- # assign variable
+  dat %>% 
+  filter(TreeID!="New") %>% #remove new trees
+  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
+  filter(Species %in% c("Sx","Bl")) %>% # filter birch and fdi out of dataset
+  filter(Status=="Live") %>% # filter for live trees
+  filter(Year %in% c(1992,2019)) %>% #filter for start and end dates
+  group_by(TreeID) %>% 
+  summarise(N=n()) %>% 
+  filter(N==2) %>% #filter for trees with 5 measurements
+  pull(TreeID)
+
+# derive periodic increment (period diameter or basal area increment)
+piDat<-
+  dat %>% 
+  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
+  filter(TreeID %in% treeList) %>% 
+  mutate(BA=(DBH/2/100)^2*pi) %>% # convert DBH to basal area
+  dplyr::select(Plot,TreeID,Species,Year,BA) %>% 
+  pivot_wider(names_from=Year,values_from=BA) %>% 
+  mutate(PI=`2019`-`1992`) %>% # 
+  filter(PI>0) %>% # one tree has a PI of 0, has to be removed
+  mutate(PIpct=PI/`1992`*100) %>% # express periodic increment a a percentage
+  dplyr::select(Plot,TreeID,Species,PI,PIpct,Init=`1992`) %>% 
+  left_join(plotDat,by="Plot") %>% 
+  mutate(BA.Target=paste("TU:",BA.Target,sep="")) # create facet column
+
+# Create figures of raw data to put into Appendix
+pi.Init.rawPlot<-  
+  piDat %>% 
+  ggplot(aes(x=log(Init),y=log(PI),color=Species))+
+  geom_point()+
+  geom_smooth(method="lm",se=FALSE,color="black")+
+  facet_grid(Species~BA.Target)  +
+   ylab("log(BA increase, m²)")+
+  xlab("log(Tree BA in 1992, m²)")+
+    theme_bw()+
+  theme(legend.position = "none")
+
+piPCT.Init.rawPlot<-  
+  piDat %>% 
+  ggplot(aes(x=log(Init),y=log(PIpct),color=Species))+
+  geom_point()+
+  geom_smooth(method="lm",se=FALSE,color="black")+
+  facet_grid(Species~BA.Target) +
+  ylab("log(% BA increase)")+
+  xlab("log(Tree BA in 1992, m²)")+
+  theme_bw()+
+  theme(legend.position="none")
+
+# fixed effects model with species pooled, using PBAI
+fitBA<-
+  piDat %>% 
+  lmer(log(PI)~BA.Target:log(Init):Species+(1|Plot),data=.)
+
+# fixed effects model with species pooled, using PBAIpct
+fitBApct<-
+  piDat %>% 
+  lmer(log(PIpct)~BA.Target:log(Init):Species+(1|Plot),data=.)
+
+# Create figure showing marginal means from model
+fitBA.plot<-
+  ggpredict(fitBA,terms=c("Init [exp]","Species","BA.Target")) %>% 
+  dplyr::rename(Species=group) %>% 
+  ggplot(aes(x,predicted,color=Species))+
+  geom_line()+
+  facet_wrap(~facet)+
+  ylab("Predicted 27-year basal area increment (m²)")+
+  xlab("Tree BA in 1992 (m²)")+
+  theme_bw()+
+  theme(legend.position="bottom")
+
+# Create figure showing marginal means from model
+fitBApct.plot<-
+  ggpredict(fitBApct,terms=c("Init [exp]","Species","BA.Target")) %>% 
+  dplyr::rename(Species=group) %>% 
+  ggplot(aes(x,predicted,color=Species))+
+  geom_line()+
+  facet_wrap(~facet)+
+  ylab("Predicted 27-year basal area % increase")+
+  xlab("Tree BA in 1992 (m²)")+
+  theme_bw()+
+  theme(legend.position="bottom")
+
+# Captions
+appNums(name="piInitRaw",display=FALSE)
+appNums(name="piPCTInitRaw",display=FALSE)
+figNums(name="fitBA.plot",display=FALSE)
+figNums(name="fitBApct.plot",display=FALSE)
+
+```
+
+A comparison of tree-level 1992-2019 basal area increment to tree size at the time of treatment showed that basal area increment over 27 years increased with tree size, and that this relationship varied between species and among treatment units (`r appNums("piInitRaw",display="cite")`).
+A linear mixed effects model using
+`r nrow(piDat)`
+ trees (
+`r nrow(filter(piDat,Species=="Sx"))`
+Spruce, 
+`r nrow(filter(piDat,Species=="Bl"))` Fir) confirmed that slopes and intercepts varied with species and residual basal area (p<0.001 for all interactions).  This model explained 
+`r round(r.squaredGLMM(fitBA)[1]*100,1)`
+% variation (marginal R^2^) in diameter increment over the 27 years of the study, and showed that diameter increment increased with initial tree size, and increment was higher in the treatment units with basal area removed, compared to the control, and that spruce trees had higher basal area increment than fir (`r figNums("fitDBH.plot",display="cite")`).
+A similar model predicting basal area increment as a percentage of tree size at the time of treatment showed that smaller trees grew proportionally more than larger trees, with a strong negative exponential decrease with initial tree size (`r figNums("fitDBHpct.plot",display="cite")`), however, the model fit was poorer, explaining
+`r round(r.squaredGLMM(fitBApct)[1]*100,1)`
+% of variation in % basal area increase.
+
+```{r tree level time lag response}
+treeList<- # assign variable
+  dat %>% 
+  filter(TreeID!="New") %>% #remove new trees
+  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
+  filter(Species %in% c("Sx","Bl")) %>% # filter birch and fdi out of dataset
+  filter(Status=="Live") %>% # filter for live trees
+  group_by(TreeID) %>% 
+  summarise(N=n()) %>% 
+  filter(N==5) %>% #filter for trees with 5 measurements
+  pull(TreeID)
+
+# derive mean annual increment by period
+
+piDat<-
+  dat %>% 
+  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
+  filter(TreeID %in% treeList) %>% 
+   left_join(plotDat,by="Plot") %>% 
+  mutate(BA=(DBH/2/100)^2*pi) %>% # convert DBH to basal area
+  dplyr::select(BA.Target,Plot,TreeID,Species,Year,val=BA) %>% 
+  pivot_wider(names_from=Year,values_from=val) %>% 
+  mutate(BA.Target=paste("TU:",BA.Target,sep="")) %>%  # create facet column
+
+  # Create growth columns
+  mutate(`2019`=(`2019`-`2009`)/11) %>% # divide by 11 to get annual increment
+  mutate(`2009`=(`2009`-`1997`)/11) %>% # divide by 11 to get annual increment
+  mutate(`1997`=(`1997`-`1994`)/3) %>% # divide by 3 to get annual increment
+  mutate(`1994`=(`1994`-`1992`)/3) %>% # divide by 3 to get annual increment
+  mutate(`Init`=`1992`) %>% # create initial size column
+  pivot_longer(cols=c(`1992`,`1994`,`1997`,`2009`,`2019`),values_to="Inc",names_to="Year") %>% 
+  filter(Year!=1992) %>% 
+  filter(Inc>0) %>% # need to remove any trees with increment of 0- %>% 
+  mutate(initClass=cut(Init,breaks=seq(from=7.5,to=90,by=20))) %>% 
+  mutate(incPCT=Inc/Init*100)
+
+# Raw plots
+baAnn.raw.plot<-
+  # Plot boxplot of increment by year
+  piDat %>% 
+  ggplot(aes(x=Year,y=Inc,color=BA.Target))+
+  geom_boxplot()+
+  facet_grid(Species~BA.Target)+
+  xlab("Measurement year")+
+  ylab("Annualized basal area increment (m²)")+
+  theme(legend.position = "none")+
+    theme_bw()+
+  theme(legend.position = "none")
+
+# Fit lmer
+# fixed effects model
+fitBA.Ann<-
+  piDat %>% 
+  lmer(log(Inc)~Year+Species+BA.Target+ (1|Plot/TreeID) ,data=.)
+
+ # Create figure showing marginal means from model
+fitBA.Ann.plot<-
+  ggpredict(fitBA.Ann,terms=c("Year","Species","BA.Target"))  %>%
+  plot()+
+  theme_bw()+
+  ylab("Predicated annualized basal area increment (m²)")+
+  xlab("Measurement year")+
+  ggtitle("")+
+  theme(legend.position = "bottom")
+
+# Captions 
+appNums(name="baAnnRaw",display=FALSE)
+figNums(name="fitBA.Ann.plot",display=FALSE)
+
+```
+
+To detect differences in growth rates over the 27-year period of this study, we compared growth rates between the four measurement periods, by species and treatment unit (`r appNums(name="baAnnRaw",display="cite")`).
+Growth rates were lowest in the three years immediately after treatment, and increased over the subsequent two measurement periods (1994-1997 and 1997-2009), before decreasing again in the last measurement perid (2009-2019).
+This temporal pattern appeared consistent between species and among treatment units; as well, spruce trees had higher growth rates than fir over all periods and treatment units, and growth rates were higher in the high-removal treatment unit, compared to the low-removal treatment and control units.
+A linear mixed effects model with species, measurement year and treatment unit as fixed effects (not shown) explained `r MuMIn::r.squaredGLMM(fitBA.Ann)[1] %>% round(3)*100`% variation (marginal R^2^) in log-transformed basal area increment (m^2^ at breast height) over different periods (`r figNums("fitBA.Ann.plot",display="cite")`).
+
+# Discussion
+
+# Tables
+
+##### page break
+
+# Figures
+
+```{r plot marginal means DBH increment by tree size}
+fitBA.plot
+```
+
+`r figNums("fitBA.plot",caption=paste("Predicted basal area (m^2^ at breast height) increment (cm) by tree basal area at time of treatment, species and treatment unit.
+Interactions between initial tree size, treatment unit and species were significant (p<0.001) in a linear model explaining ", MuMIn::r.squaredGLMM(fitBA)[1] %>% round(3)*100, "% variation in basal area increment.
+Predictions are back-transformed to original response scale.",sep=""))`
+
+```{r plot marginal means DBH increment pct}
+
+fitBApct.plot
+
+```
+
+`r figNums("fitBApct.plot",caption=paste("Predicted basal area (m^2^ at breast height) increment (cm), as a percentage of tree basal area at the time of treatment, by initial tree size.
+Interactions between initial tree size and (i) treatment unit and (ii) species factors were significant (p<0.001) in a linear model explaining ", MuMIn::r.squaredGLMM(fitBApct)[1] %>% round(3)*100, "% variation in basal area increment.
+Predictions are back-transformed to original response scale.",sep=""))`
+
+##### page break
+
+```{r plot marginal means BA increment by year}
+
+fitBA.Ann.plot
+
+```
+
+`r figNums("fitBA.Ann.plot",caption=paste("Predicted annualized basal area increment (m^2^ at breast height) over four periods.
+Basal area increment are average for growing seasons within each period.
+For example, values at x-axis value of 1994 are mean for three growing seasons (between 1992 and 1994).
+Species and treatment units were significant factors (p<0.001 for both, not shown) in a linear mixed effects model explaining ", MuMIn::r.squaredGLMM(fitBA.Ann)[1] %>% round(3)*100, "% variation in log-transformed basal area increment.
+Predictions are back-transformed to original response scale.
+Error bars represent 95% confidence intervals",sep=""))`
+
+##### page break
+
+# Appendices
+```{r plot raw PBAI by species and TU}
+
+pi.Init.rawPlot
+
+```
+
+`r appNums("piInitRaw",caption="Log-transformed basal area increment (m²) over 1992-2019 period by log-transformed tree size (basal area, m²) at the time of treatment, faceted by treatment unit and species.
+Both axes have been log-transformed.
+Linear regression lines added to emphasize slope.")`
+
+##### page break
+
+```{r plot raw PBAI as pct initial tree size by species and TU}
+
+piPCT.Init.rawPlot
+
+```
+
+`r appNums("piPCTInitRaw",caption="Log-transformed basal area increment (m²), as a pecentage of log-transformed initial tree size, over 1992-2019 period by tree size (basal area, m²) at the time of treatment, faceted by treatment unit and species.
+Both axes have been log-transformed.
+Linear regression lines added to emphasize slope.")`
+
+##### page break
+
+```{r plot raw basal area increment data by year}
+baAnn.raw.plot
+
+```
+
+`r appNums("baAnnRaw",caption="Annualized basal area increment (m² at breast height) over four measurement periods.
+Basal area increment are averaged at the tree-level for growing seasons within each period.
+For example, values at x-axis value of 1994 are mean for three growing seasons (between 1992 and 1994).
+Data are faceted by species and treatment unit.")`

### 2020-05-01 manuscript/EP1162_manuscript_draft.Rmd
more manny work

@@
 ### Tree-level growth responses
-To test the hypothesis that tree-level growth response over the 1992-2019 period varied with interactions between tree basal area at the time of treatment (1992) and (i) residual basal area; and (ii) species, we fit linear mixed effects models with log-transformed 1992-2019 basal area increment as the response variable, and treatment, species and log-transformed 1992 basal area as interacting fixed effects, with plot as a random intercept effect.
-We also fit a similar model with the 1992-2019 basal area increment expressed as a percentage of 1992 basal.
+To test the hypothesis that tree-level growth response over the 1992-2019 period varied with interactions between (i) tree basal area at the time of treatment (1992); (ii) residual basal area; and (iii) species, we fit linear mixed effects models with log-transformed 1992-2019 basal area increment as the response variable, and treatment, species and log-transformed 1992 basal area as interacting fixed effects, with plot as a random intercept effect.
+We also fit a similar model with the 1992-2019 basal area increment expressed as a percentage of 1992 basal area.
 Plots of residuals and fit lines with orginal data were examined to evaluate model goodness of fit.
@@
 We also examined how basal area increment at the tree-level varied over the four measurement periods, by fitting a similar model as described above, with an additional fixed effect of measurement period.
-This allowed us to examine potential lag responses in growth.
+This allowed us to examine temporal changes in growth responses, including evaluating for a potential lag effect.
 
@@
 `r nrow(piDat)`
- trees (
-`r nrow(filter(piDat,Species=="Sx"))`
-Spruce, 
-`r nrow(filter(piDat,Species=="Bl"))` Fir) confirmed that slopes and intercepts varied with species and residual basal area (p<0.001 for all interactions).  This model explained 
+ trees (`r nrow(filter(piDat,Species=="Sx"))`
+spruce, 
+`r nrow(filter(piDat,Species=="Bl"))` fir) confirmed that slopes and intercepts varied with species and residual basal area (p<0.001 for all interactions).  This model explained 
 `r round(r.squaredGLMM(fitBA)[1]*100,1)`
-% variation (marginal R^2^) in diameter increment over the 27 years of the study, and showed that diameter increment increased with initial tree size, and increment was higher in the treatment units with basal area removed, compared to the control, and that spruce trees had higher basal area increment than fir (`r figNums("fitDBH.plot",display="cite")`).
-A similar model predicting basal area increment as a percentage of tree size at the time of treatment showed that smaller trees grew proportionally more than larger trees, with a strong negative exponential decrease with initial tree size (`r figNums("fitDBHpct.plot",display="cite")`), however, the model fit was poorer, explaining
+% variation (marginal R^2^) in diameter increment over the 27 years of the study, and showed that diameter increment increased with initial tree size, and increment was higher in the treatment units with basal area removed, compared to the control, and that spruce trees had higher basal area increment than fir (`r figNums("fitBA.plot",display="cite")`).
+A similar model predicting basal area increment as a percentage of tree size at the time of treatment showed that smaller trees grew proportionally more than larger trees, with a strong negative exponential decrease with initial tree size (`r appNums("piPCTInitRaw",display="cite")` and `r figNums("fitBApct.plot",display="cite")`), however, the model fit was relatively poorer, explaining
 `r round(r.squaredGLMM(fitBApct)[1]*100,1)`
@@
 To detect differences in growth rates over the 27-year period of this study, we compared growth rates between the four measurement periods, by species and treatment unit (`r appNums(name="baAnnRaw",display="cite")`).
-Growth rates were lowest in the three years immediately after treatment, and increased over the subsequent two measurement periods (1994-1997 and 1997-2009), before decreasing again in the last measurement perid (2009-2019).
+Growth rates were lowest in the three years immediately after treatment, and increased over the subsequent two measurement periods (1994-1997 and 1997-2009), before decreasing again in the last measurement period (2009-2019).
 This temporal pattern appeared consistent between species and among treatment units; as well, spruce trees had higher growth rates than fir over all periods and treatment units, and growth rates were higher in the high-removal treatment unit, compared to the low-removal treatment and control units.
@@
 
-`r figNums("fitBA.plot",caption=paste("Predicted basal area (m^2^ at breast height) increment (cm) by tree basal area at time of treatment, species and treatment unit.
-Interactions between initial tree size, treatment unit and species were significant (p<0.001) in a linear model explaining ", MuMIn::r.squaredGLMM(fitBA)[1] %>% round(3)*100, "% variation in basal area increment.
+`r figNums("fitBA.plot",caption=paste("Predicted basal area (m^2^ at breast height) increment by tree basal area at time of treatment, species and treatment unit.
+Interactions between initial tree size, treatment unit and species were significant (p<0.001) in a linear model explaining ", MuMIn::r.squaredGLMM(fitBA)[1] %>% round(3)*100, "% variation in basal area increment (not shown).
 Predictions are back-transformed to original response scale.",sep=""))`
@@
 `r figNums("fitBApct.plot",caption=paste("Predicted basal area (m^2^ at breast height) increment (cm), as a percentage of tree basal area at the time of treatment, by initial tree size.
-Interactions between initial tree size and (i) treatment unit and (ii) species factors were significant (p<0.001) in a linear model explaining ", MuMIn::r.squaredGLMM(fitBApct)[1] %>% round(3)*100, "% variation in basal area increment.
+Interactions between initial tree size and (i) treatment unit and (ii) species factors were significant (p<0.001) in a linear model explaining ", MuMIn::r.squaredGLMM(fitBApct)[1] %>% round(3)*100, "% variation in basal area increment (not shown).
 Predictions are back-transformed to original response scale.",sep=""))`
@@
 
-`r figNums("fitBA.Ann.plot",caption=paste("Predicted annualized basal area increment (m^2^ at breast height) over four periods.
+`r figNums("fitBA.Ann.plot",caption=paste("Predicted annualized basal area increment (m^2^ at breast height) over four periods from a linear mixed effects model explaining ", MuMIn::r.squaredGLMM(fitBA.Ann)[1] %>% round(3)*100, "% variation in log-transformed basal area increment (not shown) with period, treatment unit and species as fixed effects (p<0.001 for all fixed effects, now shown).
 Basal area increment are average for growing seasons within each period.
 For example, values at x-axis value of 1994 are mean for three growing seasons (between 1992 and 1994).
-Species and treatment units were significant factors (p<0.001 for both, not shown) in a linear mixed effects model explaining ", MuMIn::r.squaredGLMM(fitBA.Ann)[1] %>% round(3)*100, "% variation in log-transformed basal area increment.
 Predictions are back-transformed to original response scale.
-Error bars represent 95% confidence intervals",sep=""))`
+Error bars represent 95% confidence intervals.",sep=""))`
 
@@
 
-`r appNums("piPCTInitRaw",caption="Log-transformed basal area increment (m²), as a pecentage of log-transformed initial tree size, over 1992-2019 period by tree size (basal area, m²) at the time of treatment, faceted by treatment unit and species.
+`r appNums("piPCTInitRaw",caption="Log-transformed basal area increment (m²), as a percentage of log-transformed initial tree size, over 1992-2019 period by tree size (basal area, m²) at the time of treatment, faceted by treatment unit and species.
 Both axes have been log-transformed.
@@
 `r appNums("baAnnRaw",caption="Annualized basal area increment (m² at breast height) over four measurement periods.
-Basal area increment are averaged at the tree-level for growing seasons within each period.
-For example, values at x-axis value of 1994 are mean for three growing seasons (between 1992 and 1994).
+Boxplots show dispersion of mean growing season tree-level basal area increment within each period.
 Data are faceted by species and treatment unit.")`

### 2020-05-03 manuscript/EP1162_manuscript_draft.Rmd
more manny work

@@
 
+# Set ggplot theme
+theme_set(theme_bw())
+
 ```
@@
 ### Stand-level basal area and volume growth
+
+#### Volume estimates
+Tree volumes were estimated using species- and region-specific equations [@Nigh2016] based on tree height and DBH.
+We estimated tree heights by fitting diameter-height models developed separately for fir and spruce, and for each period.
+Trees with broken tops were omitted from diameter-height models.
 
@@
   xlab("log(Tree BA in 1992, m²)")+
-    theme_bw()+
   theme(legend.position = "none")
@@
   xlab("log(Tree BA in 1992, m²)")+
-  theme_bw()+
   theme(legend.position="none")
@@
   xlab("Tree BA in 1992 (m²)")+
-  theme_bw()+
   theme(legend.position="bottom")
@@
   theme(legend.position = "none")+
-    theme_bw()+
   theme(legend.position = "none")
@@
 A linear mixed effects model with species, measurement year and treatment unit as fixed effects (not shown) explained `r MuMIn::r.squaredGLMM(fitBA.Ann)[1] %>% round(3)*100`% variation (marginal R^2^) in log-transformed basal area increment (m^2^ at breast height) over different periods (`r figNums("fitBA.Ann.plot",display="cite")`).
+
+### Height-diameter models
+
+```{r height diameter raw plot}
+
+heightDiamRaw.plot<-
+  dat %>% 
+    filter(Species %in% c("Bl","Sx")) %>% # filter for Bl and Sx for now
+    filter(Status=="Live") %>% # filter for live trees
+    filter(is.na(DBT)) %>% # filter for trees with no broken tops
+    drop_na(Height) %>% # remove trees where height was not measured
+
+    # Plotting
+  ggplot()+
+  aes(x=DBH,y=Height,fill=Species)+
+  geom_point()+
+  facet_grid(Year~Species)+
+  theme(legend.position = "none")+
+  ylab("Height (m)")+
+  xlab("DBH (cm)")+
+  geom_smooth(formula = y~log(x),method="lm",se=FALSE)
+
+# Fit model
+htDiamFit<- 
+  dat %>% 
+    filter(Species %in% c("Bl","Sx")) %>% # filter for Bl and Sx for now
+    filter(Status=="Live") %>% # filter for live trees
+    filter(is.na(DBT)) %>% # filter for trees with no broken tops
+    drop_na(Height) %>% # remove trees where height was not measured
+  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create tree ID column
+  lmer(Height~log(DBH)+(1|Plot/TreeID),data=.)
+
+appNums("heightDiamRawPlot",display="FALSE")
+```
+
+Tree heights had a positive nonlinear relationship with diameter, for both species and across all time periods (`r appNums("heightDiamRawPlot",display="cite")`).
+A comparison of height-diameter models showed that intercepts and slopes of log-transformed heights did not vary significantly (p<0.05) between species or time periods (not shown), therefore a single height-diameter linear mixed-effects model was developed using `r dat %>% filter(Species %in% c("Bl","Sx")) %>% filter(Status=="Live") %>% filter(is.na(DBT)) %>% drop_na(Height) %>% nrow()` tree height-diameter observations in the dataset, with log-transformed diameter as the fixed effect, and tree nested within plot as random effect.
+This model explained `r round(MuMIn::r.squaredGLMM(htDiamFit)[1]*100,1)`% of variation in log-transformed heights, with an intercept of `r round(fixef(htDiamFit)[1],3)` and slope coefficient of  `r round(fixef(htDiamFit)[2],3)` * log(DBH).
+Residual plots were assessed to ensure goodness of fit.
+This model was then applied to predict heights in trees without height measurements, and tree-volume estimates generated as per Nigh [-@Nigh2016].
 
@@
 
-`r figNums("fitBA.Ann.plot",caption=paste("Predicted annualized basal area increment (m^2^ at breast height) over four periods from a linear mixed effects model explaining ", MuMIn::r.squaredGLMM(fitBA.Ann)[1] %>% round(3)*100, "% variation in log-transformed basal area increment (not shown) with period, treatment unit and species as fixed effects (p<0.001 for all fixed effects, now shown).
+`r figNums("fitBA.Ann.plot",caption=paste("Predicted annualized basal area increment (m^2^ at breast height) over four periods from a linear mixed effects model explaining ", MuMIn::r.squaredGLMM(fitBA.Ann)[1] %>% round(3)*100, "% variation in log-transformed basal area increment with period, treatment unit and species as fixed effects (p<0.001 for all fixed effects, now shown).
 Basal area increment are average for growing seasons within each period.
@@
 Data are faceted by species and treatment unit.")`
+
+##### page break
+
+```{r plot height to diameter models by species and year, fig.width=6, fig.asp=1.2}
+
+heightDiamRaw.plot
+
+```
+
+`r appNums("heightDiamRawPlot",caption="Height and diameter comparisons for 870 tree measurements, faceted by species and measurement year.")`
+
+##### page break

### 2020-05-06 manuscript/EP1162_manuscript_draft.Rmd
manny work

@@
 bibliography: ep1162Bib.bib
+editor_options: 
+  chunk_output_type: console
 ---
@@
 library(captioner)
+library(lsmeans)
 
@@
 
+# Some additional data prep
+
+# set up diameter classes
+diamClass=seq(from=7.5,to=95,by=5)
+dat<- # assign variable
+  dat %>% 
+  mutate(DBH.class=cut(DBH,diamClass,ordered_result = TRUE,right=F))
+
 ```
@@
 Plots of residuals and fit lines with orginal data were examined to evaluate model goodness of fit.
-Variance explained for each model are reported using the approach described in Nakagaw et al.
+Variance explained for each model are reported using the approach described in Nakagawa et al.
 [-@Nakagawa2013], as implemented in the MuMIn package for R [@add later].
@@
 #### Volume estimates
-Tree volumes were estimated using species- and region-specific equations [@Nigh2016] based on tree height and DBH.
+Tree volumes were estimated using species- and region-specific equations [@nigh2016] based on tree height and DBH.
 We estimated tree heights by fitting diameter-height models developed separately for fir and spruce, and for each period.
@@
 
+### Tree density
+```{r sph by diam class species and time}
+
+# need to create  a list of plots with five measurements
+plotDenom<-
+  dat %>% 
+  left_join(plotDat,by="Plot") %>% # join with plot data
+  group_by(BA.Target,Year,Plot) %>% 
+  summarise(Num.Plots=n()) %>% 
+  ungroup() %>% 
+  dplyr::select(BA.Target:Plot) %>% 
+  group_by(BA.Target,Year) %>% 
+  summarise(Num.Plots=n())
+
+# PLOTS
+# Raw plot - examine SPH in each plot to detect where mortality occurred
+# sphPlot.raw<-
+#   dat %>% 
+#   left_join(plotDat,by="Plot") %>% # join with plot data
+#   filter(Status=="Live") %>% # filter for live trees only
+#   group_by(BA.Target,Species,Plot,Year,DBH.class) %>% 
+#   summarise(SPH=n()*20)  %>% 
+#   filter(Species %in% c("Bl","Sx")) %>% # remove Fdi and Ep
+#   ggplot()+
+#   aes(x=DBH.class,y=SPH,group=Species,color=Species)+
+#   geom_point()+
+#   facet_grid(Plot~Year)+
+#   theme(axis.text.x = element_text(angle = 90, hjust = 1),
+#         legend.position="bottom")+
+#   xlab("DBH Class (cm)")
+# 
+# appNums(name="sphRawPlot",display=FALSE)
+
+# 1. SPH by diameter class and species for 1992 and 2019 periods
+sphPlot1<-
+  dat %>% 
+  filter(Year %in% c(1992,2019)) %>% # Select for two time periods
+    filter(Species %in% c("Bl","Sx")) %>% 
+  left_join(plotDat,by="Plot") %>% # join with plot data
+  group_by(BA.Target,Year,Species,DBH.class) %>% 
+  summarise(SPH.Sum=n()*20) %>% # plot multiplier
+  left_join(plotDenom,by=c("BA.Target","Year")) %>% 
+  mutate(SPH=SPH.Sum/Num.Plots) %>%
+
+  ggplot()+
+  aes(x=DBH.class,y=SPH,group=Species,fill=Species,color=Species)+
+  geom_point()+
+  geom_path()+
+    # geom_density(alpha=0.4)+
+  facet_grid(Year~BA.Target)+
+  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5),
+        legend.position = "bottom")+
+  xlab("DBH Class (cm)")
+
+  figNums(name="sphSpeciesPlot",display=FALSE)
+```
+
+```{r plot pct change in SPH by diam class between 92 and 19}
+
+# % change in diameter class between 1992 and 2019 time periods by species
+ sphPCTPlot<- 
+  dat %>% 
+  filter(Year %in% c(1992,2019)) %>% 
+  filter(Status=="Live") %>% 
+  filter(Species %in% c("Bl","Sx")) %>% 
+  group_by(Plot,Species,Year,DBH.class) %>% 
+  summarise(SPH.Total=n()) %>% 
+  pivot_wider(names_from="Year",values_from="SPH.Total") %>% 
+  mutate(Pct.Change=(`2019`-`1992`)/`1992`*100) %>% 
+  mutate(Pct.Change=round(Pct.Change,3)) %>%
+  drop_na(Pct.Change) %>% 
+  left_join(plotDat,by="Plot") %>% # join with plot data
+
+  # Plotting
+  ggplot()+
+  aes(x=DBH.class,y=Pct.Change,fill=Species)+
+  geom_boxplot()+
+  geom_hline(yintercept=0, linetype="dashed", color="darkgrey")+
+  facet_grid(Species~BA.Target)+
+  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5),
+        legend.position="bottom")+
+  xlab("DBH Class (cm)")+
+  ylab("% change in SPH between 1992 and 2019")
+
+  figNums(name="sphPCTPlot",display=FALSE)
+
+# ggsave(filename="baDiamClass.png")
+
+```
+
+Mean plot tree density by diameter class diagrams showed that across all treatment levels, subalpine fir comprised the majority of stems in the smaller diameter classes, both in the year after treatment, and in 2019 (`r figNums("sphSpeciesPlot",display="cite")`).
+The diameter class distribution of fir in all treatment units was negative exponential.
+Mean spruce density was lower than fir, with a relatively flatter diameter distribution.
+Within the high-removal treatment unit, fir density increased mostly in the smaller diameter classes between 1992 and 2019; in the low-removal treatment unit, fir density increased most in trees between 12.5 to 27.5cm DBH; and in the control treatment unit, fir density increased mostly in trees above 22.5cm DBH ( `r figNums("sphPCTPlot",display="cite")`).
+Spruce tree density increased less than fir, with the exception of an increase of over 250% in SPH in the smallest diameter class in the light-removal treatment unit.
+
+## Stand-level changes
+
+### Volume estimates
+
 ### Height-diameter models
@@
   mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create tree ID column
-  lmer(Height~log(DBH)+(1|Plot/TreeID),data=.)
+  lmer(log(Height)~log(DBH)+(1|Plot/TreeID),data=.)
 
+# Apply fit model to data and create new height column
+dat<-
+  dat %>% 
+  mutate(Height.Pred=coalesce(Height,exp(fixef(htDiamFit)[1]+fixef(htDiamFit)[2]*log(DBH))))
+
+# Height model fit plot
+htModelFit.plot<-
+  dat %>% 
+  ggplot(aes(x=Height,y=Height.Pred))+
+  geom_point()+
+  xlim(0,40)+
+  ylim(0,40)+
+  geom_smooth(method="lm",se=F)+
+  ylab("Predicted height (m) from diameter-height model")+
+  xlab("Measured height (m)")
+
+# create captions
 appNums("heightDiamRawPlot",display="FALSE")
+appNums("htModelFit",display="FALSE")
+
 ```
@@
 This model explained `r round(MuMIn::r.squaredGLMM(htDiamFit)[1]*100,1)`% of variation in log-transformed heights, with an intercept of `r round(fixef(htDiamFit)[1],3)` and slope coefficient of  `r round(fixef(htDiamFit)[2],3)` * log(DBH).
-Residual plots were assessed to ensure goodness of fit.
-This model was then applied to predict heights in trees without height measurements, and tree-volume estimates generated as per Nigh [-@Nigh2016].
+Residual and predicted vs actual height plots were assessed to ensure goodness of fit (`appNums("htModelFit",display="cite")`).
+The model slightly overestimated tree heights in shorter trees, and underestimated tree heights in taller trees.
+This model was applied to predict heights in trees without height measurements, and tree-volume estimates generated as per Nigh [-@Nigh2016].
+
+```{r hardcode volume coefficients per Nigh}
+
+volCoef<-
+
+  ## TOTAL VOLUME
+  #  Subalpine fir 
+  data.frame(Species="Bl",VolType="total",b0=-9.79,b1=1.813,b2=1.033) %>%
+
+  # Douglas-fir
+  add_row(Species="Fd",VolType="total",b0=-10.19,b1=1.73,b2=1.206) %>%
+
+  # Add spruce
+  add_row(Species="Sx",VolType="total",b0=-10.024,b1= 1.801,b2= 1.107) %>%
+
+  ## MERCHANTABLE VOLUME
+  # Subalpine fir
+  add_row(Species="Bl",VolType="merch",b0=-9.980,b1= 1.792,b2= 1.102) %>%
+
+  # Douglas-fir
+  add_row(Species="Fd",VolType="merch",b0=-10.503,b1= 1.724,b2= 1.296) %>%
+
+  # Spruce
+  add_row(Species="Sx",VolType="merch",b0=-10.280,b1= 1.769,b2= 1.206 ) %>%
+
+  mutate(BGC.Zone="SBS") %>%
+
+  # convert species to factor
+  mutate(Species=factor(Species)) %>% 
+  dplyr::select(Species,BGC.Zone,everything())
+
+```
+
+```{r generate volume estimates}
+
+dat<-
+
+  # Total volumes
+  dat %>% 
+  left_join(filter(volCoef,VolType=="total"),by="Species") %>% 
+  mutate(VolTot=exp(1)^b0*DBH^b1*Height.Pred^b2) %>% 
+  dplyr::select(-c(BGC.Zone,VolType,b0,b1,b2)) %>%
+
+  # Merchantable volume
+  left_join(filter(volCoef,VolType=="merch"),by="Species") %>% 
+  mutate(VolMerch=exp(1)^b0*DBH^b1*Height.Pred^b2) %>% 
+  dplyr::select(-c(BGC.Zone,VolType,b0,b1,b2)) %>%
+
+  # basal area
+  mutate(BA=(DBH/100/2)^2*pi)
+
+# Create stand data
+standDat<-
+  dat %>% 
+  group_by(Plot,Year) %>% 
+  summarise(VolTot=sum(VolTot,na.rm=T)*20, # notice that all variables are by hectare
+            VolMerch=sum(VolMerch,na.rm=T)*20,
+            BA=sum(BA,na.rm=T)*20,
+            k=n()*20) %>% 
+  mutate(QMD=sqrt(BA/(k*0.0000785))) %>% 
+  left_join(plotDat,by="Plot") %>%  # joinwith plot Data
+   mutate(TU=paste("TU",BA.Target,sep=":"))  # create a TU column for plotting
+
+```
+
+```{r boxplot stand volumes by year}
+
+volPlot<-
+  standDat %>% 
+  # plot
+  ggplot()+
+  aes(x=Year,y=BA,fill=BA.Target)+
+  geom_boxplot()+
+  facet_wrap(~TU)+
+  theme(legend.position="none")+
+  ylab("Mean plot volume (m/hectare)")
+
+volFit<-
+  standDat %>% 
+  lm(VolTot~BA.Target*Year,data=.)
+
+```
 
@@
 
+```{r plot density by diameter class, fig.width=6}
+sphPlot1
+
+```
+
+`r figNums(name="sphSpeciesPlot",caption="Mean density of stems per hectare (SPH) by diameter class and species, faceted by treatment unit and two measurement periods.
+Lowest value is included in each diameter class, highest value is not.")`
+
+##### page break
+
+```{r plot density change PCT by diameter class, fig.width=6}
+sphPCTPlot
+
+```
+
+`r figNums(name="sphSpeciesPlot",caption="Percentage change between 1992 and 2019 in stems per hectare (SPH) by diameter class, faceted by treatment unit and species.
+tsphPCTPlotwo measurement periods.
+Lowest value is included in each diameter class, highest value is not.")`
+
+##### page break
+
 # Appendices
@@
 
-```{r plot height to diameter models by species and year, fig.width=6, fig.asp=1.2}
+```{r plot height to diameter models by species and year, fig.asp=1.2, fig.width=6}
 
@@
 ##### page break
+```{r plot diam to height model fit}
+
+htModelFit.plot
+```
+
+`r appNums("htModelFit",caption="Predicted height against measured height for 870 tree measurements.  Outlier represents a tree with a broken top.")`

### 2020-05-07 manuscript/EP1162_manuscript_draft.Rmd
manny work

@@
 Tree volumes were estimated using species- and region-specific equations [@nigh2016] based on tree height and DBH.
-We estimated tree heights by fitting diameter-height models developed separately for fir and spruce, and for each period.
+We estimated tree heights by fitting a single diameter-height models developed with a linear mixed effects model, with log-transformed DBH as a fixed effect, and tree nested within plot as a random effect.
 Trees with broken tops were omitted from diameter-height models.
@@
 The model slightly overestimated tree heights in shorter trees, and underestimated tree heights in taller trees.
-This model was applied to predict heights in trees without height measurements, and tree-volume estimates generated as per Nigh [-@Nigh2016].
+This model was applied to predict heights in trees without height measurements, and tree-volume estimates generated as per Nigh [-@nigh2016].
 
@@
   group_by(Plot,Year) %>% 
-  summarise(VolTot=sum(VolTot,na.rm=T)*20, # notice that all variables are by hectare
-            VolMerch=sum(VolMerch,na.rm=T)*20,
+  summarise(Vol.Tot=sum(VolTot,na.rm=T)*20, # notice that all variables are by hectare
+            Vol.Merch=sum(VolMerch,na.rm=T)*20,
             BA=sum(BA,na.rm=T)*20,
@@
   theme(legend.position="none")+
-  ylab("Mean plot volume (m/hectare)")
+  ylab("Plot volume (m³/hectare)")
+
+# caption
+figNums("volPlot.plot",display=FALSE)
 
@@
   standDat %>% 
-  lm(VolTot~BA.Target*Year,data=.)
+  lm(Vol.Tot~BA.Target*Year,data=.)
 
 ```
+
+### Stand increment
+```{r function to compute stand increment}
+
+standIncFunc<-function(varName) {
+
+  standDat %>% 
+  dplyr::select(Plot,Year,varName,BA.Target) %>% 
+  pivot_wider(names_from="Year",values_from=varName) %>% 
+  mutate(`1994Inc`=`1994`-`1992`) %>% # stand periodic increment 
+  mutate(`1997Inc`=`1997`-`1994`) %>% # stand periodic increment 
+  mutate(`2009Inc`=`2009`-`1997`) %>% # stand periodic increment 
+  mutate(`2019Inc`=`2019`-`2009`) %>%  # stand periodic increment
+
+  # divide by # growing seassons in each period
+  mutate(`1994Inc`=`1994Inc`/3) %>% # 3 growing seasons in this period 
+  mutate(`1997Inc`=`1997Inc`/3) %>% # 3 growing seasons in this period 
+  mutate(`2009Inc`=`2009Inc`/11) %>% # 11 growing seasons in this period 
+  mutate(`2019Inc`=`2019Inc`/11) %>% 
+    dplyr::select(Plot,BA.Target,`1994`=`1994Inc`,`1997`=`1997Inc`,
+                  `2009`=`2009Inc`,`2019`=`2019Inc`) %>% 
+    pivot_longer(cols=`1994`:`2019`,names_to="Year",values_to="Inc") %>% 
+    mutate(varname=varName) %>% 
+    dplyr::select(Plot,BA.Target,varname,Year,Inc) %>% 
+    return() # return
+}
+
+```
+
+```{r apply stand increment function}
+
+incPlot<-
+  standIncFunc("Vol.Tot") %>%   # extract total volume 
+  rbind(standIncFunc("Vol.Merch")) %>%   # extract total volume 
+  rbind(standIncFunc("BA")) %>% # extract basal area
+  drop_na(Inc) %>%
+
+  # Create group summaries
+  group_by(BA.Target,Year,varname) %>% 
+  summarise(IncMean=mean(Inc),
+            se=sd(Inc)/sqrt(length(Inc))) %>%
+
+  # create lookup column
+  mutate(TU=paste("TU",BA.Target,sep=":")) %>% # create lookup column
+
+  # plotting
+  ggplot()+
+  aes(x=Year,y=IncMean,color=TU,group=1)+
+  geom_errorbar(aes(ymin=IncMean-se,ymax=IncMean+se,width=.2))+
+  geom_point(size=2)+
+  geom_path()+
+  facet_grid(varname~TU,scales="free_y")+
+  theme(legend.position="none")+
+  ylab("Annualized increment by period")
+
+# caption
+figNums("incPlot",display=FALSE)
+
+```
+
+Annual increment between the four periods differed between treatment units (`r figNums("incPlot",display="cite")`).
+In the high-removal treatment unit, annual basal area increment increased in each subsequent measurement period, with the highest mean plot values of 0.9m^2^ per hectare.
+In contrast, mean plot basal area annual increment was highest in the low-removal treatment unit during the 1994-1997 period, and decreased in subsequent periods.
+A similar pattern was observed in the control treamtent unit.
+Annual increment patterns in total and merchantable volumes were similar to basal area, with the exception of the high-removal treatment unit.
+Here, maximum annual volume increment was highest between 1997 and 2009, with a mean plot value of 8m^3^ per year per hectare; this increment decreased slightly from 2009-2019 period.
+In the control unit, volume increment from 2009-2019 was lower than from 1992-1994.
 
@@
 
-`r figNums(name="sphSpeciesPlot",caption="Percentage change between 1992 and 2019 in stems per hectare (SPH) by diameter class, faceted by treatment unit and species.
+`r figNums(name="sphPCTPlot",caption="Percentage change between 1992 and 2019 in stems per hectare (SPH) by diameter class, faceted by treatment unit and species.
 tsphPCTPlotwo measurement periods.
@@
 ##### page break
+
+```{r volume stand plots}
+volPlot
+
+```
+
+`r figNums("volPlot.plot",caption="Volume per plot (m³ per hectare) at five measurement years, faceted by treatment unit.")`
+
+##### page break
+
+```{r plot stand increment}
+
+incPlot
+
+```
+
+`r figNums("incPlot",caption="Mean annual increment per plot over four measurement periods.
+Top panel row shows basal area (BA), with units of m2 per hecatre.
+Middle panel row shows merchantable volume (Vol.Merch), and bottom panel row shows total volume (Vol.Tot).
+Volume units are m³ per hectare.")`
 
@@
 `r appNums("htModelFit",caption="Predicted height against measured height for 870 tree measurements.  Outlier represents a tree with a broken top.")`
+
+# References

### 2020-05-13 manuscript/EP1162_manuscript_draft.Rmd
add diam class work

@@
   dat %>% 
-  mutate(DBH.class=cut(DBH,diamClass,ordered_result = TRUE,right=F))
+  mutate(DBH.class=cut(DBH,diamClass,ordered_result = TRUE,right=F)) %>% 
+  mutate(Plot=as.factor(Plot))
+
+# formatting plotDat
+plotDat<-
+  plotDat %>% 
+  mutate(Plot=as.factor(Plot))
 
@@
 ### Tree-level growth responses
-To test the hypothesis that tree-level growth response over the 1992-2019 period varied with interactions between (i) tree basal area at the time of treatment (1992); (ii) residual basal area; and (iii) species, we fit linear mixed effects models with log-transformed 1992-2019 basal area increment as the response variable, and treatment, species and log-transformed 1992 basal area as interacting fixed effects, with plot as a random intercept effect.
+To test the hypothesis that tree-level growth response over the 1992-2019 period varied with interactions between (i) tree basal area at the time of treatment (1992); (ii) residual basal area; and (iii) species, we fit linear mixed effects models with log-transformed 1992-2019 basal area increment as the response variable, and treatment, species and log-transformed 1992 basal area as interacting fixed effects, and plot as a random intercept effect.
 We also fit a similar model with the 1992-2019 basal area increment expressed as a percentage of 1992 basal area.
@@
 This allowed us to examine temporal changes in growth responses, including evaluating for a potential lag effect.
+We do note that because we did not measure increment before treatment, we cannot address questions regarding short-term growth declines associated with treatment.
 
@@
   dat %>% 
-  ggplot(aes(x=Height,y=Height.Pred))+
+  mutate(h2=exp(fixef(htDiamFit)[1]+fixef(htDiamFit)[2]*log(DBH))) %>% 
+  ggplot(aes(x=Height,y=h2))+
   geom_point()+
@@
 This model explained `r round(MuMIn::r.squaredGLMM(htDiamFit)[1]*100,1)`% of variation in log-transformed heights, with an intercept of `r round(fixef(htDiamFit)[1],3)` and slope coefficient of  `r round(fixef(htDiamFit)[2],3)` * log(DBH).
-Residual and predicted vs actual height plots were assessed to ensure goodness of fit (`appNums("htModelFit",display="cite")`).
+Residual and predicted vs actual height plots were assessed to ensure goodness of fit (`r appNums("htModelFit",display="cite")`).
 The model slightly overestimated tree heights in shorter trees, and underestimated tree heights in taller trees.
@@
 
+### Species composition over time
+
+```{r create stand data by species}
+# Create stand data
+sppDat<-
+  dat %>% 
+  group_by(Plot,Species,Year) %>% 
+  summarise(Vol.Tot=sum(VolTot,na.rm=T)*20, # notice that all variables are by hectare
+            Vol.Merch=sum(VolMerch,na.rm=T)*20,
+            BA=sum(BA,na.rm=T)*20,
+            SPH=n()*20) %>% 
+  mutate(QMD=sqrt(BA/(SPH*0.0000785))) %>% 
+  left_join(plotDat,by="Plot") %>%  # joinwith plot Data
+  dplyr::select(Plot:BA.Target) %>% 
+  pivot_longer(cols=Vol.Tot:QMD,names_to="varName",values_to="Quantity") %>% 
+   mutate(TU=paste("TU",BA.Target,sep=":"))  # create a TU column for plotting
+
+sppIncFunc<-function(varName) {
+
+  sppDat %>% 
+  dplyr::select(Plot,BA.Target,varName,Year) %>% 
+  pivot_wider(names_from="Year",values_from=varName) %>% 
+  mutate(`1994Inc`=`1994`-`1992`) %>% # stand periodic increment 
+  mutate(`1997Inc`=`1997`-`1994`) %>% # stand periodic increment 
+  mutate(`2009Inc`=`2009`-`1997`) %>% # stand periodic increment 
+  mutate(`2019Inc`=`2019`-`2009`) %>%  # stand periodic increment
+
+  # divide by # growing seassons in each period
+  mutate(`1994Inc`=`1994Inc`/3) %>% # 3 growing seasons in this period 
+  mutate(`1997Inc`=`1997Inc`/3) %>% # 3 growing seasons in this period 
+  mutate(`2009Inc`=`2009Inc`/11) %>% # 11 growing seasons in this period 
+  mutate(`2019Inc`=`2019Inc`/11) %>% 
+    dplyr::select(Plot,Species,BA.Target,`1994`=`1994Inc`,`1997`=`1997Inc`,
+                  `2009`=`2009Inc`,`2019`=`2019Inc`) %>% 
+    pivot_longer(cols=`1994`:`2019`,names_to="Year",values_to="Inc") %>% 
+    mutate(varname=varName) %>% 
+    dplyr::select(Plot,Species,BA.Target,varname,Year,Inc) %>%
+    mutate(TU=paste("TU",BA.Target,sep=":")) %>% 
+    return() # return
+}
+
+```
+
+```{r plot species breakdown}
+sppPlot<-
+  sppDat %>% 
+  filter(Species %in% c("Bl","Sx")) %>% 
+  ggplot()+
+  aes(x=Year,y=Quantity,fill=Species)+
+  geom_boxplot()+
+  facet_grid(varName~TU,scales="free_y")+
+  theme(legend.position="bottom")
+
+# caption
+figNums("sppPlot",display=FALSE)
+
+```
+
+#### Volume and basal area
+Fir basal area and volume exceeded that of spruce in both the control and light-removal treatment unit; these differences increased in each subsequent period after ca.1997 (`r figNums("sppPlot",display="cite")`).
+From 1997-2019, fir basal area and volume increased faster than spruce in the light-removal and control units.
+The differences were not as clear in the high-removal treatment unit, as there was a lot of variation in spruce among plots.
+*we could test this for significance*
+
+Among treatments, spruce volumes and basal were similar across the three treatment units, especially from 1997-2019.
+Spruce volume and basal area varied substantially among plots within the high-removal treatment unit.
+Fir basal area and volumes in the control unit exceeded other treatment units, with the largest difference between control and high-removal treatment unit.
+
+#### Tree density
+In the control and light-removal treatment unit, there were more fir trees than spruce across all time periods.
+The difference between fir and spruce density appeared to remain stable in the control unit over the period of this study.
+In the high-removal treatment unit, fir density began to noticeably exceed that of spruce starting ca.1997, although the variation in fir tree density among plots in this treatment was relatively high.
+
+#### Quadratic mean diameter
+Quadratic mean diameter in both fir and spruce remained relatively stable over all measurement periods in the control unit.
+In the low-removal treatment unit, spruce QMD exceeded that of fir across all periods, however, the differences between species apeared to be diminishing over time.
+In the high-removal treatment unit, spruce QMD increased over time, and both species had approximately similar mean QMD.
+
+```{r compare sph and BA differences between periods and diam classes}
+perComp<-
+  dat %>%
+  filter(Year %in% c(1992,2019)) %>% # filter for start and end years
+  group_by(Plot,Year) %>% 
+  summarise(N.Plot=n()) %>% # summarise number of years per plot
+  group_by(Plot) %>% 
+  summarise(N.Plot=n()) %>% 
+  filter(N.Plot==2) %>% 
+  pull(Plot) %>% 
+  data.frame(Plot=.) %>%
+
+  # now filter dat for these plots
+  right_join(dat,by="Plot") %>% 
+  filter(Year %in% c(1992,2019)) %>% # filter for years
+  filter(Status=="Live") %>% #filter for live trees
+
+  # break trees into diameter classes
+  mutate(DBH.class=cut(DBH,breaks=c(7.5,12.5,18.5,30,90),include.lowest = TRUE,right=FALSE)) %>% 
+  left_join(plotDat,by="Plot")
+
+# fit models
+fit1<-  
+  perComp %>%
+
+  # come up with a denominator (number of plots per BA.Target per year)
+  distinct(Plot,.keep_all = TRUE) %>% 
+  group_by(BA.Target) %>% 
+  summarise(Num.Plot=n()) %>% 
+  right_join(perComp,by="BA.Target") %>%
+
+  # Plot averages
+  group_by(Plot,Year,BA.Target,Num.Plot,Species,DBH.class) %>% 
+  summarise(BA=sum(BA)*20,# Notice the plot multiplier
+            Vol=sum(VolTot)*20,
+            SPH=n()*20) %>% 
+  filter(DBH.class=="[30,90]") %>% 
+  filter(Species=="Sx") %>% 
+  lm(BA~BA.Target+Year,data=.)
+
+# plotting
+perCompPlot<- # assign variable  
+  perComp %>%
+
+  # come up with a denominator (number of plots per BA.Target per year)
+  distinct(Plot,.keep_all = TRUE) %>% 
+  group_by(BA.Target) %>% 
+  summarise(Num.Plot=n()) %>% 
+  right_join(perComp,by="BA.Target") %>%
+
+  # Plot averages
+  group_by(Plot,Year,BA.Target,Num.Plot,Species,DBH.class) %>% 
+  summarise(BA=sum(BA)*20,# Notice the plot multiplier
+            Vol=sum(VolTot)*20,
+            SPH=n()*20) %>%
+
+  #BA Target averages
+  group_by(BA.Target,Year,Species,Num.Plot,DBH.class) %>% 
+  summarise(BA.sum=sum(BA),
+            BA.se=sd(BA)/sqrt(length(BA)),
+            Vol.sum=sum(Vol),
+            Vol.se=sd(Vol)/sqrt(length(Vol)),
+            SPH.sum=sum(SPH),
+            SPH.se=sd(SPH)/sqrt(length(SPH))) %>% 
+  ungroup() %>%
+
+  # Select only for Sx and Bl
+  filter(Species %in% c("Bl","Sx")) %>%
+
+  # Compute means
+  mutate(BA.mean=BA.sum/Num.Plot,
+         Vol.mean=Vol.sum/Num.Plot,
+         SPH.mean=SPH.sum/Num.Plot) %>%
+
+  # format
+  dplyr::select(BA.Target,Year,Species,DBH.class,BA=BA.mean,Vol=Vol.mean,SPH=SPH.mean) %>% 
+  pivot_longer(cols=BA:SPH,names_to="var",values_to="value")
+
+vol.plot2<-  
+  perCompPlot %>% 
+  filter(var=="Vol") %>%
+
+  # Plotting
+  ggplot()+
+  aes(x=DBH.class,y=value,fill=Species,group=Species)+
+  geom_bar(stat="Identity",position="dodge")+
+  facet_grid(Year~BA.Target)
+
+BA.plot2<-  
+  perCompPlot %>% 
+  filter(var=="BA") %>%
+
+  # Plotting
+  ggplot()+
+  aes(x=DBH.class,y=value,fill=Species,group=Species)+
+  geom_bar(stat="Identity",position="dodge")+
+  facet_grid(Year~BA.Target)
+
+SPH.plot2<-  
+  perCompPlot %>% 
+  filter(var=="SPH") %>%
+
+  # Plotting
+  ggplot()+
+  aes(x=DBH.class,y=value,fill=Species,group=Species)+
+  geom_bar(stat="Identity",position="dodge")+
+  facet_grid(Year~BA.Target)
+
+```
+
 # Discussion
+Here are the main results that emerge to me from our data:
+
+1.
+All treatment units maintained a stand structure analagous to an uneven-aged stand (ie reverse-J diameter distribution).
+In the high-removal TU, most recruitment was into the smallest size class (pole), whereas recruitment was into larger size classes in the light-removal and control.
+This suggests that understory response may increase with level of change in the canopy.
+\newLine
+\newLine
+
+2. High-level of basal area removal stimulated a stronger and more sustained growth response
+Basal area increment has higher in the high-removal treatment units compared to other two, and peaked ~20 years after treatment, compared to ~5 years in control/light-removal.
+What does this mean for stand development over the next 20-30 years?
+- By 2019, the high-removal and light-removal TU converged in terms of volume and basal area.
+This seems remarkable to me, and perhaps runs counter to what we learned in silviculture II: *one will never recoup thinned volume!*  Growth rates in the high-removal TU exceeded those in the light-removal; what does this mean for stand dynamics over the next 20 years?
+Will the high-removal TU actually exceed light-removal in terms of volume?
+- Most basal area increases are from the larger trees, however, the smaller trees had stronger responses when considered as a % of initial tree size.
+
+\newLine
+\newLine
+
+3. Partial cutting in this uneven-aged spruce-fir forest affected spruce-fir composition differently depending on harvest level
+- Stand development after harvest showed that spruce volume, basal area, and QMD increased more in the high-removal treatment compared to the other two treatments.
+In many plots in the high-removal TU, spruce volumes exceeds control/light-removal spruce.
+- High removal stimulated stronger growth responses in spruce trees, compared to fir.
+In the light-removal and control units, spruce basal area and volume changed much less over the 27 years of the trial, and in fact, have similar levels.
+- Within a time period, spruce quadratic means diameters were inversely related to harvest level.
+QMD is highest in the control unit, and lowest in the high-removal treatment unit, suggesting that there are fewer but larger trees in the control, whereas recuitment into sapling class trees has been highest in high-removal TU.
+Does this suggest that spruce are well adapted to take advantage of canopy gaps?
+- Would we see this response in even more removal, say down to 5m^2^ per hectare?
+- An important note: fir stems per hectare were low in the high-removal TU until around 1997, and then increased more rapidly relative to other units from ca. 1997-2019.
+- Note that Figure 3 and 7 are similar - Figure 3 shows predicted BAI from model, and 7 shows actual data.
+Recommend deleting Figure 3 and using 7, b/c prefer to use real data!
+Appendix three is similar to Figure 7 - just uses boxplots instead of mean/se.
+Recommend Figure 7 and delete Appendix 3.
+
+\newLine
+\newLine
+
+4. What other results emerge from the data?
+
+\newLine
+\newLine
+
+**Discussion structure from Mike below**
+
+1.  comparisons of findings to the related scientific literature
+2.  comparisons of findings to biological principles and current ecological understandings - e.g. do our results make sense based on our understandings of stand dynamics, or example?
+3.
+discussion of findings that seem anomalous, counter-intuitive, or contrary to prevailing literature or theory, AND consideration of whether our results and data may be an advancement in the science.
+4.  future research directions
+5.
+commentary and direction on our findings and current management practices relative to: (a) similar practices globally, (b) similar practices in similar Picea-Abies types elsewhere, and finally (c) practices in ESSF-SBS spruce fir types in BC.
+6.
+Regarding commentary on BC management:  Keep this squarely focused on what the scientific foundation of these types of practices SHOULD be, based on the scientific evidence (and comparative literature) we have assembled.
+We can then comment on current practices and standards and whether they do or do not meet these tests.
+We must acknowledge that management standards for silviculture incorporate multiple considerations including best available science, but also other factors (e.g.
+simplicity, enforceability, etc).
+7.
+Extrapolation and interpretation of our results beyond the bounds of this particular study and study site:  this is where the comparative literature and reference to established principles of stand dynamics and productivity will aid us - i.e.
+if our results are consistent with broad-based silvicultural and ecological understandings and principles, then we can have good confidence that our results are robust and sound for extrapolation of other similar stands and forest types.
+8.   Opportunities for modelling etc in future?
 
@@
 `r figNums(name="sphPCTPlot",caption="Percentage change between 1992 and 2019 in stems per hectare (SPH) by diameter class, faceted by treatment unit and species.
-tsphPCTPlotwo measurement periods.
 Lowest value is included in each diameter class, highest value is not.")`
@@
 
-`r figNums("volPlot.plot",caption="Volume per plot (m³ per hectare) at five measurement years, faceted by treatment unit.")`
+`r figNums("volPlot.plot",caption="Variation in volume per plot (m³ per hectare) at five measurement years, faceted by treatment unit.")`
 
@@
 `r figNums("incPlot",caption="Mean annual increment per plot over four measurement periods.
-Top panel row shows basal area (BA), with units of m2 per hecatre.
+Top panel row shows basal area (BA), with units of m² per hectare.
 Middle panel row shows merchantable volume (Vol.Merch), and bottom panel row shows total volume (Vol.Tot).
 Volume units are m³ per hectare.")`
+
+##### page break
+```{r plot stand attributes by species,fig.height=8,fig.width=6}
+sppPlot
+
+```
+
+`r figNums("sppPlot",caption="Among-plot variation in stand attributes, by species and faceted by treatment unit and attribute.
+Units are as follows: basal area (BA, m² per hectare); QMD (quadratic mean diameter,cm), trees per hectare (SPH, count), Merchantable volume (Vol.Merch, m³ per hectare); Total volume (Vol.Merch, m³ per hectare)")`
 
@@
 
-`r appNums("htModelFit",caption="Predicted height against measured height for 870 tree measurements.  Outlier represents a tree with a broken top.")`
+`r appNums("htModelFit",caption="Predicted height against measured height for 870 tree measurements.  Outliers above the line may represent a tree with a broken top.")`
 

### 2020-05-14 manuscript/EP1162_manuscript_draft.Rmd
more manny work

@@
 library(lsmeans)
+library(cowplot)
 
@@
   ggplot()+
-  aes(x=Year,y=BA,fill=BA.Target)+
+  aes(x=Year,y=Vol.Tot,fill=BA.Target)+
   geom_boxplot()+
@@
 Annual increment between the four periods differed between treatment units (`r figNums("incPlot",display="cite")`).
-In the high-removal treatment unit, annual basal area increment increased in each subsequent measurement period, with the highest mean plot values of 0.9m^2^ per hectare.
-In contrast, mean plot basal area annual increment was highest in the low-removal treatment unit during the 1994-1997 period, and decreased in subsequent periods.
-A similar pattern was observed in the control treamtent unit.
-Annual increment patterns in total and merchantable volumes were similar to basal area, with the exception of the high-removal treatment unit.
-Here, maximum annual volume increment was highest between 1997 and 2009, with a mean plot value of 8m^3^ per year per hectare; this increment decreased slightly from 2009-2019 period.
-In the control unit, volume increment from 2009-2019 was lower than from 1992-1994.
+In the first three growing seasons after treatment, basal area and volume increments in the two treatment units with basal area removal were lower than increment in the control unit.
+In the high-removal treatment unit, basal area increased in each subsequent measurement period, and exceeded the other units between 1997 and 2019.
+In the low-removal and control treatment units, increment either declined or stabilized after ca.1997.
+Volume increment declined in all units from 2009-2019, and in the control unit, annual volume increment was lower in 2019 than in 1992.
 
@@
 
-```{r compare sph and BA differences between periods and diam classes}
-perComp<-
+```{r compare species proportions as per Deal 2002}
+
+sppProp.Dat<-
   dat %>%
@@
   mutate(DBH.class=cut(DBH,breaks=c(7.5,12.5,18.5,30,90),include.lowest = TRUE,right=FALSE)) %>%
-  left_join(plotDat,by="Plot")
 
-# fit models
-fit1<-  
-  perComp %>%
-
-  # come up with a denominator (number of plots per BA.Target per year)
-  distinct(Plot,.keep_all = TRUE) %>% 
-  group_by(BA.Target) %>% 
-  summarise(Num.Plot=n()) %>% 
-  right_join(perComp,by="BA.Target") %>%
-
-  # Plot averages
-  group_by(Plot,Year,BA.Target,Num.Plot,Species,DBH.class) %>% 
-  summarise(BA=sum(BA)*20,# Notice the plot multiplier
-            Vol=sum(VolTot)*20,
-            SPH=n()*20) %>% 
-  filter(DBH.class=="[30,90]") %>% 
-  filter(Species=="Sx") %>% 
-  lm(BA~BA.Target+Year,data=.)
-
-# plotting
-perCompPlot<- # assign variable  
-  perComp %>%
-
-  # come up with a denominator (number of plots per BA.Target per year)
-  distinct(Plot,.keep_all = TRUE) %>% 
-  group_by(BA.Target) %>% 
-  summarise(Num.Plot=n()) %>% 
-  right_join(perComp,by="BA.Target") %>%
-
-  # Plot averages
-  group_by(Plot,Year,BA.Target,Num.Plot,Species,DBH.class) %>% 
+  # Calculate plot BA, Volume and SPH by species, plot and DBH.class 
+  group_by(Plot,Species,Year,DBH.class) %>%
    summarise(BA=sum(BA)*20,# Notice the plot multiplier
@@
 
-  #BA Target averages
-  group_by(BA.Target,Year,Species,Num.Plot,DBH.class) %>% 
-  summarise(BA.sum=sum(BA),
-            BA.se=sd(BA)/sqrt(length(BA)),
-            Vol.sum=sum(Vol),
-            Vol.se=sd(Vol)/sqrt(length(Vol)),
-            SPH.sum=sum(SPH),
-            SPH.se=sd(SPH)/sqrt(length(SPH))) %>% 
-  ungroup() %>%
-
-  # Select only for Sx and Bl
-  filter(Species %in% c("Bl","Sx")) %>%
-
-  # Compute means
-  mutate(BA.mean=BA.sum/Num.Plot,
-         Vol.mean=Vol.sum/Num.Plot,
-         SPH.mean=SPH.sum/Num.Plot) %>%
-
-  # format
-  dplyr::select(BA.Target,Year,Species,DBH.class,BA=BA.mean,Vol=Vol.mean,SPH=SPH.mean) %>% 
-  pivot_longer(cols=BA:SPH,names_to="var",values_to="value")
-
-vol.plot2<-  
-  perCompPlot %>% 
-  filter(var=="Vol") %>%
+  # Calculate species proportions
+    filter(Species %in% c("Bl","Sx")) %>% # filter for Sx and Bl
+    pivot_longer(cols=BA:SPH,names_to="Var",values_to="Value") %>% 
+ pivot_wider(names_from=Species,values_from=Value) %>% 
+  replace_na(list(Bl=0,Sx=0))      # replace NA with 0
 
 # Plotting
+sppProp.Plot<-
+  sppProp.Dat %>% 
+  mutate(Total=Bl+Sx) %>% 
+  mutate(Bl=Bl/Total,
+         Sx=Sx/Total) %>% 
+  pivot_longer(cols=Bl:Sx,names_to="Species",values_to="Prop") %>% 
+  left_join(plotDat,by="Plot") %>% 
+  group_by(BA.Target,Year,DBH.class,Var,Species) %>% 
+  summarise(Prop=mean(Prop)) %>%
+  mutate(TU=paste("TU",BA.Target,sep=":")) %>% 
+  filter(Var=="Vol") %>% 
   ggplot()+
-  aes(x=DBH.class,y=value,fill=Species,group=Species)+
-  geom_bar(stat="Identity",position="dodge")+
-  facet_grid(Year~BA.Target)
+  aes(x=DBH.class,y=Prop,group=Species,fill=Species)+
+  geom_bar(stat="Identity")+
+  facet_grid(Year~TU)+
+  ylab("Proportion")+
+  xlab("Diameter class (cm)")+
+  theme(legend.position="bottom",
+        axis.text.x = element_text(angle = 270,vjust=0.2,hjust=0))
 
-BA.plot2<-  
-  perCompPlot %>% 
-  filter(var=="BA") %>%
+sppVol.Plot<-
+  sppProp.Dat %>% 
+  pivot_longer(cols=Bl:Sx,names_to="Species",values_to="Val") %>% 
+  left_join(plotDat,by="Plot") %>% 
+  group_by(BA.Target,Year,DBH.class,Var,Species) %>% 
+  summarise(Val=mean(Val)) %>%
+  mutate(TU=paste("TU",BA.Target,sep=":")) %>% 
+  filter(Var=="Vol") %>% 
+  ggplot()+
+  aes(x=DBH.class,y=Val,group=Species,fill=Species)+
+  geom_bar(stat="Identity")+
+  facet_grid(Year~TU)+
+  ylab("Volume (m³/ha)")+
+  xlab("Diameter class (cm)")+
+  theme(legend.position="none",
+        axis.title.x=element_blank(),
+        axis.text.x=element_blank())
 
-  # Plotting
-  ggplot()+
-  aes(x=DBH.class,y=value,fill=Species,group=Species)+
-  geom_bar(stat="Identity",position="dodge")+
-  facet_grid(Year~BA.Target)
+sppProp.plot<-plot_grid(sppVol.Plot,sppProp.Plot,nrow=2,rel_heights=c(1,1.5),axis="lr",align="h")
 
-SPH.plot2<-  
-  perCompPlot %>% 
-  filter(var=="SPH") %>%
-
-  # Plotting
-  ggplot()+
-  aes(x=DBH.class,y=value,fill=Species,group=Species)+
-  geom_bar(stat="Identity",position="dodge")+
-  facet_grid(Year~BA.Target)
-
+figNums("sppProp.plot",display=FALSE)
 ```
@@
 
-`r figNums("volPlot.plot",caption="Variation in volume per plot (m³ per hectare) at five measurement years, faceted by treatment unit.")`
+`r figNums("volPlot.plot",caption="Variation in total volume per plot (m³ per hectare) at five measurement years, faceted by treatment unit.")`
 
@@
 Units are as follows: basal area (BA, m² per hectare); QMD (quadratic mean diameter,cm), trees per hectare (SPH, count), Merchantable volume (Vol.Merch, m³ per hectare); Total volume (Vol.Merch, m³ per hectare)")`
+
+##### page break
+```{r plot total volume species proportion,fig.height=6}
+
+sppProp.plot
+
+```
+
+`r figNums("sppProp.plot",caption="Proportion of mean plot volume by four diameter classes of fir and spruce, expressed as volume (m³ per hectare, top panel) and proportion in bottom panel.
+Almost identical proportions were obtained when comparing basal area and tree density.")`
 

### 2020-06-05 manuscript/EP1162_manuscript_draft.Rmd
manny work

@@
 # Discussion
+Partial cutting silviculture systems aim to affect the growth and structure of a stand primarily through the manipulation of stand density, tree spatial distribution, tree size distribution and species composition.
+The growth and regeneration response of the residual stand will vary with type of tree removal (ie thinning from below or above, or mixed), residual stand structure, silvical characteristics of the residual species, site productivity, tree distribution, stand age and health, and other site-level factors.
+We note that the literature contains ample studies of commercial thinning in even aged stands (references), however, studies of selection harvest systems in uneven-aged stands such as this study are relatively rare.
+
+Consistent with other studies, we found that tree-level growth increment (basal area and volume) and recruitment of new trees into various size classes varied by species, tree size at the time of treatment, and level of basal area removal.
+These changes in growth and stand structure resulted in stand-level changes that suggest the following, specific to uneven-aged spruce-fir forests in our study region:
+
+1. Higher levels of basal area removal can stimulate a stronger and more sustained growth response, resulting in overall more cumulative volume gains; 
+2. Partial cutting can be effectively applied to maintain similar proportions of spruce and fir;
+and 
+3. Different levels of partial cutting can effectively maintain a typical uneven-aged stand structure.
+
+## Growth responses to partial cutting
+Partial cutting in mixed-species uneven-aged stands requires an understanding of species- and size-specific growth responses to different levels of stand manipulation [@Deal2002].
+Our results suggest that a higher level of basal area removal (i.e., a residual basal area (RBA) of 10m^2^ in 1992) stimulated a stronger and more sustained growth (both volume and basal area increment) response than the high RBA treatment unit and the control unit.
+Over each period, basal area increment was higher in the low RBA treatment units compared to other two, and peaked ca.
+20 years after treatment, compared to  ca.
+5 years in high RBA and control units.
+At the tree-level, spruce trees grew faster than fir, and in both species, larger trees responded with more growth increment than smaller trees (although this was reversed when increment was expressed as a percentage of original tree size).
+
+### Species- and tree-size factors influencing growth responses
+
+Our results show that both species had initial delayed growth response after treatment.
+Trees acclimated to shady conditions may suffer a shock after release due to relatively rapid changes in light, temperatures and humidity which can increase evapotranspiration and respiration stress [get sources from @Wang2011].
+A post-release lag in height growth response may also reflect the tree's reallocation of of carbon to below-ground structures at the expense of aboveground, as per the functional balance theory [@Kneeshaw2002; @Brouwer1962]; a relatively immediate response in basal area increment response after treatment may represent a species strategy to thicken stem afer release in response to increased wind stress [@Wang2011].
+
+### Spruce
+
+### Fir
+In an overstory removal experiment near our study region, understorey subalpine fir height growth responded best over three years to partial canopy removal, whereas radial growth responded strongest under complete canopy removal [@Wang2011].
+Subalpine fir height growth response after overstory thinning varied with tree size and level of removal; trees of all sizes responded with positive height growth responses after partial canopy removal, whereas small tree also responded with positive growth after complete removal, which conversely caused height growth reductions in both intermediate and large trees [@Wang2011].
+Radial growth was highest under a complete overstory removal, especially in intermediate trees, after an initial lag [@Wang2011].
+
+### Tree size
+Across both species and all treatments, post-harvest growth increment increased with residual tree size.
+This likely reflects that larger trees simply put on more growth in terms of basal area or volume than smaller trees.
+This also suggests that the trees in this study were not so large that they could not respond to changes in stand conditions [@reference].
+Intermediate crown-class trees may respond better to partial cutting, because the treatment will result in a greater shift in light conditions for these trees compared to larger trees occupying a dominant or co-dominant crown class[@Wang2011].
+Intermediate trees also may have higher vigour than suppressed trees, with a greater corresponding ability to release after treatment [@Wang2011].
+
+## Stand structure dynamics over time
+
+All treatment units maintained a stand structure analagous to an uneven-aged stand (ie reverse-J diameter distribution).
+In the low RBA unit, most recruitment was into the smallest size classes, whereas recruitment was into larger size classes in the high RBA and control units.
+This suggests that understory response may increase with level of change in the canopy, and that  more substantial changes to the stand conditions can create more favourable conditions for understory suppressed and intermediate trees to release and increase size.
+We did not account for trees smaller than 7.5cm DBH in the study, therefore, we do not have data on seedling germination and establishment after treatment.
+THe increase of tree density in the lower size classes in the low RBA may have been a result of increased seedling establishment after harvest.
+However, our data show that tree density increased in the smaller size classes in the low RBA unit.
+Natural regeneration is complex and in some cases can be largely stochastic, which can make its use difficult to achieve management objectives [@Bataineh2013].
+
+Within a time period, spruce quadratic means diameters were inversely related to harvest level.
+QMD was highest in the control unit, and lowest in the high-removal treatment unit, suggesting that there are fewer but larger trees in the control, whereas recuitment into sapling class (>7.5cm DBH) trees was higher in high-removal TU.
+
+## Management implications
+Partial harvesting
+
+***
+Draft material below this line
+
 Here are the main results that emerge to me from our data:
 
-1.
-All treatment units maintained a stand structure analagous to an uneven-aged stand (ie reverse-J diameter distribution).
-In the high-removal TU, most recruitment was into the smallest size class (pole), whereas recruitment was into larger size classes in the light-removal and control.
-This suggests that understory response may increase with level of change in the canopy.
+Silviculture treatments aim to affect the growth of stands through the manipulation of stand density, composition, distribution.
+Effects of stand density and composition management on growth and yield will vary with type of treatment (from below or above), site productivity, silvics of the residual species, tree distribution (clumped, dispersed, etc.), stand age, forest health factors and other factors.
+
+Partial removal of the overstorey can allow forest managers to manipulate light levels in the stand, however, the successful application of this treatment requires an understanding of how the understory trees will respond [@Wang2011].
+
+Across species, removal intensity and size classes, trees showed a growth response after treatment.
+
 \newLine
@@
 
+## Density-growth relationship
 2. High-level of basal area removal stimulated a stronger and more sustained growth response
@@
 
-3. Partial cutting in this uneven-aged spruce-fir forest affected spruce-fir composition differently depending on harvest level
+## Species differences in growth response
+
+### Spruce
+
+## Species composition
+In a retrospective analysis of partially cut hemlock-sitka spruce stands in Alaska, partial cutting had little effect on species composition after treatment [@Deal2002].  
+Partial cutting in this uneven-aged spruce-fir forest affected spruce-fir composition differently depending on harvest level, as well as tree sizes.
 - Stand development after harvest showed that spruce volume, basal area, and QMD increased more in the high-removal treatment compared to the other two treatments.
-In many plots in the high-removal TU, spruce volumes exceeds control/light-removal spruce.
+In several plots in the high-removal TU, spruce volumes exceeds control/light-removal spruce.
 - High removal stimulated stronger growth responses in spruce trees, compared to fir.
 In the light-removal and control units, spruce basal area and volume changed much less over the 27 years of the trial, and in fact, have similar levels.
-- Within a time period, spruce quadratic means diameters were inversely related to harvest level.
-QMD is highest in the control unit, and lowest in the high-removal treatment unit, suggesting that there are fewer but larger trees in the control, whereas recuitment into sapling class trees has been highest in high-removal TU.
-Does this suggest that spruce are well adapted to take advantage of canopy gaps?
+- 
 - Would we see this response in even more removal, say down to 5m^2^ per hectare?
@@
 \newLine
+
+5. Management implications
 

### 2020-06-11 manuscript/EP1162_manuscript_draft.Rmd
discussion work

@@
 
-1. Higher levels of basal area removal can stimulate a stronger and more sustained growth response, resulting in overall more cumulative volume gains; 
-2. Partial cutting can be effectively applied to maintain similar proportions of spruce and fir;
+1. Higher levels of basal area removal can stimulate a stronger and more sustained growth response; 
+2. Partial cutting can be used to maintain pre-harvest species composition (in other words, maintain proportions of spruce and fir that are similar to pre-harvest stand composition);
 and 
-3. Different levels of partial cutting can effectively maintain a typical uneven-aged stand structure.
+3. Different levels of partial cutting can maintain a stand structure with trees well represented in different size classes (commonly referred to as uneven-aged stand structure).
 
@@
 Partial cutting in mixed-species uneven-aged stands requires an understanding of species- and size-specific growth responses to different levels of stand manipulation [@Deal2002].
-Our results suggest that a higher level of basal area removal (i.e., a residual basal area (RBA) of 10m^2^ in 1992) stimulated a stronger and more sustained growth (both volume and basal area increment) response than the high RBA treatment unit and the control unit.
+Our results suggest that a higher level of basal area removal (i.e., a residual basal area (RBA) of 10m^2^ in 1992) stimulated a stronger and more sustained tree-level growth (both volume and basal area increment) response than the high RBA treatment unit and the control unit.
 Over each period, basal area increment was higher in the low RBA treatment units compared to other two, and peaked ca.
-20 years after treatment, compared to  ca.
+20 years after treatment, compared to an increment peak of ca.
 5 years in high RBA and control units.
 At the tree-level, spruce trees grew faster than fir, and in both species, larger trees responded with more growth increment than smaller trees (although this was reversed when increment was expressed as a percentage of original tree size).
+
+## Stand-level changes
+
+A stand's cumulative volume over time can be estimated with stand density and periodic increment [@Pretzsch2005].
 
@@
 Draft material below this line
+
+-	Our study was conducted in a relatively homogenous site (most plots are located on sites with circum-mesic soil edaphics) which represents the majority of managed forests in this region.
+The application of partial harvesting on azonal sites should be guided by site-specific data, as site productivity was found to have an important (and potentially opposite) effect on stand dynamics after harvesting [@Pretzch2005].
+-	Stand-level periodic increment follows a unimodal density optimum pattern [@Pretzsch2005], which can be used by silviculturists to manipulate density in order to maximize periodic increment and cumulative volume.
+Our results suggest that higher levels of basal area removal resulted in tree-level growth rate increases that were higher and more sustained than the high RBA treaetment unit and control.
+After approximately 20 years, the low RBA unit had converged with the high RBA unit in terms of stand-level basal area and volume, suggesting a higher cumulative volume in the low RBA unit.
+Presumably, an even higher level of removal (e.g., residual basal area of 5m2/ha) could have resulted in a reduced stand-level periodic increment and subsequent lower cumulative volume, as increased growth rates would be offset by lower stand density, reflecting the unimodal density optimum pattern [@Pretzsch2005].
+This density optimum may also vary with stand age [@Pretzsch2005].
+-	By three decades after treatment, both treatment units still had lower volume than the control, however, the differences between the three units diminished over time.
+Given enough time, stand volumes may converge.
+A study in thinned Norway spruce stands found that cumulative volumes in thinned treatments exceeded unthinned volumes by 30-39 years after treatment [@Pretzsch2005]; however, it may not be suitable to compare these fiindings to our study, as the treatment was a thin from below.
 

### 2020-06-22 manuscript/EP1162_manuscript_draft.Rmd
c

@@
 Our results suggest that a higher level of basal area removal (i.e., a residual basal area (RBA) of 10m^2^ in 1992) stimulated a stronger and more sustained tree-level growth (both volume and basal area increment) response than the high RBA treatment unit and the control unit.
+
 Over each period, basal area increment was higher in the low RBA treatment units compared to other two, and peaked ca.
@@
 At the tree-level, spruce trees grew faster than fir, and in both species, larger trees responded with more growth increment than smaller trees (although this was reversed when increment was expressed as a percentage of original tree size).
+Longer periods (i.e., over 40 years) of increased growth rates have been documented after understorey release by insect attack in subalpine spruce-fir forests in Colorado, US [@Veblen1991].
 
@@
 A stand's cumulative volume over time can be estimated with stand density and periodic increment [@Pretzsch2005].
+
+Spruce trees in the understorey can have a clumpier distribution than subalpine fir, which may indicate that spruce has a narrower range of suitable microsites for regeneration [@Shea1985; @Klinka2000].
+On the other hand, subalpine fir can regenerate across a wider range of substrates, and therefore may be expected to proliferate more uniformly in the understorey [@Klinka2000].
+Thus, partial harvesting systems that aim to maintain spruce composition in the stand should aim to protect spruce understorey as much as possible, as well as create suitable microsites for seed germination such as exposed mineral soil [@Coates1994].
+Once established, the performance of planted spruce and subalpine may be more or less equal [@references].
+Understorey Engelmann spruce and subalpine fir displayed similar growth responses to release after bark beetle attack in a subalpine forests in Colorado [@Veblen1991].
 
@@
 Trees acclimated to shady conditions may suffer a shock after release due to relatively rapid changes in light, temperatures and humidity which can increase evapotranspiration and respiration stress [get sources from @Wang2011].
-A post-release lag in height growth response may also reflect the tree's reallocation of of carbon to below-ground structures at the expense of aboveground, as per the functional balance theory [@Kneeshaw2002; @Brouwer1962]; a relatively immediate response in basal area increment response after treatment may represent a species strategy to thicken stem afer release in response to increased wind stress [@Wang2011].
+A post-release lag in height growth response may also reflect the tree's reallocation of of carbon to below-ground structures at the expense of aboveground, as per the functional balance theory [@Kneeshaw2002; @Brouwer1962], whereas a relatively immediate response in basal area increment response after treatment may represent a species strategy to thicken stem afer release in response to increased wind stress [@Wang2011].
+
+Hybrid spruce and subalpine fir share similar silvics and life histories [@Klinka2000], and have demonstrated similar growth and survival in managed forests in British Columbia [@Lajzerowicz2006] and elsewhere [@Veblen1991].
+Thus, we may have anticipated similar growth responses between these species.
 
@@
 Natural regeneration is complex and in some cases can be largely stochastic, which can make its use difficult to achieve management objectives [@Bataineh2013].
+We note that in high-elevation spruce-fir stands in British Columbia, understorey tree age and diameter may be uncorrelated [@Puttonen1998].
 
@@
 
+We note that our data do not provide information on the effects of partial harvest removal on spatial distribution of tree stems, which may be an important consideration when applying silvicultural treatments [@Saunders2008].
+Spatial characteristics include spatial distribution of tree stems (e.g., clumped or even distribution), the degree of species intermingling, and size differences among neighbours, all of which can influence resource availablility, regeneration success and competition [cf @Saunders2008].
+
 ## Management implications
 Partial harvesting
+
+Both treatments resulted in a stand that would be considered fully stocked immediately after harvest according to the BC free growing guidelines [@reference].
 

### 2020-10-17 manuscript/EP1162_manuscript_draft.Rmd
manny edits

@@
 
+# Create a table for the appendix with growth rates
+piDat.table<-
+  dat %>% 
+  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
+  filter(TreeID %in% treeList) %>%
+  filter(Year %in% c(1992,2019)) %>% 
+  # mutate(BA=(DBH/2/100)^2*pi) %>% # convert DBH to basal area
+  dplyr::select(Plot,TreeID,Species,Year,DBH) %>% 
+  pivot_wider(names_from=Year,values_from=DBH) %>% 
+  mutate(BA92=(`1992`/2/100)^2*pi) %>% 
+  mutate(BA19=(`2019`/2/100)^2*pi) %>%
+  mutate(PI=BA19-BA92) %>% # 
+  filter(PI>0) %>%   # one tree has a PI of 0, has to be removed
+  mutate(initDiam=`1992`) %>% 
+  mutate(diamClass=cut(initDiam,breaks=c(7.5,12.5,18.5,30,90),include.lowest=T,right=F)) %>% 
+  left_join(plotDat,by="Plot") %>% 
+  mutate(BA.Target=paste("TU:",BA.Target,sep="")) %>%  # create facet column
+  group_by(BA.Target,Species,diamClass) %>% 
+  summarise(BAImean=mean(PI),
+            BAIse=sd(PI)/sqrt(length(PI))) %>% 
+  arrange(BA.Target,diamClass,Species)
+
 # Create figures of raw data to put into Appendix
@@
 appNums(name="piPCTInitRaw",display=FALSE)
+appNums(name="piDataRaw",display=FALSE)
+
 figNums(name="fitBA.plot",display=FALSE)

### 2020-11-05 analysis/dataExplore_PrelimResults.Rmd
init

---
title: "EP1162 Data and Prelim Results"
author: "Hardy Griesbauer"
date: "19/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load libraries
library(tidyverse)
library(lme4)
library(emmeans)
library(flextable)
library(officer)

# Load data
load(here::here("data","ep1162_Data.RData"))
load(here::here("data","ep1162_plotData.RData"))

# Merge data
X<-
  dat %>% 
  left_join(plotDat,by="Plot") %>%

  # replace species codes with words
  mutate(Species=recode_factor(Species,Bl="Fir",Sx="Spruce"))

# Set against scientific notation
options(scipen=999)

```

## Data exploration

### Question: How did tree diameter distribution vary by treatment and measurement period?

``` {r create histogram of diameter data}

# Create histogram of plots of DBH by species and basal area treatments
X %>% 
  filter(Species %in% c("Fir","Spruce")) %>% # drop Fdi and Ep trees
  ggplot()+
  aes(x=DBH,color=Species,fill=Species)+
  geom_histogram(fill="white", alpha=0.5, position="identity")+

  facet_grid(Year~BA.Target)

```

``` {r create density diagram of diameter data}

# Create histogram of plots of DBH by species and basal area treatments
X %>% 
  filter(Species %in% c("Fir","Spruce")) %>% # drop Fdi and Ep trees
  ggplot()+
  aes(x=DBH,color=Species,fill=Species)+
  geom_density(alpha=0.6)+
  facet_grid(Year~BA.Target)

```

*Note*  
- These are density plots, not true histograms.  
- They can be used to compare how the distribution of diameters changes among measurement periods within a single species.  
- They cannot be used to compare frequency of diameters between species nor among treatments.
- The number of plots by measurement period varies, which will explain some of the variability in this plot.

**Of interest**
- As might be expected, we see that trees got larger over time, and proportionally, more occupied larger diameter classes over time.
- We can see that spruce occupied the larger diameter classes (40+) more than fir.
- Subalpine fir generally maintained a reverse-J diameter distribution across all measurement periods and treatments.
- In the low-removal and control units, spruce maintained a more even diameter distribution.
- In the low-removal unit, we see that spruce started to develop a bimodal distribution in the last two measurment periods.  Does this mean more recruitment into the smaller size classes?
- In the high-removal treatment unit, spruce maintained a more reverse-J distribution.  More trees concentrated in the smaller size classes.

Implications:
- Subalpine fir seems to continuously recruit into smaller size classes
- Spruce trees form the largest trees in all treatment units
- Do the bimodal spruce diameter distributions in the low-removal and control units suggest some spruce seedling establishment and subsequent recruitment into the smaller size classes?

### Question: how did tree-level growth response vary with initial tree size, treatment and species?

To test the hypothesis that tree-level growth response over the 1992-2019 period varied with interactions between (i) tree diameter at the time of treatment (1992); (ii) residual basal area; and (iii) species, we fit linear mixed effects models with log-transformed 1992-2019 diameter increment as the response variable, and the interaction between treatment, species and log-transformed 1992 diameter as a fixed effect, and plot as a random intercept effect.  The figure below shows the output from the model:

```{r}

# Let's filter only for trees that have measurements in 1992 and 2019 
treeList<- # assign variable
  X %>% 
  filter(TreeID!="New") %>% #remove new trees
  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
  filter(Species %in% c("Spruce","Fir")) %>% # filter birch and fdi out of dataset
  filter(Status=="Live") %>% # filter for live trees
  filter(Year %in% c(1992,2019)) %>% #filter for start and end dates
  group_by(TreeID) %>% 
  summarise(N=n()) %>% 
  filter(N==2) %>% #filter for trees with 5 measurements
  pull(TreeID)

# derive periodic increment (period diameter or basal area increment)
piDat<-
  X %>% 
  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
  filter(TreeID %in% treeList) %>% 
  mutate(BA=(DBH/2/100)^2*pi) %>% # convert DBH to basal area
  dplyr::select(Plot,TreeID,Species,Year,DBH) %>% 
  pivot_wider(names_from=Year,values_from=DBH) %>% 
  mutate(PI=`2019`-`1992`) %>% # 
  filter(PI>0) %>% # one tree has a PI of 0, has to be removed
  mutate(PIpct=PI/`1992`*100) %>% # express periodic increment a a percentage
  dplyr::select(Plot,TreeID,Species,PI,PIpct,Init=`1992`) %>% 
  left_join(plotDat,by="Plot") %>% 
  mutate(BA.Target=paste("TU:",BA.Target,sep="")) # create facet column

# fit lmer model
# fixed effects model with species pooled, using PBAI
fitBA<-
  piDat %>% 
  lmer(log(PI)~BA.Target:log(Init):Species+(1|Plot),data=.)

# create estimates of marginal means
# We'll estimate marginal means at four different diameter classes (10,20,30,40cm DBH)
rg<-ref_grid(fitBA,at = list(Init = c(10, 20, 30, 40)),cov.keep=4)

```

```{r did spruce outgrow fir figure}

emmeans(rg,~Init|BA.Target+Species) %>% 
  as.data.frame() %>%

  # plotting
  ggplot()+
  aes(x=Init,y=emmean,fill=Species,color=Species)+
  geom_point()+
  geom_line()+
  geom_errorbar(aes(ymin=emmean-SE, ymax=emmean+SE), width=.2,
                 position=position_dodge(0.05))+
  xlab("1992 DBH (cm)")+
  ylab("Diameter growth from 1992-2019 (cm)")+
  facet_wrap(~BA.Target)

```

In this figure, we can see a lot of interesting relationships.  The question is, which ones are significant?  We can look at this in more detail with some contrasts (below):

### Question: At the tree-level, how did 27-year growth responses vary by species and initial tree size?

```{r did spruce outgrow fir table}

pairs(rg,by=c("BA.Target","Init"),type="response") %>% 
  as.data.frame() %>% 
  arrange(BA.Target,Init) %>% 
  rename(DBH.1992="Init") %>% 
  mutate_if(is.double,round,5) %>%  # round to five digits

  # format table

  # flextable
  flextable(col_keys = c(colnames(.)[c(1:2,6:8,3:5)])) %>%

  align(align="left",part = "all") %>% 
  vline( border = fp_border(color="#2c3a51" , width = 1), part = "all" ) %>% 
  hline( border = fp_border(color="#2c3a51" , width = 1), part = "all" ) %>% 
  merge_v(j=c(1:2,6:8)) %>% 
  valign(valign="top") %>%
 border_outer(part="all", border = fp_border(color="#2c3a51" , width = 2)) %>% 
   fix_border_issues()  # border(border.bottom = fp_border(color="black"))

```
This table contrasts 1992-2019 diameter growth between fir and spruce trees by (a) treatment and (b) initial DBH.  The contrast is expressed in the 'ratio' column, and is the ratio of modeled fir growth to spruce growth.  Values are based on estimated marginal means from the mixed effects model fitted previously.

Results indicate that spruce radial growth exceeded that of fir across all diameter classes (p=0.0001), but only in the high-removal unit.  Spruce and fir radial growth rates were close to equal in the low-removal unit (p=0.5849) and control unit (fir slightly less than spruce, p=0.09890).

### Question: At the species level, did tree growth vary with treatment?

```{r how did growth vary by treatment}

pairs(rg,by=c("Species","Init"),type="response") %>% 
  as.data.frame() %>% 
  arrange(Species,contrast,Init) %>% 
  rename(DBH.1992="Init") %>% 
  mutate_if(is.double,round,5) %>%

  # flextable
  flextable(col_keys = c(colnames(.)[c(2,1,6:8,3:5)])) %>% 
  border_outer(part="all", border = fp_border(color="#2c3a51" , width = 2)) %>% 
  align(align="left",part = "all") %>% 
  vline( border = fp_border(color="#2c3a51" , width = 1), part = "all" ) %>% 
  hline( border = fp_border(color="#2c3a51" , width = 1), part = "all" ) %>% 
  merge_v(j=c(1:2,6:8)) %>% 
  valign(valign="top") %>%
  hline_bottom() %>% 
  fix_border_issues()# border(border.bottom = fp_border(color="black"))

```

This table indicates that:

1. Spruce in the high-removal treatment grew faster (radial growth) than spruce in the other treatment units.  Our model suggests that spruce growth in the high-removal unit was almost four times higher than spruce radial growth in the control, for trees that were 40cm at the time of treatment.  However, this seems high, I would interpret this with caution.  Spruce growth in the low-removal and control units were close to equal (p=0.8).
2. Fir radial growth in both high- and low-removal units exceeded growth in the control unit (p=).  Fir diameter growth did not differ between high- and low-removal units.

# Stand-level changes

The following stand-level changes:

1. Quadratic mean diameters by species, TU and time
2. Volume and basal area by species, TU and time
3. Species diameter distributions
4. Species composition by: SPH and basal area, by diameter class

### 2020-11-05 analysis/questionforPeter.Rmd
init

---
title: "Question for Peter"
author: "Hardy Griesbauer"
date: "03/11/2020"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# load data
library(tidyverse)
library(emmeans)
library(lme4)

# Load data
load("~/Documents/Git/EP1162/analysis/standData.RData")

options(contrasts=c(factor="contr.sum", ordered="contr.poly"))

```

## Study background

The data were taken from a partial cutting silviculture trial established in 1992 near Prince George (EP1162).  The study is set up as follows:

1. Nineteen 50m x 50m sites were harvested with three levels of basal area retention (treatments were randomly assigned to each site);
2. The three levels of basal area retention are as follows: (i) control (no harvest, around 30m2/hectare in 1992); (ii) 10m2/hectare residual basal area (relatively high level of partial harvest); (iii) 20m2/hectare residual basal area (relatively light level of partial harvest);
3. Each of the 19 sites has one 11.28m fixed-radius permanent sample plot established in the centre;
4. Data were collected at the 19 plots over 5 years: 1992, 1994, 1997, 2009 and 2019.  Not all plots were measured at each period!

Below are the number of plots measured in each of the three treatment units and five years:

```{r}

standData2.X %>% 
  group_by(BA.Target,Year) %>% 
  summarise(Obs=n()) %>% 
  rename(Treatment="BA.Target") %>% 
  pivot_wider(names_from="Year",values_from="Obs") %>% 
  knitr::kable()
```

This table shows that in a given measurement year, data were collected in between 17-19 plots.  Most plots were visited in all five measurement years.

Here is a screenshot of some of the stand-level data:

```{r show data}
ep1162.dat<-
  standData2.X %>% 
  setNames(c("Treatment","Plot","Year","Volume","Basal.Area","SPH","QMD")) %>% 
  # mutate(Year=as.integer(as.character(Year))) %>% 
  as.tibble()

ep1162.dat %>% 
  head(10)

```

Here is a plot of mean plot-level basal area trend over time:
```{r plot stand basal area}

load("~/Documents/Git/EP1162/analysis/standBA.plot.RData")

standBA.plot
```

*Whiskers are standard error of the mean*

To me, it looks like the plots in TU:10 increased their basal area faster than plots in the other two units. This was the one of the study's hypotheses (opening the stand up more will stimulate faster growth).  To test that hypothesis, I would like to contrast the slope of the lines between treatments.

I believe I can do this in R with the following code:

```{r lmer, echo=TRUE}

fit.BA<-lmer(Basal.Area~Year*Treatment+(1|Plot),data=ep1162.dat)

summary(fit.BA)

```

## Contrasts between slopes
Using the `emmeans` package, I contrast the slopes as follows:

```{r test the interaction, echo=TRUE}
# emtrends(fit.BA, pairwise ~ Treatment, var = "Year")

library(lmerTest)
anova(fit.BA, type="3")

```

-----------------

## Generate cell means

```{r cell means,echo=TRUE}

# joint_tests(fit.BA) #another way to see the ANOVA table for the overall tests

cell.means <- emmeans(fit.BA, ~ Treatment*Year)

cell.means

```

```{r plot cell means with confint,echo=TRUE}
plot(cell.means, by="Year")

```

---------------------------

## Contrasts

This is the code with interaction="consec".
```{r contrasts,echo=TRUE}

year.by.trt <- contrast(cell.means, interaction = "consec", by = NULL) #might need to change the by = option here.

```

Need a slight change in code so that all treatments are contrasted.  Right now it's only contrasting consecutive treatments.  Let's change the interaction to "pairwise":

```{r pairwise, echo=TRUE}

year.by.trt <- contrast(cell.means,interaction="pairwise") #might need to change the

year.by.trt

```

------------------------------

## Test

```{r, echo=TRUE}
test(year.by.trt, joint=TRUE)
```

### 2020-11-05 book/bookdown-minimal-master/1_Methods.Rmd
init

# Methods

```{r load libraries for analyses, echo=FALSE, message=FALSE, warning=FALSE}

knitr::opts_chunk$set(echo = FALSE,message=FALSE,warning=FALSE)

# Load libraries
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom)
library(MuMIn)
library(ggeffects)
library(car)
library(flextable)
library(captioner)
library(lsmeans)
library(cowplot)
library(officer)
library(sjPlot)
library(emmeans)

load("~/Documents/Git/EP1162/data/ep1162_Data.RData")
load("~/Documents/Git/EP1162/data/ep1162_plotData.RData")

# SOME PRELIM DATA FORMATTING
# formatting plotDat
plotDat<-
  plotDat %>% 
  mutate(Plot=as.factor(Plot))

# set up diameter classes
diamClass=seq(from=7.5,to=100,by=5)

# Set against scientific notation
options(scipen=999)

# Merge data
X<-
  dat %>% 
  left_join(plotDat,by="Plot") %>% 
  mutate(DBH.class=cut(DBH,diamClass,include.lowest=TRUE,ordered_result = TRUE,right=F,
                       labels=diamClass[1:(length(diamClass)-1)])) %>% 
  mutate(Plot=as.factor(Plot)) %>%

  # replace species codes with words
  mutate(Species=recode_factor(Species,Bl="Fir",Sx="Spruce")) %>% 
  mutate(Species=fct_drop(Species)) %>% # have an extra level in here for some reason

  # Create TU lookup
  mutate(BA.Target=paste0("TU:",BA.Target))

# set up captions

# Code to create captions for tables
tableNums <- captioner(prefix = "Table")

# Code to create captions for figures
figNums <- captioner(prefix="Figure")

# Code to create captions for appendices
appNums <- captioner(prefix="Appendix")

# Set ggplot theme
theme_set(theme_bw())

# Define function to convert diameter (cm) to basal area (m2), and vice versa
d2ba<-function(diam) {return((diam/2/100)^2*pi)}
ba2d<-function(ba) {return(sqrt(ba/pi)*200)}

```

This section will describe the methods we used for data analyses, not field data collection.

From our technical report submitted to FCI earlier:

Our data analyses and summaries examined the effects of the different levels of stand-level basal area density treatments on three main attributes of stand development following treatment. These include, in order:

1. Post-treatment stand basal area dynamics, including the rate and pattern of basal area re-growth and recovery over time following different levels of initial basal area density reductions;
2. Stand structural outcomes, including treatment influences on the overall abundance of trees (in sph) by diameter class, within the different treatment types, and;
3. Tree species composition outcomes, including treatment influences on the relative abundance and size class distributions of subalpine fir and hybrid white spruce within the different treatment types.

Analyses are grouped into three main categories (below).  These are treated as separate categories in the Results section.

1. Tree-level responses to treatment;
2. Stand-level responses to treatment:

```{r define captions to be used in methods results and discussion, echo=FALSE}

# Captions (put these in order)

  # Introductory 
  appNums(name="paiTable",caption="Number of growing seasons between each measurement period",display=FALSE) # Table showing number of growing seasons between each measurement period
  appNums(name="standVol.plot",caption="Plot total volume for five measurement periods",display=FALSE) # Plot volume figure
  appNums(name="standBA.plot",caption="Plot basal area for five measurement periods",display=FALSE)  # Plot basal area figure
  appNums(name="standQMD.plot",caption="Plot quadratic mean diameter for five measurement periods",display=FALSE) # QMD figure  
  appNums(name="standK.plot",caption="Plot tree density for five measurement periods", display=FALSE) # SPH figure

  # Tree BAI model
  figNums(name="treeBAI.plot",display=FALSE)
  tableNums("fitBA.table",display=FALSE) # Summary of mixed effects model for BAI
  appNums("sppContrast.table",display=FALSE) # Species contrasts from treeBAI model
  appNums("trtContrast.table",display=FALSE) # Treatment contracts from treeBAI model

  # Tree heights model
  appNums("heightDiamRawPlot",display=FALSE) # Plot of diameter/heights`

  # Diameter class analysis
  figNums("diamVol.plot", caption="Mean plot volume distribution by diameter class and treatment, for two periods.  Black line shows total volume.  Volume calculated on trees meeting 18.5cm DBH limit.  Spruce and fir contribution to volume by diameter class are shown with bar plots.  Douglas-fir volume contribution by diameter class is not shown.",display="FALSE")

  figNums("diamBA.plot", caption="Mean plot basal area distribution by diameter class and treatment, for two periods.  Black line shows total basal area.  Spruce and fir contribution by diameter class are shown ith bar plots.  Douglas-fir contribution by diameter class is not shown.",display="FALSE")

  figNums("diamK.plot", caption="Mean tree density distribution by diameter class and treatment, for two periods.  Black line shows total tree density.  Spruce and fir contribution by diameter class are shown ith bar plots.  Douglas-fir contribution by diameter class is not shown.",display="FALSE")

```

## Tree-level growth responses

To test the hypothesis that tree-level radial growth response over the 1992-2019 period varied with interactions between (i) tree diameter at the time of treatment (1992); (ii) treatment; and (iii) species, we fit linear mixed effects models with log-transformed 1992-2019 diameter increment as the response variable, and treatment, species and log-transformed 1992 diameter as interacting fixed effects, and plot as a random intercept effect.  We also fit a similar model with the 1992-2019 diameter increment expressed as a percentage of 1992 diameter.  Plots of residuals and fit lines with orginal data were examined to evaluate model goodness of fit.  Variance explained for each model are reported using the approach described in Nakagawa et al. [-@Nakagawa2013], as implemented in the MuMIn package for R [@add later].

## Stand-level growth and structure

### Stand-level volume and basal area estimates
Total and merchantable volumes were estimated for trees using species- and region-specific equations [@nigh2016] based on tree height and DBH.  Consistent with forest management practices in British Columbia, we estimated volumes for trees meeting a minimum diameter utilization limit of 18.5cm DBH.  For trees lacking a tree height measurement, we estimated height by fitting a single diameter-height model developed with a linear mixed effects model, with log-transformed DBH as a fixed effect, and tree nested within plot as a random effect.  Trees with broken tops were omitted from diameter-height models. We summed tree-level estimates of volume and basal area to generate stand-level estimates for each plot. To test the hypothesis that changes in stand attributes over 27 years varied by treatment, we fit separate linear models with stand total volume, basal area, and density as response variables, and the interaction between treatment and year as the independent variable.

----------------

`r appNums("paiTable",caption="Number of growing seasons between measurement periods")`
```{r show paiTable, ft.align="left"}

  tibble(`Measurement Year`=c(1992,1994,1997,2009,2019),`Growing Seasons`=c(NA,3,3,11,11)) %>% 
  flextable() %>% 
  align(align="left",part="all") %>% 
  # set_caption(caption="Number of growing seasons between measurement periods") %>% 
  autofit()

```

----------------------

### 2020-11-05 book/bookdown-minimal-master/2_results.Rmd
init

# Results

## Tree-level

### Basal area increment from 1992-2019

```{r mixed effects model of tree growth}

# Let's filter only for trees that have measurements in 1992 and 2019 
treeList<- # assign variable
  X %>% 
  filter(TreeID!="New") %>% #remove new trees
  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
  filter(Species %in% c("Spruce","Fir")) %>% # filter birch and fdi out of dataset
  filter(Status=="Live") %>% # filter for live trees
  filter(Year %in% c(1992,2019)) %>% #filter for start and end dates
  group_by(TreeID) %>% 
  summarise(N=n()) %>% 
  filter(N==2) %>% #filter for trees with 5 measurements
  pull(TreeID)

# derive periodic increment (period diameter or basal area increment)
piDat<-
  dat %>% 
  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
  filter(TreeID %in% treeList) %>% 
  mutate(BA=d2ba(DBH)) %>% # convert DBH to basal area
  dplyr::select(Plot,TreeID,Species,Year,BA) %>% 
  pivot_wider(names_from=Year,values_from=BA) %>% 
  mutate(PI=`2019`-`1992`) %>% # 
  filter(PI>0) %>% # one tree has a PI of 0, has to be removed
  mutate(PIpct=PI/`1992`*100) %>% # express periodic increment a a percentage
  dplyr::select(Plot,TreeID,Species,PBAI=PI,PBAIpct=PIpct,Init=`1992`) %>% 
  left_join(plotDat,by="Plot")

# fit lmer model
  # fixed effects model with species pooled, using PBAI
  fitBA<-
    piDat %>% 
    lmer(log(PBAI)~BA.Target:log(Init):Species+(1|Plot),data=.)

  fitBA.summ<-
    list(table=tab_model(fitBA),
         caption="Summary of mixed effects model of tree-level growth")

  # create a reference grid for five different diameter classes (converted to BA)
  rg<-ref_grid(fitBA,at=list(Init=d2ba(c(10,20,30,40,50))),type="response")

  # Using PBAIpct
  fitBApct<-
    piDat %>% 
    lmer(log(PBAIpct)~BA.Target:log(Init):Species+(1|Plot),data=.)

  fitBApct.summ<-
    list(table=tab_model(fitBApct),
         caption="Summary of mixed effects model of tree-level periodic basal area % increment (PBAI %")

  # create a reference grid for five different diameter classes (converted to BA)
  rg.pbaipct<-ref_grid(fitBApct,at=list(Init=d2ba(c(10,20,30,40,50))),type="response")

```

```{r tree BAI by treatment species and diameter class}

# Create plot of periodic increment

treeBAI.plot<-

  emmeans(rg,~Init|BA.Target+Species,type="response") %>% 
  as.data.frame() %>%
  mutate(Init=ba2d(Init)) %>% # convert initial basal area back to diameter

  # plotting
  ggplot()+
  aes(x=Init,y=response,fill=Species,color=Species)+
  geom_point()+
  geom_line()+
  geom_errorbar(aes(ymin=response-SE, ymax=response+SE), width=.2,
                 position=position_dodge(0.05))+
  xlab("1992 DBH (cm)")+
  ylab("Basal area increment from 1992-2019 (m²)")+
  facet_wrap(~BA.Target)

```

```{r tree level fir spruce contrasts}

# set variable
sppContrast<-
  list(caption="Contrasts between predicted fir and spruce growth across treatment units and initial diameter classes.  Ratio between predicted fir and growth is in 'ratio' column.",

  table=

  pairs(rg,by=c("BA.Target","Init"),type="response") %>% 
  as.data.frame() %>%
    mutate(Init=as.numeric(as.character(Init))) %>% # change from factor to numeric
    mutate(Init=ba2d(Init)) %>% # convert basal area to diameter
  arrange(BA.Target,Init) %>% 
  rename(DBH.1992="Init") %>% 
  mutate_if(is.double,round,5) %>%  # round to five digits

  # format table

  # flextable
  flextable(col_keys = c(colnames(.)[c(1:2,6:8,3:5)])) %>%

  align(align="left",part = "all") %>% 
  vline( border = fp_border(color="#2c3a51" , width = 1), part = "all" ) %>% 
  hline( border = fp_border(color="#2c3a51" , width = 1), part = "all" ) %>% 
  merge_v(j=c(1:2,6:8)) %>% 
  valign(valign="top") %>%
 border_outer(part="all", border = fp_border(color="#2c3a51" , width = 2)) %>% 
   fix_border_issues()  # border(border.bottom = fp_border(color="black"))
) # close list

```

```{r treatment contrasts}

trtContrast<-list(
  caption="Contrasts of predicted 1992-2019 tree growth among treatments within each species.  Ratio of predicted growth between treatments is in 'ratio' column.",

# set up the table and place within list  
table=  
  pairs(rg,by=c("Species","Init"),type="response") %>% 
  as.data.frame() %>% 
  mutate(Init=as.numeric(as.character(Init))) %>% # change from factor to numeric
  mutate(Init=ba2d(Init)) %>% # convert basal area to diameter
  arrange(Species,contrast,Init) %>% 
  rename(DBH.1992="Init") %>% 
  mutate_if(is.double,round,5) %>%

  # flextable
  flextable(col_keys = c(colnames(.)[c(2,1,6:8,3:5)])) %>% 
  border_outer(part="all", border = fp_border(color="#2c3a51" , width = 2)) %>% 
  align(align="left",part = "all") %>% 
  vline( border = fp_border(color="#2c3a51" , width = 1), part = "all" ) %>% 
  hline( border = fp_border(color="#2c3a51" , width = 1), part = "all" ) %>% 
  merge_v(j=c(1:2,6:8)) %>% 
  valign(valign="top") %>%
  hline_bottom() %>% 
  fix_border_issues()# border(border.bottom = fp_border(color="black"))

# Close list
) # close list

```

The tree-level basal area increment model indicated that interactions between log-transformed tree size, species and treatment explained `r round(r.squaredGLMM(fitBA)[1]*100,1)`% (marginal R²) of the variation in tree log-transformed 1992-2019 BAI (`r tableNums("fitBA.table",display="cite")`).  Estimated marginal means from the model showed that between species and among treatments, larger trees increased basal area more than smaller trees (`r figNums("treeGrowth.plot",display="cite")`).  Across diameter classes, spruce trees had more basal area increment than fir in the high-removal treatment unit, whereas growth between the two species was close to equal in the low-removal and control units (`r appNums("sppContrast.table",display="cite")`).  Spruce basal area increment in the high-removal unit exceeded spruce growth in both other units (`r appNums("trtContrast.table",display="cite")`).  Spruce growth was similar between the low-removal and control units.  Fir basal area increment in the high- and low-removal units was higher than the control unit, and there were no significant differences between the two harvested units (`r appNums("trtContrast.table",display="cite")`).

----------------

`r tableNums("fitBA.table",caption=fitBA.summ$caption)`

```{r present tree growth model summary}

tab_model(fitBA)

```

---------------------------

```{r display tree growth figure}

treeBAI.plot

```

`r figNums("treeBAI.plot",caption="Estimated marginal means of spruce and fir basal area increment (m²) across five inital DBH classes and three treatments.  Whiskers are standard error of the mean.  Responses are back-transformed from the model.")`

--------------------

`r appNums("sppContrast.table",caption=sppContrast$caption)`

```{r print fir spruce contrasts, ft.align="left"}

# print to screen
sppContrast$table
```

-----------------------

`r appNums("trtContrast.table",caption=trtContrast$caption)`

```{r print treatment contrasts, ft.align="left"}

# print to screen
trtContrast$table

```

--------------------------

### Mortality

```{r mortality graphs}

# Create dead dataset

# Dead species totals
deadList<-
  X %>% 
  filter(Status=="Dead") %>% 
  mutate(ID=paste(Plot,TreeID,sep="-")) %>% 
  group_by(ID) %>% 
  summarise_all(first)

dead<-
  X %>% 
  filter(Status=="Dead") %>% 
  mutate(ID=paste(Plot,TreeID,sep="-")) %>% 
  mutate(Species=fct_recode(Species,`Douglas-fir`="Fd")) %>% 
  group_by(ID) %>% 
  summarise_all(first) %>% 
  ungroup() %>% 
  mutate(BA=d2ba(DBH)) %>% # convert to basal area per hectare 
  group_by(BA.Target,DBH.class,Year) %>% 
  summarise(BA=sum(BA),
            k=n())

dead.Summ<-
  dead %>% 
  ungroup() %>% 
  group_by(BA.Target) %>% 
  summarise(BA=sum(BA)/sum(dead$BA)*100,
            k=sum(k)/sum(dead$k)*100) %>% 
  mutate_if(is.numeric,round,1)

p1<-
  dead %>% 
  ggplot(aes(x=DBH.class,y=k,fill=Year))+
  geom_bar(stat="identity")+
  ylab("Total tree mortality (stems)")+
  facet_wrap(~BA.Target)+
  xlab("")+
  theme(axis.text.x = element_blank())

p2<-
  dead %>% 
  ggplot(aes(x=DBH.class,y=BA,fill=Year))+
  geom_bar(stat="identity")+
  ylab("Total basal area mortality (m²)")+
  facet_wrap(~BA.Target)

mort.plot<-cowplot::plot_grid(p1, p2, ncol = 1, align = "v")
figNums("mort.plot",display=FALSE)

```

A total of `r deadList %>% nrow()` trees (`r deadList %>% filter(Species=="Fir") %>% nrow()` fir, `r deadList %>% filter(Species=="Spruce") %>% nrow()` spruce, and `r deadList %>% filter(Species=="Fd") %>% nrow()` Douglas-fir) died from 1992 to 2019.  No tree mortality was noted during the 1992 and 1994 data collection.  By treatment unit, the control unit had the highest mortality, in terms of trees and basal area (
`r dead.Summ %>% filter(BA.Target=="TU:Control") %>% pull("BA")` and 
`r dead.Summ %>% filter(BA.Target=="TU:Control") %>% pull("BA")`% of tree mortality, respectively).  The high-removal treatment unit had the lowest mortality by number of trees and basal area (
`r dead.Summ %>% filter(BA.Target=="TU:10") %>% pull("BA")` and 
`r dead.Summ %>% filter(BA.Target=="TU:10") %>% pull("BA")`% of tree mortality, respectively).  Tree mortality was concentrated in the smaller diameter classes in the control and low-removal treatment unit, and most mortality occured during the most recent measurement period of 2009-2019 (
`r figNums("mort.plot",display="cite")`).

```{r display mortality plot}
mort.plot
```

`r figNums("mort.plot",caption="Total tree mortality recorded in plots from 1992-2019, presented as total stems (top panel) and basal area (bottom panel).")`

---------------------------

From Mike's email:
Eventually as BA stand density increases, density dependent mortality kicks in, because there is only so much growing space or access to site resources to go around. This can occur with the stand (conceptually or hypothetically)  in two possible ways: (a) “classic” suppression of smaller tree classes through asymmetric competition of bigger trees shading out smaller ones, or (b) high stand densities suppressing the vigor of the whole stand through various means, suggesting more symmetric competition – i.e. the idea that small trees compete for growing space with big ones as well. Asymmetric competition is generally accepted by most foresters and silviculturists; symmetric competition less so, and is often harder to demonstrate.

*For the discussion: it seems that we're seeing mortality increasing in the plots, especially the control unit over the last 11 years.  Does this reflect the density-dependent mortality that Mike refers to above?.*

More from Mike on mortality:

The thing about unevenaged or complex stand management, esp in ‘wilder’ BC stands, is that we are working in stands with a long legacy of pre-existing ecosystem processes before we enter the stand to manipulate them. Clearcutting liquidates these legacies for the most part, and traditional silviculture and GY has not had to deal with mortality much, other than clearly density dependent suppression of smaller dbh classes. [Side issue: most BC experience and data with unevenaged stands has been in IDF, which have been fire excluded on dry sites, and unevenaged structures allowed to develop. We must be careful about comparing Summit data to IDF precedents, because of substantial differences in ecology.]

And we are working a more extensive level of management in the SBS and ESSF,  where our stands are older, and not manicured nor vacuumed up for mortality and blowdown. From an ecosystem management point of view, that Is probably a good thing.

So warts and all, I think mortality is an inherent characteristic of the types of spruce-balsam stands and management we are studying. Expecting and documenting mortality was built into the EP 1162 working plan. The treatment effects we anticipate are the net stand growth = gross growth minus mortality.

Conceptually, tree mortality following the 1991/92 harvest (and subsequent sanitation felling)  functions as a BA removal from the live-tree pool. In most plots, BA of trees that die within a measurement period is subsumed or overridden by the total BA growth of the rest of the trees during that period.

Where the periodic data trends get messier is where single big trees or a cluster of trees gets killed by something, resulting in a pulse of BA mortality. Examples: The great big Fdi in Plot 18, or notably, about 3-4 big Sx in Plot 3 that got nailed by spruce beetle.

In our results and discussion:
1.	We can report on what stand growth and development trends seem to be clearly and unambiguously  influenced by initial BA conditions, and the magnitude of these treatment effects. 
2.	A component of tree mortality (e.g. certain size classes) may be density dependent. 
3.	A component of tree mortality may be density-independent (a.k.a stochastic or chaotic). 
4.	By reporting on all the above, we are providing managers with results that integrate silvicultural expectations observations from real-life “messy” stands which will experience all of the above. By keeping our eye on overall stand growth and performance following treatment, we can inform / reassure managers that individual tree mortality events are not the end of the world, but rather need to be viewed in the context of overall stand dynamics and growth expectations.

## Stand-level analyses

```{r height to diameter models}

heightDiamRaw.plot<-
  X %>% 
    filter(Species %in% c("Bl","Sx")) %>% # filter for Bl and Sx for now
    filter(Status=="Live") %>% # filter for live trees
    filter(is.na(DBT)) %>% # filter for trees with no broken tops
    drop_na(Height) %>% # remove trees where height was not measured

    # Plotting
  ggplot()+
  aes(x=DBH,y=Height,fill=Species)+
  geom_point()+
  facet_grid(Year~Species)+
  theme(legend.position = "none")+
  ylab("Height (m)")+
  xlab("DBH (cm)")+
  geom_smooth(formula = y~log(x),method="lm",se=FALSE)

# Fit model
htDiamFit<- 
  X %>% 
    filter(Species %in% c("Fir","Spruce")) %>% # filter for Bl and Sx for now
    filter(Status=="Live") %>% # filter for live trees
    filter(is.na(DBT)) %>% # filter for trees with no broken tops
    drop_na(Height) %>% # remove trees where height was not measured
  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create tree ID column
  lmer(log(Height)~log(DBH)+(1|Plot/TreeID),data=.)

# Apply fit model to data and create new height column
X<-
  X %>% 
  mutate(Height.Pred=coalesce(Height,exp(fixef(htDiamFit)[1]+fixef(htDiamFit)[2]*log(DBH))))

# Height model fit plot
htModelFit.plot<-
  X %>% 
  filter(is.na(DBT)) %>% # filter for trees with no broken tops
  mutate(h2=exp(fixef(htDiamFit)[1]+fixef(htDiamFit)[2]*log(DBH))) %>% 
  ggplot(aes(x=Height,y=h2))+
  geom_point()+
  xlim(0,40)+
  ylim(0,40)+
  geom_smooth(method="lm",se=F)+
  ylab("Predicted height (m) from diameter-height model")+
  xlab("Measured height (m)")

appNums("htModelFit.plot",display="FALSE") # model output

```

### Estimating tree height from diameter
Tree heights had a positive nonlinear relationship with diameter, for both species and across all time periods (`r appNums("heightDiamRawPlot",display="cite")`).  A comparison of height-diameter models showed that intercepts and slopes of log-transformed heights did not vary significantly (i.e., p>0.05) between species or time periods (not shown), therefore a single height-diameter linear mixed-effects model was developed using `r X %>% filter(Species %in% c("Fir","Spruce")) %>% filter(Status=="Live") %>% filter(is.na(DBT)) %>% drop_na(Height) %>% nrow()` tree height-diameter observations in the dataset, with log-transformed diameter as the fixed effect, and tree nested within plot as random effects. This model explained `r round(MuMIn::r.squaredGLMM(htDiamFit)[1]*100,1)`% of variation in log-transformed heights, with an intercept of `r round(fixef(htDiamFit)[1],3)` and slope coefficient of  `r round(fixef(htDiamFit)[2],3)` * log-transformed DBH (not shown).  Residual and predicted vs actual height plots were assessed to ensure goodness of fit (`r appNums("htModelFit.plot",display="cite")`). This model was applied to predict heights in trees without height measurements, and tree-volume estimates generated as per Nigh [-@nigh2016].

----------------------------
```{r height model fit}

htModelFit.plot

```

---------------

`r appNums("htModelFit.plot",caption="Scatterplot of predicted height(m) vs actual height(m) from diameter-height model.")`

```{r hardcode volume coefficients per Nigh}

volCoef<-

  ## TOTAL VOLUME
  #  Subalpine fir 
  data.frame(Species="Fir",VolType="total",b0=-9.79,b1=1.813,b2=1.033) %>%

  # Douglas-fir
  add_row(Species="Fd",VolType="total",b0=-10.19,b1=1.73,b2=1.206) %>%

  # Add spruce
  add_row(Species="Spruce",VolType="total",b0=-10.024,b1= 1.801,b2= 1.107) %>%

  ## MERCHANTABLE VOLUME
  # Subalpine fir
  add_row(Species="Fir",VolType="merch",b0=-9.980,b1= 1.792,b2= 1.102) %>%

  # Douglas-fir
  add_row(Species="Fd",VolType="merch",b0=-10.503,b1= 1.724,b2= 1.296) %>%

  # Spruce
  add_row(Species="Spruce",VolType="merch",b0=-10.280,b1= 1.769,b2= 1.206 ) %>%

  mutate(BGC.Zone="SBS") %>%

  # convert species to factor
  mutate(Species=factor(Species)) %>% 
  dplyr::select(Species,BGC.Zone,everything())

```

```{r summarize stand attributes by diameter class}

# We will produce diameter class stand attributes for two time periods 1992 and 2019.

standDiam.X<-

  # Determine which plots have measurements in 1992 and 2019
  X %>% 
   filter(Year %in% c(1992,2019)) %>% 
   filter(Status=="Live") %>%

  group_by(Plot,Year) %>% 
  summarise(n=n()) %>% 
  ungroup() %>% 
  group_by(Plot) %>% 
  summarise(n=length(Year)) %>% 
  filter(n>1) %>% # pick plots with measurements in both years
  pull(Plot) %>% 
  as.data.frame() %>% 
  setNames("Plot") %>%

  # Now filter original dataset 
  left_join(X,by="Plot") %>% 
  mutate(Plot=fct_drop(Plot)) %>%  # drop unused levels

  # Filter for years of interest
  filter(Year %in% c(1992,2019)) %>% 
  mutate(Year=fct_drop(Year)) %>% 
  filter(Status=="Live") %>%

    # Calculate volume
  left_join(filter(volCoef,VolType=="total"),by="Species") %>% 
  mutate(vol=exp(1)^b0*DBH^b1*Height.Pred^b2) %>% 
  mutate(BA=d2ba(DBH)) %>% 
  dplyr::select(BA.Target,Plot:Species,Year,DBH,BA,vol,DBH.class,Height.Pred) %>% 
  mutate(Species=factor(Species)) %>%

 # Now create plot-level stats
  ungroup() %>% 
  group_by(BA.Target,Plot,Species,Year,DBH.class) %>% 
  summarise(BA=sum(BA)*20,
            Vol=sum(vol)*20,
            SPH=n()*20) %>% 
  ungroup()

# Now summarize periodic diameter class stand attributes by species

# # Create total summaries
#   diamClass.total<-
#     standDiam.X %>% 
#     
#     # create plot totals first
#     group_by(BA.Target,Plot,Year,DBH.class) %>% 
#     summarise(BA=sum(BA),
#               Vol=sum(Vol),
#               SPH=sum(SPH)) %>% 
#     mutate(Species="Total") %>% 
#     ungroup() %>% 
#     
#     # now we have to complete the data frame to create BA.Target means
#     complete(nesting(BA.Target,Plot),Year,DBH.class,fill=list(BA=0,SPH=0,Vol=0)) %>% 
#     group_by(BA.Target,Year,DBH.class) %>% 
# 
#     # Plot means
#     summarise(BA=mean(BA),
#               Vol=mean(Vol),
#               SPH=mean(SPH)) %>% 
#     ungroup() %>% 
#     mutate(Species="Total") %>% 
#     filter(BA>0) %>% 
#     dplyr::select(BA.Target,Species,Year,DBH.class,BA,Vol,SPH)

  diamClass.spp<-
     standDiam.X %>% 
    # filter(Species %in% c("Fir","Spruce")) %>%  
    # create plot totals first
    filter(Species!="Ep") %>% # remove birch to clarify
    mutate(Species=fct_drop(Species)) %>% # drop Ep from levels

    # now we have to complete the data frame to create BA.Target means
    complete(nesting(BA.Target,Plot),Species,Year,DBH.class,fill=list(BA=0,SPH=0,Vol=0)) %>% 
    group_by(BA.Target,Species,Year,DBH.class) %>%

    # Plot means
    summarise(BA=mean(BA),
              Vol=mean(Vol),
              SPH=mean(SPH)) %>% 
    ungroup() %>% 
    filter(BA>0) %>% 
    dplyr::select(BA.Target,Species,Year,DBH.class,BA,Vol,SPH) %>%

    # reorder Species levels for plotting
    mutate(Species=fct_relevel(Species, "Fd", after = Inf)) %>% 
    mutate(Species=fct_recode(Species,`Douglas-fir`="Fd"))

diamClass.plot<-function(standVar,y.label) {

  diamClass.spp %>% 
   ggplot()+
  geom_bar(aes(x=DBH.class,y=get(standVar),group=Species,fill=Species),stat="identity")+
  # geom_point(data=filter(diamClass.spp,Species=="Total"),
  #              aes(x=DBH.class,y=get(standVar)))+
  #   geom_line(data=filter(diamClass.spp,Species=="Total"),
  #              aes(x=DBH.class,y=get(standVar),group=Species))+  
   scale_fill_manual(values=c("#F8766D", "#00BFCA","azure4")) +  # control color
     theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5),
        legend.position = "bottom")+
    ylab(y.label)+
    xlab("DBH class (cm)")+

    # coord_flip()+
    facet_grid(Year~BA.Target) %>% 
    return()
}

# create stand-level plots by diameter distribution
diamVol.plot<-diamClass.plot("Vol",y.label="Mean plot volume (m³/ha)")
diamBA.plot<-diamClass.plot("BA",y.label="Mean plot basal area (m²/ha)")  
diamK.plot<-diamClass.plot("SPH",y.label="Mean plot tree density (sph)")

# diamClass.plot<-cowplot::plot_grid(diamVol.plot,diamBA.plot,diamK.plot,ncol=1)

```

### Plot means

```{r generate stand level plot means}

plotData.X<-
# Determine which plots have measurements in 1992 and 2019
  X %>% 
  #  filter(Year %in% c(1992,2019)) %>% 
  #  filter(Status=="Live") %>% 
  # 
  # group_by(Plot,Year) %>% 
  # summarise(n=n()) %>% 
  # ungroup() %>% 
  # group_by(Plot) %>% 
  # summarise(n=length(Year)) %>% 
  # filter(n>1) %>% # pick plots with measurements in both years
  # pull(Plot) %>% 
  # as.data.frame() %>% 
  # setNames("Plot") %>% 
  # 
  # # Now filter original dataset 
  # left_join(X,by="Plot") %>% 
  # mutate(Plot=fct_drop(Plot)) %>%  # drop unused levels
  # 
  # # Filter for years of interest
  # filter(Year %in% c(1992,2019)) %>% 
  # mutate(Year=fct_drop(Year)) %>% 
  filter(Status=="Live") %>%

    # Calculate volume
  left_join(filter(volCoef,VolType=="total"),by="Species") %>% 
  mutate(vol=exp(1)^b0*DBH^b1*Height.Pred^b2) %>% 
  mutate(BA=d2ba(DBH)) %>% 
  dplyr::select(BA.Target,Plot:Species,Year,DBH,BA,vol,DBH.class,Height.Pred) %>% 
  mutate(Species=factor(Species)) %>%

  # Remove volumes below 18.5cm DBH
  mutate(vol=replace(vol,DBH<18.5,0)) %>%

  # now create plot sums grid so that you can calculate means
  group_by(BA.Target,Plot,Year) %>% 
    summarise(Vol.sum=sum(vol,na.rm=T)*20,
              BA.sum=sum(BA)*20,
              SPH.sum=n()*20) %>%

  ungroup() %>%

  # Create QMD
  mutate(QMD.mean=sqrt(BA.sum/(SPH.sum*0.0000785)))

# Now summarise plot data at the treatment unit level
standData.X<-
  plotData.X %>% 
  group_by(BA.Target,Year) %>% 
  summarise(Vol=mean(Vol.sum),
            Vol.se=sd(Vol.sum)/sqrt(length(Vol.sum)),
            BA=mean(BA.sum),
            BA.se=sd(BA.sum)/sqrt(length(BA.sum)),
            SPH=mean(SPH.sum),
            SPH.se=sd(SPH.sum)/sqrt(length(SPH.sum)),
            QMD=mean(QMD.mean),
            QMD.se=sd(QMD.mean)/sqrt(length(QMD.mean)))

# PLotting

# Create function
standPlot<-function(standVar,y.label) {

    standData.X %>% 
    mutate(Year=as.numeric(as.character(Year))) %>% 
    dplyr::select(BA.Target,Year,standVar,paste0(standVar,".se")) %>% 
    setNames(c("BA.Target","Year","Mean","SE")) %>%

    ggplot(aes(x=Year,y=Mean,color=BA.Target,group=BA.Target))+
    geom_point()+
    geom_line()+
    geom_errorbar(aes(ymin=Mean-SE, ymax=Mean+SE, width=.2),
                   position=position_dodge(.9))+
    facet_wrap(~BA.Target)+
      scale_x_continuous(limits=c(1992,2019),breaks=c(1992,1994,1997,2009,2019),
                     minor_breaks = c(1992,1994,1997,2009,2019),
                     name="Measurement year")+ # stretch out x-axis
  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5),
        legend.position = "none")+
    ylab(y.label)

}

standVol.plot<-
  standPlot("Vol",y.label="Vol. (m³/ha)")+
  theme(axis.text.x=element_blank(),axis.title.x=element_blank())

standBA.plot<-
  standPlot("BA",y.label="BA (m²/ha)")+
  theme(axis.text.x=element_blank(),axis.title.x=element_blank())

standK.plot<-
  standPlot("SPH",y.label="Tree dens. (sph)")+
  theme(axis.text.x=element_blank(),axis.title.x=element_blank())

standQMD.plot<-standPlot("QMD",y.label="QMD (cm)")

standPlots<-cowplot::plot_grid(standVol.plot,standBA.plot,standK.plot,standQMD.plot,ncol=1,
                      align="v")

figNums("standPlots",display=FALSE)
```

```{r fit model to plot data}

# Rename columns in plotData.X
plotData.X<-
  plotData.X %>% 
  rename(Vol="Vol.sum",
         BA="BA.sum",
         SPH="SPH.sum",
         QMD="QMD.mean")

# fit linear models
fit.vol<-lmer(Vol~BA.Target*Year+(1|Plot),data=plotData.X)
fit.ba<-lmer(BA~BA.Target*Year+(1|Plot),data=plotData.X)
fit.sph<-lmer(SPH~BA.Target*Year+(1|Plot),data=plotData.X)

# Summarize models
library(lmerTest)
anova(fit.vol, type="3") 
  rbind(anova(fit.ba, type="3"))
anova(fit.sph, type="3")

# Create estimated marginal means
ba.means <- emmeans(fit.ba, ~ BA.Target*Year)
sph.means <- emmeans(fit.sph, ~ BA.Target*Year)

# Contrasts
ba.contrast <- contrast(ba.means,interaction=c("pairwise","revpairwise")) 
sph.contrast <- contrast(sph.means,interaction=c("pairwise","revpairwise"))
```

Plot means of volume and basal area show that from 2009-2019, stand volume and basal area converged between both harvested treatment units, and were less than the control (
`r figNums("standPlots",display="cite")`).  The slope of all attributes showed that the low RBA accrued volume, density and diameter faster than the other units over the 27-year period of this study.  A linear model

-----------

```{r fig.height=10}

standPlots
```

`r figNums("standPlots",caption="Mean plot attributes by treatment unit and year.  Whiskers are one standard error of the mean.  Vol., volume; BA, basal area; Tree dens., tree density; QMD, quadratic mean diameter")`

*Note that sample size varies with year.  Some plots were not measured in 1992.  This likely explains the dip in basal area and volume in TU:20 from 1992 to 1994, despite no recorded mortality in 1992/1994.*

---------------

### Stand attribute by diameter class

Among treatments and throughout the stand development after treatment, the stands have maintained a reverse-J diameter class distribution (`r figNums("diamK.plot",display="cite")`).  
Immediately after the partial harvest, the diameter class distribution of volume differed between the treatment.  In the 27 years since partial harvest, volume in all three treatment units was mostly concentrated in the 22.5-27.5cm DBH class, and mostly comprised of fir (`r figNums("diamVol.plot",display="cite")`).  Over both time periods shown in the figure and all three treatments, most of the spruce volume was concentrated in DBH classes spanning 37.5 to 47.5cm.  Basal area trends were very similar to volume, and are not shown.

-----------

```{r}
diamK.plot

```

`r figNums("diamVol.plot",caption="Diameter-class distribution of tree density (stems/hectare) by species, treatment and year.")`

---------------

```{r}
diamVol.plot

```

`r figNums("diamVol.plot",caption="Diameter-class distribution of volume (m³/hectare) by species, treatment and year.")`

*Note that in this figure, we show volume for all diameter classes.  This is different than the stand-level estimate of volume, where we only estimate volume for trees that would be considered merchantable under conventional harvest in British Columbia (i.e., minimum 18.5cm DBH).*

---------------

### Species composition

```{r summarize changes in species composition}

sppComp.plot<-function(standVar){

    diamClass.spp %>% 
    filter(Species %in% c("Fir","Spruce")) %>% 
    dplyr::select(BA.Target,Species,Year,DBH.class,standVar=standVar) %>% # select columns
    pivot_wider(names_from="Species",values_from="standVar") %>%
    mutate(Fir=replace_na(Fir,0),
           Spruce=replace_na(Spruce,0)) %>%

    mutate(Total=Fir+Spruce) %>% 
    mutate(Fir.prop=Fir/Total) %>% 
    mutate(Spruce.prop=Spruce/Total) %>%

    mutate(Fir.prop=replace_na(Fir.prop,0),
           Spruce.prop=replace_na(Spruce.prop,0)) %>% 
    mutate(Spruce.prop=-Spruce.prop) %>% # change to neg value for plotting
    dplyr::select(BA.Target,Year,DBH.class,Spruce.prop,Fir.prop) %>% 
    pivot_longer(cols=contains(".prop"),names_to="Species",values_to="Prop") %>% 
    mutate(Species=str_remove(Species,".prop")) %>% 
    filter(Prop!=0) %>%

    ggplot()+
    aes(x=DBH.class,y=Prop,group=Species,fill=Species)+
      geom_bar(stat="identity")+
    xlab("DBH class (cm)")+
    scale_y_continuous(breaks=c(-1,-0.5,0,0.5,1),
                       labels=c(1,0.5,0,0.5,1))+
    ylab("Proportion")+
    coord_flip()+
    facet_grid(Year~BA.Target) %>% 
    return()
} # end function

# create stand-level plots by diameter distribution
sppVol.plot<-sppComp.plot("Vol")

figNums("sppVol.plot",display=FALSE)

# no need to provide other figures, as they're very similar

```

A comparison of proportional species composition by diameter class, treatment and year showed that fir and spruce were organized similarly between the high RBA and control units (`r figNums("sppVol.plot",display="cite")`).  In 1992, the largest diameter classes were dominated by spruce, whereas fir was the dominant species in the smallest diameter classes.  This pattern was found for volume, basal area, and tree density.  We only show volume in this document.  By 2019, the fir-spruce composition by distribution class was maintained in these two units, with small increases in spruce composition in the smallest diameter class.

In the low RBA, the species composition distribution was different.  Here, both species maintained a similar proportion of fir-leading across diameter classes smaller than 37.5cm DBH, whereas spruce formed the leading species in diameter classes above 37.5cm DBH, with the exception of the largest class.

```{r species composition plot}
sppVol.plot

```

`r figNums("sppVol.plot",caption="Spruce-fir volume proportional composition by diameter class, treatment and year.")`

*Note this figure is almost identical if we compare stems, basal area or volume. For brevity, we only show volume here.*

### 2020-11-05 manuscript/EP1162_manuscript_draft.Rmd
init

@@
 
+  # piDat.table %>% 
+  #   ggplot(aes(x=diamClass,y=BAImean,color=BA.Target))+
+  #   geom_point()+
+  #   facet_wrap(~Species)
+
 # Create figures of raw data to put into Appendix
@@
   theme_bw()+
-  ylab("Predicated annualized basal area increment (m²)")+
+  ylab("Predicted annualized basal area increment (m²)")+
   xlab("Measurement year")+

### 2020-11-05 manuscript/EP1162_manuscript_draft_October2020.Rmd
init

--- 2020-11-05 book/bookdown-minimal-master/2_results.Rmd
@@
-# Results
+---
+title: Analyses of 27 years of stand structure development and basal area growth response to a range of initial basal area density levels, 1992-2019 
+author: "Mike Jull and Hardy Griesbauer"
+date: "29/03/2020"
+output:
+  html_document:
+    df_print: paged
+    toc: true
+    toc_float: true
+    collapsed: false
+    theme: united
+    number_sections: true
+  word_document:
+    reference_docx: wordTemplate2.docx
+csl: cjfr.csl
+bibliography: ep1162Bib.bib
+editor_options: 
+  chunk_output_type: console
+---
 
-## Tree-level
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = FALSE,message=FALSE,warning=FALSE)
 
-### Basal area increment from 1992-2019
+# Load libraries
+library(tidyverse)
+library(here)
+library(lme4)
+library(lmerTest)
+library(broom)
+library(MuMIn)
+library(ggeffects)
+library(car)
+library(flextable)
+library(captioner)
+library(lsmeans)
+library(cowplot)
+library(officer)
+library(sjPlot)
+library(emmeans)
+
+# Load data
+load(here::here("data","ep1162_Data.RData"))
+load(here::here("data","ep1162_plotData.RData"))
+
+# SOME PRELIM DATA FORMATTING
+# formatting plotDat
+plotDat<-
+  plotDat %>% 
+  mutate(Plot=as.factor(Plot))
+
+# set up diameter classes
+diamClass=seq(from=7.5,to=100,by=5)
+
+# Set against scientific notation
+options(scipen=999)
+
+# Merge data
+X<-
+  dat %>% 
+  left_join(plotDat,by="Plot") %>% 
+  mutate(DBH.class=cut(DBH,diamClass,include.lowest=TRUE,ordered_result = TRUE,right=F,
+                       labels=diamClass[1:(length(diamClass)-1)])) %>% 
+  mutate(Plot=as.factor(Plot)) %>%
+
+  # replace species codes with words
+  mutate(Species=recode_factor(Species,Bl="Fir",Sx="Spruce")) %>% 
+  mutate(Species=fct_drop(Species)) %>% # have an extra level in here for some reason
+
+  # Create TU lookup
+  mutate(BA.Target=paste0("TU:",BA.Target))
+
+# set up captions
+
+# Code to create captions for tables
+tableNums <- captioner(prefix = "Table")
+
+# Code to create captions for figures
+figNums <- captioner(prefix="Figure")
+
+# Code to create captions for appendices
+appNums <- captioner(prefix="Appendix")
+
+# Set ggplot theme
+theme_set(theme_bw())
+
+# Define function to convert diameter (cm) to basal area (m2), and vice versa
+d2ba<-function(diam) {return((diam/2/100)^2*pi)}
+ba2d<-function(ba) {return(sqrt(ba/pi)*200)}
+
+```
+
+# Methods
+
+```{r define captions to be used in methods results and discussion}
+
+# Captions (put these in order)
+
+  # Introductory 
+  appNums(name="paiTable",caption="Number of growing seasons between each measurement period",display=FALSE) # Table showing number of growing seasons between each measurement period
+  appNums(name="standVol.plot",caption="Plot total volume for five measurement periods",display=FALSE) # Plot volume figure
+  appNums(name="standBA.plot",caption="Plot basal area for five measurement periods",display=FALSE)  # Plot basal area figure
+  appNums(name="standQMD.plot",caption="Plot quadratic mean diameter for five measurement periods",display=FALSE) # QMD figure  
+  appNums(name="standK.plot",caption="Plot tree density for five measurement periods", display=FALSE) # SPH figure
+
+  # Tree BAI model
+  figNums(name="treeBAI.plot",display=FALSE)
+  tableNums("fitBA.table",display=FALSE) # Summary of mixed effects model for BAI
+  appNums("sppContrast.table",display=FALSE) # Species contrasts from treeBAI model
+  appNums("trtContrast.table",display=FALSE) # Treatment contracts from treeBAI model
+
+  # Tree heights model
+  appNums("heightDiamRawPlot",display=FALSE) # Plot of diameter/heights`
+  appNums("htModelFit",display="FALSE") # model output
+
+  # Diameter class analysis
+  figNums("diamVol.plot", caption="Mean plot volume distribution by diameter class and treatment, for two periods.
+Black line shows total volume.
+Volume calculated on trees meeting 18.5cm DBH limit.
+Spruce and fir contribution to volume by diameter class are shown with bar plots.
+Douglas-fir volume contribution by diameter class is not shown.",display="FALSE")
+
+  figNums("diamBA.plot", caption="Mean plot basal area distribution by diameter class and treatment, for two periods.
+Black line shows total basal area.
+Spruce and fir contribution by diameter class are shown ith bar plots.
+Douglas-fir contribution by diameter class is not shown.",display="FALSE")
+
+  figNums("diamK.plot", caption="Mean tree density distribution by diameter class and treatment, for two periods.
+Black line shows total tree density.
+Spruce and fir contribution by diameter class are shown ith bar plots.
+Douglas-fir contribution by diameter class is not shown.",display="FALSE")
+
+```
+
+## Analyses
+
+### Tree-level growth responses
+To test the hypothesis that tree-level radial growth response over the 1992-2019 period varied with interactions between (i) tree diameter at the time of treatment (1992); (ii) treatment; and (iii) species, we fit linear mixed effects models with log-transformed 1992-2019 diameter increment as the response variable, and treatment, species and log-transformed 1992 diameter as interacting fixed effects, and plot as a random intercept effect.
+We also fit a similar model with the 1992-2019 diameter increment expressed as a percentage of 1992 diameter.
+Plots of residuals and fit lines with orginal data were examined to evaluate model goodness of fit.
+Variance explained for each model are reported using the approach described in Nakagawa et al.
+[-@Nakagawa2013], as implemented in the MuMIn package for R [@add later].
+
+### Stand structure and growth dynamics
+
+#### Stand-level volume and basal area estimates
+Total and merchantable volumes were estimated for trees using species- and region-specific equations [@nigh2016] based on tree height and DBH.
+Consistent with forest management practices in British Columbia, we estimated volumes for trees meeting a minimum diameter utilization limit of 18.5cm DBH.
+For trees lacking a tree height measurement, we estimated height by fitting a single diameter-height model developed with a linear mixed effects model, with log-transformed DBH as a fixed effect, and tree nested within plot as a random effect.
+Trees with broken tops were omitted from diameter-height models.
+We summed tree-level estimates of volume and basal area to generate stand-level estimates.
+
+#### Stand growth
+
+To test the hypothesis that stand volume and basal area growth from 1992-2019 varied by treatment, we fit separate linear mixed effects models using periodic annual volume and basal area increments as response variables, treatment as a fixed effects, and plot as a random effect.
+Periodic annual increment was calculated for the four measurement periods by dividing increment over the measurement period by the number of complete growing seasons within that period (`r appNums("paiTable",display="cite")`).
+
+---------------
+
+## Results
 
@@
 
+### Tree-level basal area increment from 1992-2019
+
 The tree-level basal area increment model indicated that interactions between log-transformed tree size, species and treatment explained `r round(r.squaredGLMM(fitBA)[1]*100,1)`% (marginal R²) of the variation in tree log-transformed 1992-2019 BAI (`r tableNums("fitBA.table",display="cite")`).
@@
 
-----------------
-
-`r tableNums("fitBA.table",caption=fitBA.summ$caption)`
-
-```{r present tree growth model summary}
-
-tab_model(fitBA)
-
-```
-
----------------------------
-
-```{r display tree growth figure}
-
-treeBAI.plot
-
-```
-
-`r figNums("treeBAI.plot",caption="Estimated marginal means of spruce and fir basal area increment (m²) across five inital DBH classes and three treatments.
-Whiskers are standard error of the mean.
-Responses are back-transformed from the model.")`
-
---------------------
-
-`r appNums("sppContrast.table",caption=sppContrast$caption)`
-
-```{r print fir spruce contrasts, ft.align="left"}
-
-# print to screen
-sppContrast$table
-```
-
------------------------
-
-`r appNums("trtContrast.table",caption=trtContrast$caption)`
-
-```{r print treatment contrasts, ft.align="left"}
-
-# print to screen
-trtContrast$table
-
-```
-
---------------------------
-
-### Mortality
-
-```{r mortality graphs}
-
-# Create dead dataset
-
-# Dead species totals
-deadList<-
-  X %>% 
-  filter(Status=="Dead") %>% 
-  mutate(ID=paste(Plot,TreeID,sep="-")) %>% 
-  group_by(ID) %>% 
-  summarise_all(first)
-
-dead<-
-  X %>% 
-  filter(Status=="Dead") %>% 
-  mutate(ID=paste(Plot,TreeID,sep="-")) %>% 
-  mutate(Species=fct_recode(Species,`Douglas-fir`="Fd")) %>% 
-  group_by(ID) %>% 
-  summarise_all(first) %>% 
-  ungroup() %>% 
-  mutate(BA=d2ba(DBH)) %>% # convert to basal area per hectare 
-  group_by(BA.Target,DBH.class,Year) %>% 
-  summarise(BA=sum(BA),
-            k=n())
-
-dead.Summ<-
-  dead %>% 
-  ungroup() %>% 
-  group_by(BA.Target) %>% 
-  summarise(BA=sum(BA)/sum(dead$BA)*100,
-            k=sum(k)/sum(dead$k)*100) %>% 
-  mutate_if(is.numeric,round,1)
-
-p1<-
-  dead %>% 
-  ggplot(aes(x=DBH.class,y=k,fill=Year))+
-  geom_bar(stat="identity")+
-  ylab("Total tree mortality (stems)")+
-  facet_wrap(~BA.Target)+
-  xlab("")+
-  theme(axis.text.x = element_blank())
-
-p2<-
-  dead %>% 
-  ggplot(aes(x=DBH.class,y=BA,fill=Year))+
-  geom_bar(stat="identity")+
-  ylab("Total basal area mortality (m²)")+
-  facet_wrap(~BA.Target)
-
-mort.plot<-cowplot::plot_grid(p1, p2, ncol = 1, align = "v")
-figNums("mort.plot",display=FALSE)
-
-```
-
-A total of `r deadList %>% nrow()` trees (`r deadList %>% filter(Species=="Fir") %>% nrow()` fir, `r deadList %>% filter(Species=="Spruce") %>% nrow()` spruce, and `r deadList %>% filter(Species=="Fd") %>% nrow()` Douglas-fir) died from 1992 to 2019.
-No tree mortality was noted during the 1992 and 1994 data collection.
-By treatment unit, the control unit had the highest mortality, in terms of trees and basal area (
-`r dead.Summ %>% filter(BA.Target=="TU:Control") %>% pull("BA")` and 
-`r dead.Summ %>% filter(BA.Target=="TU:Control") %>% pull("BA")`% of tree mortality, respectively).  The high-removal treatment unit had the lowest mortality by number of trees and basal area (
-`r dead.Summ %>% filter(BA.Target=="TU:10") %>% pull("BA")` and 
-`r dead.Summ %>% filter(BA.Target=="TU:10") %>% pull("BA")`% of tree mortality, respectively).
-Tree mortality was concentrated in the smaller diameter classes in the control and low-removal treatment unit, and most mortality occured during the most recent measurement period of 2009-2019 (
-`r figNums("mort.plot",display="cite")`).
-
-```{r display mortality plot}
-mort.plot
-```
-
-`r figNums("mort.plot",caption="Total tree mortality recorded in plots from 1992-2019, presented as total stems (top panel) and basal area (bottom panel).")`
-
----------------------------
-
-From Mike's email:
-Eventually as BA stand density increases, density dependent mortality kicks in, because there is only so much growing space or access to site resources to go around.
-This can occur with the stand (conceptually or hypothetically)  in two possible ways: (a) “classic” suppression of smaller tree classes through asymmetric competition of bigger trees shading out smaller ones, or (b) high stand densities suppressing the vigor of the whole stand through various means, suggesting more symmetric competition – i.e.
-the idea that small trees compete for growing space with big ones as well.
-Asymmetric competition is generally accepted by most foresters and silviculturists; symmetric competition less so, and is often harder to demonstrate.
-
-*For the discussion: it seems that we're seeing mortality increasing in the plots, especially the control unit over the last 11 years.
-Does this reflect the density-dependent mortality that Mike refers to above?.*
-
-More from Mike on mortality:
-
-The thing about unevenaged or complex stand management, esp in ‘wilder’ BC stands, is that we are working in stands with a long legacy of pre-existing ecosystem processes before we enter the stand to manipulate them.
-Clearcutting liquidates these legacies for the most part, and traditional silviculture and GY has not had to deal with mortality much, other than clearly density dependent suppression of smaller dbh classes.
-[Side issue: most BC experience and data with unevenaged stands has been in IDF, which have been fire excluded on dry sites, and unevenaged structures allowed to develop.
-We must be careful about comparing Summit data to IDF precedents, because of substantial differences in ecology.]
-
-And we are working a more extensive level of management in the SBS and ESSF,  where our stands are older, and not manicured nor vacuumed up for mortality and blowdown.
-From an ecosystem management point of view, that Is probably a good thing.
-
-So warts and all, I think mortality is an inherent characteristic of the types of spruce-balsam stands and management we are studying.
-Expecting and documenting mortality was built into the EP 1162 working plan.
-The treatment effects we anticipate are the net stand growth = gross growth minus mortality.
-
-Conceptually, tree mortality following the 1991/92 harvest (and subsequent sanitation felling)  functions as a BA removal from the live-tree pool.
-In most plots, BA of trees that die within a measurement period is subsumed or overridden by the total BA growth of the rest of the trees during that period.
-
-Where the periodic data trends get messier is where single big trees or a cluster of trees gets killed by something, resulting in a pulse of BA mortality.
-Examples: The great big Fdi in Plot 18, or notably, about 3-4 big Sx in Plot 3 that got nailed by spruce beetle.
-
-In our results and discussion:
-1.	We can report on what stand growth and development trends seem to be clearly and unambiguously  influenced by initial BA conditions, and the magnitude of these treatment effects. 
-2.	A component of tree mortality (e.g. certain size classes) may be density dependent. 
-3.	A component of tree mortality may be density-independent (a.k.a stochastic or chaotic). 
-4.	By reporting on all the above, we are providing managers with results that integrate silvicultural expectations observations from real-life “messy” stands which will experience all of the above.
-By keeping our eye on overall stand growth and performance following treatment, we can inform / reassure managers that individual tree mortality events are not the end of the world, but rather need to be viewed in the context of overall stand dynamics and growth expectations.
-
-## Stand-level analyses
+## Stand-level development over 27 years
 
@@
 
-appNums("htModelFit.plot",display="FALSE") # model output
-
 ```
 
-### Estimating tree height from diameter
+### Estimating tree heights
 Tree heights had a positive nonlinear relationship with diameter, for both species and across all time periods (`r appNums("heightDiamRawPlot",display="cite")`).
-A comparison of height-diameter models showed that intercepts and slopes of log-transformed heights did not vary significantly (i.e., p>0.05) between species or time periods (not shown), therefore a single height-diameter linear mixed-effects model was developed using `r X %>% filter(Species %in% c("Fir","Spruce")) %>% filter(Status=="Live") %>% filter(is.na(DBT)) %>% drop_na(Height) %>% nrow()` tree height-diameter observations in the dataset, with log-transformed diameter as the fixed effect, and tree nested within plot as random effects.
-This model explained `r round(MuMIn::r.squaredGLMM(htDiamFit)[1]*100,1)`% of variation in log-transformed heights, with an intercept of `r round(fixef(htDiamFit)[1],3)` and slope coefficient of  `r round(fixef(htDiamFit)[2],3)` * log-transformed DBH (not shown).
-Residual and predicted vs actual height plots were assessed to ensure goodness of fit (`r appNums("htModelFit.plot",display="cite")`).
+A comparison of height-diameter models showed that intercepts and slopes of log-transformed heights did not vary significantly (i.e., p>0.05) between species or time periods (not shown), therefore a single height-diameter linear mixed-effects model was developed using `r dat %>% filter(Species %in% c("Fir","Spruce")) %>% filter(Status=="Live") %>% filter(is.na(DBT)) %>% drop_na(Height) %>% nrow()` tree height-diameter observations in the dataset, with log-transformed diameter as the fixed effect, and tree nested within plot as random effect.
+This model explained `r round(MuMIn::r.squaredGLMM(htDiamFit)[1]*100,1)`% of variation in log-transformed heights, with an intercept of `r round(fixef(htDiamFit)[1],3)` and slope coefficient of  `r round(fixef(htDiamFit)[2],3)` * log(DBH).
+Residual and predicted vs actual height plots were assessed to ensure goodness of fit (`r appNums("htModelFit",display="cite")`).
 This model was applied to predict heights in trees without height measurements, and tree-volume estimates generated as per Nigh [-@nigh2016].
-
-----------------------------
-```{r height model fit}
-
-htModelFit.plot
-
-```
-
----------------
-
-`r appNums("htModelFit.plot",caption="Scatterplot of predicted height(m) vs actual height(m) from diameter-height model.")`
 
@@
 ```
+
+```{r generate stand attributes}
+
+standDatspp<-
+
+# Create tree volumes
+  X %>% 
+  filter(Status=="Live") %>% # filter for live trees
+  filter(DBH>=18.5) %>% # filter for trees with minimum diameter utilization limits
+  left_join(filter(volCoef,VolType=="total"),by="Species") %>% 
+  mutate(vol=exp(1)^b0*DBH^b1*Height.Pred^b2) %>% 
+  dplyr::select(Plot,TreeID,Year,vol) %>%
+
+# Create stand data, by species
+  right_join(X,by=c("Plot","TreeID","Year")) %>%
+  group_by(BA.Target,Plot,Species,Year) %>% 
+  mutate(BA=d2ba(DBH)) %>% # create BA column
+  summarise(Vol=sum(vol,na.rm=T)*20, # notice that all variables are by hectare
+            BA=sum(BA,na.rm=T)*20,
+            k=n()*20) %>% 
+   mutate(QMD=sqrt(BA/(k*0.0000785))) %>% 
+  arrange(BA.Target,Plot,Species,Year) %>% 
+  ungroup()
+
+# Create stand data with totals
+standDat<-
+# Create tree volumes
+  X %>% 
+  filter(Status=="Live") %>% # filter for live trees
+  filter(DBH>=18.5) %>% # filter for trees with minimum diameter utilization limits
+  left_join(filter(volCoef,VolType=="total"),by="Species") %>% 
+  mutate(vol=exp(1)^b0*DBH^b1*Height.Pred^b2) %>% 
+  dplyr::select(Plot,TreeID,Year,vol) %>%
+
+# Create stand data, by species
+  right_join(X,by=c("Plot","TreeID","Year")) %>%
+  group_by(BA.Target,Plot,Year) %>% 
+  mutate(BA=d2ba(DBH)) %>% # create BA column
+  summarise(Vol=sum(vol,na.rm=T)*20, # notice that all variables are by hectare
+            BA=sum(BA,na.rm=T)*20,
+            k=n()*20) %>% 
+   mutate(QMD=sqrt(BA/(k*0.0000785))) %>% 
+  mutate(Species="Total") %>% 
+  dplyr::select(BA.Target,Plot,Species,Year,Vol,BA,k,QMD) %>% 
+  ungroup() %>% 
+  rbind(standDatspp) %>% 
+  arrange(BA.Target,Plot,Species,Year) %>% 
+  filter(Species %in% c("Total","Spruce","Fir")) # limit to certain species
+
+####
+# Basic plotting function
+standDat.plot<-function(standVar,y.label) {
+  standDat %>% 
+  mutate(Plot=paste0("Plot:",Plot)) %>% 
+  mutate(Year=as.numeric(as.character(Year))) %>% 
+  ggplot(aes(x=Year,y=get(standVar),color=Species,group=Species))+
+  geom_point()+
+  geom_line()+
+    ylab(y.label) +
+   facet_wrap(BA.Target~Plot,ncol=5)+
+   scale_x_continuous(limits=c(1992,2019),breaks=c(1992,1994,1997,2009,2019),
+                     minor_breaks = c(1992,1994,1997,2009,2019),
+                     name="Measurement year")+ # stretch out x-axis 
+  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5),
+        legend.position = "bottom")+
+   scale_color_manual(values=c("#E69F00", "#56B4E9","#000000")) %>%  # control color
+
+  return()
+} # close function
+
+# create stand-level plots
+standVol.plot<-standDat.plot("Vol",y.label="Total volume (m³/ha)")
+standBA.plot<-standDat.plot("BA",y.label="Basal area (m²/ha)")  
+standQMD.plot<-standDat.plot("QMD",y.label="Quadratic mean diameter (cm)")  
+standK.plot<-standDat.plot("k",y.label="Total tree density (sph)")
+
+```
+
+### Stand-level attributes by diameter class
+To elucidate how stand structure varied among treatments, we also evaluated stand attributes across diameter classes.  Diameter classes were produced in 5cm increments.
 
@@
 
-# diamClass.plot<-cowplot::plot_grid(diamVol.plot,diamBA.plot,diamK.plot,ncol=1)
-
 ```
 
-### Plot means
-
-```{r generate stand level plot means}
-
-plotData.X<-
-# Determine which plots have measurements in 1992 and 2019
-  X %>% 
-  #  filter(Year %in% c(1992,2019)) %>% 
-  #  filter(Status=="Live") %>% 
-  # 
-  # group_by(Plot,Year) %>% 
-  # summarise(n=n()) %>% 
-  # ungroup() %>% 
-  # group_by(Plot) %>% 
-  # summarise(n=length(Year)) %>% 
-  # filter(n>1) %>% # pick plots with measurements in both years
-  # pull(Plot) %>% 
-  # as.data.frame() %>% 
-  # setNames("Plot") %>% 
-  # 
-  # # Now filter original dataset 
-  # left_join(X,by="Plot") %>% 
-  # mutate(Plot=fct_drop(Plot)) %>%  # drop unused levels
-  # 
-  # # Filter for years of interest
-  # filter(Year %in% c(1992,2019)) %>% 
-  # mutate(Year=fct_drop(Year)) %>% 
-  filter(Status=="Live") %>%
-
-    # Calculate volume
-  left_join(filter(volCoef,VolType=="total"),by="Species") %>% 
-  mutate(vol=exp(1)^b0*DBH^b1*Height.Pred^b2) %>% 
-  mutate(BA=d2ba(DBH)) %>% 
-  dplyr::select(BA.Target,Plot:Species,Year,DBH,BA,vol,DBH.class,Height.Pred) %>% 
-  mutate(Species=factor(Species)) %>%
-
-  # Remove volumes below 18.5cm DBH
-  mutate(vol=replace(vol,DBH<18.5,0)) %>%
-
-  # now create plot sums grid so that you can calculate means
-  group_by(BA.Target,Plot,Year) %>% 
-    summarise(Vol.sum=sum(vol,na.rm=T)*20,
-              BA.sum=sum(BA)*20,
-              SPH.sum=n()*20) %>%
-
-  ungroup() %>%
-
-  # Create QMD
-  mutate(QMD.mean=sqrt(BA.sum/(SPH.sum*0.0000785)))
-
-# Now summarise plot data at the treatment unit level
-standData.X<-
-  plotData.X %>% 
-  group_by(BA.Target,Year) %>% 
-  summarise(Vol=mean(Vol.sum),
-            Vol.se=sd(Vol.sum)/sqrt(length(Vol.sum)),
-            BA=mean(BA.sum),
-            BA.se=sd(BA.sum)/sqrt(length(BA.sum)),
-            SPH=mean(SPH.sum),
-            SPH.se=sd(SPH.sum)/sqrt(length(SPH.sum)),
-            QMD=mean(QMD.mean),
-            QMD.se=sd(QMD.mean)/sqrt(length(QMD.mean)))
-
-# PLotting
-
-# Create function
-standPlot<-function(standVar,y.label) {
-
-    standData.X %>% 
-    mutate(Year=as.numeric(as.character(Year))) %>% 
-    dplyr::select(BA.Target,Year,standVar,paste0(standVar,".se")) %>% 
-    setNames(c("BA.Target","Year","Mean","SE")) %>%
-
-    ggplot(aes(x=Year,y=Mean,color=BA.Target,group=BA.Target))+
-    geom_point()+
-    geom_line()+
-    geom_errorbar(aes(ymin=Mean-SE, ymax=Mean+SE, width=.2),
-                   position=position_dodge(.9))+
-    facet_wrap(~BA.Target)+
-      scale_x_continuous(limits=c(1992,2019),breaks=c(1992,1994,1997,2009,2019),
-                     minor_breaks = c(1992,1994,1997,2009,2019),
-                     name="Measurement year")+ # stretch out x-axis
-  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5),
-        legend.position = "none")+
-    ylab(y.label)
-
-}
-
-standVol.plot<-
-  standPlot("Vol",y.label="Vol. (m³/ha)")+
-  theme(axis.text.x=element_blank(),axis.title.x=element_blank())
-
-standBA.plot<-
-  standPlot("BA",y.label="BA (m²/ha)")+
-  theme(axis.text.x=element_blank(),axis.title.x=element_blank())
-
-standK.plot<-
-  standPlot("SPH",y.label="Tree dens. (sph)")+
-  theme(axis.text.x=element_blank(),axis.title.x=element_blank())
-
-standQMD.plot<-standPlot("QMD",y.label="QMD (cm)")
-
-standPlots<-cowplot::plot_grid(standVol.plot,standBA.plot,standK.plot,standQMD.plot,ncol=1,
-                      align="v")
-
-figNums("standPlots",display=FALSE)
-```
-
-```{r fit model to plot data}
-
-# Rename columns in plotData.X
-plotData.X<-
-  plotData.X %>% 
-  rename(Vol="Vol.sum",
-         BA="BA.sum",
-         SPH="SPH.sum",
-         QMD="QMD.mean")
-
-# fit linear models
-fit.vol<-lmer(Vol~BA.Target*Year+(1|Plot),data=plotData.X)
-fit.ba<-lmer(BA~BA.Target*Year+(1|Plot),data=plotData.X)
-fit.sph<-lmer(SPH~BA.Target*Year+(1|Plot),data=plotData.X)
-
-# Summarize models
-library(lmerTest)
-anova(fit.vol, type="3") 
-  rbind(anova(fit.ba, type="3"))
-anova(fit.sph, type="3")
-
-# Create estimated marginal means
-ba.means <- emmeans(fit.ba, ~ BA.Target*Year)
-sph.means <- emmeans(fit.sph, ~ BA.Target*Year)
-
-# Contrasts
-ba.contrast <- contrast(ba.means,interaction=c("pairwise","revpairwise")) 
-sph.contrast <- contrast(sph.means,interaction=c("pairwise","revpairwise"))
-```
-
-Plot means of volume and basal area show that from 2009-2019, stand volume and basal area converged between both harvested treatment units, and were less than the control (
-`r figNums("standPlots",display="cite")`).
-The slope of all attributes showed that the low RBA accrued volume, density and diameter faster than the other units over the 27-year period of this study.
-A linear model
-
------------
-
-```{r fig.height=10}
-
-standPlots
-```
-
-`r figNums("standPlots",caption="Mean plot attributes by treatment unit and year.
-Whiskers are one standard error of the mean.
-Vol., volume; BA, basal area; Tree dens., tree density; QMD, quadratic mean diameter")`
-
-*Note that sample size varies with year.
-Some plots were not measured in 1992.
-This likely explains the dip in basal area and volume in TU:20 from 1992 to 1994, despite no recorded mortality in 1992/1994.*
-
----------------
-
-### Stand attribute by diameter class
-
-Among treatments and throughout the stand development after treatment, the stands have maintained a reverse-J diameter class distribution (`r figNums("diamK.plot",display="cite")`).  
-Immediately after the partial harvest, the diameter class distribution of volume differed between the treatment.
-In the 27 years since partial harvest, volume in all three treatment units was mostly concentrated in the 22.5-27.5cm DBH class, and mostly comprised of fir (`r figNums("diamVol.plot",display="cite")`).
-Over both time periods shown in the figure and all three treatments, most of the spruce volume was concentrated in DBH classes spanning 37.5 to 47.5cm.
-Basal area trends were very similar to volume, and are not shown.
-
------------
-
-```{r}
-diamK.plot
-
-```
-
-`r figNums("diamVol.plot",caption="Diameter-class distribution of tree density (stems/hectare) by species, treatment and year.")`
-
----------------
-
-```{r}
-diamVol.plot
-
-```
-
-`r figNums("diamVol.plot",caption="Diameter-class distribution of volume (m³/hectare) by species, treatment and year.")`
-
-*Note that in this figure, we show volume for all diameter classes.
-This is different than the stand-level estimate of volume, where we only estimate volume for trees that would be considered merchantable under conventional harvest in British Columbia (i.e., minimum 18.5cm DBH).*
-
----------------
-
 ### Species composition
-
 ```{r summarize changes in species composition}
@@
 
-figNums("sppVol.plot",display=FALSE)
-
 # no need to provide other figures, as they're very similar
@@
 
-A comparison of proportional species composition by diameter class, treatment and year showed that fir and spruce were organized similarly between the high RBA and control units (`r figNums("sppVol.plot",display="cite")`).
-In 1992, the largest diameter classes were dominated by spruce, whereas fir was the dominant species in the smallest diameter classes.
-This pattern was found for volume, basal area, and tree density.
-We only show volume in this document.
-By 2019, the fir-spruce composition by distribution class was maintained in these two units, with small increases in spruce composition in the smallest diameter class.
+### Mortality
 
-In the low RBA, the species composition distribution was different.
-Here, both species maintained a similar proportion of fir-leading across diameter classes smaller than 37.5cm DBH, whereas spruce formed the leading species in diameter classes above 37.5cm DBH, with the exception of the largest class.
+```{r mortality graphs}
 
-```{r species composition plot}
-sppVol.plot
+# Create dead dataset
+dead<-
+  X %>% 
+  filter(Status=="Dead") %>% 
+  mutate(ID=paste(Plot,TreeID,sep="-")) %>% 
+  group_by(ID) %>% 
+  summarise_all(first) %>% 
+  ungroup() %>% 
+  mutate(BA=d2ba(DBH)) %>% # convert to basal area per hectare 
+  group_by(BA.Target,DBH.class) %>% 
+  summarise(BA=sum(BA),
+            k=n())
+
+p1<-
+  dead %>% 
+  ggplot(aes(x=DBH.class,y=k))+
+  geom_bar(stat="identity")+
+  ylab("Total tree mortality (stems)")+
+  facet_wrap(~BA.Target)+
+  xlab("")+
+  theme(axis.text.x = element_blank())
+
+p2<-
+  dead %>% 
+  ggplot(aes(x=DBH.class,y=BA))+
+  geom_bar(stat="identity")+
+  ylab("Total basal area mortality (m²)")+
+  facet_wrap(~BA.Target)
+
+cowplot::plot_grid(p1, p2, ncol = 1, align = "v")
 
@@
 
-`r figNums("sppVol.plot",caption="Spruce-fir volume proportional composition by diameter class, treatment and year.")`
+---------------------------
 
-*Note this figure is almost identical if we compare stems, basal area or volume. For brevity, we only show volume here.*
+## TABLES
+
+`r tableNums("fitBA.table",caption=fitBA.summ$caption)`
+
+```{r present tree growth model summary}
+
+tab_model(fitBA)
+
+```
+
+---------------------------
+
+## FIGURES
+
+```{r display tree growth figure}
+
+treeBAI.plot
+
+```
+
+`r figNums("treeBAI.plot",caption="Estimated marginal means of spruce and fir basal area increment (m²) across five inital DBH classes and three treatments.
+Whiskers are standard error of the mean.
+Responses are back-transformed from the model.")`
+
+--------------------
+
+```{r plot volume by diameter class}
+
+diamVol.plot
+```
+
+`r figNums("diamVol.plot")`
+
+----------------------
+
+```{r plot basal area by diameter class}
+
+diamBA.plot
+
+```
+
+`r figNums("diamBA.plot")`
+
+--------------------
+
+```{r plot tree density by diameter class}
+diamK.plot
+```
+
+`r figNums("diamK.plot")`
+
+---------------------
+
+## APPENDICES/SUPPLEMENTARY MATERIALS
+
+`r appNums("paiTable",caption="Number of growing seasons between measurement periods")`
+```{r show paiTable, ft.align="left"}
+
+  tibble(`Measurement Year`=c(1992,1994,1997,2009,2019),`Growing Seasons`=c(NA,3,3,11,11)) %>% 
+  flextable() %>% 
+  align(align="left",part="all") %>% 
+  # set_caption(caption="Number of growing seasons between measurement periods") %>% 
+  autofit()
+
+```
+
+------------------------
+
+```{r set figure options for Appendix figs}
+# set options for figures in the next few sections
+knitr::opts_chunk$set(fig.width=8, fig.height=8)
+```
+
+```{r plot volume}
+
+standVol.plot
+```
+
+`r appNums("standVol.plot")`
+
+----------------------
+
+```{r plot basal area}
+standBA.plot
+```
+
+`r appNums("standBA.plot")`
+
+-----------------------
+
+```{r qmd plot}
+standQMD.plot
+```
+
+`r appNums("standQMD.plot")`
+
+-----------------
+
+```{r SPH plot}
+standK.plot
+```
+
+`r appNums("standK.plot")`
+
+------------------------
+
+`r appNums("sppContrast.table",caption=sppContrast$caption)`
+
+```{r print fir spruce contrasts, ft.align="left"}
+
+# print to screen
+sppContrast$table
+```
+
+-----------------------
+
+`r appNums("trtContrast.table",caption=trtContrast$caption)`
+
+```{r print treatment contrasts, ft.align="left"}
+
+# print to screen
+trtContrast$table
+
+```
+
+--------------------------
+
+## REFERENCES

### 2020-11-06 book/bookdown-minimal-master/2_results.Rmd
Update 2_results.Rmd

@@
 
-standPlots<-cowplot::plot_grid(standVol.plot,standBA.plot,standK.plot,standQMD.plot,ncol=1,
+standPlots<-cowplot::plot_grid(standVol.plot,standBA.plot,standK.plot,ncol=1,
                       align="v")
+
+# could put in standQMD.plot as well
 
@@
 library(lmerTest)
-anova(fit.vol, type="3") 
-  rbind(anova(fit.ba, type="3"))
-anova(fit.sph, type="3")
 
 # Create estimated marginal means
+vol.means <- emmeans(fit.vol, ~ BA.Target*Year)
 ba.means <- emmeans(fit.ba, ~ BA.Target*Year)
@@
 # Contrasts
-ba.contrast <- contrast(ba.means,interaction=c("pairwise","revpairwise")) 
-sph.contrast <- contrast(sph.means,interaction=c("pairwise","revpairwise"))
+vol.contrast <- contrast(vol.means,interaction="pairwise")
+ba.contrast <- contrast(ba.means,interaction=c("pairwise")) 
+sph.contrast <- contrast(sph.means,interaction=c("pairwise"))
+
+# CREATE TABLES
+
+# create the data frame of ANOVA
+standFit.X<-
+  data.frame(Variable="Volume",anova(fit.vol, type="3")) %>% 
+  rbind(data.frame(Variable="Basal area",anova(fit.ba, type="3"))) %>% 
+   rbind(data.frame(Variable="Stand density",anova(fit.sph, type="3"))) %>%
+
+  # formatting
+  rownames_to_column("Fixed effect") %>% 
+  dplyr::select(Variable,`Fixed effect`,Sum.Sq:F.value,P.value="Pr..F.") %>% 
+  mutate_at(vars("F.value","P.value"),round,5) %>% 
+  mutate(`Fixed effect`=rep(c("Treatment","Year","Trt.*Year"),times=3)) %>% 
+  rename(`Sum sq.`="Sum.Sq",
+         `Mean sq.`="Mean.Sq",
+         `Num. DF`="NumDF",
+         `Den. DF`="DenDF",
+         `F value`="F.value",
+         `p value`="P.value")
+
+# Create contrasts table
+standFit.contrasts<-
+  data.frame(Variable="Volume",vol.contrast) %>% 
+  rbind(data.frame(Variable="Basal area",ba.contrast)) %>% 
+  rbind(data.frame(Variable="Stand density",ba.contrast)) %>% 
+  filter(Year_pairwise %in% c("1992 - 2019","1997 - 2019")) %>% 
+  dplyr::select(Variable,`Year contrast`="Year_pairwise",`Treatment contrast`="BA.Target_pairwise",everything()) %>% 
+  rename(`t ratio`="t.ratio",
+         `p value`="p.value") %>% 
+  mutate_at(vars(estimate:`t ratio`),round,3) %>% 
+  mutate_at(vars(`p value`),round,5)
+
+figNums("standFit.X",display=FALSE)
+
 ```
@@
 ---------------
+
+`r tableNums("standFit.X",caption="Output from three separate linear models testing the effect of the interaction between treatment and year on plot-level volume, basal area and density." )`
+
+```{r stand variable model output}
+
+standFit.X %>% 
+  flextable() %>% 
+  align(align="left",part="all") %>% 
+  merge_v(j="Variable") %>% 
+  valign(valign="top",part="body") %>% 
+    hline(i=c(3,6,9),border=fp_border(color="black")) %>% 
+    fix_border_issues() %>% 
+  autofit()
+```
+*Trt, treatment; Sq., squares, Num., numerator; Den., denominator; DF, degrees of freedom*
+
+```{r show contrasts between plot means}
+
+```
 

### 2020-11-06 book/bookdown-minimal-master/2_results.Rmd
Update 2_results.Rmd

@@
   rbind(data.frame(Variable="Basal area",ba.contrast)) %>% 
-  rbind(data.frame(Variable="Stand density",ba.contrast)) %>% 
+  rbind(data.frame(Variable="Stand density",sph.contrast)) %>% 
   filter(Year_pairwise %in% c("1992 - 2019","1997 - 2019")) %>% 
@@
 figNums("standFit.X",display=FALSE)
+appNums("standFit.contrasts",display=FALSE)
 
@@
 ```{r show contrasts between plot means}
+
+standFit.contrasts %>% 
+  flextable() %>% 
+   align(align="left",part="all") %>% 
+  merge_v(j=c("Variable","Year contrast")) %>% 
+  valign(valign="top",part="body") %>% 
+    hline(i=c(3,6,9,12,15),border=fp_border(color="black")) %>% 
+    fix_border_issues() %>% 
+  autofit()
 

### 2020-11-30 book/bookdown-minimal-master/1_Methods.Rmd
Remove nested structure from emmeans::ref_grid

@@
 library(emmeans)
-
-load("~/Documents/Git/EP1162/data/ep1162_Data.RData")
-load("~/Documents/Git/EP1162/data/ep1162_plotData.RData")
 
@@
 Black line shows total volume.
-Volume calculated on trees meeting 18.5cm DBH limit.
+Volume calculated on trees meeting 17.5cm DBH limit.
 Spruce and fir contribution to volume by diameter class are shown with bar plots.
@@
 Total and merchantable volumes were estimated for trees using species- and region-specific equations [@nigh2016] based on tree height and DBH.
-Consistent with forest management practices in British Columbia, we estimated volumes for trees meeting a minimum diameter utilization limit of 18.5cm DBH.
+Consistent with forest management practices in British Columbia, we estimated volumes for trees meeting a minimum diameter utilization limit of 17.5cm DBH.
 For trees lacking a tree height measurement, we estimated height by fitting a single diameter-height model developed with a linear mixed effects model, with log-transformed DBH as a fixed effect, and tree nested within plot as a random effect.

### 2020-11-30 book/bookdown-minimal-master/2_results.Rmd
Remove nested structure from emmeans::ref_grid

--- 2020-11-06 book/bookdown-minimal-master/2_results.Rmd
@@
   # create a reference grid for five different diameter classes (converted to BA)
-  rg<-ref_grid(fitBA,at=list(Init=d2ba(c(10,20,30,40,50))),type="response")
+  rg<-ref_grid(fitBA,at=list(Init=d2ba(c(10,20,30,40,50))),type="response",nesting=NULL)
 
@@
   # create a reference grid for five different diameter classes (converted to BA)
-  rg.pbaipct<-ref_grid(fitBApct,at=list(Init=d2ba(c(10,20,30,40,50))),type="response")
+  rg.pbaipct<-ref_grid(fitBApct,at=list(Init=d2ba(c(10,20,30,40,50))),type="response",nesting=NULL)
 
@@
 Estimated marginal means from the model showed that between species and among treatments, larger trees increased basal area more than smaller trees (`r figNums("treeGrowth.plot",display="cite")`).
-Across diameter classes, spruce trees had more basal area increment than fir in the high-removal treatment unit, whereas growth between the two species was close to equal in the low-removal and control units (`r appNums("sppContrast.table",display="cite")`).
-Spruce basal area increment in the high-removal unit exceeded spruce growth in both other units (`r appNums("trtContrast.table",display="cite")`).
-Spruce growth was similar between the low-removal and control units.
-Fir basal area increment in the high- and low-removal units was higher than the control unit, and there were no significant differences between the two harvested units (`r appNums("trtContrast.table",display="cite")`).
+Species contrasts showed that spruce trees increased  basal area more than than fir in the high-removal treatment unit, whereas basal growth between the two species was close to equal in the low-removal and control units (`r appNums("sppContrast.table",display="cite")`).
+Within each species, treatment contrasts showed that spruce basal area increment in the high-removal unit exceeded spruce growth in both other units (`r appNums("trtContrast.table",display="cite")`), whereas fir basal area increment in the high- and low-removal units was higher than the control unit, and there were no significant differences between the two harvested units (`r appNums("trtContrast.table",display="cite")`).
 
@@
 
-  # Remove volumes below 18.5cm DBH
-  mutate(vol=replace(vol,DBH<18.5,0)) %>%
+  # Remove volumes below 17.5cm DBH
+  mutate(vol=replace(vol,DBH<17.5,0)) %>%
 
@@
 
-# Contrasts
+# Slope Contrasts
 vol.contrast <- contrast(vol.means,interaction="pairwise")
@@
 sph.contrast <- contrast(sph.means,interaction=c("pairwise"))
+
+# Mean contrasts
+volM.contrast <- pairs(vol.means,by="Year")
+baM.contrast <- pairs(ba.means,by="Year")
+sphM.contrast <- pairs(sph.means,by="Year")
 
@@
 
-# Create contrasts table
+# Create slope contrasts table
 standFit.contrasts<-
@@
   rbind(data.frame(Variable="Basal area",ba.contrast)) %>% 
-  rbind(data.frame(Variable="Stand density",ba.contrast)) %>% 
-  filter(Year_pairwise %in% c("1992 - 2019","1997 - 2019")) %>% 
+  rbind(data.frame(Variable="Stand density",sph.contrast)) %>% 
+  filter(Year_pairwise %in% c("1992 - 2019")) %>% 
   dplyr::select(Variable,`Year contrast`="Year_pairwise",`Treatment contrast`="BA.Target_pairwise",everything()) %>% 
@@
 
+# Create mean contrasts table
+standFitM.contrasts<-
+  data.frame(Variable="Volume",volM.contrast) %>% 
+  rbind(data.frame(Variable="Basal area",baM.contrast)) %>% 
+  rbind(data.frame(Variable="Stand density",sphM.contrast)) %>% 
+  filter(Year %in% c(1992,2019))
+
 figNums("standFit.X",display=FALSE)
+appNums("standFit.contrasts",display=FALSE)
 
@@
 
-Plot means of volume and basal area show that from 2009-2019, stand volume and basal area converged between both harvested treatment units, and were less than the control (
-`r figNums("standPlots",display="cite")`).
-The slope of all attributes showed that the low RBA accrued volume, density and diameter faster than the other units over the 27-year period of this study.
-A linear model
+_Observations:_
+
+1. All units increased volume, basal area and stand density over the 1992-2019 period (`r figNums("standPlots",display="cite")`).
+
+2.
+The rate of change differed between units.
+The low RBA unit accrued basal area faster than the other two units over the 1992-2019 period, and this difference was significant (`r appNums("standFit.contrasts",display="cite")`).
+The estimated rate of basal area increment in the low RBA unit was higher by by 8.4 and 7.8m² per hectare more than the high RBA and control units, respectively.
+
+3.
+The low RBA unit also significantly increased its stand density by an estimated 340 stems per hectare more than the control unit.
+The rate of change in tree density did not differ significantly between the two harvested units.
+
+4. The rate of change in volume in trees over 17.5cm DBH was barely significant between the two harvested units, and there were no significant differences between the control and harvested units.
 
@@
 
+--------------------
+
+`r appNums("standFit.contrasts",caption="Contrasts of 1992-2019 stand structure development over 1992-2019 between treatments.
+Contrasts are shown for three variables: volume, basal area and stand density.")`
+
 ```{r show contrasts between plot means}
+
+standFit.contrasts %>% 
+  dplyr::select(-`Year contrast`) %>% 
+  flextable() %>% 
+   align(align="left",part="all") %>% 
+  merge_v(j=c("Variable")) %>% 
+  valign(valign="top",part="body") %>% 
+    hline(i=c(3,6,9),border=fp_border(color="black")) %>% 
+    fix_border_issues() %>% 
+  autofit()
 
@@
 ```{r}
-diamVol.plot
+diamBA.plot
 
@@
 *Note that in this figure, we show volume for all diameter classes.
-This is different than the stand-level estimate of volume, where we only estimate volume for trees that would be considered merchantable under conventional harvest in British Columbia (i.e., minimum 18.5cm DBH).*
+This is different than the stand-level estimate of volume, where we only estimate volume for trees that would be considered merchantable under conventional harvest in British Columbia (i.e., minimum 17.5cm DBH).*
 
@@
 # create stand-level plots by diameter distribution
-sppVol.plot<-sppComp.plot("Vol")
+sppVol.plot<-sppComp.plot("BA")
 

### 2020-12-12 manuscript_draft.Rmd
start final manuscript process

--- 2020-11-30 book/bookdown-minimal-master/2_results.Rmd
@@
+---
+title: "Manuscript_draft"
+author: "Michael Jull and Hardy Griesbauer"
+date: "10/12/2020"
+output:
+  word_document:
+    reference_docx: wordTemplate2.docx
+  html_document:
+    df_print: paged
+csl: cjfr.csl
+bibliography: ep1162.bib
+
+---
+
+```{r load libraries for analyses, echo=FALSE, message=FALSE, warning=FALSE}
+
+knitr::opts_chunk$set(echo = FALSE,message=FALSE,warning=FALSE)
+
+# Load libraries
+library(tidyverse)
+library(lme4)
+library(lmerTest)
+library(broom)
+library(MuMIn)
+library(ggeffects)
+library(car)
+library(flextable)
+library(captioner)
+library(lsmeans)
+library(cowplot)
+library(officer)
+library(sjPlot)
+library(emmeans)
+
+load("data/ep1162_Data.RData")
+load("data/ep1162_plotData.RData")
+
+# SOME PRELIM DATA FORMATTING
+# formatting plotDat
+plotDat<-
+  plotDat %>% 
+  mutate(Plot=as.factor(Plot))
+
+# set up diameter classes
+diamClass=seq(from=7.5,to=100,by=5)
+
+# Set against scientific notation
+options(scipen=999)
+
+# Merge data
+X<-
+  dat %>% 
+  left_join(plotDat,by="Plot") %>% 
+  mutate(DBH.class=cut(DBH,diamClass,include.lowest=TRUE,ordered_result = TRUE,right=F,
+                       labels=diamClass[1:(length(diamClass)-1)])) %>% 
+  mutate(Plot=as.factor(Plot)) %>%
+
+  # replace species codes with words
+  mutate(Species=recode_factor(Species,Bl="Fir",Sx="Spruce")) %>% 
+  mutate(Species=fct_drop(Species)) %>% # have an extra level in here for some reason
+
+  # Create TU lookup
+  mutate(BA.Target=paste0("TU:",BA.Target))
+
+# set up captions
+
+# Code to create captions for tables
+tableNums <- captioner(prefix = "Table")
+
+# Code to create captions for figures
+figNums <- captioner(prefix="Figure")
+
+# Code to create captions for appendices
+appNums <- captioner(prefix="Appendix")
+
+# Set ggplot theme
+theme_set(theme_bw())
+
+# Define function to convert diameter (cm) to basal area (m2), and vice versa
+d2ba<-function(diam) {return((diam/2/100)^2*pi)}
+ba2d<-function(ba) {return(sqrt(ba/pi)*200)}
+
+```
+
+This section will describe the methods we used for data analyses, not field data collection.
+
+From our technical report submitted to FCI earlier:
+
+Our data analyses and summaries examined the effects of the different levels of stand-level basal area density treatments on three main attributes of stand development following treatment.
+These include, in order:
+
+1. Post-treatment stand basal area dynamics, including the rate and pattern of basal area re-growth and recovery over time following different levels of initial basal area density reductions;
+2. Stand structural outcomes, including treatment influences on the overall abundance of trees (in sph) by diameter class, within the different treatment types, and;
+3.
+Tree species composition outcomes, including treatment influences on the relative abundance and size class distributions of subalpine fir and hybrid white spruce within the different treatment types.
+
+Analyses are grouped into three main categories (below).  These are treated as separate categories in the Results section.
+
+1. Tree-level responses to treatment;
+2. Stand-level responses to treatment:
+
+```{r define captions to be used in methods results and discussion, echo=FALSE}
+
+# Captions (put these in order)
+
+  # Introductory 
+  appNums(name="paiTable",caption="Number of growing seasons between each measurement period",display=FALSE) # Table showing number of growing seasons between each measurement period
+  appNums(name="standVol.plot",caption="Plot total volume for five measurement periods",display=FALSE) # Plot volume figure
+  appNums(name="standBA.plot",caption="Plot basal area for five measurement periods",display=FALSE)  # Plot basal area figure
+  appNums(name="standQMD.plot",caption="Plot quadratic mean diameter for five measurement periods",display=FALSE) # QMD figure  
+  appNums(name="standK.plot",caption="Plot tree density for five measurement periods", display=FALSE) # SPH figure
+
+  # Tree heights model
+  appNums("heightDiamRawPlot",display=FALSE) # Plot of diameter/heights`
+
+  # Diameter class analysis
+  figNums("diamVol.plot", caption="Mean plot volume distribution by diameter class and treatment, for two periods.
+Black line shows total volume.
+Volume calculated on trees meeting 17.5cm DBH limit.
+Spruce and fir contribution to volume by diameter class are shown with bar plots.
+Douglas-fir volume contribution by diameter class is not shown.",display="FALSE")
+
+  figNums("diamBA.plot", caption="Mean plot basal area distribution by diameter class and treatment, for two periods.
+Black line shows total basal area.
+Spruce and fir contribution by diameter class are shown ith bar plots.
+Douglas-fir contribution by diameter class is not shown.",display="FALSE")
+
+  figNums("diamK.plot", caption="Mean tree density distribution by diameter class and treatment, for two periods.
+Black line shows total tree density.
+Spruce and fir contribution by diameter class are shown ith bar plots.
+Douglas-fir contribution by diameter class is not shown.",display="FALSE")
+
+```
+
+## Tree-level growth responses
+
+To test the hypothesis that tree-level radial growth response over the 1992-2019 period varied with interactions between (i) tree diameter at the time of treatment (1992); (ii) treatment; and (iii) species, we fit linear mixed effects models with log-transformed 1992-2019 diameter increment as the response variable, and treatment, species and log-transformed 1992 diameter as interacting fixed effects, and plot as a random intercept effect.
+We also fit a similar model with the 1992-2019 diameter increment expressed as a percentage of 1992 diameter.
+Plots of residuals and fit lines with orginal data were examined to evaluate model goodness of fit.
+Variance explained for each model are reported using the approach described in Nakagawa et al.
+[-@Nakagawa2013], as implemented in the MuMIn package for R [@add later].
+
+## Stand-level growth and structure
+
+### Stand-level volume and basal area estimates
+Total and merchantable volumes were estimated for trees using species- and region-specific equations [@nigh2016] based on tree height and DBH.
+Consistent with forest management practices in British Columbia, we estimated volumes for trees meeting a minimum diameter utilization limit of 17.5cm DBH.
+For trees lacking a tree height measurement, we estimated height by fitting a single diameter-height model developed with a linear mixed effects model, with log-transformed DBH as a fixed effect, and tree nested within plot as a random effect.
+Trees with broken tops were omitted from diameter-height models.
+We summed tree-level estimates of volume and basal area to generate stand-level estimates for each plot.
+To test the hypothesis that changes in stand attributes over 27 years varied by treatment, we fit separate linear models with stand total volume, basal area, and density as response variables, and the interaction between treatment and year as the independent variable.
+
 # Results
@@
 
-# Let's filter only for trees that have measurements in 1992 and 2019 
+# Filter only for trees that have measurements in 1992 and 2019 
 treeList<- # assign variable
   X %>% 
-  filter(TreeID!="New") %>% #remove new trees
+  filter(TreeID!="New") %>% # remove new trees
   mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
@@
 piDat<-
-  dat %>% 
+  X %>% 
   mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
@@
 
-  fitBA.summ<-
-    list(table=tab_model(fitBA),
-         caption="Summary of mixed effects model of tree-level growth")
-
   # create a reference grid for five different diameter classes (converted to BA)
@@
 
-  fitBApct.summ<-
-    list(table=tab_model(fitBApct),
-         caption="Summary of mixed effects model of tree-level periodic basal area % increment (PBAI %")
-
 # create a reference grid for five different diameter classes (converted to BA)
@@
 
-```{r tree BAI by treatment species and diameter class}
-
-# Create plot of periodic increment
-
-treeBAI.plot<-
-
-  emmeans(rg,~Init|BA.Target+Species,type="response") %>% 
-  as.data.frame() %>%
-  mutate(Init=ba2d(Init)) %>% # convert initial basal area back to diameter
-
-  # plotting
-  ggplot()+
-  aes(x=Init,y=response,fill=Species,color=Species)+
-  geom_point()+
-  geom_line()+
-  geom_errorbar(aes(ymin=response-SE, ymax=response+SE), width=.2,
-                 position=position_dodge(0.05))+
-  xlab("1992 DBH (cm)")+
-  ylab("Basal area increment from 1992-2019 (m²)")+
-  facet_wrap(~BA.Target)
-
-```
-
 ```{r tree level fir spruce contrasts}
 
-# set variable
 sppContrast<-
-  list(caption="Contrasts between predicted fir and spruce growth across treatment units and initial diameter classes.  Ratio between predicted fir and growth is in 'ratio' column.",
-
-  table=
 
@@
   arrange(BA.Target,Init) %>% 
-  rename(DBH.1992="Init") %>% 
-  mutate_if(is.double,round,5) %>%  # round to five digits
-
-  # format table
-
-  # flextable
-  flextable(col_keys = c(colnames(.)[c(1:2,6:8,3:5)])) %>%
-
-  align(align="left",part = "all") %>% 
-  vline( border = fp_border(color="#2c3a51" , width = 1), part = "all" ) %>% 
-  hline( border = fp_border(color="#2c3a51" , width = 1), part = "all" ) %>% 
-  merge_v(j=c(1:2,6:8)) %>% 
-  valign(valign="top") %>%
- border_outer(part="all", border = fp_border(color="#2c3a51" , width = 2)) %>% 
-   fix_border_issues()  # border(border.bottom = fp_border(color="black"))
-) # close list
+  rename(DBH.1992="Init")
 
@@
 
-The tree-level basal area increment model indicated that interactions between log-transformed tree size, species and treatment explained `r round(r.squaredGLMM(fitBA)[1]*100,1)`% (marginal R²) of the variation in tree log-transformed 1992-2019 BAI (`r tableNums("fitBA.table",display="cite")`).
+```{r set up tree BAI references}
+
+  figNums(name="treeBAI.plot",display=FALSE)
+
+  tableNums("fitBA.table",display=FALSE) # Summary of mixed effects model for BAI
+  tableNums("fitBApct.table",display=FALSE) # Summary of mixed effects model for BAI
+
+  appNums("sppContrast.table",display=FALSE) # Species contrasts from treeBAI model
+  appNums("trtContrast.table",display=FALSE) # Treatment contracts from treeBAI model
+
+```
+
+The tree-level basal area increment model indicated that interactions between log-transformed tree size, species and treatment explained `r round(r.squaredGLMM(fitBA)[1]*100,1)`% (marginal R²) of the variation in log-transformed 1992-2019 tree-level BAI (`r tableNums("fitBA.table",display="cite")`).
 Estimated marginal means from the model showed that between species and among treatments, larger trees increased basal area more than smaller trees (`r figNums("treeGrowth.plot",display="cite")`).
@@
 --------------------------
-
-### Mortality
-
-```{r mortality graphs}
-
-# Create dead dataset
-
-# Dead species totals
-deadList<-
-  X %>% 
-  filter(Status=="Dead") %>% 
-  mutate(ID=paste(Plot,TreeID,sep="-")) %>% 
-  group_by(ID) %>% 
-  summarise_all(first)
-
-dead<-
-  X %>% 
-  filter(Status=="Dead") %>% 
-  mutate(ID=paste(Plot,TreeID,sep="-")) %>% 
-  mutate(Species=fct_recode(Species,`Douglas-fir`="Fd")) %>% 
-  group_by(ID) %>% 
-  summarise_all(first) %>% 
-  ungroup() %>% 
-  mutate(BA=d2ba(DBH)) %>% # convert to basal area per hectare 
-  group_by(BA.Target,DBH.class,Year) %>% 
-  summarise(BA=sum(BA),
-            k=n())
-
-dead.Summ<-
-  dead %>% 
-  ungroup() %>% 
-  group_by(BA.Target) %>% 
-  summarise(BA=sum(BA)/sum(dead$BA)*100,
-            k=sum(k)/sum(dead$k)*100) %>% 
-  mutate_if(is.numeric,round,1)
-
-p1<-
-  dead %>% 
-  ggplot(aes(x=DBH.class,y=k,fill=Year))+
-  geom_bar(stat="identity")+
-  ylab("Total tree mortality (stems)")+
-  facet_wrap(~BA.Target)+
-  xlab("")+
-  theme(axis.text.x = element_blank())
-
-p2<-
-  dead %>% 
-  ggplot(aes(x=DBH.class,y=BA,fill=Year))+
-  geom_bar(stat="identity")+
-  ylab("Total basal area mortality (m²)")+
-  facet_wrap(~BA.Target)
-
-mort.plot<-cowplot::plot_grid(p1, p2, ncol = 1, align = "v")
-figNums("mort.plot",display=FALSE)
-
-```
-
-A total of `r deadList %>% nrow()` trees (`r deadList %>% filter(Species=="Fir") %>% nrow()` fir, `r deadList %>% filter(Species=="Spruce") %>% nrow()` spruce, and `r deadList %>% filter(Species=="Fd") %>% nrow()` Douglas-fir) died from 1992 to 2019.
-No tree mortality was noted during the 1992 and 1994 data collection.
-By treatment unit, the control unit had the highest mortality, in terms of trees and basal area (
-`r dead.Summ %>% filter(BA.Target=="TU:Control") %>% pull("BA")` and 
-`r dead.Summ %>% filter(BA.Target=="TU:Control") %>% pull("BA")`% of tree mortality, respectively).  The high-removal treatment unit had the lowest mortality by number of trees and basal area (
-`r dead.Summ %>% filter(BA.Target=="TU:10") %>% pull("BA")` and 
-`r dead.Summ %>% filter(BA.Target=="TU:10") %>% pull("BA")`% of tree mortality, respectively).
-Tree mortality was concentrated in the smaller diameter classes in the control and low-removal treatment unit, and most mortality occured during the most recent measurement period of 2009-2019 (
-`r figNums("mort.plot",display="cite")`).
-
-```{r display mortality plot}
-mort.plot
-```
-
-`r figNums("mort.plot",caption="Total tree mortality recorded in plots from 1992-2019, presented as total stems (top panel) and basal area (bottom panel).")`
-
----------------------------
-
-From Mike's email:
-Eventually as BA stand density increases, density dependent mortality kicks in, because there is only so much growing space or access to site resources to go around.
-This can occur with the stand (conceptually or hypothetically)  in two possible ways: (a) “classic” suppression of smaller tree classes through asymmetric competition of bigger trees shading out smaller ones, or (b) high stand densities suppressing the vigor of the whole stand through various means, suggesting more symmetric competition – i.e.
-the idea that small trees compete for growing space with big ones as well.
-Asymmetric competition is generally accepted by most foresters and silviculturists; symmetric competition less so, and is often harder to demonstrate.
-
-*For the discussion: it seems that we're seeing mortality increasing in the plots, especially the control unit over the last 11 years.
-Does this reflect the density-dependent mortality that Mike refers to above?.*
-
-More from Mike on mortality:
-
-The thing about unevenaged or complex stand management, esp in ‘wilder’ BC stands, is that we are working in stands with a long legacy of pre-existing ecosystem processes before we enter the stand to manipulate them.
-Clearcutting liquidates these legacies for the most part, and traditional silviculture and GY has not had to deal with mortality much, other than clearly density dependent suppression of smaller dbh classes.
-[Side issue: most BC experience and data with unevenaged stands has been in IDF, which have been fire excluded on dry sites, and unevenaged structures allowed to develop.
-We must be careful about comparing Summit data to IDF precedents, because of substantial differences in ecology.]
-
-And we are working a more extensive level of management in the SBS and ESSF,  where our stands are older, and not manicured nor vacuumed up for mortality and blowdown.
-From an ecosystem management point of view, that Is probably a good thing.
-
-So warts and all, I think mortality is an inherent characteristic of the types of spruce-balsam stands and management we are studying.
-Expecting and documenting mortality was built into the EP 1162 working plan.
-The treatment effects we anticipate are the net stand growth = gross growth minus mortality.
-
-Conceptually, tree mortality following the 1991/92 harvest (and subsequent sanitation felling)  functions as a BA removal from the live-tree pool.
-In most plots, BA of trees that die within a measurement period is subsumed or overridden by the total BA growth of the rest of the trees during that period.
-
-Where the periodic data trends get messier is where single big trees or a cluster of trees gets killed by something, resulting in a pulse of BA mortality.
-Examples: The great big Fdi in Plot 18, or notably, about 3-4 big Sx in Plot 3 that got nailed by spruce beetle.
-
-In our results and discussion:
-1.	We can report on what stand growth and development trends seem to be clearly and unambiguously  influenced by initial BA conditions, and the magnitude of these treatment effects. 
-2.	A component of tree mortality (e.g. certain size classes) may be density dependent. 
-3.	A component of tree mortality may be density-independent (a.k.a stochastic or chaotic). 
-4.	By reporting on all the above, we are providing managers with results that integrate silvicultural expectations observations from real-life “messy” stands which will experience all of the above.
-By keeping our eye on overall stand growth and performance following treatment, we can inform / reassure managers that individual tree mortality events are not the end of the world, but rather need to be viewed in the context of overall stand dynamics and growth expectations.
 
@@
 *Note this figure is almost identical if we compare stems, basal area or volume. For brevity, we only show volume here.*
+
+\newpage
+
+# Tables
+
+`r tablesNums("fitBA.table",caption="Summary of mixed effects model of tree-level basal area increment")`
+
+```{r summarize tree level BAI model}
+
+```
+
+\newpage
+
+`r tablesNums("fitBApct.table",caption="Summary of mixed effects model of tree-level basal area increment as percentage of initial tree size")`
+
+```{r summarize tree level BAI as percentage model}
+
+```
+
+\newpage
+
+# Figures
+
+```{r tree BAI by treatment species and diameter class}
+
+# Create plot of periodic increment
+
+# Using estimated marginal means
+  emmeans(rg,~Init|BA.Target+Species,type="response") %>% 
+  as.data.frame() %>%
+  mutate(Init=ba2d(Init)) %>% # convert initial basal area back to diameter
+
+  # plotting
+  ggplot()+
+  aes(x=Init,y=response,fill=Species,color=Species)+
+  geom_point()+
+  geom_line()+
+  geom_errorbar(aes(ymin=response-SE, ymax=response+SE), width=.2,
+                 position=position_dodge(0.05))+
+  xlab("1992 DBH (cm)")+
+  ylab("Basal area increment from 1992-2019 (m²)")+
+  facet_wrap(~BA.Target)
+
+```
+
+\newpage
+
+# Appendices
+
+`r appNums("paiTable",caption="Number of growing seasons between measurement periods")`
+
+```{r show paiTable, ft.align="left"}
+
+  tibble(`Measurement Year`=c(1992,1994,1997,2009,2019),`Growing Seasons`=c(NA,3,3,11,11)) %>% 
+  flextable() %>% 
+  align(align="left",part="all") %>% 
+  # set_caption(caption="Number of growing seasons between measurement periods") %>% 
+  autofit()
+
+```
+
+\newpage
+
+`r appNums("sppContrast.table",caption="Contrasts between predicted fir and spruce growth across treatment units and initial diameter classes.
+Ratio between predicted fir and growth is in 'ratio' column.")`
+
+```{r treelevel species contrast table}
+
+# Tree-level species contrast table
+  sppContrast %>%
+
+  mutate_if(is.double,round,5) %>%  # round to five digits
+
+  # format table
+  flextable(col_keys = c(colnames(.)[c(1:2,6:8,3:5)])) %>% 
+  align(align="left",part = "all") %>% 
+  vline( border = fp_border(color="#2c3a51" , width = 1), part = "all" ) %>% 
+  hline( border = fp_border(color="#2c3a51" , width = 1), part = "all" ) %>% 
+  merge_v(j=c(1:2,6:8)) %>% 
+  valign(valign="top") %>%
+ border_outer(part="all", border = fp_border(color="#2c3a51" , width = 2)) %>% 
+   fix_border_issues()  # border(border.bottom = fp_border(color="black"))
+
+```
+
+\newpage
+
+`r appNums("trtContrast.table",caption="Contrasts of predicted 1992-2019 tree growth among treatments within each species.  Ratio of predicted growth between treatments is in 'ratio' column.")`
+
+```{r tree level treatment contrasts}
+
+# Treatment contrast
+
+trtContrast %>%
+
+  mutate_if(is.double,round,5) %>%
+
+  # flextable
+  flextable(col_keys = c(colnames(.)[c(2,1,6:8,3:5)])) %>% 
+  border_outer(part="all", border = fp_border(color="#2c3a51" , width = 2)) %>% 
+  align(align="left",part = "all") %>% 
+  vline( border = fp_border(color="#2c3a51" , width = 1), part = "all" ) %>% 
+  hline( border = fp_border(color="#2c3a51" , width = 1), part = "all" ) %>% 
+  merge_v(j=c(1:2,6:8)) %>% 
+  valign(valign="top") %>%
+  hline_bottom() %>% 
+  fix_border_issues() # border(border.bottom = fp_border(color="black"))
+
+```
+
+\newpage

### 2020-12-12 docs/2_results.Rmd

--- 2020-11-30 book/bookdown-minimal-master/2_results.Rmd
@@
   # create a reference grid for five different diameter classes (converted to BA)
-  rg<-ref_grid(fitBA,at=list(Init=d2ba(c(10,20,30,40,50))),type="response",nesting=NULL)
+  rg<-ref_grid(fitBA,at=list(Init=d2ba(c(10,20,30,40,50))),type="response")
 
@@
   # create a reference grid for five different diameter classes (converted to BA)
-  rg.pbaipct<-ref_grid(fitBApct,at=list(Init=d2ba(c(10,20,30,40,50))),type="response",nesting=NULL)
+  rg.pbaipct<-ref_grid(fitBApct,at=list(Init=d2ba(c(10,20,30,40,50))),type="response")
 
@@
 Estimated marginal means from the model showed that between species and among treatments, larger trees increased basal area more than smaller trees (`r figNums("treeGrowth.plot",display="cite")`).
-Species contrasts showed that spruce trees increased  basal area more than than fir in the high-removal treatment unit, whereas basal growth between the two species was close to equal in the low-removal and control units (`r appNums("sppContrast.table",display="cite")`).
-Within each species, treatment contrasts showed that spruce basal area increment in the high-removal unit exceeded spruce growth in both other units (`r appNums("trtContrast.table",display="cite")`), whereas fir basal area increment in the high- and low-removal units was higher than the control unit, and there were no significant differences between the two harvested units (`r appNums("trtContrast.table",display="cite")`).
+Across diameter classes, spruce trees had more basal area increment than fir in the high-removal treatment unit, whereas growth between the two species was close to equal in the low-removal and control units (`r appNums("sppContrast.table",display="cite")`).
+Spruce basal area increment in the high-removal unit exceeded spruce growth in both other units (`r appNums("trtContrast.table",display="cite")`).
+Spruce growth was similar between the low-removal and control units.
+Fir basal area increment in the high- and low-removal units was higher than the control unit, and there were no significant differences between the two harvested units (`r appNums("trtContrast.table",display="cite")`).
 
@@
 
-  # Remove volumes below 17.5cm DBH
-  mutate(vol=replace(vol,DBH<17.5,0)) %>%
+  # Remove volumes below 18.5cm DBH
+  mutate(vol=replace(vol,DBH<18.5,0)) %>%
 
@@
 
-4. The rate of change in volume in trees over 17.5cm DBH was barely significant between the two harvested units, and there were no significant differences between the control and harvested units.
+4. The rate of change in volume in trees over 18.5cm DBH was barely significant between the two harvested units, and there were no significant differences between the control and harvested units.
 
@@
 ```{r}
-diamBA.plot
+diamVol.plot
 
@@
 *Note that in this figure, we show volume for all diameter classes.
-This is different than the stand-level estimate of volume, where we only estimate volume for trees that would be considered merchantable under conventional harvest in British Columbia (i.e., minimum 17.5cm DBH).*
+This is different than the stand-level estimate of volume, where we only estimate volume for trees that would be considered merchantable under conventional harvest in British Columbia (i.e., minimum 18.5cm DBH).*
 
@@
 # create stand-level plots by diameter distribution
-sppVol.plot<-sppComp.plot("BA")
+sppVol.plot<-sppComp.plot("Vol")
 

### 2020-12-16 manuscript_draft.Rmd
manny work

@@
 
-  # Tree heights model
-  appNums("heightDiamRawPlot",display=FALSE) # Plot of diameter/heights`
-
   # Diameter class analysis
@@
 
-```{r tree level fir spruce contrasts}
+```{r tree level species contrasts}
 
@@
 
-trtContrast<-list(
-  caption="Contrasts of predicted 1992-2019 tree growth among treatments within each species.  Ratio of predicted growth between treatments is in 'ratio' column.",
-
-# set up the table and place within list  
-table=  
+trtContrast<-
   pairs(rg,by=c("Species","Init"),type="response") %>% 
@@
   arrange(Species,contrast,Init) %>% 
-  rename(DBH.1992="Init") %>% 
-  mutate_if(is.double,round,5) %>%
-
-  # flextable
-  flextable(col_keys = c(colnames(.)[c(2,1,6:8,3:5)])) %>% 
-  border_outer(part="all", border = fp_border(color="#2c3a51" , width = 2)) %>% 
-  align(align="left",part = "all") %>% 
-  vline( border = fp_border(color="#2c3a51" , width = 1), part = "all" ) %>% 
-  hline( border = fp_border(color="#2c3a51" , width = 1), part = "all" ) %>% 
-  merge_v(j=c(1:2,6:8)) %>% 
-  valign(valign="top") %>%
-  hline_bottom() %>% 
-  fix_border_issues()# border(border.bottom = fp_border(color="black"))
-
-# Close list
-) # close list
+  rename(DBH.1992="Init")
 
@@
 
-  figNums(name="treeBAI.plot",display=FALSE)
+  figNums(name="treeBAI.plot",display=FALSE) # Estimated marginal means from Lme
 
@@
 
-----------------
-
-`r tableNums("fitBA.table",caption=fitBA.summ$caption)`
-
-```{r present tree growth model summary}
-
-tab_model(fitBA)
-
-```
-
----------------------------
-
-```{r display tree growth figure}
-
-treeBAI.plot
-
-```
-
-`r figNums("treeBAI.plot",caption="Estimated marginal means of spruce and fir basal area increment (m²) across five inital DBH classes and three treatments.
-Whiskers are standard error of the mean.
-Responses are back-transformed from the model.")`
-
---------------------
-
-`r appNums("sppContrast.table",caption=sppContrast$caption)`
-
-```{r print fir spruce contrasts, ft.align="left"}
-
-# print to screen
-sppContrast$table
-```
-
------------------------
-
-`r appNums("trtContrast.table",caption=trtContrast$caption)`
-
-```{r print treatment contrasts, ft.align="left"}
-
-# print to screen
-trtContrast$table
-
-```
-
---------------------------
-
 ## Stand-level analyses
@@
 ```{r height to diameter models}
-
-heightDiamRaw.plot<-
-  X %>% 
-    filter(Species %in% c("Bl","Sx")) %>% # filter for Bl and Sx for now
-    filter(Status=="Live") %>% # filter for live trees
-    filter(is.na(DBT)) %>% # filter for trees with no broken tops
-    drop_na(Height) %>% # remove trees where height was not measured
-
-    # Plotting
-  ggplot()+
-  aes(x=DBH,y=Height,fill=Species)+
-  geom_point()+
-  facet_grid(Year~Species)+
-  theme(legend.position = "none")+
-  ylab("Height (m)")+
-  xlab("DBH (cm)")+
-  geom_smooth(formula = y~log(x),method="lm",se=FALSE)
 
@@
 
+# Tree heights model
+  appNums("heightDiam.plot",display=FALSE) # Plot of diameter/heights`
+
 ```
@@
 ### Estimating tree height from diameter
-Tree heights had a positive nonlinear relationship with diameter, for both species and across all time periods (`r appNums("heightDiamRawPlot",display="cite")`).
+Tree heights had a positive nonlinear relationship with diameter, for both species and across all time periods (`r appNums("heightDiam.plot",display="cite")`).
 A comparison of height-diameter models showed that intercepts and slopes of log-transformed heights did not vary significantly (i.e., p>0.05) between species or time periods (not shown), therefore a single height-diameter linear mixed-effects model was developed using `r X %>% filter(Species %in% c("Fir","Spruce")) %>% filter(Status=="Live") %>% filter(is.na(DBT)) %>% drop_na(Height) %>% nrow()` tree height-diameter observations in the dataset, with log-transformed diameter as the fixed effect, and tree nested within plot as random effects.
@@
 
+`r figNums("treeBAI.plot",caption="Estimated marginal means of spruce and fir basal area increment (m²) across five inital DBH classes and three treatments.
+Whiskers are standard error of the mean.
+Responses are back-transformed from the model.")`
+
 \newpage
@@
 \newpage
+
+```{r height to diameter plot}
+
+  X %>% 
+    filter(Species %in% c("Fir","Spruce")) %>% # filter for Bl and Sx for now
+    filter(Status=="Live") %>% # filter for live trees
+    filter(is.na(DBT)) %>% # filter for trees with no broken tops
+    drop_na(Height) %>% # remove trees where height was not measured
+
+    # Plotting
+  ggplot()+
+  aes(x=DBH,y=Height,fill=Species)+
+  geom_point()+
+  facet_grid(Year~Species)+
+  theme(legend.position = "none")+
+  ylab("Height (m)")+
+  xlab("DBH (cm)")+
+  geom_smooth(formula = y~log(x),method="lm",se=FALSE)
+
+```
+
+`r appNums("heightDiam.plot",caption="Relationship between height (m) and diameter (cm) for ")`

### 2020-12-17 manuscript_draft.Rmd
manuscript work

@@
 
-# Height model fit plot
-htModelFit.plot<-
-  X %>% 
-  filter(is.na(DBT)) %>% # filter for trees with no broken tops
-  mutate(h2=exp(fixef(htDiamFit)[1]+fixef(htDiamFit)[2]*log(DBH))) %>% 
-  ggplot(aes(x=Height,y=h2))+
-  geom_point()+
-  xlim(0,40)+
-  ylim(0,40)+
-  geom_smooth(method="lm",se=F)+
-  ylab("Predicted height (m) from diameter-height model")+
-  xlab("Measured height (m)")
-
 appNums("htModelFit.plot",display="FALSE") # model output
@@
 This model was applied to predict heights in trees without height measurements, and tree-volume estimates generated as per Nigh [-@nigh2016].
-
-----------------------------
-```{r height model fit}
-
-htModelFit.plot
-
-```
-
----------------
-
-`r appNums("htModelFit.plot",caption="Scatterplot of predicted height(m) vs actual height(m) from diameter-height model.")`
 
@@
 # diamClass.plot<-cowplot::plot_grid(diamVol.plot,diamBA.plot,diamK.plot,ncol=1)
-
 ```
@@
   X %>%
-  #  filter(Year %in% c(1992,2019)) %>% 
-  #  filter(Status=="Live") %>% 
-  # 
-  # group_by(Plot,Year) %>% 
-  # summarise(n=n()) %>% 
-  # ungroup() %>% 
-  # group_by(Plot) %>% 
-  # summarise(n=length(Year)) %>% 
-  # filter(n>1) %>% # pick plots with measurements in both years
-  # pull(Plot) %>% 
-  # as.data.frame() %>% 
-  # setNames("Plot") %>% 
-  # 
-  # # Now filter original dataset 
-  # left_join(X,by="Plot") %>% 
-  # mutate(Plot=fct_drop(Plot)) %>%  # drop unused levels
-  # 
-  # # Filter for years of interest
-  # filter(Year %in% c(1992,2019)) %>% 
-  # mutate(Year=fct_drop(Year)) %>% 
+
   filter(Status=="Live") %>%
@@
             QMD.se=sd(QMD.mean)/sqrt(length(QMD.mean)))
-
-# PLotting
-
-# Create function
-standPlot<-function(standVar,y.label) {
-
-    standData.X %>% 
-    mutate(Year=as.numeric(as.character(Year))) %>% 
-    dplyr::select(BA.Target,Year,standVar,paste0(standVar,".se")) %>% 
-    setNames(c("BA.Target","Year","Mean","SE")) %>%
-
-    ggplot(aes(x=Year,y=Mean,color=BA.Target,group=BA.Target))+
-    geom_point()+
-    geom_line()+
-    geom_errorbar(aes(ymin=Mean-SE, ymax=Mean+SE, width=.2),
-                   position=position_dodge(.9))+
-    facet_wrap(~BA.Target)+
-      scale_x_continuous(limits=c(1992,2019),breaks=c(1992,1994,1997,2009,2019),
-                     minor_breaks = c(1992,1994,1997,2009,2019),
-                     name="Measurement year")+ # stretch out x-axis
-  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5),
-        legend.position = "none")+
-    ylab(y.label)
-
-}
-
-standVol.plot<-
-  standPlot("Vol",y.label="Vol. (m³/ha)")+
-  theme(axis.text.x=element_blank(),axis.title.x=element_blank())
-
-standBA.plot<-
-  standPlot("BA",y.label="BA (m²/ha)")+
-  theme(axis.text.x=element_blank(),axis.title.x=element_blank())
-
-standK.plot<-
-  standPlot("SPH",y.label="Tree dens. (sph)")+
-  theme(axis.text.x=element_blank(),axis.title.x=element_blank())
-
-standQMD.plot<-standPlot("QMD",y.label="QMD (cm)")
-
-standPlots<-cowplot::plot_grid(standVol.plot,standBA.plot,standK.plot,ncol=1,
-                      align="v")
-
-# could put in standQMD.plot as well
 
@@
   mutate_at(vars("F.value","P.value"),round,5) %>% 
-  mutate(`Fixed effect`=rep(c("Treatment","Year","Trt.*Year"),times=3)) %>% 
+  mutate(`Fixed effect`=rep(c("Treatment","Year","Trt*Year"),times=3)) %>% 
   rename(`Sum sq.`="Sum.Sq",
@@
 
-```{r fig.height=10}
-
-standPlots
-```
-
-`r figNums("standPlots",caption="Mean plot attributes by treatment unit and year.
-Whiskers are one standard error of the mean.
-Vol., volume; BA, basal area; Tree dens., tree density; QMD, quadratic mean diameter")`
-
-*Note that sample size varies with year.
-Some plots were not measured in 1992.
-This likely explains the dip in basal area and volume in TU:20 from 1992 to 1994, despite no recorded mortality in 1992/1994.*
-
 ---------------
 
-`r tableNums("standFit.X",caption="Output from three separate linear models testing the effect of the interaction between treatment and year on plot-level volume, basal area and density." )`
-
-```{r stand variable model output}
-
-standFit.X %>% 
-  flextable() %>% 
-  align(align="left",part="all") %>% 
-  merge_v(j="Variable") %>% 
-  valign(valign="top",part="body") %>% 
-    hline(i=c(3,6,9),border=fp_border(color="black")) %>% 
-    fix_border_issues() %>% 
-  autofit()
-```
-*Trt, treatment; Sq., squares, Num., numerator; Den., denominator; DF, degrees of freedom*
-
 --------------------
-
-`r appNums("standFit.contrasts",caption="Contrasts of 1992-2019 stand structure development over 1992-2019 between treatments.
-Contrasts are shown for three variables: volume, basal area and stand density.")`
-
-```{r show contrasts between plot means}
-
-standFit.contrasts %>% 
-  dplyr::select(-`Year contrast`) %>% 
-  flextable() %>% 
-   align(align="left",part="all") %>% 
-  merge_v(j=c("Variable")) %>% 
-  valign(valign="top",part="body") %>% 
-    hline(i=c(3,6,9),border=fp_border(color="black")) %>% 
-    fix_border_issues() %>% 
-  autofit()
-
-```
 
@@
 
-```{r summarize changes in species composition}
+```{r caption for species composition plots}
+
+figNums("sppComp.plot",display=FALSE)
+
+```
+
+A comparison of proportional species composition by diameter class, treatment and year showed that fir and spruce were organized similarly between the high RBA and control units (`r figNums("sppComp.plot",display="cite")`).
+In 1992, the largest diameter classes were dominated by spruce, whereas fir was the dominant species in the smallest diameter classes.
+This pattern was found for volume, basal area, and tree density.
+We only show volume in this document.
+By 2019, the fir-spruce composition by distribution class was maintained in these two units, with small increases in spruce composition in the smallest diameter class.
+
+In the low RBA, the species composition distribution was different.
+Here, both species maintained a similar proportion of fir-leading across diameter classes smaller than 37.5cm DBH, whereas spruce formed the leading species in diameter classes above 37.5cm DBH, with the exception of the largest class.
+
+\newpage
+
+# Tables
+
+`r tableNums("fitBA.table",caption="Summary of mixed effects model of tree-level basal area increment")`
+
+```{r summarize tree level BAI model}
+
+```
+
+\newpage
+
+`r tableNums("fitBApct.table",caption="Summary of mixed effects model of tree-level basal area increment as percentage of initial tree size")`
+
+```{r summarize tree level BAI as percentage model}
+
+```
+
+\newpage
+
+`r tableNums("standFit.X",caption="Output from three separate linear models testing the effect of the interaction between treatment and year on plot-level volume, basal area and density." )`
+
+```{r stand variable model output}
+
+standFit.X %>% 
+  flextable() %>% 
+  align(align="left",part="all") %>% 
+  merge_v(j="Variable") %>% 
+  valign(valign="top",part="body") %>% 
+    hline(i=c(3,6,9),border=fp_border(color="black")) %>% 
+    fix_border_issues() %>% 
+  autofit()
+```
+*Trt, treatment; Sq., squares, Num., numerator; Den., denominator; DF, degrees of freedom*
+
+\newpage
+
+# Figures
+
+```{r tree BAI by treatment species and diameter class}
+
+# Create plot of periodic increment
+
+# Using estimated marginal means
+  emmeans(rg,~Init|BA.Target+Species,type="response") %>% 
+  as.data.frame() %>%
+  mutate(Init=ba2d(Init)) %>% # convert initial basal area back to diameter
+
+  # plotting
+  ggplot()+
+  aes(x=Init,y=response,fill=Species,color=Species)+
+  geom_point()+
+  geom_line()+
+  geom_errorbar(aes(ymin=response-SE, ymax=response+SE), width=.2,
+                 position=position_dodge(0.05))+
+  xlab("1992 DBH (cm)")+
+  ylab("Basal area increment from 1992-2019 (m²)")+
+  facet_wrap(~BA.Target)
+
+```
+
+`r figNums("treeBAI.plot",caption="Estimated marginal means of spruce and fir basal area increment (m²) across five inital DBH classes and three treatments.
+Whiskers are standard error of the mean.
+Responses are back-transformed from the model.")`
+
+\newpage
+
+```{r function to create figure of plot means, fig.height=10}
+
+# Plotting
+
+# Create function
+standPlot<-function(standVar,y.label) {
+
+    standData.X %>% 
+    mutate(Year=as.numeric(as.character(Year))) %>% 
+    dplyr::select(BA.Target,Year,standVar,paste0(standVar,".se")) %>% 
+    setNames(c("BA.Target","Year","Mean","SE")) %>%
+
+    ggplot(aes(x=Year,y=Mean,color=BA.Target,group=BA.Target))+
+    geom_point()+
+    geom_line()+
+    geom_errorbar(aes(ymin=Mean-SE, ymax=Mean+SE, width=.2),
+                   position=position_dodge(.9))+
+    facet_wrap(~BA.Target)+
+      scale_x_continuous(limits=c(1992,2019),breaks=c(1992,1994,1997,2009,2019),
+                     minor_breaks = c(1992,1994,1997,2009,2019),
+                     name="Measurement year")+ # stretch out x-axis
+  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5),
+        legend.position = "none")+
+    ylab(y.label)
+
+}
+
+standVol.plot<-
+  standPlot("Vol",y.label="Vol. (m³/ha)")+
+  theme(axis.text.x=element_blank(),axis.title.x=element_blank())
+
+standBA.plot<-
+  standPlot("BA",y.label="BA (m²/ha)")+
+  theme(axis.text.x=element_blank(),axis.title.x=element_blank())
+
+standK.plot<-
+  standPlot("SPH",y.label="Tree dens. (sph)")
+
+standQMD.plot<-standPlot("QMD",y.label="QMD (cm)")
+
+standPlots<-cowplot::plot_grid(standVol.plot,standBA.plot,standK.plot,ncol=1,
+                      align="v")
+
+# display plot
+standPlots
+```
+
+`r figNums("standPlots",caption="Mean plot attributes by treatment unit and year.  Whiskers are one standard error of the mean.  Vol., volume; BA, basal area; Tree dens., tree density")`
+
+\newpage
+
+```{r species composition plots by diameter class}
 
@@
 # create stand-level plots by diameter distribution
-sppVol.plot<-sppComp.plot("BA")
-
-figNums("sppVol.plot",display=FALSE)
-
-# no need to provide other figures, as they're very similar
+sppComp.plot("BA")
 
@@
 
-A comparison of proportional species composition by diameter class, treatment and year showed that fir and spruce were organized similarly between the high RBA and control units (`r figNums("sppVol.plot",display="cite")`).
-In 1992, the largest diameter classes were dominated by spruce, whereas fir was the dominant species in the smallest diameter classes.
-This pattern was found for volume, basal area, and tree density.
-We only show volume in this document.
-By 2019, the fir-spruce composition by distribution class was maintained in these two units, with small increases in spruce composition in the smallest diameter class.
+`r figNums("sppComp.plot",caption="Spruce-fir volume proportional composition by diameter class, treatment and year.")`
 
-In the low RBA, the species composition distribution was different.
-Here, both species maintained a similar proportion of fir-leading across diameter classes smaller than 37.5cm DBH, whereas spruce formed the leading species in diameter classes above 37.5cm DBH, with the exception of the largest class.
-
-```{r species composition plot}
-sppVol.plot
-
-```
-
-`r figNums("sppVol.plot",caption="Spruce-fir volume proportional composition by diameter class, treatment and year.")`
-
-*Note this figure is almost identical if we compare stems, basal area or volume. For brevity, we only show volume here.*
-
-\newpage
-
-# Tables
-
-`r tablesNums("fitBA.table",caption="Summary of mixed effects model of tree-level basal area increment")`
-
-```{r summarize tree level BAI model}
-
-```
-
-\newpage
-
-`r tablesNums("fitBApct.table",caption="Summary of mixed effects model of tree-level basal area increment as percentage of initial tree size")`
-
-```{r summarize tree level BAI as percentage model}
-
-```
-
-\newpage
-
-# Figures
-
-```{r tree BAI by treatment species and diameter class}
-
-# Create plot of periodic increment
-
-# Using estimated marginal means
-  emmeans(rg,~Init|BA.Target+Species,type="response") %>% 
-  as.data.frame() %>%
-  mutate(Init=ba2d(Init)) %>% # convert initial basal area back to diameter
-
-  # plotting
-  ggplot()+
-  aes(x=Init,y=response,fill=Species,color=Species)+
-  geom_point()+
-  geom_line()+
-  geom_errorbar(aes(ymin=response-SE, ymax=response+SE), width=.2,
-                 position=position_dodge(0.05))+
-  xlab("1992 DBH (cm)")+
-  ylab("Basal area increment from 1992-2019 (m²)")+
-  facet_wrap(~BA.Target)
-
-```
-
-`r figNums("treeBAI.plot",caption="Estimated marginal means of spruce and fir basal area increment (m²) across five inital DBH classes and three treatments.
-Whiskers are standard error of the mean.
-Responses are back-transformed from the model.")`
-
-\newpage
+*Note this figure is almost identical for compairsons of basal area, volume or tree density. For brevity, we only show basal area.*
 
@@
 `r appNums("heightDiam.plot",caption="Relationship between height (m) and diameter (cm) for ")`
+
+\newpage
+
+```{r height model fit}
+
+# Height model fit plot
+
+  X %>% 
+  filter(is.na(DBT)) %>% # filter for trees with no broken tops
+  mutate(h2=exp(fixef(htDiamFit)[1]+fixef(htDiamFit)[2]*log(DBH))) %>% 
+  ggplot(aes(x=Height,y=h2))+
+  geom_point()+
+  xlim(0,40)+
+  ylim(0,40)+
+  geom_smooth(method="lm",se=F)+
+  ylab("Predicted height (m) from diameter-height model")+
+  xlab("Measured height (m)")
+
+```
+
+`r appNums("htModelFit.plot",caption="Scatterplot of predicted height(m) vs actual height(m) from diameter-height model.")`
+
+\newpage
+
+`r appNums("standFit.contrasts",caption="Contrasts of 1992-2019 stand structure development over 1992-2019 between treatments.
+Contrasts are shown for three variables: volume, basal area and stand density.")`
+
+```{r show contrasts between plot means}
+
+standFit.contrasts %>% 
+  dplyr::select(-`Year contrast`) %>% 
+  flextable() %>% 
+   align(align="left",part="all") %>% 
+  merge_v(j=c("Variable")) %>% 
+  valign(valign="top",part="body") %>% 
+    hline(i=c(3,6,9),border=fp_border(color="black")) %>% 
+    fix_border_issues() %>% 
+  autofit()
+
+```
+
+\newpage
+
+# References

### 2020-12-19 manuscript_draft.Rmd
manuscript work

@@
 
-  diamClass.plot<-function(standVar,y.label) {
-
-  diamClass.spp %>% 
-   ggplot()+
-  geom_bar(aes(x=DBH.class,y=get(standVar),group=Species,fill=Species),stat="identity")+
-  # geom_point(data=filter(diamClass.spp,Species=="Total"),
-  #              aes(x=DBH.class,y=get(standVar)))+
-  #   geom_line(data=filter(diamClass.spp,Species=="Total"),
-  #              aes(x=DBH.class,y=get(standVar),group=Species))+  
-   scale_fill_manual(values=c("#F8766D", "#00BFCA","azure4")) +  # control color
-     theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5),
-        legend.position = "bottom")+
-    ylab(y.label)+
-    xlab("DBH class (cm)")+
-
-    # coord_flip()+
-    facet_grid(Year~BA.Target) %>% 
-    return()
-}
-
-# create stand-level plots by diameter distribution
-diamVol.plot<-diamClass.plot("Vol",y.label="Mean plot volume (m³/ha)")
-diamBA.plot<-diamClass.plot("BA",y.label="Mean plot basal area (m²/ha)")  
-diamK.plot<-diamClass.plot("SPH",y.label="Mean plot tree density (sph)")
-
-# diamClass.plot<-cowplot::plot_grid(diamVol.plot,diamBA.plot,diamK.plot,ncol=1)
 ```
@@
 
------------
+### Stand attributes by diameter class
 
----------------
-
---------------------
-
-### Stand attribute by diameter class
-
-Among treatments and throughout the stand development after treatment, the stands have maintained a reverse-J diameter class distribution (`r figNums("diamK.plot",display="cite")`).  
+Among treatments and throughout the stand development after treatment, the stands maintained a reverse-J diameter class distribution (`r figNums("diamK.plot",display="cite")`).  
 Immediately after the partial harvest, the diameter class distribution of volume differed between the treatment.
@@
 
+```{r plot species composition by diameter class}
+
+# Create plotting function
+
+diamClass.plot<-function(standVar,y.label) {
+diamClass.spp %>% 
+   ggplot()+
+  geom_bar(aes(x=DBH.class,y=get(standVar),group=Species,fill=Species),stat="identity")+
+  # geom_point(data=filter(diamClass.spp,Species=="Total"),
+  #              aes(x=DBH.class,y=get(standVar)))+
+  #   geom_line(data=filter(diamClass.spp,Species=="Total"),
+  #              aes(x=DBH.class,y=get(standVar),group=Species))+  
+   scale_fill_manual(values=c("#F8766D", "#00BFCA","azure4")) +  # control color
+     theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5),
+        legend.position = "bottom")+
+    ylab(y.label)+
+    xlab("DBH class (cm)")+
+
+    # coord_flip()+
+    facet_grid(Year~BA.Target) %>% 
+    return()
+}
+
+# create stand-level plots by diameter distribution
+diamVol.plot<-diamClass.plot("Vol",y.label="Mean plot volume (m³/ha)")
+diamBA.plot<-diamClass.plot("BA",y.label="Mean plot basal area (m²/ha)")  
+diamK.plot<-diamClass.plot("SPH",y.label="Mean plot tree density (sph)")
+
+# diamClass.plot<-cowplot::plot_grid(diamVol.plot,diamBA.plot,diamK.plot,ncol=1)
+```
+
 ```{r species composition plots by diameter class}
@@
 ```
+
+\newpage
 

### 2020-12-21 manuscript_draft.Rmd
j

@@
 
-### Estimating tree height from diameter
+### Tree height modeling
 Tree heights had a positive nonlinear relationship with diameter, for both species and across all time periods (`r appNums("heightDiam.plot",display="cite")`).
-A comparison of height-diameter models showed that intercepts and slopes of log-transformed heights did not vary significantly (i.e., p>0.05) between species or time periods (not shown), therefore a single height-diameter linear mixed-effects model was developed using `r X %>% filter(Species %in% c("Fir","Spruce")) %>% filter(Status=="Live") %>% filter(is.na(DBT)) %>% drop_na(Height) %>% nrow()` tree height-diameter observations in the dataset, with log-transformed diameter as the fixed effect, and tree nested within plot as random effects.
+We modeled tree height using a linear mixed effects model with log-transformed diameter as the fixed effect, and tree nested within plot as a random effect.
+We originally developed separate models for each species and time period, however, a comparison of height-diameter models showed that intercepts and slopes odid not vary significantly (i.e., p>0.05) between species or time periods (not shown).
+Therefore, we developed a single height-diameter linear mixed-effects model was developed using `r X %>% filter(Species %in% c("Fir","Spruce")) %>% filter(Status=="Live") %>% filter(is.na(DBT)) %>% drop_na(Height) %>% nrow()` tree height-diameter observations in the dataset.
 This model explained `r round(MuMIn::r.squaredGLMM(htDiamFit)[1]*100,1)`% of variation in log-transformed heights, with an intercept of `r round(fixef(htDiamFit)[1],3)` and slope coefficient of  `r round(fixef(htDiamFit)[2],3)` * log-transformed DBH (not shown).
@@
 
-_Observations:_
-
-1. All units increased volume, basal area and stand density over the 1992-2019 period (`r figNums("standPlots",display="cite")`).
-
-2.
+All units increased volume, basal area and stand density over the 1992-2019 period (`r figNums("standPlots",display="cite")`).
 The rate of change differed between units.
 The low RBA unit accrued basal area faster than the other two units over the 1992-2019 period, and this difference was significant (`r appNums("standFit.contrasts",display="cite")`).
-The estimated rate of basal area increment in the low RBA unit was higher by by 8.4 and 7.8m² per hectare more than the high RBA and control units, respectively.
+The estimated rate of basal area increment in the low RBA unit was higher by
+`r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Basal area") %>%   pull(estimate) %>% round(.,1) %>%  abs() %>%   .[1]`
+and
+`r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Basal area") %>%   pull(estimate) %>% round(.,1) %>%  abs() %>%   .[2]`
+m² per hectare more than the high RBA and control units, respectively (p<0.01 for both). The low RBA unit also significantly increased its stand density by an estimated 
+`r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Stand density") %>%   pull(estimate) %>% round(.,1) %>%  abs() %>%   .[2]`
+stems per hectare more than the control unit (p<0.001). The rate of change in tree density did not differ significantly between the two harvested units.
 
-3.
-The low RBA unit also significantly increased its stand density by an estimated 340 stems per hectare more than the control unit.
-The rate of change in tree density did not differ significantly between the two harvested units.
-
-4. The rate of change in volume in trees over 17.5cm DBH was barely significant between the two harvested units, and there were no significant differences between the control and harvested units.
+The rate of change in volume in trees over 18.5cm DBH was different between barely significant between the two harvested units, and there were no significant differences between the control and harvested units.
 
@@
 Basal area trends were very similar to volume, and are not shown.
-
------------
-
-```{r}
-diamK.plot
-
-```
-
-`r figNums("diamVol.plot",caption="Diameter-class distribution of tree density (stems/hectare) by species, treatment and year.")`
-
----------------
-
-```{r}
-diamBA.plot
-
-```
-
-`r figNums("diamVol.plot",caption="Diameter-class distribution of volume (m³/hectare) by species, treatment and year.")`
-
-*Note that in this figure, we show volume for all diameter classes.
-This is different than the stand-level estimate of volume, where we only estimate volume for trees that would be considered merchantable under conventional harvest in British Columbia (i.e., minimum 18.5cm DBH).*
-
----------------
 
@@
 ```
+
+```{r show volume by diameter class}
+diamVol.plot
+
+```
+
+`r figNums("diamVol.plot",caption="Diameter-class distribution of volume (m³/hectare) by species, treatment and year.")`
+
+*Note that in this figure, we show volume for all diameter classes.
+This is different than the stand-level estimate of volume, where we only estimate volume for trees that would be considered merchantable under conventional harvest in British Columbia (i.e., minimum 18.5cm DBH).*
+
+\newpage
 

### 2020-12-31 manuscript_draft.Rmd
sdfds

@@
 
-Among treatments and throughout the stand development after treatment, the stands maintained a reverse-J diameter class distribution (`r figNums("diamK.plot",display="cite")`).  
-Immediately after the partial harvest, the diameter class distribution of volume differed between the treatment.
+Among treatments and throughout the stand development after treatment, the stands maintained a negative exponential (also referred to as 'reverse-J') diameter class distribution (`r figNums("diamK.plot",display="cite")`).
+Immediately after the partial harvest, the diameter class distribution of volume differed among treatments.
 In the 27 years since partial harvest, volume in all three treatment units was mostly concentrated in the 22.5-27.5cm DBH class, and mostly comprised of fir (`r figNums("diamVol.plot",display="cite")`).
 Over both time periods shown in the figure and all three treatments, most of the spruce volume was concentrated in DBH classes spanning 37.5 to 47.5cm.
-Basal area trends were very similar to volume, and are not shown.
+Basal area trends were very similar to volume (not shown).
 
@@
 
-A comparison of proportional species composition by diameter class, treatment and year showed that fir and spruce were organized similarly between the high RBA and control units (`r figNums("sppComp.plot",display="cite")`).
+A comparison of proportional species composition by diameter class, treatment and year showed that fir and spruce proportion by diameter class varied similarly between the high RBA and control units (`r figNums("sppComp.plot",display="cite")`).
 In 1992, the largest diameter classes were dominated by spruce, whereas fir was the dominant species in the smallest diameter classes.
@@
 
-In the low RBA, the species composition distribution was different.
-Here, both species maintained a similar proportion of fir-leading across diameter classes smaller than 37.5cm DBH, whereas spruce formed the leading species in diameter classes above 37.5cm DBH, with the exception of the largest class.
+In the low RBA, the species composition distribution by diameter class was different.
+Here, both fir and spruce maintained similar proportions across diameter classes smaller than 37.5cm DBH.
+Spruce formed the leading species in diameter classes above 37.5cm DBH, with the exception of the largest class.
+
+# Discussion
+
+## Species differences in growth response to partial harvest
+
+At the tree-level, our results suggest that while both species respond to thinning with faster radial growth, heavy thinning can elicit a stronger radial growth response in spruce, compared to fir.
+This coincides with our finding of proportionally higher spruce volume and basal area in the heavy thinning treatment in 2019, compared to the other treatments.
+Where thinning treatments in this region retain both spruce and fir, high-removal treatments may favour spruce release.
+Other studies have shown that residual spruce growth responds favourably to heavy thinning.
+A 10-year analysis of white spruce growth underneath an aspen canopy found that a complete overstory removal and heavy basal area removal resulted in immediate and significant spruce radial and height growth responses, compared to control and partial cutting treatments (Smith et al.
+2016).
+In a 35-year old white spruce stand, diameter increment was maximized with the heaviest thinning intensity, however, the treatment in this study was thinning from below (Stiell 1970).
+Frank (1973) also showed that 70-year old white spruce basal increment increased with the amount of thinning.
+Competition for light is a key factor that limits white spruce growth, and opening up the stand to allow for increased light penetration enhances white spruce growth responses [@Lieffers1994].
+High basal area removal may also increase stand air and soil temperatures, which might preferentially favour spruce growth responses compared to fir (Lacjcerowicz et al.
+2004).
+
+Other comparative studies of spruce and fir growth rates present differing evidence regarding species differences in release potential to thinning.
+In mature Engelmann spruce (Picea engelmannii) – subalpine fir forests in western North America, understory spruce growth rates exceeded those of fir, allowing spruce to achieve canopy position faster, including after canopy gaps were created through disturbance (Antos et al.
+2000, Andrus et al.
+2018).
+In mature forest in central BC, spruce growth exceeded that of subalpine fir after age 80 (Eis and Craigdallie 1983).
+Because spruce tends to live longer than subalpine fir, higher growth rates at later ages may allow this species to become larger than fir.
+In contrast, thinning studies in spruce-fir forests in eastern North America report that balsam fir (Abies balsamifera) responds to release with faster height and diameter growth than co-occurring red spruce (Picea rubens) (Dumais and Prevost, 2007; Dumais et al.
+2014, Messier et al.
+1999).
+Although both species are similar in shade tolerance, firs are considered to be better adapted to low light conditions in the understory, allowing them to persist longer in undisturbed forest  (Antos 2000) and potentially able to respond faster to changes in understory light (Dumais et al 2014, Messier et al.
+1999).
+Other studies have found no difference in growth response between spruce and fir species after treatment (MCaughey and Smith1982).
+Our study also suggests that at lower levels of basal area removal, spruce and fir growth responses will be similar.
 

### 2021-01-01 manuscript_draft.Rmd
more manny changes

@@
 
+```{r periodic increment by treatment level}
+
+# Create a function to compute standard error for vectors with NA
+incSE<-function(x) {
+
+  x2<-
+    x %>% 
+    .[!is.na(.)]
+
+    sd(x2)/sqrt(length(x2)) %>% 
+    return()
+  }
+
+perInc.X<-
+  plotData.X %>% 
+  pivot_longer(cols=Vol.sum:QMD.mean,names_to="var",values_to = "val") %>% 
+  pivot_wider(names_from = Year,values_from = val) %>% 
+  mutate(Period1=(`1994`-`1992`)/3,
+         Period2=(`1997`-`1994`)/3,
+         Period3=(`2009`-`1997`)/11,
+         Period4=(`2019`-`2009`)/11) %>% 
+  dplyr::select(BA.Target:var,Period1:Period4) %>% 
+  pivot_longer(cols=Period1:Period4,names_to="Period",values_to="Values") %>% 
+  pivot_wider(names_from=var,values_from=Values) %>%
+
+  # now summarize at treatment unit level
+  group_by(BA.Target,Period) %>% 
+   summarise(Vol=mean(Vol.sum,na.rm=T),
+            Vol.se=incSE(Vol.sum),
+            BA=mean(BA.sum,na.rm=T),
+            BA.se=incSE(BA.sum),
+            SPH=mean(SPH.sum,na.rm=T),
+            SPH.se=incSE(SPH.sum),
+            QMD=mean(QMD.mean,na.rm=T),
+            QMD.se=incSE(QMD.mean))
+
+# create captions for figures and table
+figNums("perInc.plot",display=FALSE)
+appNums("perInc.table",display=FALSE)
+
+```
+
 ```{r fit model to plot data}
@@
 
+```{r plot periodic increment}
+
+periodInc.plot<-function(standVar,y.label){
+
+  perInc.X %>%
+  dplyr::select(BA.Target,Period,standVar,paste0(standVar,".se")) %>% 
+    setNames(c("BA.Target","Period","Mean","SE")) %>%
+
+  ggplot(aes(x=Period,y=Mean,group=BA.Target,color=BA.Target))+
+  geom_point()+
+  geom_line()+
+     geom_errorbar(aes(ymin=Mean-SE, ymax=Mean+SE, width=.2),
+                   position=position_dodge(0)) %>% 
+    return()
+}
+
+periodInc.plot("BA",y.label="BA (m²/ha)")
+periodInc.plot("Vol",y.label="BA (m²/ha)")
+
+```
+
 ```{r plot species composition by diameter class}

### 2021-01-03 manuscript_draft.Rmd
manny commit

@@
 Total and merchantable volumes were estimated for trees using species- and region-specific equations [@nigh2016] based on tree height and DBH.
-Consistent with forest management practices in British Columbia, we estimated volumes for trees meeting a minimum diameter utilization limit of 17.5cm DBH.
+Consistent with forest management practices in British Columbia, we estimated volumes for trees meeting a minimum diameter utilization limit of 18.5cm DBH.
 For trees lacking a tree height measurement, we estimated height by fitting a single diameter-height model developed with a linear mixed effects model, with log-transformed DBH as a fixed effect, and tree nested within plot as a random effect.
@@
             QMD=mean(QMD.mean,na.rm=T),
-            QMD.se=incSE(QMD.mean))
+            QMD.se=incSE(QMD.mean)) %>%
+
+  # rename periods
+  mutate(Period=dplyr::recode(Period,
+                       Period1="1992-1994",
+                       Period2="1994-1997",
+                       Period3="1997-2009",
+                       Period4="2009-2019"))
 
@@
 All units increased volume, basal area and stand density over the 1992-2019 period (`r figNums("standPlots",display="cite")`).
-The rate of change differed between units.
+Mean annual basal area and volume increment peaked during the 1997-2009 period in the two harvested treatments, and peaked in the 1994-1997 period in the control treatment unit (`r figNums("perInc.plot",display="cite")`.
 The low RBA unit accrued basal area faster than the other two units over the 1992-2019 period, and this difference was significant (`r appNums("standFit.contrasts",display="cite")`).
@@
 `r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Stand density") %>%   pull(estimate) %>% round(.,1) %>%  abs() %>%   .[2]`
-stems per hectare more than the control unit (p<0.001). The rate of change in tree density did not differ significantly between the two harvested units.
-
+stems per hectare more than the control unit (p<0.001).
+The rate of change in tree density did not differ significantly between the two harvested units.
 The rate of change in volume in trees over 18.5cm DBH was different between barely significant between the two harvested units, and there were no significant differences between the control and harvested units.
@@
 Spruce formed the leading species in diameter classes above 37.5cm DBH, with the exception of the largest class.
+
+## Mortality
 
@@
 
+## Management considerations
+1.
+Both species responded with increased growth after partial harvest.
+High removal of basal area in these forests stimulated stronger spruce growth.
+This information can help silviculturists manipulate stand structure to preferentially ...
+2.
+Partial harvesting at both intensities delayed peak volume and basal area increment by approximately 10 years.
+Annual increment in the control unit declined faster over the last measurement period compared to previous periods, perhaps suggesting that the stand is becoming mature and losses due to mortality may substantially offset growth increment.
+
 \newpage
@@
 
-periodInc.plot<-function(standVar,y.label){
+# Note this function performs the operation for basal area and volume only
 
   perInc.X %>%
-  dplyr::select(BA.Target,Period,standVar,paste0(standVar,".se")) %>% 
-    setNames(c("BA.Target","Period","Mean","SE")) %>%
-
-  ggplot(aes(x=Period,y=Mean,group=BA.Target,color=BA.Target))+
+  dplyr::select(BA.Target:BA.se) %>% 
+    setNames(c("BA.Target","Period","Vol_mean","Vol_SE","BA_mean","BA_SE")) %>% 
+    pivot_longer(cols=-c(BA.Target,Period),
+                 names_to=c("Var","SE"),
+                 names_sep="_",
+                 values_to="Value") %>% 
+    pivot_wider(names_from="SE",values_from="Value") %>%
+
+  ggplot(aes(x=Period,y=mean,group=BA.Target,color=BA.Target))+
   geom_point()+
   geom_line()+
-     geom_errorbar(aes(ymin=Mean-SE, ymax=Mean+SE, width=.2),
-                   position=position_dodge(0)) %>% 
-    return()
-}
-
-periodInc.plot("BA",y.label="BA (m²/ha)")
-periodInc.plot("Vol",y.label="BA (m²/ha)")
-
-```
+
+     geom_errorbar(aes(ymin=mean-SE, ymax=mean+SE, width=.2),
+                   position=position_dodge(0))+
+    facet_wrap(~Var,nrow=2,
+               scales="free_y", strip.position = "left",
+               labeller=as_labeller(c(BA="Basal area inc. (m²/ha/year)",Vol="Volume inc. (m²/ha/year)")))+
+    ylab(NULL)+
+     theme(strip.background = element_blank(),
+           strip.placement = "outside")
+
+```
+
+`r figNums("perInc.plot",caption="Mean plot annual increment over four measurement periods for basal area and volume, by treatment.  Whiskers are one standard error of the mean. Inc, increment.")`
 

### 2021-01-04 manuscript_draft.Rmd
manny work

@@
 
+```{r generate species composition data}
+
+sppComp.X<-
+  X %>% 
+  filter(Status=="Live") %>% # filter for live trees
+  filter(Species %in% c("Spruce","Fir")) %>% # filter for spruce and fir trees
+
+  # Compute basal area and volume
+ left_join(filter(volCoef,VolType=="total"),by="Species") %>% 
+  mutate(vol=exp(1)^b0*DBH^b1*Height.Pred^b2) %>% 
+  mutate(BA=d2ba(DBH)) %>% 
+  dplyr::select(BA.Target,Plot:Species,Year,DBH,BA,vol) %>% 
+  mutate(Species=factor(Species)) %>%
+
+  # Need to add zero values for spruce in Plot 19
+  add_row(BA.Target="TU:10",Plot=19,Year=c(1992,1994,1997,2009,2019),Species="Spruce",DBH=20,vol=0,BA=0) %>%
+
+  # Remove volumes below 18.5cm DBH
+  mutate(vol=replace(vol,DBH<18.5,0)) %>%
+
+  # summarise 
+  group_by(BA.Target,Plot,Species,Year) %>% 
+  summarise(sph=n()*20,
+            BA=sum(BA)*20,
+            Vol=sum(vol)*20) %>% 
+  ungroup() %>%
+
+  # pivot_longer
+  pivot_longer(cols=sph:Vol,names_to="Var",values_to = "Value")
+
+plotTotal.X<-
+  sppComp.X %>% 
+  pivot_wider(names_from="Var",values_from = "Value") %>% 
+  group_by(BA.Target,Plot,Year) %>% 
+   summarise(sph=sum(sph),
+            BA=sum(BA),
+            Vol=sum(Vol)) %>% 
+  pivot_longer(cols=sph:Vol,names_to="Var",values_to="Plot.Total") %>% 
+  inner_join(sppComp.X,by=c("BA.Target","Plot","Year","Var")) %>%
+
+  # create proportion column
+  mutate(Prop=Value/Plot.Total) %>% 
+  ungroup() %>%
+
+  # a Few plots have no values in some years
+  drop_na(Prop)
+
+plotTotal.table<-
+  plotTotal.X %>% 
+  group_by(BA.Target,Year,Species,Var) %>% 
+  summarise(Prop=mean(Prop,na.rm=T))
+
+```
+
 ```{r caption for species composition plots}
@@
 figNums("sppComp.plot",display=FALSE)
-
-```
-
-A comparison of proportional species composition by diameter class, treatment and year showed that fir and spruce proportion by diameter class varied similarly between the high RBA and control units (`r figNums("sppComp.plot",display="cite")`).
+figNums("sppCompDiam.plot",display=FALSE)
+
+```
+
+A comparison of proportional species composition by diameter class, treatment and year showed that fir and spruce proportion by diameter class varied similarly between the high RBA and control units (`r figNums("sppCompDiam.plot",display="cite")`).
 In 1992, the largest diameter classes were dominated by spruce, whereas fir was the dominant species in the smallest diameter classes.
@@
 
+\newpage
+
+```{r plot species composition by attribute}
+
+```
+
 ```{r plot species composition by diameter class}
@@
 
-`r figNums("sppComp.plot",caption="Spruce-fir volume proportional composition by diameter class, treatment and year.")`
+`r figNums("sppCompDiam.plot",caption="Spruce-fir volume proportional composition by diameter class, treatment and year.")`
 

### 2021-01-05 manuscript_draft.Rmd
manny work

@@
 Variance explained for each model are reported using the approach described in Nakagawa et al.
-[-@Nakagawa2013], as implemented in the MuMIn package for R [@add later].
+[-@Nakagawa2013], as implemented in the MuMIn package for R [@barton2009mumin].
 
@@
 We modeled tree height using a linear mixed effects model with log-transformed diameter as the fixed effect, and tree nested within plot as a random effect.
-We originally developed separate models for each species and time period, however, a comparison of height-diameter models showed that intercepts and slopes odid not vary significantly (i.e., p>0.05) between species or time periods (not shown).
-Therefore, we developed a single height-diameter linear mixed-effects model was developed using `r X %>% filter(Species %in% c("Fir","Spruce")) %>% filter(Status=="Live") %>% filter(is.na(DBT)) %>% drop_na(Height) %>% nrow()` tree height-diameter observations in the dataset.
+We originally developed separate models for each species and time period, however, a comparison of height-diameter models showed that intercepts and slopes did not vary significantly (i.e., p>0.05) between species or time periods (not shown).
+Therefore, we developed a single height-diameter linear mixed-effects model using `r X %>% filter(Species %in% c("Fir","Spruce")) %>% filter(Status=="Live") %>% filter(is.na(DBT)) %>% drop_na(Height) %>% nrow()` tree height-diameter observations in the dataset.
 This model explained `r round(MuMIn::r.squaredGLMM(htDiamFit)[1]*100,1)`% of variation in log-transformed heights, with an intercept of `r round(fixef(htDiamFit)[1],3)` and slope coefficient of  `r round(fixef(htDiamFit)[2],3)` * log-transformed DBH (not shown).
 Residual and predicted vs actual height plots were assessed to ensure goodness of fit (`r appNums("htModelFit.plot",display="cite")`).
-This model was applied to predict heights in trees without height measurements, and tree-volume estimates generated as per Nigh [-@nigh2016].
+Using this model, we generated height estimates for trees without height measurements, and then estimated tree total volume as per Nigh [-@nigh2016].
 
@@
 # Slope Contrasts
-vol.contrast <- contrast(vol.means,interaction="pairwise")
-ba.contrast <- contrast(ba.means,interaction=c("pairwise")) 
-sph.contrast <- contrast(sph.means,interaction=c("pairwise"))
+vol.contrast <- contrast(vol.means,interaction=c("pairwise","revpairwise"))
+ba.contrast <- contrast(ba.means,interaction=c("pairwise","revpairwise")) 
+sph.contrast <- contrast(sph.means,interaction=c("pairwise","revpairwise"))
 
@@
   rbind(data.frame(Variable="Stand density",sph.contrast)) %>% 
-  filter(Year_pairwise %in% c("1992 - 2019")) %>% 
-  dplyr::select(Variable,`Year contrast`="Year_pairwise",`Treatment contrast`="BA.Target_pairwise",everything()) %>% 
+  filter(Year_revpairwise %in% c("2019 - 1992")) %>% 
+  dplyr::select(Variable,`Year contrast`="Year_revpairwise",`Treatment contrast`="BA.Target_pairwise",everything()) %>% 
   rename(`t ratio`="t.ratio",
@@
   rbind(data.frame(Variable="Stand density",sphM.contrast)) %>% 
-  filter(Year %in% c(1992,2019))
+  dplyr::select(Contrast=contrast,Variable,Year,everything())
 
@@
 appNums("standFit.contrasts",display=FALSE)  
+appNums("standFitM.contrasts",display=FALSE)
 
@@
 Mean annual basal area and volume increment peaked during the 1997-2009 period in the two harvested treatments, and peaked in the 1994-1997 period in the control treatment unit (`r figNums("perInc.plot",display="cite")`.
-The low RBA unit accrued basal area faster than the other two units over the 1992-2019 period, and this difference was significant (`r appNums("standFit.contrasts",display="cite")`).
-The estimated rate of basal area increment in the low RBA unit was higher by
-`r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Basal area") %>%   pull(estimate) %>% round(.,1) %>%  abs() %>%   .[1]`
+The low RBA unit accrued basal area significantly faster than the other two units over the 1992-2019 period (`r appNums("standFit.contrasts",display="cite")`).
+The estimated rate of basal area increment in the low RBA unit from 1992 to 2019 was higher by
+`r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Basal area") %>%   pull(estimate) %>% round(.,1) %>% .[1]`
 and
-`r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Basal area") %>%   pull(estimate) %>% round(.,1) %>%  abs() %>%   .[2]`
-m² per hectare more than the high RBA and control units, respectively (p<0.01 for both). The low RBA unit also significantly increased its stand density by an estimated 
-`r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Stand density") %>%   pull(estimate) %>% round(.,1) %>%  abs() %>%   .[2]`
+`r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Basal area") %>%   pull(estimate) %>% round(.,1) %>% .[2]`
+m² per hectare more than the high RBA and control units, respectively (p<0.01 for both).
+
+The low RBA unit also significantly increased its stand density by an estimated 
+`r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Stand density") %>%   pull(estimate) %>% round(.,0)  %>%   .[2]`
 stems per hectare more than the control unit (p<0.001).
@@
 ## Mortality
+
+```{r mortality data}
+
+# Create dead dataset
+
+# Dead species totals
+deadList<-
+  X %>% 
+  filter(Status=="Dead") %>% 
+  mutate(ID=paste(Plot,TreeID,sep="-")) %>% 
+  group_by(ID) %>% 
+  summarise_all(first)
+
+# Summarize mortality by treatment unit
+dead<-
+  X %>% 
+  filter(Status=="Dead") %>% 
+  mutate(ID=paste(Plot,TreeID,sep="-")) %>% 
+  mutate(Species=fct_recode(Species,`Douglas-fir`="Fd")) %>% 
+  group_by(ID) %>% 
+  summarise_all(first) %>% 
+  ungroup() %>% 
+  mutate(BA=d2ba(DBH)) %>% # convert to basal area per hectare 
+  group_by(BA.Target,DBH.class,Year) %>% 
+  summarise(BA=sum(BA),
+            k=n())
+
+# Summarize mortality by plot - this will be used for Figure
+deadPlot.X<-
+  X %>% 
+  filter(Status=="Dead") %>% 
+  mutate(ID=paste(Plot,TreeID,sep="-")) %>% 
+  mutate(Species=fct_recode(Species,`Douglas-fir`="Fd")) %>% 
+  group_by(ID) %>% 
+  summarise_all(first) %>% 
+  ungroup() %>% 
+  mutate(BA=d2ba(DBH)) %>% # convert to basal area per hectare 
+  group_by(BA.Target,Plot) %>% 
+  summarise(BA=sum(BA),
+            k=n()) %>% 
+  full_join(plotData.X,by=c("BA.Target","Plot")) %>% 
+  filter(Year==1997) %>% 
+  rename("BA.dead"=BA.x,"Num.dead"=k) %>% 
+  mutate(BA.dead=replace_na(BA.dead,0),
+         Num.dead=replace_na(Num.dead,0)) %>% 
+  mutate(prop.K=Num.dead/SPH,
+         prop.BA=BA.dead/BA.y)
+
+# linear model to analyze mortality by plot volume
+deadPlot.ANOVA<-
+  lm(Num.dead~Vol,data=deadPlot.X)
+
+dead.Summ<-
+  dead %>% 
+  ungroup() %>% 
+  group_by(BA.Target) %>% 
+  summarise(BA=sum(BA)/sum(dead$BA)*100,
+            k=sum(k)/sum(dead$k)*100) %>% 
+  mutate_if(is.numeric,round,1)
+
+figNums("mort.plot",display=FALSE)
+figNums("mortDiam.plot",display=FALSE)
+
+```
+
+A total of `r deadList %>% nrow()` trees (`r deadList %>% filter(Species=="Fir") %>% nrow()` fir, `r deadList %>% filter(Species=="Spruce") %>% nrow()` spruce, and `r deadList %>% filter(Species=="Fd") %>% nrow()` Douglas-fir) died in `r length(unique(deadList$Plot))` plots from 1997 to 2019.
+By treatment unit, the control unit had the highest mortality, in terms of trees and basal area (
+`r dead.Summ %>% filter(BA.Target=="TU:Control") %>% pull("k")` and 
+`r dead.Summ %>% filter(BA.Target=="TU:Control") %>% pull("BA")`% of tree mortality, respectively).  The high-removal treatment unit had the lowest mortality by number of trees and basal area (
+`r dead.Summ %>% filter(BA.Target=="TU:10") %>% pull("k")` and 
+`r dead.Summ %>% filter(BA.Target=="TU:10") %>% pull("BA")`% of tree mortality, respectively).
+Tree mortality was concentrated in the smaller diameter classes in the control and low-removal treatment unit, and most mortality occured during the most recent measurement period of 2009-2019 (
+`r figNums("mort.plot",display="cite")`).
 
@@
                scales="free_y", strip.position = "left",
-               labeller=as_labeller(c(BA="Basal area inc. (m²/ha/year)",Vol="Volume inc. (m²/ha/year)")))+
+               labeller=as_labeller(c(BA="Basal area inc. (m²/ha/year)",Vol="Volume inc. (m³/ha/year)")))+
     ylab(NULL)+
@@
 
-```
+plotTotal.table %>%
+  mutate(Var=fct_relevel(Var,"BA","Vol","sph")) %>% 
+  mutate(Var=dplyr::recode(Var,BA="Basal area",
+                    Vol="Volume",
+                    sph="Tree density")) %>%
+
+  ggplot()+
+  aes(x=Year,y=Prop,fill=Species)+
+  ylab("Proportion")+
+   geom_bar(position="fill",stat="identity")+
+   geom_hline(yintercept=0.5, linetype="dashed", color = "black")+
+  theme(legend.position = "bottom")+
+  facet_grid(Var~BA.Target)
+
+```
+
+`r figNums("sppComp.plot",caption="Mean proportion of spruce and fir basal area, volume, and tree density by treatment and year.")`
 
@@
 
+```{r show tree density by diameter class}
+diamK.plot
+
+```
+
+`r figNums("diamK.plot",caption="Diameter-class distribution of tree density (stems/hectare) by species, treatment and year.")`
+
+\newpage
+
 ```{r species composition plots by diameter class}
@@
 
+\newpage
+
+```{r plot mortality as a function of 1997 stand attributes}
+
+deadPlot.X %>% 
+  ggplot()+
+  aes(x=Vol,y=Num.dead,color=BA.Target,shape=BA.Target)+
+  geom_point(size=4)+
+  stat_smooth(method = lm, formula = y ~ x,se=FALSE,aes(group=1))+
+  ylab("Tree mortality from 1997-2019")+
+  xlab("Plot 1997 volume (m³/hectare) ")+
+  theme(legend.position = "bottom")
+
+```
+
+```{r plot mortality by treatment}
+
+dead %>% 
+  pivot_longer(cols=BA:k,names_to = "Var",values_to = "Value") %>%
+
+  mutate(Var=factor(Var)) %>% 
+  mutate(Var=fct_relevel(Var,"k","BA")) %>%
+
+  mutate(Var=dplyr::recode(Var,BA="Basal area (m²)",k="Number of trees")) %>% 
+  ggplot()+
+  aes(x=DBH.class,y=Value,fill=Year)+
+  geom_bar(stat="identity")+
+  facet_grid(Var~BA.Target,
+               scales="free_y",
+               switch="y")+
+    ylab(NULL)+
+  xlab("DBH class (cm)")+
+     theme( strip.placement = "outside",
+            strip.background.y = element_blank(),
+            legend.position = "bottom",
+            axis.text.x = element_text(angle=270,vjust = 0.3))
+
+```
+
+`r figNums("mort.plot",caption="Total tree mortality recorded in plots from 1997-2019, presented as number of trees (top panel) and basal area (bottom panel).")`
+
+\newpage
+
+```{r plot mortality by year and DBH class}
+
+dead %>% 
+  pivot_longer(cols=BA:k,names_to = "Var",values_to = "Value") %>% 
+  mutate(Var=factor(Var)) %>% 
+  mutate(Var=fct_relevel(Var,"k","BA")) %>% 
+  mutate(Var=dplyr::recode(Var,BA="Basal area (m²)",k="Number of trees")) %>% 
+  ggplot()+
+  aes(x=DBH.class,y=Value,fill=Year)+
+  geom_bar(stat="identity")+
+  facet_grid(Var~BA.Target,
+               scales="free_y",
+               switch="y")+
+    ylab(NULL)+
+  xlab("DBH class (cm)")+
+     theme( strip.placement = "outside",
+            strip.background.y = element_blank(),
+            legend.position = "bottom",
+            axis.text.x = element_text(angle=270,vjust = 0.3))
+
+```
+
+`r figNums("mortDiam.plot",caption="Total tree mortality recorded in plots from 1997-2019, presented as number of trees (top panel) and basal area (bottom panel).")`
+
 # Appendices
@@
 
+`r appNums("standFitM.contrasts",caption="Contrasts of stand attributes (volume, basal area and tree density) between treatments.
+Contrasts are presented for five measurement years.
+SE, standard error, df, degrees freedom")`
+
+```{r table contrasts between plot means}
+
+standFitM.contrasts %>% 
+  arrange(Contrast,Variable,Year) %>% 
+  mutate_if(is.numeric,round,5) %>%
+
+  flextable() %>% 
+   align(align="left",part="all") %>% 
+  merge_v(j=c("Contrast","Variable","Year")) %>% 
+  valign(valign="top",part="body") %>% 
+    hline(i=seq(from=5,to=40,by=5),border=fp_border(color="black")) %>% 
+    fix_border_issues() %>% 
+  autofit()
+
+```
+
+\newpage
+
 `r appNums("standFit.contrasts",caption="Contrasts of 1992-2019 stand structure development over 1992-2019 between treatments.

### 2021-01-06 manuscript_draft.Rmd
Add text from Mike's emails

@@
 
+# To-do list
+
+Comments from Mike - we can put text below each comment to describe how it was addressed, and then delete comments for any issues once they are resolved.
+
+From November 6 e-mail:
+
+Use of terminology
+
+  *   The term "units" is used in several places, where it seems that you mean "treatments" (or treatment types, treatment unit).
+
+  *   The use of the term "plot" or "plots" is used appropriately as far as I can see.
+
+  *   The term "fir" is used to describe subalpine fir or 'balsam' in several parts of the text, including possibly figure and table captions.
+I know what species you are referring to, of course, but others, especially cursory readers, may misconstrue it.
+The colloquial use of 'fir" is spread across both Abies and Douglas-fir.
+"Balsam" as a term has well-documented ambiguities, and while Canadians tend to use balsam for Abies, Americans often call Abies "firs", as in the various species of true fir.
+I would recommend using Abies as the shorthand descriptor in the text and captions for subalpine fir.
+It is a short word, and we have only 1 Abies species at this site.
+Also, my experience in the on-line-search literature world and with things like Google Scholar is that Abies is a much better and more diagnostic search word for this subject matter than fir or balsam.
+"Spruce-Abies" is very unambiguous as a descriptor for this forest type.
+
+  *   Rate of change (increase, decrease) in vol, BA, and sph: In both in your stats tables and figure captions, and in the text narrative, ensure that what you are describing is clearly and accurately described as either:  (a) the incremental change in values between two measurement periods, or (b)  the difference in absolute value in certain parameters (e.g.
+standing volume) between two measurement periods, or between two treatments.
+Also, the annualized growth rate of something averaged over a measurement period (e.g.
+mean PBAI, like BA growth in m2/yr over a period) is different than the rate of change in a growth rate over time (e.g.
+is a growth rate increasing or decreasing).
+I got confused a few times on this issue.
+
+Commentary on sections and sub-sections
+
+Section 3.1.1 BA increment from 1992 to 2019
+
+  *   Great: Very clear and straightforward. i.e. - log-transformed treatment (initial BA) was a very statistically significant predictor of tree response across dbh classes for both Sx and Bl.
+  *   Don't know what all the random effects outputs are. I am sure they are meaningful, but I just dont know what they mean.
+  *   Number of observations of 510: Should ensure it is clear that only trees > 7.5 cm dbh are counted.
+  *   Would be good to define the dbh class limits somewhere. In Appendix 6 table, we describe the class mid-point.
+  *   For Appendix 6, I assume that the null hypothesis for the ratio is = 1.
+
+Section 3.1.2 Mortality
+
+  1.
+I am thinking that mortality should / could be analysed as a frequency response - i.e.
+- expected vs actual distribution of mortality by size class and/or species.
+e.g.
+a Chi-squared or allied type of analysis.
+Null hypothesis is that mortality of stems is randomly distributed across different sizes and species of trees.
+H1 is that it is non-random.
+  2.
+I'd suggest that we don't compare total mortality (e.g # of stems) across treatment units, but rather frequency relative to initial number of stems (i.e.
+- at beginning of observation period).
+That way we normalize mortality rates relative to initial stem densities.
+Otherwise, results could be mis-leading - e.g.
+Control treatments could have more mortality in overall sph, simply because they have more initial stems.
+  3.  ACTION ITEM FOR ME:  I need to check my original datasets and data sorting iterations to see how I handled mortality observations for the 1992 and 1994 plot remeasurements.
+  4.
+Note for you:  All of the mortality you've analysed and identified so far is mortality that is incremental following 92/94, so it is still solid analysis data, just with that qualification.
+I will clarify the 92/94 mortalilty observations for the Methods section; we did observe dead trees at these times, but these mortality pre-dated our treatments and/or first measurements.
+Anyway, probably not a big deal, but this needs examination and clarification to explain this.
+If necessary, I will quantify observations of dead trees observed on these years, because they are actually > 0.
+
+Testing for density-dependence of mortality responses... some ideas or hypotheses..  Note that f (___) means "a function of"...
+
+  *   MORT RATE of a size class = f (Sum of BA in that plot > than that size class).
+Would assume asymmetric competition where only bigger trees exert competition on smaller trees.
+  *   MORT RATE of a size class = f ( Sum of BA in that plot that is either > or < that size class).
+Assumes symmetric competition where both big and small trees exert competition on their neighbors.
+
+Caveats:   (a) our plots and tree distributions might be too spatially variable to get a reliable test of these hypotheses, and (b) BA would be the BA at the beginning of a measurement period.
+And (c) a tree that died of suppression would probably show reduced growth rates prior to death; but trees that had abiotic damage and mortality - e.g.
+- snow or wind breakage - might not.
+Complicated stuff !!
+
+3.2.1  Estimating tree height from diameter
+
+  *   Really solid stats and rationale for a single ht:dbh model. Great !
+  *   What is really interesting and useful about this is that the ht:dbh relationship is seemingly statistically independent of treatment or, that is, initial density,  A very useful simplifying assumption proven for future modellers of these stands.
+We should highlight this in either our results or discussion.
+
+3.2.2 Plot means
+
+  *   See notes above about use of term "units" vs treatment (or similar).
+  *   Clarify effects of treatment and/or species on growth rate vs standing vol / BA at each measurement period.
+  *   A Figure 7b or equivalent could visually present changes in growth rate over time. (e.g. previous graphs we've developed)
+  *   Under Observations, # 2 - the 2nd sentence needs work to edit and clarify.
+
+3.2.3 Stand attributes by diameter class
+
+  *   We should define what we mean by the term inverse J or inverse J-shaped dbh distribution. Just for clarity.
+  *   Should we quantitatively describe these distributions? Don't have to, just asking the question.
+  *   Is the merch limit for min merch tree size 18.5 cm dbh, or 17.5 cm dbh? I remember the latter from various cruises and cruise compilations.
+  *   There are two Figure 2's, one is for vol and one is for sph.
+  *   Clarify in figures and text that sph is based on trees 7.5 cm dbh and greater. Increases in sph can be from ingress of trees growing from size classes below this limit.
+  *   A suggestion:  Have a table comparing total volume to merch volume by treatment, etc?  Would be useful for some readers, and communicate MAI and volume growth rates too.
+
+3.2.4  Species composition
+
+  *   First sentence:  Clarify what you mean by "organized similarly". Does this refer to frequency distribution ? What constitutes similarity?
+
+# Introduction
+
+# Methods
+
 This section will describe the methods we used for data analyses, not field data collection.
@@
 2. Stand-level responses to treatment:
-
-```{r define captions to be used in methods results and discussion, echo=FALSE}
-
-# Captions (put these in order)
-
-  # Introductory 
-  appNums(name="paiTable",caption="Number of growing seasons between each measurement period",display=FALSE) # Table showing number of growing seasons between each measurement period
-  appNums(name="standVol.plot",caption="Plot total volume for five measurement periods",display=FALSE) # Plot volume figure
-  appNums(name="standBA.plot",caption="Plot basal area for five measurement periods",display=FALSE)  # Plot basal area figure
-  appNums(name="standQMD.plot",caption="Plot quadratic mean diameter for five measurement periods",display=FALSE) # QMD figure  
-  appNums(name="standK.plot",caption="Plot tree density for five measurement periods", display=FALSE) # SPH figure
-
-  # Diameter class analysis
-  figNums("diamVol.plot", caption="Mean plot volume distribution by diameter class and treatment, for two periods.
-Black line shows total volume.
-Volume calculated on trees meeting 17.5cm DBH limit.
-Spruce and fir contribution to volume by diameter class are shown with bar plots.
-Douglas-fir volume contribution by diameter class is not shown.",display="FALSE")
-
-  figNums("diamBA.plot", caption="Mean plot basal area distribution by diameter class and treatment, for two periods.
-Black line shows total basal area.
-Spruce and fir contribution by diameter class are shown ith bar plots.
-Douglas-fir contribution by diameter class is not shown.",display="FALSE")
-
-  figNums("diamK.plot", caption="Mean tree density distribution by diameter class and treatment, for two periods.
-Black line shows total tree density.
-Spruce and fir contribution by diameter class are shown ith bar plots.
-Douglas-fir contribution by diameter class is not shown.",display="FALSE")
-
-```
 
@@
 To test the hypothesis that changes in stand attributes over 27 years varied by treatment, we fit separate linear models with stand total volume, basal area, and density as response variables, and the interaction between treatment and year as the independent variable.
+
+### Mortality
+
+From Mike's e-mail:
+
+Last winter, as well as keeping reference copies of the initial “raw” dataset following data entry and related stuff, I kept copies each stage of data cleaning and prep (about 10 iterations).
+
+We did have “dead tree” observations in the first ~ 2 year of monitoring. The key thing to see is : if I can determine is when the trees were dead, and/or when death was initiated, and why.  
+Timing of tree death and observations of causal factors of tree death in these first two years will help inform our interpretation and data handling of these observations.
+
+If trees were dead or moribund at the first “initial” observation or two (as soon after the logging as possible), this is – arguably -  different than mortality that occurs over time in response to the basal area treatment and/or could be considered to be density- or treatment-dependent in some way.
+
+The 1992 data was in mid-March really early after logging, and before snowmelt (in retrospect not a great idea but we had resources available before the end of the fiscal year at the time).
+The 1994 data was from when the plots were re-visited under snow-free conditions.
+We did a thinning and sanitation culling of trees in the treatment units and PSP’s in 1994 to remove logging-damaged trees and most deciduous, to bring the units and PSP’s down to their final treatment BA.
+1994 remeasurements were completed after that.
+
+A few plots, as you know, were established in 1994 not 1992.
+
+Interpretive implications:
+
+1.	The very-early observed mortality may be a result of pre-existing stand conditions not the created, treatment-induced stand conditions. “stochastic” vs potentially treatment induced. 
+2.	It can be argued that trees dead at time of treatment or very soon  thereafter were never taking up growing space in the post-harvest stand, and therefore would not contribute to the initial “green” basal area going forward.
+3.	We do have 1992 data in most plots, but it was 1994 when the BA in each trmt unit and embedded PSP was finalized. 1994 is the most homogeneous start point for all PSP’s.
+
+Regarding tree mortality in the period from 1992 to 1994 at Summit Lake EP 1162:
+
+I reviewed our data regarding this, any relevant project records, and my own recollection of our monitoring of tree mortality during this period.
+Here is my train of thought, and what I have concluded on this topic:
+
+1992 Observations and PSP establishment limitations
+
+Trees that were dead at the time of the 1992 PSP establishment were not tagged or recorded.
+The rationale for this was that these trees were dead and therefore not taking up up growing space in the stands.
+Dead trees in 1992 would have been either snags / dead trees pre-existing the harvest, or trees fatally damaged or killed by logging activities (e.g.
+– trees with stems snapped and crowns broken off, no live limbs) or trees smashed to the ground by felled trees).
+
+Summer 1994 sanitation treatment and Fall 1994 PSP Remeasurement
+
+EP 1162 treatment units, including the PSP area contained therein, were assessed in Spring 1994 to review stand density / basal area, and stand condition, in order to:  
+a)	identify and mark-to-cut residual trees that were considered too badly scarred or damaged to constitute good growing stock for trial purposes, 
+b)	where needed, to lower the initial BA closer to the target level, 
+c)	to either fell or girdle birch trees, especially in the PSP’s themselves, but usually also within the TU as a whole, and 
+d)	to fell any dead, dangerous, and/or moribund trees within the treatment units and PSP’s. 
+Generally most if not all of these trees felled in the sanitation felling were understory or subdominant trees, and occasionally a lower codominant tree.
+Tagged trees felled by the sanitation felling were documented on field sheets and our Excel datasets, so we can retrieve the species and size of these trees as needed.
+However, sanitation-felled trees included both live, moribund, and dead trees – generally all poorer quality or vigor though.
+
+After felling, the felled trees were bucked to a 2 to 3 m length so as to lay these felled trees to as close to the ground as possibly, maximize decay, and minimize physical obstructions within the TU’s and PSP’s.
+
+As a result of the sanitation felling, potentially some trees which died or were damaged between 1992 and Summer 1994 were felled without recording these mortality events and precise tree vigor beforehand.
+Some minor mortality occurred between 1992 and 94, but was obscured by the post-harvest sanitation treatment.
+
+The 1994 Remeasurement occurred in Fall 1994, with a few plot remeasurements occurring in Winter 1994/95 (i.e. no intervening growing season).
+
+For scientific rigour, we should not state that there was zero mortality between 1992 and 1994.
+Rather, I suggest, it is more accurate to state that data on tree mortalilty between 1992 and 1994 was not collected, and that the Summer 1994 sanitation felling treatment did not allow accurate documentation of tree mortality in this initial period on a retrospective basis.
+
+We don’t, therefore, have data on mortality between 1992 and 1994, though my recollection is that it was minor in nature.
+
+The difference in basal area between 1992 and 1994 is predominantly a product of the sanitation felling of some basal area, with minor natural mortality.
+
+Mortality occurring between Fall 1994 and Fall 1997 remeasurements was captured accurately by these two sequential remeasurements.
+
+“The difference in basal area between 1992 and 1994 is a result of basal area reductions through the sanitation felling of some stems and some minor natural mortality, combined with basal area increases through the growth of remaining live stems.”
+
+A review of the various plots indicates that between 1992 and 1994, some PSP’s decreased in BA (as sanitation removals and minor mortality exceeded growth of live stems), while others increased in BA as tree growth exceeded BA removals.
 

### 2021-01-07 manuscript_draft.Rmd
manny work

@@
 
+\newpage
+
 # Introduction
@@
 A review of the various plots indicates that between 1992 and 1994, some PSP’s decreased in BA (as sanitation removals and minor mortality exceeded growth of live stems), while others increased in BA as tree growth exceeded BA removals.
+
+  *   Is there any change in which sizes or size classes of trees die as stand volume increases?
+- e.g.
+is competition asymmetrical or symmetrical?
+Currently the y axis is total number of trees, I believe.
+  *   Not sure if we have enough sample size of trees to look at species dynamics (Sx vs Bl).
+  *   I am wondering if the same relationship that appears between volume and tree mortality, is mirrored in the relationship between BA and tree mortality. A bit more on this below...
+
+Foresters (like me) do like to use BA per ha as a density measure for field planning and stand assessment because it seems to be broadly useful and is easily assessed via stand walkthroughs or plots.
+Sapwood BA is pretty well correlated with foliar biomass of the stand - a biological linkage.
+Tree BA is often correlated well with things like crown width.
+
+I hadn't thought alot previously about vol per ha as a density measurement, but it does have the characteristic of integrating not just BA, but also tree height, so it integrates tree growing space occupancy in both in horizontal dimensions (crown width and probably root spread) but also vertical.
+Vol ha theoretically provides more of a 3D measure of how the trees are occupying a given land area.
+The fact that volume is of commercial interest is a bonus.
+Key assumptions here: That trees are healthy and most importantly, crowns are health and growing relatively optimally relative to their growing space opportunities.
+Once crown vigor etc starts to decline with age, pests or pathogens, then the relationship could start to break down.
 
@@
   filter(Status=="Dead") %>% 
+  # filter(Species=="Spruce") %>% #dot this if you want to summarize by species
   mutate(ID=paste(Plot,TreeID,sep="-")) %>% 
@@
   full_join(plotData.X,by=c("BA.Target","Plot")) %>% 
-  filter(Year==1997) %>% 
-  rename("BA.dead"=BA.x,"Num.dead"=k) %>% 
+  filter(Year==1994) %>% 
+  rename("BA.dead"=BA.x,"Num.dead"=k,"BA"=BA.y) %>% 
   mutate(BA.dead=replace_na(BA.dead,0),
@@
   mutate(prop.K=Num.dead/SPH,
-         prop.BA=BA.dead/BA.y)
+         prop.BA=BA.dead/BA)
 
@@
 deadPlot.ANOVA<-
-  lm(Num.dead~Vol,data=deadPlot.X)
+  lm(Num.dead~BA,data=deadPlot.X)
 
@@
 figNums("mortDiam.plot",display=FALSE)
-
-```
-
-A total of `r deadList %>% nrow()` trees (`r deadList %>% filter(Species=="Fir") %>% nrow()` fir, `r deadList %>% filter(Species=="Spruce") %>% nrow()` spruce, and `r deadList %>% filter(Species=="Fd") %>% nrow()` Douglas-fir) died in `r length(unique(deadList$Plot))` plots from 1997 to 2019.
-By treatment unit, the control unit had the highest mortality, in terms of trees and basal area (
-`r dead.Summ %>% filter(BA.Target=="TU:Control") %>% pull("k")` and 
-`r dead.Summ %>% filter(BA.Target=="TU:Control") %>% pull("BA")`% of tree mortality, respectively).  The high-removal treatment unit had the lowest mortality by number of trees and basal area (
-`r dead.Summ %>% filter(BA.Target=="TU:10") %>% pull("k")` and 
-`r dead.Summ %>% filter(BA.Target=="TU:10") %>% pull("BA")`% of tree mortality, respectively).
-Tree mortality was concentrated in the smaller diameter classes in the control and low-removal treatment unit, and most mortality occured during the most recent measurement period of 2009-2019 (
-`r figNums("mort.plot",display="cite")`).
+figNums("mortSppDiam.plot",display=FALSE)
+
+```
+
+A total of `r deadList %>% nrow()` trees (`r deadList %>% filter(Species=="Fir") %>% nrow()` fir, `r deadList %>% filter(Species=="Spruce") %>% nrow()` spruce, and `r deadList %>% filter(Species=="Fd") %>% nrow()` Douglas-fir) died in `r length(unique(deadList$Plot))` plots from 1994 to 2019.
+The number of trees that died per plot from 1994 to 2019 increased significantly with plot basal area in 1994 (`r figNums("mort.plot",display="cite")`, R^2 = `r glance(deadPlot.ANOVA) %>% .$r.squared %>% round(.,3)*100`, p = `r glance(deadPlot.ANOVA) %>% pull(p.value)`).
+Similar linear models fitted separately for fir and spruce showed that fir mortality had a similar relationship to plot basal area, whereas spruce mortality did not (not shown).
+The diameter distribution of fir mortality followed a negative exponential distribution, with mortality concentrated in smaller diameter classes (`r figNums("mortSppDiam.plot",display="cite")`).
+Spruce mortality was concentrated in larger diameter classes.
+By treatment unit, the control unit had the highest mortality, whereas the high-removal treatment unit had the lowest mortality (`r figNums("mortDiam.plot",display="cite")`).
+The majority of mortality occured during the most recent measurement period of 2009-2019.
 
@@
 
-At the tree-level, our results suggest that while both species respond to thinning with faster radial growth, heavy thinning can elicit a stronger radial growth response in spruce, compared to fir.
+At the tree-level, our results suggest that while both hybrid white sruce and subalpine fir respond to thinning with faster radial growth, heavy thinning can elicit a stronger radial growth response in spruce, compared to fir.
 This coincides with our finding of proportionally higher spruce volume and basal area in the heavy thinning treatment in 2019, compared to the other treatments.
-Where thinning treatments in this region retain both spruce and fir, high-removal treatments may favour spruce release.
+Where thinning treatments in this region retain both species, removing a relatively high amount of the stand may favour spruce release.
 Other studies have shown that residual spruce growth responds favourably to heavy thinning.
-A 10-year analysis of white spruce growth underneath an aspen canopy found that a complete overstory removal and heavy basal area removal resulted in immediate and significant spruce radial and height growth responses, compared to control and partial cutting treatments (Smith et al.
+A 10-year analysis of white spruce growth underneath an aspen canopy found that a complete or high overstory removal resulted in immediate and significant spruce radial and height growth responses, compared to control and partial cutting treatments (Smith et al.
 2016).
-In a 35-year old white spruce stand, diameter increment was maximized with the heaviest thinning intensity, however, the treatment in this study was thinning from below (Stiell 1970).
+In a 35-year old white spruce stand, diameter increment was maximized with the heaviest thinning intensity, however, the treatment thinning from below (Stiell 1970).
 Frank (1973) also showed that 70-year old white spruce basal increment increased with the amount of thinning.
-Competition for light is a key factor that limits white spruce growth, and opening up the stand to allow for increased light penetration enhances white spruce growth responses [@Lieffers1994].
-High basal area removal may also increase stand air and soil temperatures, which might preferentially favour spruce growth responses compared to fir (Lacjcerowicz et al.
-2004).
+Competition for light is a key factor that limits white spruce growth, and opening up the stand to allow for increased light penetration may be a primary mechanism to enhance white spruce growth responses [@Lieffers1999].
+High basal area removal may also increase stand air and soil temperatures, which can preferentially favour spruce growth responses compared to fir [@Lajzerowicz2004].
 
@@
 
-```{r plot mortality as a function of 1997 stand attributes}
+```{r plot mortality as a function of 1994 stand attributes}
+
+mortPlot.text<-paste("y = ",round(tidy(deadPlot.ANOVA)[1,"estimate"],4)," + ",round(tidy(deadPlot.ANOVA)[2,"estimate"],4),"x ",
+              "\np = ",round(tidy(deadPlot.ANOVA)[2,"p.value"],4),
+              "\nR² = ",round(glance(deadPlot.ANOVA)$r.squared, 3),sep="")
 
@@
   ggplot()+
-  aes(x=Vol,y=Num.dead,color=BA.Target,shape=BA.Target)+
+  aes(x=BA,y=Num.dead,color=BA.Target,shape=BA.Target)+
   geom_point(size=4)+
   stat_smooth(method = lm, formula = y ~ x,se=FALSE,aes(group=1))+
-  ylab("Tree mortality from 1997-2019")+
-  xlab("Plot 1997 volume (m³/hectare) ")+
-  theme(legend.position = "bottom")
-
-```
-
-```{r plot mortality by treatment}
-
-dead %>% 
-  pivot_longer(cols=BA:k,names_to = "Var",values_to = "Value") %>%
-
-  mutate(Var=factor(Var)) %>% 
-  mutate(Var=fct_relevel(Var,"k","BA")) %>%
-
-  mutate(Var=dplyr::recode(Var,BA="Basal area (m²)",k="Number of trees")) %>% 
-  ggplot()+
-  aes(x=DBH.class,y=Value,fill=Year)+
-  geom_bar(stat="identity")+
-  facet_grid(Var~BA.Target,
-               scales="free_y",
-               switch="y")+
-    ylab(NULL)+
-  xlab("DBH class (cm)")+
-     theme( strip.placement = "outside",
-            strip.background.y = element_blank(),
-            legend.position = "bottom",
-            axis.text.x = element_text(angle=270,vjust = 0.3))
-
-```
-
-`r figNums("mort.plot",caption="Total tree mortality recorded in plots from 1997-2019, presented as number of trees (top panel) and basal area (bottom panel).")`
+  ylab("Number of trees that died (1994-2019)")+
+  xlab("Plot 1994 basal area (m²/hectare) ")+
+  theme(legend.position = "bottom")+
+  geom_text(aes(x=7,y=10,label=mortPlot.text,hjust=0),color="black",check_overlap = TRUE,size=4)
+
+```
+
+`r figNums("mort.plot",caption="Total tree mortality over 1994-2019 period vs plot basal area in 1994.  Plots are symbolized to show three treatment types.")`
 
@@
 
+\newpage
+
+```{r plot mortality by species and diameter class}
+
+  deadList %>% 
+  filter(Species %in% c("Spruce","Fir")) %>% 
+  group_by(Species,DBH.class,BA.Target) %>% 
+  summarise(Num.Dead=n()) %>%
+
+  # plot
+  ggplot()+
+  aes(x=DBH.class,y=Num.Dead,fill=BA.Target) %>% 
+  geom_bar(stat="identity")+
+  facet_wrap(~Species)+
+  ylab("Number dead 1994-2019")+
+  xlab("DBH class (cm")+
+  theme(legend.position="bottom")
+```
+
+`r figNums("mortSppDiam.plot",caption="Fir and spruce mortality recorded in plots from 1997-2019, presented as number of trees, by diameter breast height class.
+Numbers identifying each diameter classes represent the lower size limit for that class.
+For example, trees in the 7.5cm DBH class are greater or equal to 7.5 and less than 12.5cm DBH.")`
+
+\newpage
+
 # Appendices

### 2021-01-07 manuscript_draft.Rmd
manuscript work

@@
 
-Comments from Mike - we can put text below each comment to describe how it was addressed, and then delete comments for any issues once they are resolved.
-
-From November 6 e-mail:
-
-Use of terminology
-
-  *   The term "units" is used in several places, where it seems that you mean "treatments" (or treatment types, treatment unit).
-
-  *   The use of the term "plot" or "plots" is used appropriately as far as I can see.
-
-  *   The term "fir" is used to describe subalpine fir or 'balsam' in several parts of the text, including possibly figure and table captions.
+### Comments from Mike's emails
+
+we can put text below each comment to describe how it was addressed, and then delete comments for any issues once they are resolved.
+
+#### Use of terminology
+
+- The term "units" is used in several places, where it seems that you mean "treatments" (or treatment types, treatment unit).
+
+- The use of the term "plot" or "plots" is used appropriately as far as I can see.
+
+- The term "fir" is used to describe subalpine fir or 'balsam' in several parts of the text, including possibly figure and table captions.
 I know what species you are referring to, of course, but others, especially cursory readers, may misconstrue it.
@@
 
-  *   Rate of change (increase, decrease) in vol, BA, and sph: In both in your stats tables and figure captions, and in the text narrative, ensure that what you are describing is clearly and accurately described as either:  (a) the incremental change in values between two measurement periods, or (b)  the difference in absolute value in certain parameters (e.g.
+- Rate of change (increase, decrease) in vol, BA, and sph: In both in your stats tables and figure captions, and in the text narrative, ensure that what you are describing is clearly and accurately described as either:  (a) the incremental change in values between two measurement periods, or (b)  the difference in absolute value in certain parameters (e.g.
 standing volume) between two measurement periods, or between two treatments.
@@
 
-Commentary on sections and sub-sections
-
-Section 3.1.1 BA increment from 1992 to 2019
-
-  *   Great: Very clear and straightforward. i.e. - log-transformed treatment (initial BA) was a very statistically significant predictor of tree response across dbh classes for both Sx and Bl.
-  *   Don't know what all the random effects outputs are. I am sure they are meaningful, but I just dont know what they mean.
-  *   Number of observations of 510: Should ensure it is clear that only trees > 7.5 cm dbh are counted.
-  *   Would be good to define the dbh class limits somewhere. In Appendix 6 table, we describe the class mid-point.
-  *   For Appendix 6, I assume that the null hypothesis for the ratio is = 1.
-
-Section 3.1.2 Mortality
-
-  1.
-I am thinking that mortality should / could be analysed as a frequency response - i.e.
+### Commentary on sections and sub-sections
+
+#### Section 3.1.1 BA increment from 1992 to 2019
+
+- Great: Very clear and straightforward. i.e. - log-transformed treatment (initial BA) was a very statistically significant predictor of tree response across dbh classes for both Sx and Bl.
+- Don't know what all the random effects outputs are. I am sure they are meaningful, but I just dont know what they mean.
+- Number of observations of 510: Should ensure it is clear that only trees > 7.5 cm dbh are counted.
+- Would be good to define the dbh class limits somewhere. In Appendix 6 table, we describe the class mid-point.
+- For Appendix 6, I assume that the null hypothesis for the ratio is = 1.
+
+#### Section 3.1.2 Mortality
+
+- I am thinking that mortality should / could be analysed as a frequency response - i.e.
 - expected vs actual distribution of mortality by size class and/or species.
@@
 H1 is that it is non-random.
-  2.
-I'd suggest that we don't compare total mortality (e.g # of stems) across treatment units, but rather frequency relative to initial number of stems (i.e.
+- I'd suggest that we don't compare total mortality (e.g # of stems) across treatment units, but rather frequency relative to initial number of stems (i.e.
 - at beginning of observation period).
@@
 Control treatments could have more mortality in overall sph, simply because they have more initial stems.
-  3.  ACTION ITEM FOR ME:  I need to check my original datasets and data sorting iterations to see how I handled mortality observations for the 1992 and 1994 plot remeasurements.
-  4.
-Note for you:  All of the mortality you've analysed and identified so far is mortality that is incremental following 92/94, so it is still solid analysis data, just with that qualification.
+- ACTION ITEM FOR ME:  I need to check my original datasets and data sorting iterations to see how I handled mortality observations for the 1992 and 1994 plot remeasurements.
+- Note for you:  All of the mortality you've analysed and identified so far is mortality that is incremental following 92/94, so it is still solid analysis data, just with that qualification.
 I will clarify the 92/94 mortalilty observations for the Methods section; we did observe dead trees at these times, but these mortality pre-dated our treatments and/or first measurements.
@@
 
-3.2.1  Estimating tree height from diameter
-
-  *   Really solid stats and rationale for a single ht:dbh model. Great !
-  *   What is really interesting and useful about this is that the ht:dbh relationship is seemingly statistically independent of treatment or, that is, initial density,  A very useful simplifying assumption proven for future modellers of these stands.
+#### 3.2.1  Estimating tree height from diameter
+
+- Really solid stats and rationale for a single ht:dbh model. Great !
+- What is really interesting and useful about this is that the ht:dbh relationship is seemingly statistically independent of treatment or, that is, initial density,  A very useful simplifying assumption proven for future modellers of these stands.
 We should highlight this in either our results or discussion.
 
-3.2.2 Plot means
-
-  *   See notes above about use of term "units" vs treatment (or similar).
-  *   Clarify effects of treatment and/or species on growth rate vs standing vol / BA at each measurement period.
-  *   A Figure 7b or equivalent could visually present changes in growth rate over time. (e.g. previous graphs we've developed)
-  *   Under Observations, # 2 - the 2nd sentence needs work to edit and clarify.
-
-3.2.3 Stand attributes by diameter class
-
-  *   We should define what we mean by the term inverse J or inverse J-shaped dbh distribution. Just for clarity.
-  *   Should we quantitatively describe these distributions? Don't have to, just asking the question.
-  *   Is the merch limit for min merch tree size 18.5 cm dbh, or 17.5 cm dbh? I remember the latter from various cruises and cruise compilations.
-  *   There are two Figure 2's, one is for vol and one is for sph.
-  *   Clarify in figures and text that sph is based on trees 7.5 cm dbh and greater. Increases in sph can be from ingress of trees growing from size classes below this limit.
-  *   A suggestion:  Have a table comparing total volume to merch volume by treatment, etc?  Would be useful for some readers, and communicate MAI and volume growth rates too.
-
-3.2.4  Species composition
-
-  *   First sentence:  Clarify what you mean by "organized similarly". Does this refer to frequency distribution ? What constitutes similarity?
+#### 3.2.2 Plot means
+
+- See notes above about use of term "units" vs treatment (or similar).
+- Clarify effects of treatment and/or species on growth rate vs standing vol / BA at each measurement period.
+- A Figure 7b or equivalent could visually present changes in growth rate over time. (e.g. previous graphs we've developed)
+- Under Observations, # 2 - the 2nd sentence needs work to edit and clarify.
+
+#### 3.2.3 Stand attributes by diameter class
+
+- We should define what we mean by the term inverse J or inverse J-shaped dbh distribution. Just for clarity.
+- Should we quantitatively describe these distributions? Don't have to, just asking the question.
+- Is the merch limit for min merch tree size 18.5 cm dbh, or 17.5 cm dbh? I remember the latter from various cruises and cruise compilations.
+- There are two Figure 2's, one is for vol and one is for sph.
+- Clarify in figures and text that sph is based on trees 7.5 cm dbh and greater. Increases in sph can be from ingress of trees growing from size classes below this limit.
+- A suggestion:  Have a table comparing total volume to merch volume by treatment, etc?  Would be useful for some readers, and communicate MAI and volume growth rates too.
+
+#### 3.2.4  Species composition
+
+- First sentence:  Clarify what you mean by "organized similarly". Does this refer to frequency distribution ? What constitutes similarity?
 
@@
 
-```{r set up tree BAI references}
-
-  figNums(name="treeBAI.plot",display=FALSE) # Estimated marginal means from Lme
-
-  tableNums("fitBA.table",display=FALSE) # Summary of mixed effects model for BAI
-  tableNums("fitBApct.table",display=FALSE) # Summary of mixed effects model for BAI
-
-  appNums("sppContrast.table",display=FALSE) # Species contrasts from treeBAI model
-  appNums("trtContrast.table",display=FALSE) # Treatment contracts from treeBAI model
-
-```
-
-The tree-level basal area increment model indicated that interactions between log-transformed tree size, species and treatment explained `r round(r.squaredGLMM(fitBA)[1]*100,1)`% (marginal R²) of the variation in log-transformed 1992-2019 tree-level BAI (`r tableNums("fitBA.table",display="cite")`).
+The tree-level basal area increment model indicated that interactions between log-transformed tree size, species and treatment explained `r round(r.squaredGLMM(fitBA)[2]*100,1)`% of the variation in log-transformed 1992-2019 tree-level BAI (`r tableNums("fitBA.table",display="cite")`).
 Estimated marginal means from the model showed that between species and among treatments, larger trees increased basal area more than smaller trees (`r figNums("treeGrowth.plot",display="cite")`).
@@
 
-figNums("standPlots",display=FALSE)
 ```
@@
 
-# create captions for figures and table
-figNums("perInc.plot",display=FALSE)
-appNums("perInc.table",display=FALSE)
-
 ```
@@
 
-figNums("standFit.X",display=FALSE)
-appNums("standFit.contrasts",display=FALSE)  
-appNums("standFitM.contrasts",display=FALSE)
-
 ```
@@
 All units increased volume, basal area and stand density over the 1992-2019 period (`r figNums("standPlots",display="cite")`).
-Mean annual basal area and volume increment peaked during the 1997-2009 period in the two harvested treatments, and peaked in the 1994-1997 period in the control treatment unit (`r figNums("perInc.plot",display="cite")`.
+Mean annual basal area and volume increment peaked during the 1997-2009 period in the two harvested treatments, and peaked in the 1994-1997 period in the control treatment unit (`r figNums("perInc.plot",display="cite")`).
 The low RBA unit accrued basal area significantly faster than the other two units over the 1992-2019 period (`r appNums("standFit.contrasts",display="cite")`).
@@
 
-```{r caption for species composition plots}
-
-figNums("sppComp.plot",display=FALSE)
-figNums("sppCompDiam.plot",display=FALSE)
-
-```
-
-A comparison of proportional species composition by diameter class, treatment and year showed that fir and spruce proportion by diameter class varied similarly between the high RBA and control units (`r figNums("sppCompDiam.plot",display="cite")`).
+A comparison of proportional species composition by diameter class, treatment and year showed that fir and spruce volume proportion by diameter class varied similarly between the high RBA and control units (`r figNums("sppCompDiam.plot",display="cite")`).
 In 1992, the largest diameter classes were dominated by spruce, whereas fir was the dominant species in the smallest diameter classes.
-This pattern was found for volume, basal area, and tree density.
-We only show volume in this document.
+This pattern was also found for basal area and tree density (not shown).
 By 2019, the fir-spruce composition by distribution class was maintained in these two units, with small increases in spruce composition in the smallest diameter class.
@@
 
-figNums("mort.plot",display=FALSE)
-figNums("mortDiam.plot",display=FALSE)
-figNums("mortSppDiam.plot",display=FALSE)
-
 ```
@@
   ylab("Basal area increment from 1992-2019 (m²)")+
-  facet_wrap(~BA.Target)
+  facet_wrap(~BA.Target)+
+  theme(legend.position="bottom")
 
@@
 
-# Create function
-standPlot<-function(standVar,y.label) {
-
 standData.X %>% 
+  setNames(c("BA.Target","Year","Vol_mean","Vol_se",
+             "BA_mean","BA_se",
+             "SPH_mean","SPH_se",
+             "QMD_mean","QMD_se")) %>%
+
+  # Need to change year to numeric for plotting
   mutate(Year=as.numeric(as.character(Year))) %>%
-    dplyr::select(BA.Target,Year,standVar,paste0(standVar,".se")) %>% 
-    setNames(c("BA.Target","Year","Mean","SE")) %>%
-
-    ggplot(aes(x=Year,y=Mean,color=BA.Target,group=BA.Target))+
+
+  # reshape for ggplot
+  pivot_longer(cols=Vol_mean:last_col(),names_to=c("Var","Metric"),names_sep="_",values_to="Values") %>%   pivot_wider(names_from="Metric",values_from="Values") %>%
+
+  # Filter to remove quadratic mean diameter
+  filter(Var!="QMD") %>%
+
+  # Reset factor levels for plotting
+  mutate(Var=fct_relevel(Var,"Vol","BA","SPH")) %>% 
+  mutate(Var=fct_recode(Var,Volume="Vol",
+                        `Basal area`="BA",
+                        `Stand density (stems per hectare)`="SPH")) %>%
+
+  ggplot()+
+  aes(x=Year,y=mean,fill=BA.Target,color=BA.Target,group=BA.Target)+
   geom_point()+
   geom_line()+
-    geom_errorbar(aes(ymin=Mean-SE, ymax=Mean+SE, width=.2),
-                   position=position_dodge(.9))+
-    facet_wrap(~BA.Target)+
+  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, width=.2))+
+                   # position=position_dodge(.9))+
    scale_x_continuous(limits=c(1992,2019),breaks=c(1992,1994,1997,2009,2019),
@@
                      name="Measurement year")+ # stretch out x-axis
-  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5),
-        legend.position = "none")+
-    ylab(y.label)
-
-}
-
-standVol.plot<-
-  standPlot("Vol",y.label="Vol. (m³/ha)")+
-  theme(axis.text.x=element_blank(),axis.title.x=element_blank())
-
-standBA.plot<-
-  standPlot("BA",y.label="BA (m²/ha)")+
-  theme(axis.text.x=element_blank(),axis.title.x=element_blank())
-
-standK.plot<-
-  standPlot("SPH",y.label="Tree dens. (sph)")
-
-standQMD.plot<-standPlot("QMD",y.label="QMD (cm)")
-
-standPlots<-cowplot::plot_grid(standVol.plot,standBA.plot,standK.plot,ncol=1,
-                      align="v")
-
-# display plot
-standPlots
-```
-
-`r figNums("standPlots",caption="Mean plot attributes by treatment unit and year.  Whiskers are one standard error of the mean.  Vol., volume; BA, basal area; Tree dens., tree density")`
+  facet_wrap(~Var,scales="free_y")+
+  xlab("Plot mean")+
+  theme(legend.position = "bottom")
+
+# Old code to produce separate grapohs below
+
+# Create function
+# standPlot<-function(standVar,y.label) {
+# 
+#     standData.X %>% 
+#     mutate(Year=as.numeric(as.character(Year))) %>% 
+#     dplyr::select(BA.Target,Year,standVar,paste0(standVar,".se")) %>% 
+#     setNames(c("BA.Target","Year","Mean","SE")) %>% 
+#   
+#     ggplot(aes(x=Year,y=Mean,color=BA.Target,group=BA.Target))+
+#     geom_point()+
+#     geom_line()+
+#     geom_errorbar(aes(ymin=Mean-SE, ymax=Mean+SE, width=.2),
+#                    position=position_dodge(.9))+
+#     # facet_wrap(~BA.Target)+
+#       scale_x_continuous(limits=c(1992,2019),breaks=c(1992,1994,1997,2009,2019),
+#                      minor_breaks = c(1992,1994,1997,2009,2019),
+#                      name="Measurement year")+ # stretch out x-axis
+#   theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5),
+#         legend.position = "none")+
+#     ylab(y.label)
+# 
+# }
+#     
+# standVol.plot<-
+#   standPlot("Vol",y.label="Vol. (m³/ha)")+
+#   theme(axis.text.x=element_blank(),axis.title.x=element_blank())
+#   
+# standBA.plot<-
+#   standPlot("BA",y.label="BA (m²/ha)")+
+#   theme(axis.text.x=element_blank(),axis.title.x=element_blank())
+#   
+# standK.plot<-
+#   standPlot("SPH",y.label="Tree dens. (sph)")
+#   
+# standQMD.plot<-standPlot("QMD",y.label="QMD (cm)") 
+# 
+# standPlots<-cowplot::plot_grid(standVol.plot,standBA.plot,standK.plot,ncol=1,
+#                       align="v")
+# 
+# # display plot
+# standPlots
+```
+
+`r figNums("standPlots",caption="Mean plot live-tree attributes by treatment unit and year.  Whiskers are one standard error of the mean.  Vol., volume; BA, basal area; Tree dens., tree density")`
 

### 2021-01-08 manuscript_draft.Rmd
manny work

@@
 
-## Tree-level growth responses
+## Data analyses
+
+### Tree-level growth responses
 
@@
 
-## Stand-level growth and structure
-
-### Stand-level volume and basal area estimates
+### Stand-level growth and structure
+
+#### Stand-level volume and basal area estimates
 Total and merchantable volumes were estimated for trees using species- and region-specific equations [@nigh2016] based on tree height and DBH.
@@
 
-### Mortality
+#### Mortality
 
@@
 
-The tree-level basal area increment model indicated that interactions between log-transformed tree size, species and treatment explained `r round(r.squaredGLMM(fitBA)[2]*100,1)`% of the variation in log-transformed 1992-2019 tree-level BAI (`r tableNums("fitBA.table",display="cite")`).
-Estimated marginal means from the model showed that between species and among treatments, larger trees increased basal area more than smaller trees (`r figNums("treeGrowth.plot",display="cite")`).
+The tree-level basal area increment model indicated that interactions between log-transformed tree size, species and treatment explained `r round(r.squaredGLMM(fitBA)[2]*100,1)`% of the variation in log-transformed 1992-2019 tree-level BAI (`r tableNums("fitBA.table",display="cite")`), and `r round(r.squaredGLMM(fitBApct)[2]*100,1)`% of the variation in basal area increment as a percentage of tree size in 1992 (`r tableNums("fitBApct.table",display="cite")`).
+Estimated marginal means from the model showed that between species and among treatments, basal area increment increased with tree size, meaning that larger trees increased basal area more than smaller trees (`r figNums("treeGrowth.plot",display="cite")`), although smaller trees increased their size proportionally more than large trees (`r figNums("treeGrowthPct.plot",display="cite")`).
 Species contrasts showed that spruce trees increased  basal area more than than fir in the high-removal treatment unit, whereas basal growth between the two species was close to equal in the low-removal and control units (`r appNums("sppContrast.table",display="cite")`).
@@
   mutate(Height.Pred=coalesce(Height,exp(fixef(htDiamFit)[1]+fixef(htDiamFit)[2]*log(DBH))))
-
-appNums("htModelFit.plot",display="FALSE") # model output
-
-# Tree heights model
-  appNums("heightDiam.plot",display=FALSE) # Plot of diameter/heights`
 
@@
 
-`r figNums("treeBAI.plot",caption="Estimated marginal means of spruce and fir basal area increment (m²) across five inital DBH classes and three treatments.
+`r figNums("treeGrowth.plot",caption="Estimated marginal means of spruce and fir basal area increment (m²) across five inital DBH classes and three treatments.
 Whiskers are standard error of the mean.
@@
 
-```{r function to create figure of plot means, fig.height=10}
+```{r tree BAI as percentage by treatment species and diameter class}
+
+# Create plot of periodic increment
+
+# Using estimated marginal means
+  emmeans(rg.pbaipct,~Init|BA.Target+Species,type="response") %>% 
+  as.data.frame() %>%
+  mutate(Init=ba2d(Init)) %>% # convert initial basal area back to diameter
+
+  # plotting
+  ggplot()+
+  aes(x=Init,y=response,fill=Species,color=Species)+
+  geom_point()+
+  geom_line()+
+  geom_errorbar(aes(ymin=response-SE, ymax=response+SE), width=.2,
+                 position=position_dodge(0.05))+
+  xlab("1992 DBH (cm)")+
+  ylab("Basal area increment from 1992-2019 (% increase)")+
+  facet_wrap(~BA.Target)+
+  theme(legend.position="bottom")
+
+```
+
+`r figNums("treeGrowthPct.plot",caption="Estimated marginal means of spruce and fir basal area increment (expressed as a percentage increase relative to 1992 DBH) across five inital DBH classes and three treatments.
+Whiskers are standard error of the mean.
+Responses are back-transformed from the model.")`
+
+```{r plot means plots, fig.height=10}
 
@@
 
+```{r plot means by species}
+
+plotSums<-
+  plotTotal.X %>% 
+  group_by(BA.Target,Plot,Year,Var) %>% 
+  summarise(Value=sum(Value)) %>% 
+  mutate(Species="Total") %>% 
+  dplyr::select(BA.Target,Plot,Year,Var,Species,Value)
+
+plotTotal.X %>%
+
+  # bring in plotSums
+  dplyr::select(BA.Target,Plot,Year,Var,Species,Value) %>% 
+  bind_rows(plotSums) %>%
+
+  # need to convert year to numeric for plotting
+  mutate(Year=as.numeric(as.character(Year))) %>%
+
+  group_by(BA.Target,Year,Species,Var) %>% 
+  summarise(Mean=mean(Value),
+            SE=sd(Value)/sqrt(length(Value))) %>%
+
+  ggplot()+
+  aes(x=Year,y=Mean,group=Species,fill=Species,color=Species)+
+  geom_point()+
+  geom_line()+
+  geom_errorbar(aes(ymin=Mean-SE, ymax=Mean+SE, width=.2),
+                   position=position_dodge(0))+
+ scale_x_continuous(limits=c(1992,2019),breaks=c(1992,1994,1997,2009,2019),
+                     minor_breaks = c(1992,1994,1997,2009,2019),
+                     name="Measurement year")+ # stretch out x-axis
+
+  facet_grid(Var~BA.Target,
+             scales="free_y")+
+  theme(legend.position = "bottom",
+  axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5))
+
+```
+
 ```{r plot species composition by attribute}

### 2021-01-09 manuscript_draft.Rmd
commit

@@
 
+```{r plot comparisons at the species level}
+
+# Summarize plot data at the species and treatment unit level
+speciesData.X<-
+  X %>%
+
+  filter(Status=="Live") %>% 
+  filter(Species %in% c("Spruce","Fir")) %>% 
+    # Calculate volume
+  left_join(filter(volCoef,VolType=="total"),by="Species") %>% 
+  mutate(vol=exp(1)^b0*DBH^b1*Height.Pred^b2) %>% 
+  mutate(BA=d2ba(DBH)) %>% 
+  dplyr::select(BA.Target,Plot:Species,Year,DBH,BA,vol,DBH.class,Height.Pred) %>% 
+  mutate(Species=factor(Species)) %>%
+
+  # Remove volumes below 18.5cm DBH
+  mutate(vol=replace(vol,DBH<18.5,0)) %>%
+
+  # now create plot sums grid so that you can calculate means
+  group_by(BA.Target,Species,Plot,Year) %>% 
+    summarise(Vol=sum(vol,na.rm=T)*20,
+              BA=sum(BA)*20,
+              SPH=n()*20) %>%
+
+  ungroup() %>%
+
+  # Create QMD
+  mutate(QMD.mean=sqrt(BA/(SPH*0.0000785)))
+
+# fit linear model
+# fit linear models
+fit.vol.spp<-lmer(Vol~BA.Target*Year*Species+(1|Plot),data=speciesData.X)
+fit.ba.spp<-lmer(BA~BA.Target*Year*Species+(1|Plot),data=speciesData.X)
+fit.sph.spp<-lmer(SPH~BA.Target*Year*Species+(1|Plot),data=speciesData.X)
+
+# Summarize models
+library(lmerTest)
+
+# Create estimated marginal means
+vol.spp.means <- emmeans(fit.vol.spp, ~ BA.Target*Year*Species)
+ba.spp.means <- emmeans(fit.ba.spp, ~ BA.Target*Year*Species)
+sph.spp.means <- emmeans(fit.sph.spp, ~ BA.Target*Year*Species)
+
+# Mean contrasts
+volM.spp.contrast <- pairs(vol.spp.means,by="Year")
+baM.spp.contrast <- pairs(ba.spp.means,by="Year")
+sphM.spp.contrast <- pairs(sph.spp.means,by="Year")
+
+# Create species mean contrasts table
+speciesFitM.contrasts<-
+
+```
+
 All units increased volume, basal area and stand density over the 1992-2019 period (`r figNums("standPlots",display="cite")`).
@@
   xlab("1992 DBH (cm)")+
-  ylab("Basal area increment from 1992-2019 (m²)")+
+  ylab("Tree BAI from 1992-2019 (m²)")+
   facet_wrap(~BA.Target)+
@@
 Whiskers are standard error of the mean.
-Responses are back-transformed from the model.")`
+Responses are back-transformed from the model.
+BAI, basal area increment")`
 
@@
   xlab("1992 DBH (cm)")+
-  ylab("Basal area increment from 1992-2019 (% increase)")+
+  ylab("Tree BAI from 1992-2019 (% increase)")+
   facet_wrap(~BA.Target)+
@@
 Whiskers are standard error of the mean.
-Responses are back-transformed from the model.")`
-
-```{r plot means plots, fig.height=10}
+Responses are back-transformed from the model.
+BAI, basal area increment")`
+
+```{r plot means plots, fig.height=10,eval=FALSE}
 
@@
 
-`r figNums("standPlots",caption="Mean plot live-tree attributes by treatment unit and year.  Whiskers are one standard error of the mean.  Vol., volume; BA, basal area; Tree dens., tree density")`
-
-\newpage
-
-```{r plot periodic increment}
-
-# Note this function performs the operation for basal area and volume only
-
-  perInc.X %>%
-  dplyr::select(BA.Target:BA.se) %>% 
-    setNames(c("BA.Target","Period","Vol_mean","Vol_SE","BA_mean","BA_SE")) %>% 
-    pivot_longer(cols=-c(BA.Target,Period),
-                 names_to=c("Var","SE"),
-                 names_sep="_",
-                 values_to="Value") %>% 
-    pivot_wider(names_from="SE",values_from="Value") %>%
-
-  ggplot(aes(x=Period,y=mean,group=BA.Target,color=BA.Target))+
-  geom_point()+
-  geom_line()+
-
-     geom_errorbar(aes(ymin=mean-SE, ymax=mean+SE, width=.2),
-                   position=position_dodge(0))+
-    facet_wrap(~Var,nrow=2,
-               scales="free_y", strip.position = "left",
-               labeller=as_labeller(c(BA="Basal area inc. (m²/ha/year)",Vol="Volume inc. (m³/ha/year)")))+
-    ylab(NULL)+
-     theme(strip.background = element_blank(),
-           strip.placement = "outside")
-
-```
-
-`r figNums("perInc.plot",caption="Mean plot annual increment over four measurement periods for basal area and volume, by treatment.  Whiskers are one standard error of the mean. Inc, increment.")`
-
-\newpage
-
-```{r plot means by species}
+```{r plot means by species, fig.height=7}
 
@@
             SE=sd(Value)/sqrt(length(Value))) %>%
+
+  # reorder and relabel some factors
+   mutate(Var=factor(Var)) %>% 
+  mutate(Var=fct_relevel(Var,"Vol","BA","sph")) %>% 
+  mutate(Var=dplyr::recode(Var,BA="BA (m²/ha)",sph="SPH",
+                           Vol="Vol. (m³/ha)")) %>%
 
@@
                      name="Measurement year")+ # stretch out x-axis
-
+  ylab("")+
+  xlab("Measurement year")+
   facet_grid(Var~BA.Target,
-             scales="free_y")+
+             scales="free_y",
+              switch="y")+
   theme(legend.position = "bottom",
-  axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5))
-
-```
+  axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5),
+  strip.placement = "outside",
+            strip.background.y = element_blank())
+
+```
+
+`r figNums("standPlots",caption="Mean plot live-tree attributes by treatment unit and year.  Whiskers are one standard error of the mean.  Vol., volume; BA, basal area; Tree dens., tree density")`
+
+\newpage
+
+```{r plot periodic increment}
+
+# Note this function performs the operation for basal area and volume only
+
+  perInc.X %>%
+  dplyr::select(BA.Target:BA.se) %>% 
+    setNames(c("BA.Target","Period","Vol_mean","Vol_SE","BA_mean","BA_SE")) %>% 
+    pivot_longer(cols=-c(BA.Target,Period),
+                 names_to=c("Var","SE"),
+                 names_sep="_",
+                 values_to="Value") %>% 
+    pivot_wider(names_from="SE",values_from="Value") %>%
+
+  ggplot(aes(x=Period,y=mean,group=BA.Target,color=BA.Target))+
+  geom_point()+
+  geom_line()+
+
+     geom_errorbar(aes(ymin=mean-SE, ymax=mean+SE, width=.2),
+                   position=position_dodge(0))+
+    facet_wrap(~Var,nrow=2,
+               scales="free_y", strip.position = "left",
+               labeller=as_labeller(c(BA="Basal area inc. (m²/ha/year)",Vol="Volume inc. (m³/ha/year)")))+
+    ylab(NULL)+
+     theme(strip.background = element_blank(),
+           strip.placement = "outside")
+
+# caption
+figNums("perInc.plot",caption="Mean plot annual increment over four measurement periods for basal area and volume, by treatment.  Whiskers are one standard error of the mean. Inc, increment.")
+
+```
+
+r figNums("perInc.plot",caption="Mean plot annual increment over four measurement periods for basal area and volume, by treatment.  Whiskers are one standard error of the mean. Inc, increment.")`
+
+\newpage
 

### 2021-01-11 manuscript_draft.Rmd
manny work

@@
 
-```{r plot comparisons at the species level}
+```{r plot comparisons at the species level, eval=FALSE}
 
@@
 
-All units increased volume, basal area and stand density over the 1992-2019 period (`r figNums("standPlots",display="cite")`).
+### Rate of change from 1992 to 2019
+All units increased volume, basal area and stand density over the 1992-2019 period (`r figNums("standPlots",display="cite")` and `r appNums("standPlot.table",display="cite")`).
 Mean annual basal area and volume increment peaked during the 1997-2009 period in the two harvested treatments, and peaked in the 1994-1997 period in the control treatment unit (`r figNums("perInc.plot",display="cite")`).
-The low RBA unit accrued basal area significantly faster than the other two units over the 1992-2019 period (`r appNums("standFit.contrasts",display="cite")`).
-The estimated rate of basal area increment in the low RBA unit from 1992 to 2019 was higher by
+From 1992 to 2019, estimated marginal means of basal area increment in the low RBA unit exceeded marginal means in the high RBA and control unit by an estimated
 `r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Basal area") %>%   pull(estimate) %>% round(.,1) %>% .[1]`
@@
 `r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Basal area") %>%   pull(estimate) %>% round(.,1) %>% .[2]`
-m² per hectare more than the high RBA and control units, respectively (p<0.01 for both).
+m² per hectare, respectively (p<0.01 for both, `r appNums("standFit.contrasts",display="cite")`).
 
@@
 stems per hectare more than the control unit (p<0.001).
-The rate of change in tree density did not differ significantly between the two harvested units.
-The rate of change in volume in trees over 18.5cm DBH was different between barely significant between the two harvested units, and there were no significant differences between the control and harvested units.
+The estimated rate of change in tree density did not differ significantly between the two harvested units.
+The rate of change in volume in trees over 18.5cm DBH was different between barely significant (p=0.049) between the two harvested units, and was not significantly different between the control and harvested units.
+
+### Mean stand attributes by year
 
@@
 
-`r appNums("sppContrast.table",caption="Contrasts between predicted fir and spruce growth across treatment units and initial diameter classes.
+`r appNums("sppContrast.table",caption="Contrasts between estimated marginal means of fir and spruce growth across treatment units and initial diameter classes.
 Ratio between predicted fir and growth is in 'ratio' column.")`
@@
 
-  mutate_if(is.double,round,5) %>%  # round to five digits
+  mutate_if(is.double,round,4) %>%  # round to four digits
+  mutate_at(vars("df","t.ratio","ratio","SE"),round,2) %>%
 
@@
 
-  mutate_if(is.double,round,5) %>%
+ mutate_if(is.double,round,4) %>%  # round to four digits
+  mutate_at(vars("df","t.ratio","ratio","SE"),round,2) %>%
 
@@
 
+`r appNums("standPlot.table",caption="Plot means of volume, basal area and stand density, by species, year and treatment.")`
+
+```{r table of plot means}
+
+standData.X %>% 
+  ungroup() %>% 
+  mutate_at(vars(Vol:last_col()),round,1) %>% 
+  mutate_at(vars(contains("SPH")),round,0) %>%
+
+  # change to character to allow pasting
+  mutate_all(as.character) %>% 
+  mutate(Vol=paste0(Vol," (",Vol.se,")")) %>% 
+  mutate(BA=paste0(BA," (",BA.se,")")) %>% 
+  mutate(SPH=paste0(SPH," (",SPH.se,")")) %>% 
+  mutate(QMD=paste0(QMD," (",QMD.se,")")) %>%
+
+  # rename volume column
+  rename(Volume="Vol") %>%
+
+  # remove SE columns
+  dplyr::select(-contains(".se")) %>%
+
+  # rearrange
+  arrange(Year,BA.Target) %>%
+
+  # format table
+  flextable() %>%
+  align(align="left",part="all") %>% 
+  merge_v("Year") %>% 
+  valign(valign="top",part="body") %>% 
+  hline(i=seq(from=3,to=15,by=3),border=fp_border(color="black")) %>% 
+  fix_border_issues() %>% 
+   autofit()
+
+```
+
+*Volume is m3/hectare, BA = basal area (m2/ha), SPH = stems per hectare, QMD = quadratic mean diameter (cm) at 1.3m height.  Values in brackets are the standard error of the mean.*
+
+\newpage
+
+```{r table of periodic mean annual increment}
+
+perInc.X %>% 
+    ungroup() %>% 
+  mutate_at(vars(Vol:last_col()),round,1) %>% 
+  mutate_at(vars(contains("SPH")),round,0) %>%
+
+  # change to character to allow pasting
+  mutate_all(as.character) %>% 
+  mutate(Vol=paste0(Vol," (",Vol.se,")")) %>% 
+  mutate(BA=paste0(BA," (",BA.se,")")) %>% 
+  mutate(SPH=paste0(SPH," (",SPH.se,")")) %>% 
+  mutate(QMD=paste0(QMD," (",QMD.se,")")) %>%
+
+  # rename volume column
+  rename(Volume="Vol") %>%
+
+  # remove SE columns
+  dplyr::select(-contains(".se")) %>%
+
+  # rearrange
+  arrange(Year,BA.Target) %>%
+
+```
+
 `r appNums("standFitM.contrasts",caption="Contrasts of stand attributes (volume, basal area and tree density) between treatments.

### 2021-01-18 manuscript_draft.Rmd
fxed plot multiplier

@@
 
-knitr::opts_chunk$set(echo = FALSE,message=FALSE,warning=FALSE)
+knitr::opts_chunk$set(echo = FALSE,message=FALSE,warning=FALSE,ft.align="left")
 
@@
 ba2d<-function(ba) {return(sqrt(ba/pi)*200)}
+
+# plot multiplier
+pm=25
 
@@
 
-\newpage
-
-# Introduction
-
-# Methods
-
-This section will describe the methods we used for data analyses, not field data collection.
-
-From our technical report submitted to FCI earlier:
-
-Our data analyses and summaries examined the effects of the different levels of stand-level basal area density treatments on three main attributes of stand development following treatment.
-These include, in order:
-
-1. Post-treatment stand basal area dynamics, including the rate and pattern of basal area re-growth and recovery over time following different levels of initial basal area density reductions;
-2. Stand structural outcomes, including treatment influences on the overall abundance of trees (in sph) by diameter class, within the different treatment types, and;
-3.
-Tree species composition outcomes, including treatment influences on the relative abundance and size class distributions of subalpine fir and hybrid white spruce within the different treatment types.
-
-Analyses are grouped into three main categories (below).  These are treated as separate categories in the Results section.
-
-1. Tree-level responses to treatment;
-2. Stand-level responses to treatment:
-
-## Data analyses
-
-### Tree-level growth responses
-
-To test the hypothesis that tree-level radial growth response over the 1992-2019 period varied with interactions between (i) tree diameter at the time of treatment (1992); (ii) treatment; and (iii) species, we fit linear mixed effects models with log-transformed 1992-2019 diameter increment as the response variable, and treatment, species and log-transformed 1992 diameter as interacting fixed effects, and plot as a random intercept effect.
-We also fit a similar model with the 1992-2019 diameter increment expressed as a percentage of 1992 diameter.
-Plots of residuals and fit lines with orginal data were examined to evaluate model goodness of fit.
-Variance explained for each model are reported using the approach described in Nakagawa et al.
-[-@Nakagawa2013], as implemented in the MuMIn package for R [@barton2009mumin].
-
-### Stand-level growth and structure
-
-#### Stand-level volume and basal area estimates
-Total and merchantable volumes were estimated for trees using species- and region-specific equations [@nigh2016] based on tree height and DBH.
-Consistent with forest management practices in British Columbia, we estimated volumes for trees meeting a minimum diameter utilization limit of 18.5cm DBH.
-For trees lacking a tree height measurement, we estimated height by fitting a single diameter-height model developed with a linear mixed effects model, with log-transformed DBH as a fixed effect, and tree nested within plot as a random effect.
-Trees with broken tops were omitted from diameter-height models.
-We summed tree-level estimates of volume and basal area to generate stand-level estimates for each plot.
-To test the hypothesis that changes in stand attributes over 27 years varied by treatment, we fit separate linear models with stand total volume, basal area, and density as response variables, and the interaction between treatment and year as the independent variable.
-
 #### Mortality
-
-From Mike's e-mail:
 
@@
 Once crown vigor etc starts to decline with age, pests or pathogens, then the relationship could start to break down.
+
+\newpage
+
+# Introduction
+
+# Methods
+
+This section will describe the methods we used for data analyses, not field data collection.
+
+From our technical report submitted to FCI earlier:
+
+Our data analyses and summaries examined the effects of the different levels of stand-level basal area density treatments on three main attributes of stand development following treatment.
+These include, in order:
+
+1. Post-treatment stand basal area dynamics, including the rate and pattern of basal area re-growth and recovery over time following different levels of initial basal area density reductions;
+2. Stand structural outcomes, including treatment influences on the overall abundance of trees (in sph) by diameter class, within the different treatment types, and;
+3.
+Tree species composition outcomes, including treatment influences on the relative abundance and size class distributions of subalpine fir and hybrid white spruce within the different treatment types.
+
+Analyses are grouped into three main categories (below).  These are treated as separate categories in the Results section.
+
+1. Tree-level responses to treatment;
+2. Stand-level responses to treatment:
+
+## Data analyses
+
+### Tree-level growth responses
+
+To test the hypothesis that tree-level radial growth response over the 1992-2019 period varied with interactions between (i) tree diameter at the time of treatment (1992); (ii) treatment; and (iii) species, we fit linear mixed effects models with log-transformed 1992-2019 diameter increment as the response variable, and treatment, species and log-transformed 1992 diameter as interacting fixed effects.
+To account for plot-level productivity differences, we included plot as a random intercept effect.
+Models were fit using the 'lme4' package for R [@].
+We also fit a similar model with the 1992-2019 diameter increment expressed as a percentage of 1992 diameter.
+Plots of residuals and fit lines with orginal data were examined to evaluate model goodness of fit.
+Variance explained for each model are reported using the approach described in Nakagawa et al.
+[-@Nakagawa2013], as implemented in the MuMIn package for R [@barton2009mumin].
+Pairwise contrasts were performed using the 'emmeans' package for R [@] and Tukey HSD method of p value adjustment was used to control for multiplicity.
+We report all significant results at p<0.05.
+
+### Stand-level growth and structure
+
+#### Stand-level volume and basal area estimates
+Total and merchantable volumes were estimated for trees using species- and region-specific equations [@nigh2016] based on tree height and DBH.
+Consistent with forest management practices in British Columbia, we estimated volumes for trees meeting a minimum diameter utilization limit of 18.5cm DBH.
+For trees lacking a tree height measurement, we estimated height by fitting a single diameter-height model developed with a linear mixed effects model, with log-transformed DBH as a fixed effect, and tree nested within plot as random intercepts.
+Trees with broken tops were omitted from diameter-height models.
+We summed tree-level estimates of volume and basal area to generate stand-level estimates for each plot.
+To test the hypothesis that changes in stand attributes over 27 years varied by treatment, we fit separate linear models with stand total volume, basal area, and density as response variables, and the interaction between treatment and year as the independent variable.
+
+#### Mortality
 
@@
 
-The tree-level basal area increment model indicated that interactions between log-transformed tree size, species and treatment explained `r round(r.squaredGLMM(fitBA)[2]*100,1)`% of the variation in log-transformed 1992-2019 tree-level BAI (`r tableNums("fitBA.table",display="cite")`), and `r round(r.squaredGLMM(fitBApct)[2]*100,1)`% of the variation in basal area increment as a percentage of tree size in 1992 (`r tableNums("fitBApct.table",display="cite")`).
+The tree-level basal area increment model indicated that interactions between log-transformed tree size, species and treatment explained `r round(r.squaredGLMM(fitBA)[2]*100,1)`% of the variation in log-transformed 1992-2019 tree-level BAI, and `r round(r.squaredGLMM(fitBApct)[2]*100,1)`% of the variation in basal area increment as a percentage of tree size in 1992 (p<0.0001 for both, `r tableNums("fitBA.table",display="cite")`).
 Estimated marginal means from the model showed that between species and among treatments, basal area increment increased with tree size, meaning that larger trees increased basal area more than smaller trees (`r figNums("treeGrowth.plot",display="cite")`), although smaller trees increased their size proportionally more than large trees (`r figNums("treeGrowthPct.plot",display="cite")`).
-Species contrasts showed that spruce trees increased  basal area more than than fir in the high-removal treatment unit, whereas basal growth between the two species was close to equal in the low-removal and control units (`r appNums("sppContrast.table",display="cite")`).
-Within each species, treatment contrasts showed that spruce basal area increment in the high-removal unit exceeded spruce growth in both other units (`r appNums("trtContrast.table",display="cite")`), whereas fir basal area increment in the high- and low-removal units was higher than the control unit, and there were no significant differences between the two harvested units (`r appNums("trtContrast.table",display="cite")`).
+Species contrasts showed that spruce trees increased  basal area significantly more than than fir in the high-removal and control treatment units, whereas basal growth between the two species was close to equal in the low-removal unit (`r appNums("sppContrast.table",display="cite")`).
+Within-species contrasts showed that spruce basal area increment in the high-removal unit exceeded spruce growth in both other units (`r appNums("trtContrast.table",display="cite")`), whereas fir basal area increment in the high- and low-removal units was higher than the control unit, and there were no significant differences between the two harvested units (`r appNums("trtContrast.table",display="cite")`).
 
@@
 Tree heights had a positive nonlinear relationship with diameter, for both species and across all time periods (`r appNums("heightDiam.plot",display="cite")`).
-We modeled tree height using a linear mixed effects model with log-transformed diameter as the fixed effect, and tree nested within plot as a random effect.
 We originally developed separate models for each species and time period, however, a comparison of height-diameter models showed that intercepts and slopes did not vary significantly (i.e., p>0.05) between species or time periods (not shown).
@@
   group_by(BA.Target,Plot,Species,Year,DBH.class) %>% 
-  summarise(BA=sum(BA)*20,
-            Vol=sum(vol)*20,
-            SPH=n()*20) %>% 
+  summarise(BA=sum(BA)*pm,
+            Vol=sum(vol)*pm,
+            SPH=n()*pm) %>% 
   ungroup()
@@
   group_by(BA.Target,Plot,Year) %>% 
-    summarise(Vol.sum=sum(vol,na.rm=T)*20,
-              BA.sum=sum(BA)*20,
-              SPH.sum=n()*20) %>%
+    summarise(Vol.sum=sum(vol,na.rm=T)*pm,
+              BA.sum=sum(BA)*pm,
+              SPH.sum=n()*pm) %>%
 
@@
   group_by(BA.Target,Species,Plot,Year) %>% 
-    summarise(Vol=sum(vol,na.rm=T)*20,
-              BA=sum(BA)*20,
-              SPH=n()*20) %>%
+    summarise(Vol=sum(vol,na.rm=T)*pm,
+              BA=sum(BA)*pm,
+              SPH=n()*pm) %>%
 
@@
 ### Rate of change from 1992 to 2019
-All units increased volume, basal area and stand density over the 1992-2019 period (`r figNums("standPlots",display="cite")` and `r appNums("standPlot.table",display="cite")`).
+All treatment units increased volume, basal area and stand density over the 1992-2019 period (`r figNums("standPlots",display="cite")` and `r appNums("standPlot.table",display="cite")`).
 Mean annual basal area and volume increment peaked during the 1997-2009 period in the two harvested treatments, and peaked in the 1994-1997 period in the control treatment unit (`r figNums("perInc.plot",display="cite")`).
@@
 The estimated rate of change in tree density did not differ significantly between the two harvested units.
-The rate of change in volume in trees over 18.5cm DBH was different between barely significant (p=0.049) between the two harvested units, and was not significantly different between the control and harvested units.
-
-### Mean stand attributes by year
+The rate of change in volume in trees over 18.5cm DBH was barely significant (p=0.049) between the two harvested units, and the difference was not significant between the control and harvested units.
+
+### Stand attributes by year
+By year, mean plot volume, basal area and stand density did not differ significantly between both harvested units (i.e., p>0.05, `r appNums("standFitM.contrasts",display="cite")` and `r figNums("standPlots",display="cite")`).
+Contrasts between the high-removal and control treatment units showed that volume, basal area and stand density differences were highest and significant (i.e., p<0.05) in the 1992 measurement year, and these differences diminshed over each subsequent measurement year.
+In the 2019 measurement, volume and stand density were not significantly different between these two treatment units (i.e., p>0.05 for both).
+Contrasts between the low-removal and the control treatment units showed a similar trend with basal area and stand density; the control unit had significanly higher basal area and stand density means in 1992, but the difference between the units diminished over each subsequent measurement, becoming non-significant (i.e., p>0.05) in 2019.
+Plot means of volume did not differ significantly between the treatment units in any measurement year.
 
@@
   group_by(BA.Target,Plot,Species,Year) %>% 
-  summarise(sph=n()*20,
-            BA=sum(BA)*20,
-            Vol=sum(vol)*20) %>% 
+  summarise(sph=n()*pm,
+            BA=sum(BA)*pm,
+            Vol=sum(vol)*pm) %>% 
   ungroup() %>%
@@
 A total of `r deadList %>% nrow()` trees (`r deadList %>% filter(Species=="Fir") %>% nrow()` fir, `r deadList %>% filter(Species=="Spruce") %>% nrow()` spruce, and `r deadList %>% filter(Species=="Fd") %>% nrow()` Douglas-fir) died in `r length(unique(deadList$Plot))` plots from 1994 to 2019.
-The number of trees that died per plot from 1994 to 2019 increased significantly with plot basal area in 1994 (`r figNums("mort.plot",display="cite")`, R^2 = `r glance(deadPlot.ANOVA) %>% .$r.squared %>% round(.,3)*100`, p = `r glance(deadPlot.ANOVA) %>% pull(p.value)`).
+The number of trees that died per plot from 1994 to 2019 increased significantly with plot basal area in 1994 (`r figNums("mort.plot",display="cite")`, p < 0.0001).
 Similar linear models fitted separately for fir and spruce showed that fir mortality had a similar relationship to plot basal area, whereas spruce mortality did not (not shown).
@@
 Because spruce tends to live longer than subalpine fir, higher growth rates at later ages may allow this species to become larger than fir.
-In contrast, thinning studies in spruce-fir forests in eastern North America report that balsam fir (Abies balsamifera) responds to release with faster height and diameter growth than co-occurring red spruce (Picea rubens) (Dumais and Prevost, 2007; Dumais et al.
+In contrast, thinning studies in spruce-fir forests in eastern North America report that balsam fir (*Abies balsamifera*) responds to release with faster height and diameter growth than co-occurring red spruce (Picea rubens) (Dumais and Prevost, 2007; Dumais et al.
 2014, Messier et al.
@@
 
-`r tableNums("fitBA.table",caption="Summary of mixed effects model of tree-level basal area increment")`
+`r tableNums("fitBA.table",caption="Summary of two mixed effects models of tree-level basal area increment")`
 
@@
 
-```
-
-\newpage
-
-`r tableNums("fitBApct.table",caption="Summary of mixed effects model of tree-level basal area increment as percentage of initial tree size")`
-
-```{r summarize tree level BAI as percentage model}
-
-```
+fitBA.random<-
+  fitBA %>% 
+  VarCorr() %>% 
+  as_tibble() %>% 
+  bind_rows(as_tibble(VarCorr(fitBApct))) %>% 
+  dplyr::select(Group=grp,Var.=vcov,St.Dev=sdcor)
+
+fitBA.fixed<-
+  fitBA %>% 
+  anova() %>% 
+  tidy() %>% 
+  bind_rows(tidy(anova(fitBApct))) %>% 
+  mutate(term=c("TU*Tree Size*Sp.")) %>% 
+  mutate(p.value=c("<0.00001","<0.00001")) %>% 
+  mutate(R2=c(round(r.squaredGLMM(fitBA)[2]*100,1),
+              round(r.squaredGLMM(fitBApct)[2]*100,1))) %>% 
+  add_column(Model=c("BAI","BAI (%)"),.before="term") %>% 
+  add_row(.after=1) %>% 
+  add_row(.after=3) %>% 
+  cbind(fitBA.random) %>%
+
+  # format table
+  setNames(c("Model","Term","Sum.Sq.","Mean.Sq.","Num.DF","Den.DF",
+             "Stat.","P value","R²","Group","Var.","St.Dev.")) %>% 
+  mutate_at(vars(Sum.Sq.:Stat.),round,3) %>% 
+  mutate_at(vars(Var.:St.Dev.),round,3)
+
+# set typology
+typology <- data.frame(
+  col_keys=names(fitBA.fixed),
+  what = c( "Model",rep("Fixed effects",8),rep("Random effects",3)),
+   what=names(fitBA.fixed),
+  stringsAsFactors = FALSE )
+
+fitBA.fixed %>%
+
+  flextable() %>%
+
+  set_header_df(mapping = typology, key = "col_keys" ) %>% 
+  colformat_num(digits = 2, na_str = "") %>% 
+  colformat_num(j=9,digits = 0, na_str = "") %>% 
+  merge_h(part = "header") %>% 
+  # merge_v(part = "header") %>%
+
+  valign(valign="top",part="all") %>% 
+   hline(i=c(2,4),border=fp_border(color="black")) %>% 
+     hline(part="header",border=fp_border(color="black")) %>% 
+  vline(j=c(1,9),part="all",border=fp_border(color="black")) %>%  
+ align(align="left",part="body") %>%
+
+  fontsize(size=7,part="all") %>% 
+    fix_border_issues() %>% 
+  autofit()
+
+```
+
+*The dependent variable in the first model (BAI) is tree basal area increment from 1992-2019.
+The dependent variable in the second model (BAI (%)) is basal area increment from 1992-2019 as a percentage of the tree's basal area in 1992.*
 
@@
     hline(i=c(3,6,9),border=fp_border(color="black")) %>% 
+   fontsize(size=7,part="all") %>% 
     fix_border_issues() %>% 
@@
 BAI, basal area increment")`
+
+\newpage
 
@@
 
-# caption
-figNums("perInc.plot",caption="Mean plot annual increment over four measurement periods for basal area and volume, by treatment.  Whiskers are one standard error of the mean. Inc, increment.")
-
-```
-
-r figNums("perInc.plot",caption="Mean plot annual increment over four measurement periods for basal area and volume, by treatment.  Whiskers are one standard error of the mean. Inc, increment.")`
+```
+
+`r figNums("perInc.plot",caption="Mean plot annual increment over four measurement periods for basal area and volume, by treatment.  Whiskers are one standard error of the mean. Inc, increment.")`
 
@@
    geom_hline(yintercept=0.5, linetype="dashed", color = "black")+
-  theme(legend.position = "bottom")+
+  theme(legend.position = "bottom",
+        axis.text.x=element_text(angle=90,vjust=0.3))+
   facet_grid(Var~BA.Target)
@@
 `r figNums("sppComp.plot",caption="Mean proportion of spruce and fir basal area, volume, and tree density by treatment and year.")`
+
+\newpage
 
@@
 
-```{r show volume by diameter class}
+\newpage
+
+```{r show volume by diameter class,fig.width=8}
 diamVol.plot
@@
 
-`r figNums("diamVol.plot",caption="Diameter-class distribution of volume (m³/hectare) by species, treatment and year.")`
+`r figNums("diamVol.plot",caption="Diameter-class distribution of volume (m³/hectare) by species, treatment and year.Numbers identifying each diameter classes represent the lower size limit for that class.
+For example, trees in the 7.5cm DBH class are greater or equal to 7.5 and less than 12.5cm DBH.")`
 
 *Note that in this figure, we show volume for all diameter classes.
-This is different than the stand-level estimate of volume, where we only estimate volume for trees that would be considered merchantable under conventional harvest in British Columbia (i.e., minimum 18.5cm DBH).*
-
-\newpage
-
-```{r show tree density by diameter class}
+This is different than the stand-level estimate of volume, where we only estimate volume for trees that would be considered merchantable under conventional harvest in British Columbia (i.e., minimum 18.5cm DBH).
+*
+
+\newpage
+
+```{r show tree density by diameter class,fig.width=8}
 diamK.plot
@@
     coord_flip()+
+    theme(legend.position="bottom")+
     facet_grid(Year~BA.Target) %>% 
@@
 ```
-
-\newpage
 
@@
             legend.position = "bottom",
-            axis.text.x = element_text(angle=270,vjust = 0.3))
-
-```
-
-`r figNums("mortDiam.plot",caption="Total tree mortality recorded in plots from 1997-2019, presented as number of trees (top panel) and basal area (bottom panel).")`
+            axis.text.x = element_text(angle=90,vjust = 0.3))
+
+```
+
+`r figNums("mortDiam.plot",caption="Total tree mortality recorded in plots from 1997-2019, presented as number of trees (top panel) and basal area (bottom panel).
+Numbers identifying each diameter classes represent the lower size limit for that class.
+For example, trees in the 7.5cm DBH class are greater or equal to 7.5 and less than 12.5cm DBH")`
 
@@
   ylab("Number dead 1994-2019")+
-  xlab("DBH class (cm")+
-  theme(legend.position="bottom")
+  xlab("DBH class (cm)")+
+  theme(legend.position="bottom",
+        axis.text.x=element_text(angle=90,vjust=0.3))
 ```
@@
 
-`r appNums("paiTable",caption="Number of growing seasons between measurement periods")`
-
-```{r show paiTable, ft.align="left"}
+```{r show paiTable, ft.align="left",eval=FALSE}
 
@@
 
-```
-
-\newpage
+appNums("paiTable",caption="Number of growing seasons between measurement periods")
+```
 
@@
 
-`r appNums("heightDiam.plot",caption="Relationship between height (m) and diameter (cm) for ")`
+`r appNums("heightDiam.plot",caption="Relationship between height (m) and diameter (cm) for trees with measured heights, by species and year.")`
 
@@
 
+`r appNums("plotPAI.table",caption="Plot means of mean annual volume, basal area and stand density increment, by species, period and treatment.")`
+
 ```{r table of periodic mean annual increment}
@@
 
+  # remove QMD column because it's somewhat nonsensical
+  dplyr::select(-QMD) %>%
+
   # rearrange
-  arrange(Year,BA.Target) %>%
-
-```
-
-`r appNums("standFitM.contrasts",caption="Contrasts of stand attributes (volume, basal area and tree density) between treatments.
+  arrange(Period,BA.Target) %>%
+
+    # Insert number of growing seasons column
+  add_column(Num.Seasons=c(rep(3,6),rep(11,6)),.after="Period") %>% 
+  mutate(Period=paste0(Period," (",Num.Seasons,")")) %>% 
+  dplyr::select(-Num.Seasons) %>%
+
+    # format table
+  flextable() %>%
+  align(align="left",part="all") %>% 
+  merge_v("Period") %>% 
+  valign(valign="top",part="body") %>% 
+  hline(i=seq(from=3,to=12,by=3),border=fp_border(color="black")) %>% 
+  fix_border_issues() %>% 
+   autofit()
+
+```
+
+*Volume is m3/hectare, BA = basal area (m2/ha), SPH = stems per hectare.
+Values in brackets in Period column are number of growing seasons in that period.
+Values in brackets in Volume, BA and SPH columns are the standard error of the mean.*
+
+\newpage
+
+`r appNums("standFitM.contrasts",caption="Contrasts of stand attributes (volume, basal area and tree density) between treatments and by year.
 Contrasts are presented for five measurement years.
@@
   arrange(Contrast,Variable,Year) %>% 
-  mutate_if(is.numeric,round,5) %>%
+  mutate_at(vars(estimate:`t.ratio`),round,2) %>% 
+  mutate(p.value=round(p.value,5)) %>%
 

### 2021-02-20 manuscript_draft.Rmd
working on manuscriot

@@
 # plot multiplier
-pm=25
+pm=20
 
@@
 ### Rate of change from 1992 to 2019
-All treatment units increased volume, basal area and stand density over the 1992-2019 period (`r figNums("standPlots",display="cite")` and `r appNums("standPlot.table",display="cite")`).
+All treatment units, including the control, increased volume, basal area and stand density over the 1992-2019 period (`r figNums("standPlots",display="cite")` and `r appNums("standPlot.table",display="cite")`).
 Mean annual basal area and volume increment peaked during the 1997-2009 period in the two harvested treatments, and peaked in the 1994-1997 period in the control treatment unit (`r figNums("perInc.plot",display="cite")`).
@@
 
-## Species differences in growth response to partial harvest
+## Tree-level growth response to partial harvest
 
@@
 Frank (1973) also showed that 70-year old white spruce basal increment increased with the amount of thinning.
-Competition for light is a key factor that limits white spruce growth, and opening up the stand to allow for increased light penetration may be a primary mechanism to enhance white spruce growth responses [@Lieffers1999].
-High basal area removal may also increase stand air and soil temperatures, which can preferentially favour spruce growth responses compared to fir [@Lajzerowicz2004].
+Competition for light is a key factor that limits white spruce growth, and opening up the stand to allow for increased light penetration may be a primary mechanism that enhances white spruce growth responses [@Lieffers1999].
+High basal area removal may also increase air and soil temperatures, which can preferentially favour spruce growth responses compared to fir [@Lajzerowicz2004].
 
 Other comparative studies of spruce and fir growth rates present differing evidence regarding species differences in release potential to thinning.
-In mature Engelmann spruce (Picea engelmannii) – subalpine fir forests in western North America, understory spruce growth rates exceeded those of fir, allowing spruce to achieve canopy position faster, including after canopy gaps were created through disturbance (Antos et al.
+In mature Engelmann spruce (*Picea engelmannii*) – subalpine fir forests in western North America, understory spruce growth rates exceeded those of fir, allowing spruce to achieve canopy position faster, including after canopy gaps were created through disturbance (Antos et al.
 2000, Andrus et al.
@@
 Because spruce tends to live longer than subalpine fir, higher growth rates at later ages may allow this species to become larger than fir.
-In contrast, thinning studies in spruce-fir forests in eastern North America report that balsam fir (*Abies balsamifera*) responds to release with faster height and diameter growth than co-occurring red spruce (Picea rubens) (Dumais and Prevost, 2007; Dumais et al.
+In contrast, thinning studies in spruce-fir forests in eastern North America report that balsam fir (*Abies balsamifera*) responds to release with faster height and diameter growth than co-occurring red spruce (*Picea rubens*) (Dumais and Prevost, 2007; Dumais et al.
 2014, Messier et al.
@@
 Other studies have found no difference in growth response between spruce and fir species after treatment (MCaughey and Smith1982).
-Our study also suggests that at lower levels of basal area removal, spruce and fir growth responses will be similar.
+Our study also suggests that at lower levels of basal area removal, spruce and fir growth responses are similar.
+
+In both species, post-treatment growth rates increased with the tree size at the time of treatment.
+This has been found in other studies as well.
+and is likely explained by the assymetric competition theory.
+Implications of this result are that growth increased from treatment will be concentrated in the larger tree size classes.
+If larger trees are able to increase their size faster than smaller trees and increase other metrics, such as crown size, to take advantage of increased growing space, then smaller trees will continue to experience competition and will not benefit as much from the thinning treatment.
+Forest managers should account for this size differential when planning thinning treatments.
+For example, a silviculture treatment that seeks to release a suppressed understorey may have to ensure that sufficient overstorey trees are removed to reduce assymetric competition after treatment.
+
+## Stand-level responses to thinning treatments
+
+Volume
+
+Diameter distributions
+All units in this stand maintained a negative-exponential diameter distribution whereby the number of trees was highest in the smallest diameter classes.
+Insofar as this diameter distribution indicates an uneven-aged stand structure, thinning in these stands maintained these characteristics and did not substantially alter the structure.
+The species composition between the low removal and control units was also similar, in that subalpine fir was the leading species in the smaller diameter classes, and spruce was leading the species in the larger classes.
+This may reflect species differences in regeneration and growth.
+In closed forests with undisturbed forest floors, Subalpine fir can have higher seedling establishment success than spruce, leading to higher densities of fir in the understorey.
+Subalpine fir can be relatively short-lived, however, and can die after 250 years in the canopy.
+Spruce, on the other hand, can live for over 400 years and thus can achieve larger sizes and dominant canopy positions than fir.
+This leads to a common stand structure in the Interior Region of BC that was found in the low-reomval and control untis.
+
+The high-removal unit had a somewhat different species composition.
+Here, fir was the leading sprecies in all diameter classes.
+This means that the high-removal thinning treatment converted the stand compsition to fir-leading, which is a primary concern for many foresters in British Columbia.
+Spruce is generally much preferred as a crop species over subalpine fir.
+However, the stands in this treatment had more spruce in the smaller diameter classes than the other treatment units and our results show that spruce increased its composition compared to fir over the 27-year period of this study.
+In contrast, the ratio of spruce to fir volume decreased over the same period in both the low-removal and control units.
+Over a timespan of several decades, it seems reasonable to anticipate that spruce:fir ratios will converge between all treatment units.
+
+Stand mortality increased with basal area after treatment.
+This suggests that the thinning treatments may have captured some volume through trees that would have died naturally.
+Thus, while volume removals from thinning are never entirely recoverable, this may not be true when mortality is considered.
 
@@
 Annual increment in the control unit declined faster over the last measurement period compared to previous periods, perhaps suggesting that the stand is becoming mature and losses due to mortality may substantially offset growth increment.
+3. GIven that larger trees responded with faster growth than smaller trees, silviculturists need to consider the implications of this on residual stand structure and response after treatment.
 

### 2021-02-22 manuscript_draft.Rmd
work on manuscript

@@
 Contrasts between the high-removal and control treatment units showed that volume, basal area and stand density differences were highest and significant (i.e., p<0.05) in the 1992 measurement year, and these differences diminshed over each subsequent measurement year.
-In the 2019 measurement, volume and stand density were not significantly different between these two treatment units (i.e., p>0.05 for both).
+For example, mean plot density in the high-removal treatment unit was approximately 600 trees per hectare less than the control in 1994 (p<0.05); by 2019, that difference had decreased to 219 trees per hectare (p>0.05).
 Contrasts between the low-removal and the control treatment units showed a similar trend with basal area and stand density; the control unit had significanly higher basal area and stand density means in 1992, but the difference between the units diminished over each subsequent measurement, becoming non-significant (i.e., p>0.05) in 2019.
@@
 This has been found in other studies as well.
-and is likely explained by the assymetric competition theory.
-Implications of this result are that growth increased from treatment will be concentrated in the larger tree size classes.
+and may be explained by the asymmetric competition theory whereby larger trees exert more competition on smaller trees.
+Implications of this result are that growth increases from treatment will be predominantly concentrated in the larger tree size classes.
 If larger trees are able to increase their size faster than smaller trees and increase other metrics, such as crown size, to take advantage of increased growing space, then smaller trees will continue to experience competition and will not benefit as much from the thinning treatment.
-Forest managers should account for this size differential when planning thinning treatments.
+Forest managers should account for potential size differential when planning thinning treatments.
 For example, a silviculture treatment that seeks to release a suppressed understorey may have to ensure that sufficient overstorey trees are removed to reduce assymetric competition after treatment.
@@
 
-Volume
-
-Diameter distributions
-All units in this stand maintained a negative-exponential diameter distribution whereby the number of trees was highest in the smallest diameter classes.
-Insofar as this diameter distribution indicates an uneven-aged stand structure, thinning in these stands maintained these characteristics and did not substantially alter the structure.
-The species composition between the low removal and control units was also similar, in that subalpine fir was the leading species in the smaller diameter classes, and spruce was leading the species in the larger classes.
-This may reflect species differences in regeneration and growth.
-In closed forests with undisturbed forest floors, Subalpine fir can have higher seedling establishment success than spruce, leading to higher densities of fir in the understorey.
-Subalpine fir can be relatively short-lived, however, and can die after 250 years in the canopy.
-Spruce, on the other hand, can live for over 400 years and thus can achieve larger sizes and dominant canopy positions than fir.
-This leads to a common stand structure in the Interior Region of BC that was found in the low-reomval and control untis.
-
+Stand density 
+The high-removal treatment unit increased stand density more than the control unit over the 27-year study period and the differences between control and treatment units decreased in each subsequent measurement period.
+In 2019, there were no significant stand density differences between all units.
+These results suggest that the thinned stands increased density faster than the control and within three decades, there are no significant differences between the three units.
+We note that these estimates are for trees with a minimum diameter of 7.5cm at breast height, and therefore our data do not account for new seedlings that may have become established over the time period of the study.
+Thus, the phenomenon of increasing stand density may simply indicate that small trees responded with faster diameter growth in the thinned treatments (as previously discussed) and were able to achieve the minimum diameter class in this study.
+Increasing densities in the smaller diameter classes also suggests that these stands will maintain a negative-exponential diameter distribution, which is considered typical of older, uneven-aged stands in this region.
+Thus, thinning treatments may be used in this context to maintain a structure typical of uneven-aged stands.
+
+The low-removal thinning treatment also maintained a species composition similar to the control, in that subalpine fir was the leading species in the smaller diameter classes, and spruce was the leading species in the larger classes.
+In closed forests with undisturbed forest floors, subalpine fir can have higher seedling establishment success than spruce, leading to higher densities of fir in the understorey.
+Compared to spruce, subalpine fir can be relatively short-lived, however, and can die after 250 years in the canopy.
+Spruce can live for over 400 years and thus can achieve larger sizes and dominant canopy positions.
+This leads to a stand structure that is common in the Interior Region of BC and also found in the low-removal and control units.
 The high-removal unit had a somewhat different species composition.
-Here, fir was the leading sprecies in all diameter classes.
-This means that the high-removal thinning treatment converted the stand compsition to fir-leading, which is a primary concern for many foresters in British Columbia.
-Spruce is generally much preferred as a crop species over subalpine fir.
-However, the stands in this treatment had more spruce in the smaller diameter classes than the other treatment units and our results show that spruce increased its composition compared to fir over the 27-year period of this study.
-In contrast, the ratio of spruce to fir volume decreased over the same period in both the low-removal and control units.
-Over a timespan of several decades, it seems reasonable to anticipate that spruce:fir ratios will converge between all treatment units.
-
+Here, fir was the leading species in all diameter classes.
+This means that the high-removal thinning treatment converted the stand composition to fir-leading, which may be a primary concern for many forest managers, as spruce is often preferred as a crop species over subalpine fir [].
+However, the stands in this treatment had more spruce in the smaller diameter classes than the other treatment units and spruce increased its composition compared to fir over the 27-year period of this study.
+Over time, we expect that spruce would continue to increase its proportion relative to fir, resulting in similar species composition - diameter distributions as the other units.
+
+Volume and basal area
+Our results showed that the low-removal treatment unit and the control unit were similar in terms of stand volume and basal area, and that increment in these variables did not differ much between the two.
+Thus, the long-term implications of this is that these stands are on similar development trajectories
+
+Over the 27-year period, the high-removal treatment unit accrued ca. 8m2 per hectare basal area more than the two other units.
+
+Volume increment did not significantly differ between the units.
+
+Stand mortality
 Stand mortality increased with basal area after treatment.

### 2021-02-25 manuscript_draft.Rmd
manny work

@@
 The number of trees that died per plot from 1994 to 2019 increased significantly with plot basal area in 1994 (`r figNums("mort.plot",display="cite")`, p < 0.0001).
+Mortality was generally not attributable to a particular forest health factor, and in the majority ofcases, we could not ascertain a casual agent.
 Similar linear models fitted separately for fir and spruce showed that fir mortality had a similar relationship to plot basal area, whereas spruce mortality did not (not shown).
@@
 # Discussion
+
+## Uneven-aged stand management
+This study describes tree- and stand-level responses to partial harvest treatments in an uneven-aged stand.  
+- THere are numerous studies documenting responses of even-aged stands to partial harvest treatments such as thinning.
+In even-aged stands, a common objective of thinning treatments is to re-allocate growing space and resources to the most valulable trees in the stand that have capacity to increase production.
+The objectives of partial harvesting in uneven-aged stands is more often to provide a sustained production of wood over multiple generations of trees.
+In this context, partial harvesting in uneven-aged stands seeks to provide adequate growing space and resources to trees across a range of sizes and densities.
+By quantifying growth responses at the tree- and stand-level, this study addresses questions regarding the manipulation of structure in an uneven-aged stand to achieve ...
 
@@
 
-In both species, post-treatment growth rates increased with the tree size at the time of treatment.
-This has been found in other studies as well.
-and may be explained by the asymmetric competition theory whereby larger trees exert more competition on smaller trees.
-Implications of this result are that growth increases from treatment will be predominantly concentrated in the larger tree size classes.
-If larger trees are able to increase their size faster than smaller trees and increase other metrics, such as crown size, to take advantage of increased growing space, then smaller trees will continue to experience competition and will not benefit as much from the thinning treatment.
+In both species, post-treatment growth rates increased with tree size at the time of treatment.
+This has been found in other studies as well (references) and may be explained by the phenomenon of size-asymmetric competition whereby larger trees are able to preempt resources, especially light, from smaller individuals [@Bose2017].
+The implications of size-assymetric competition in thinned stands is that size inequality among trees may increase [@Cordonnier2015], and growth increases from treatment will be predominantly concentrated in the larger tree size classes [@Bose2018; @Schwinning1998].
+Smaller trees will likely continue to experience competition and will not benefit as much from the thinning treatment.
+Thinning treatments may not promote growth responses in the largest docminant trees in a stand either, as these trees may not experience significant changes in resource availability or may have achieved maximal growth rates [@Ryan1997, Bose2017].
 Forest managers should account for potential size differential when planning thinning treatments.
@@
 Stand density 
-The high-removal treatment unit increased stand density more than the control unit over the 27-year study period and the differences between control and treatment units decreased in each subsequent measurement period.
-In 2019, there were no significant stand density differences between all units.
-These results suggest that the thinned stands increased density faster than the control and within three decades, there are no significant differences between the three units.
-We note that these estimates are for trees with a minimum diameter of 7.5cm at breast height, and therefore our data do not account for new seedlings that may have become established over the time period of the study.
-Thus, the phenomenon of increasing stand density may simply indicate that small trees responded with faster diameter growth in the thinned treatments (as previously discussed) and were able to achieve the minimum diameter class in this study.
-Increasing densities in the smaller diameter classes also suggests that these stands will maintain a negative-exponential diameter distribution, which is considered typical of older, uneven-aged stands in this region.
+The thinned treatments increased tree density faster than the control, and within three decades, there were no significant density differences between the three units.
+Increasing stand density reflects the growth of smaller trees that over time, achieved the minimum diameter class (i.e., 7.5cm DBH) in this study.
+The faster increases in stand density of the thinned treatments compared to the control likely reflect that smaller trees in the thinned stands responded with faster diameter increment after treatment (as previously discussed).
+We note that our study did not account for new seedlings that may have become established over the period of the study.
+Based on increasing densities in the smaller diameter classes, these stands will likely maintain a negative-exponential diameter distribution, which is considered typical of older, uneven-aged stands in this region.
 Thus, thinning treatments may be used in this context to maintain a structure typical of uneven-aged stands.
 
-The low-removal thinning treatment also maintained a species composition similar to the control, in that subalpine fir was the leading species in the smaller diameter classes, and spruce was the leading species in the larger classes.
+The low-removal thinning treatment also maintained a species composition similar to the unthinned stands, in that subalpine fir was the leading species in the smaller diameter classes, and spruce was the leading species in the larger classes.
+This is a stand structure and composition common at low- to mid-elevations in the interior of British Columbia.
 In closed forests with undisturbed forest floors, subalpine fir can have higher seedling establishment success than spruce, leading to higher densities of fir in the understorey.
-Compared to spruce, subalpine fir can be relatively short-lived, however, and can die after 250 years in the canopy.
+Compared to spruce, subalpine fir can be relatively short-lived, and can die after 250 years in the canopy.
 Spruce can live for over 400 years and thus can achieve larger sizes and dominant canopy positions.
-This leads to a stand structure that is common in the Interior Region of BC and also found in the low-removal and control units.
 The high-removal unit had a somewhat different species composition.
@@
 This means that the high-removal thinning treatment converted the stand composition to fir-leading, which may be a primary concern for many forest managers, as spruce is often preferred as a crop species over subalpine fir [].
-However, the stands in this treatment had more spruce in the smaller diameter classes than the other treatment units and spruce increased its composition compared to fir over the 27-year period of this study.
-Over time, we expect that spruce would continue to increase its proportion relative to fir, resulting in similar species composition - diameter distributions as the other units.
+However, the stands in this treatment had more spruce in the smaller diameter classes than the other treatment units and over the 27-year period of this study, spruce monotonically increased its composition (by volume and basal area) compared to fir.
+Over time, we expect that spruce would continue to increase its proportion relative to fir and eventually will achieve a similar species composition - diameter distribution as the other units.
 
 Volume and basal area
-Our results showed that the low-removal treatment unit and the control unit were similar in terms of stand volume and basal area, and that increment in these variables did not differ much between the two.
-Thus, the long-term implications of this is that these stands are on similar development trajectories
+Over the 27-year period of this study, the high-removal thinning treatments accrued basal area significantly faster than the control and low-removal thinning treatment, but not volume.
+The volume formula used in this study uses both tree diameter and height [@Nigh2016].
+Treatment-level differences in basal area increment but not volume suggest that tree growth responses to thinning might be primarily concentrated in radial growth, as opposed to height growth.
+It can be difficult to precisely measure tree height growth using field-based methods, particularly in mature trees, which may be a primary reason that studies tend to focus on radial growth responses to silviculture treatments and gap dynamics [@Vepakommaetal2011].
+Enhanced radial growth is a common repsonse to reduced stand densities [@Plauborg2014].
+
+Our results showed that the low-removal treatment unit and the control unit were similar in terms of stand volume and basal area, and that volume/basal area increment did not differ between the two.
+The gaps created by the low-removal treatment may have been too small to elicit a strong growth response as observed in the high-removal thinning treatment.
+Light levels may differ little between forest canopies that are closed or contain small gaps [@Bartemucci2002], and strong vertical light gradients may result in very little light regime changes for understorey trees.
+
+Volume and basal area increment peaked in the 1994-1997 measurement period, with the strongest decline in annual growth over the last measurement period of 2009-2019.
+Age-related declines in stand productivity may be associated with decreasing water-use efficiency and reduced photosynthesis, with the age of maximal growth rate varying with factors such as species and site conditions [@Ryan1997].
+In both thinned treatments, maximal growth rates peaked in the 1997-2009 period.
+Thus, it appears the thinning treatment may have delayed the period of maximal growth rates by approximately 10 years.
+The timing of thinning treatments can be important.
+Stands that are past their period of age-related maximal growth rate may not have as much capacity to respond to treatments as a younger stand, although this is liekly a more important factor for shade-intolerant species rather than the spruce and fir trees in this study [@Bose2017].
 
@@
 Stand mortality increased with basal area after treatment.
+The thinned stands had the lowest mortality over the time period studies.
+Thinning is considered a silviculture treatment that improves growth efficiency of the stand by redistibuting growing space and capturing mortality that would otherwise reduce net stand growth.
+Our findings suggest that this occured, and that the thinning treament captured a portion of the stand that would have been lost eventually to natural mortality.
 This suggests that the thinning treatments may have captured some volume through trees that would have died naturally.
 Thus, while volume removals from thinning are never entirely recoverable, this may not be true when mortality is considered.
+
+[@Valkonen2020] found similar mortality in uneven-aged norway spruce stands.
 

### 2021-02-27 manuscript_draft.Rmd
manuscript discussion

@@
 
-## Tree-level growth response to partial harvest
-
-At the tree-level, our results suggest that while both hybrid white sruce and subalpine fir respond to thinning with faster radial growth, heavy thinning can elicit a stronger radial growth response in spruce, compared to fir.
-This coincides with our finding of proportionally higher spruce volume and basal area in the heavy thinning treatment in 2019, compared to the other treatments.
-Where thinning treatments in this region retain both species, removing a relatively high amount of the stand may favour spruce release.
+## Tree- and stand-level growth response to partial harvest
+
+### Species
+
+At the tree-level, our results suggest that while both hybrid white spruce and subalpine fir respond to thinning with faster radial growth, heavy thinning can elicit a stronger radial growth response in spruce, compared to fir.
+Spruce trees in the low-removal harvest treatment did not respond with a significant growth release.
+THus, our results suggest that a relatively high removal of the stand will result in increased spruce growth, and this release will exceed that of co-occurring fir trees.
 Other studies have shown that residual spruce growth responds favourably to heavy thinning.
@@
 High basal area removal may also increase air and soil temperatures, which can preferentially favour spruce growth responses compared to fir [@Lajzerowicz2004].
-
-Other comparative studies of spruce and fir growth rates present differing evidence regarding species differences in release potential to thinning.
+The relatively low spruce growth response in the low-removal treatment may be explained in that the gaps created may have been too small to elicit as strong a growth response as observed in the high-removal thinning treatment.
+Light levels may differ little between forest canopies that are closed or contain small gaps [@Bartemucci2002], and strong vertical light gradients may result in relatively little light regime changes for understorey trees.
+
+In comparison, both partial harvest treatments stimulated a growth response in fir, compared to the control.
+Although both species are similar in shade tolerance, firs are considered to be better adapted to low light conditions in the understory, allowing them to persist longer in undisturbed forest  (Antos 2000) and potentially able to respond faster to even modest changes in understory light (Dumais et al 2014, Messier et al.
+1999).
+Although fir growth may respond at lower levels of overstorey removal, the growth response potential of fir may be lower than spruce.
+Comparative studies of spruce and fir growth rates present differing evidence regarding species differences in release potential to thinning.
 In mature Engelmann spruce (*Picea engelmannii*) – subalpine fir forests in western North America, understory spruce growth rates exceeded those of fir, allowing spruce to achieve canopy position faster, including after canopy gaps were created through disturbance (Antos et al.
@@
 1999).
-Although both species are similar in shade tolerance, firs are considered to be better adapted to low light conditions in the understory, allowing them to persist longer in undisturbed forest  (Antos 2000) and potentially able to respond faster to changes in understory light (Dumais et al 2014, Messier et al.
-1999).
 Other studies have found no difference in growth response between spruce and fir species after treatment (MCaughey and Smith1982).
@@
 
-In both species, post-treatment growth rates increased with tree size at the time of treatment.
-This has been found in other studies as well (references) and may be explained by the phenomenon of size-asymmetric competition whereby larger trees are able to preempt resources, especially light, from smaller individuals [@Bose2017].
-The implications of size-assymetric competition in thinned stands is that size inequality among trees may increase [@Cordonnier2015], and growth increases from treatment will be predominantly concentrated in the larger tree size classes [@Bose2018; @Schwinning1998].
+Post-treatment growth rates increased with tree size at the time of treatment, for both species.
+This has been found in other studies as well (references) and may be related to size-asymmetric competition whereby larger trees preempt resources, especially light, from smaller individuals [@Bose2017].
+Thinning treatments may exacerbate size-assymetric competition [@Cordonnier2015], and growth increases from treatment will be predominantly concentrated in the larger tree size classes [@Bose2018; @Schwinning1998].
 Smaller trees will likely continue to experience competition and will not benefit as much from the thinning treatment.
-Thinning treatments may not promote growth responses in the largest docminant trees in a stand either, as these trees may not experience significant changes in resource availability or may have achieved maximal growth rates [@Ryan1997, Bose2017].
+Thinning treatments may not promote growth responses in the largest dominant trees in a stand either, as these trees may not experience significant changes in resource availability or may have achieved maximal growth rates [@Ryan1997, Bose2017].
 Forest managers should account for potential size differential when planning thinning treatments.
@@
 Stand density 
-The thinned treatments increased tree density faster than the control, and within three decades, there were no significant density differences between the three units.
+The thinned treatments increased tree density faster than the control, and within three decades, there were no significant density differences between the two harvested treatments and control.
 Increasing stand density reflects the growth of smaller trees that over time, achieved the minimum diameter class (i.e., 7.5cm DBH) in this study.
@@
 Based on increasing densities in the smaller diameter classes, these stands will likely maintain a negative-exponential diameter distribution, which is considered typical of older, uneven-aged stands in this region.
-Thus, thinning treatments may be used in this context to maintain a structure typical of uneven-aged stands.
+Thus, partial harvest treatments may be used in this context to maintain a structure typical of uneven-aged stands.
 
@@
 
-Volume and basal area
-Over the 27-year period of this study, the high-removal thinning treatments accrued basal area significantly faster than the control and low-removal thinning treatment, but not volume.
-The volume formula used in this study uses both tree diameter and height [@Nigh2016].
-Treatment-level differences in basal area increment but not volume suggest that tree growth responses to thinning might be primarily concentrated in radial growth, as opposed to height growth.
+### Stand-level growth responses
+Over the 27-year period of this study, plots with high-removal harvest treatments accrued basal area significantly faster than the control and low-removal harvest treatment, likely due to increases in tree-level growth and stand density discussed previously.
+As a result, differences in basal area between the high-removal harvest unit and the control reduced from ca.
+17m2/hectare in 1992 to ca.9m2/hectare in 2019.
+In contrast, basal area increment between the low-removal unit and the control did not differ significantly over the 1992-2019 period.
+The difference between these two units remained relatively constant over all measurement periods at ca.
+10m2/hectare.
+The similar rates of basal area increment between the low-removal thinning treatment and the control are difficult to explain.
+Tree-level results showed that fir grew faster in the low-removal unit than the control, but spruce did not.
+As well, by 2019 there were no significant differences in stand density between the two units.
+Thus, it appears that while the fir trees may have increased growth as a response of low-removal, the growth increases were insufficient to cause a significant change in basal area increment.
+Furhter monitoring in this study will help determine the future trajectory of the low-removal thinning treatment.
+
+While basal area increment differed between the high-removal treatment unit and the control, there was little difference in volume increment between any of the units.
+A possible explanation is that tree growth responses to thinning might be primarily concentrated in radial growth, as opposed to height growth; this would be important because the volume formula used in this study is based on both tree diameter and height [@Nigh2016].
 It can be difficult to precisely measure tree height growth using field-based methods, particularly in mature trees, which may be a primary reason that studies tend to focus on radial growth responses to silviculture treatments and gap dynamics [@Vepakommaetal2011].
-Enhanced radial growth is a common repsonse to reduced stand densities [@Plauborg2014].
-
-Our results showed that the low-removal treatment unit and the control unit were similar in terms of stand volume and basal area, and that volume/basal area increment did not differ between the two.
-The gaps created by the low-removal treatment may have been too small to elicit a strong growth response as observed in the high-removal thinning treatment.
-Light levels may differ little between forest canopies that are closed or contain small gaps [@Bartemucci2002], and strong vertical light gradients may result in very little light regime changes for understorey trees.
+Commercial thinning treatments in spruce fir stands increased tree taper [@]    Enhanced radial growth is a common repsonse to reduced stand densities [@Plauborg2014].
 
@@
 Stands that are past their period of age-related maximal growth rate may not have as much capacity to respond to treatments as a younger stand, although this is liekly a more important factor for shade-intolerant species rather than the spruce and fir trees in this study [@Bose2017].
-
-Over the 27-year period, the high-removal treatment unit accrued ca. 8m2 per hectare basal area more than the two other units.
-
-Volume increment did not significantly differ between the units.
 

### 2021-03-05 manuscript_draft.Rmd
k;j

@@
 [-@Nakagawa2013], as implemented in the MuMIn package for R [@barton2009mumin].
-Pairwise contrasts were performed using the 'emmeans' package for R [@] and Tukey HSD method of p value adjustment was used to control for multiplicity.
+Pairwise contrasts were performed using the 'emmeans' package for R [@] and Tukey HSD method of *p* value adjustment was used to control for multiplicity.
 We report all significant results at p<0.05.
@@
 The thinned treatments increased tree density faster than the control, and within three decades, there were no significant density differences between the two harvested treatments and control.
-Increasing stand density reflects the growth of smaller trees that over time, achieved the minimum diameter class (i.e., 7.5cm DBH) in this study.
+This likely resulted from two processes; growth and mortality.
+Increasing stand density reflects the growth of smaller trees that achieved the minimum diameter class (i.e., 7.5cm DBH) in this study.
 The faster increases in stand density of the thinned treatments compared to the control likely reflect that smaller trees in the thinned stands responded with faster diameter increment after treatment (as previously discussed).
@@
 Based on increasing densities in the smaller diameter classes, these stands will likely maintain a negative-exponential diameter distribution, which is considered typical of older, uneven-aged stands in this region.
-Thus, partial harvest treatments may be used in this context to maintain a structure typical of uneven-aged stands.
-
-The low-removal thinning treatment also maintained a species composition similar to the unthinned stands, in that subalpine fir was the leading species in the smaller diameter classes, and spruce was the leading species in the larger classes.
+
+  Stand densities between all units likely also converged because of different mortality rates.
+Tree mortality increased with stand-level basal area, and the thinned stands had the lowest mortality over the time of the study.
+Most of the mortality was concentrated in fir trees among the smaller size classes.
+Density-dependent tree mortality is common in younger, even-aged stands, reflecting intertree competition for light and other resources, and resulting in increasing uniformity of tree distribution [@Larson2015].
+In the uneven-aged stands of our study, density-dependent mortality likely reflects indicate the effects of assymetric competition and suppression of .
+In the unthinned stands, smaller trees died more frequently as a result of suppression from the larger trees, whereas thinning reduced this competition.
+Density-dependent mortality in uneven-aged, mixed-species forests can also vary with drought and site-level conditions such as aspect and soil moisture regime, which needs to be considered when planning stand thinning treatments [@VanGunst2016].
+
+Species composition
+
+The low-removal thinning treatment also maintained a species composition by diameter class similar to the unthinned stands.
+In these units, subalpine fir was the leading species by frequency in the smaller diameter classes, and spruce was the leading species in the larger classes.
 This is a stand structure and composition common at low- to mid-elevations in the interior of British Columbia.
@@
 Spruce can live for over 400 years and thus can achieve larger sizes and dominant canopy positions.
-The high-removal unit had a somewhat different species composition.
+
+The high-removal unit had a somewhat different species composition by diameter class.
 Here, fir was the leading species in all diameter classes.
-This means that the high-removal thinning treatment converted the stand composition to fir-leading, which may be a primary concern for many forest managers, as spruce is often preferred as a crop species over subalpine fir [].
+This means that the high-removal thinning treatment converted the stand composition to fir-leading in all tree sizes.
+This result may be concern for forest managers, as spruce is often preferred as a crop species over subalpine fir [].
 However, the stands in this treatment had more spruce in the smaller diameter classes than the other treatment units and over the 27-year period of this study, spruce monotonically increased its composition (by volume and basal area) compared to fir.
-Over time, we expect that spruce would continue to increase its proportion relative to fir and eventually will achieve a similar species composition - diameter distribution as the other units.
+Over time, we expect that spruce will continue to increase its proportion relative to fir and eventually achieve a similar species composition - diameter distribution as the other units.
+
+### Tree mortality
+
+, reflecting intertree competiton for light and other resources.
+In older stands with more complex structures, tree mortality tends to be primarily caused by senescence and spatially autocorrelated forest agents.
+Given that the forests in this study are older and have a stand structure typial of an uneven-aged stand,.
+, which tends to result in spatially aggregated tree distributions but by Thinning is considered a silviculture treatment that improves growth efficiency of the stand by redistibuting growing space and capturing mortality that would otherwise reduce net stand growth.
+Our findings support suggest that this occured, and that the thinning treament captured a portion of the stand that would have been lost eventually to natural mortality.
+The thinning treatments may have captured some volume through trees that would have died naturally, potentially increasing the net recoverable volume from these stands.
+[@Valkonen2020] found similar mortality in uneven-aged norway spruce stands.
 
@@
 17m2/hectare in 1992 to ca.9m2/hectare in 2019.
-In contrast, basal area increment between the low-removal unit and the control did not differ significantly over the 1992-2019 period.
-The difference between these two units remained relatively constant over all measurement periods at ca.
+
+In contrast, basal area increment between the low-removal unit and the control did not differ significantly over the 1992-2019 period, and the difference between these two units remained relatively constant over all measurement periods at ca.
 10m2/hectare.
@@
 Thus, it appears that while the fir trees may have increased growth as a response of low-removal, the growth increases were insufficient to cause a significant change in basal area increment.
-Furhter monitoring in this study will help determine the future trajectory of the low-removal thinning treatment.
-
-While basal area increment differed between the high-removal treatment unit and the control, there was little difference in volume increment between any of the units.
+Further monitoring in this study will help determine the future trajectory of the low-removal thinning treatment.
+
+While our study found that high-removal partial harvest resulted in faster basal area growth, we did not find evidence that thinning changed volume increment in trees greater than 17.5cm DBH.
 A possible explanation is that tree growth responses to thinning might be primarily concentrated in radial growth, as opposed to height growth; this would be important because the volume formula used in this study is based on both tree diameter and height [@Nigh2016].
-It can be difficult to precisely measure tree height growth using field-based methods, particularly in mature trees, which may be a primary reason that studies tend to focus on radial growth responses to silviculture treatments and gap dynamics [@Vepakommaetal2011].
-Commercial thinning treatments in spruce fir stands increased tree taper [@]    Enhanced radial growth is a common repsonse to reduced stand densities [@Plauborg2014].
-
-Volume and basal area increment peaked in the 1994-1997 measurement period, with the strongest decline in annual growth over the last measurement period of 2009-2019.
+We note that it is often difficult to precisely capture height growth changes in mature trees using conventional field-based methods [@Vepakommaetal2011], which makes capturing height changes problematic in these sorts of studies.
+We also utilized a 17.5cm DBH limit for volume calculations; as such, our volume measures only reflect the contribution of the larger trees in the stand, and do not include volume increment in smaller trees.
+
+The partial harvest treatment appeared to change the timing of peak stand growth.
+In the control unit, Volume and basal area increment peaked in the 1994-1997 measurement period, and declined thereafter, with the strongest decline in annual growth over the last measurement period of 2009-2019.
 Age-related declines in stand productivity may be associated with decreasing water-use efficiency and reduced photosynthesis, with the age of maximal growth rate varying with factors such as species and site conditions [@Ryan1997].
@@
 Stands that are past their period of age-related maximal growth rate may not have as much capacity to respond to treatments as a younger stand, although this is liekly a more important factor for shade-intolerant species rather than the spruce and fir trees in this study [@Bose2017].
-
-Stand mortality
-Stand mortality increased with basal area after treatment.
-The thinned stands had the lowest mortality over the time period studies.
-Thinning is considered a silviculture treatment that improves growth efficiency of the stand by redistibuting growing space and capturing mortality that would otherwise reduce net stand growth.
-Our findings suggest that this occured, and that the thinning treament captured a portion of the stand that would have been lost eventually to natural mortality.
-This suggests that the thinning treatments may have captured some volume through trees that would have died naturally.
-Thus, while volume removals from thinning are never entirely recoverable, this may not be true when mortality is considered.
-
-[@Valkonen2020] found similar mortality in uneven-aged norway spruce stands.
+At the time of this study, the high-removal thinning treatment had maintianed an annual basal area and volume increment that was similar to previous measurement periods; if this high level of increment continues, differences in basal area and volume will continue to diminish compared to the control, and this stand may exceed the low-removal thinning treatment.
+
+## Management of uneven-aged stands
 

### 2021-03-06 manuscript_draft.Rmd
growth dominance statistics

@@
 # Create species mean contrasts table
-speciesFitM.contrasts<-
 
@@
 Plot means of volume did not differ significantly between the treatment units in any measurement year.
+
+### Growth dominance statistics
+
+```{r calculate growth dominance}
+
+# Filter only for trees that have measurements in 1992 and 2019 
+gd.treeList<- # assign variable
+  X %>% 
+  filter(TreeID!="New") %>% # remove new trees
+  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
+  filter(Status=="Live") %>% # filter for live trees
+  group_by(TreeID) %>% 
+  summarise(N=n()) %>% 
+  filter(N>=2) %>% #filter for trees with at least 2 measurements
+  pull(TreeID)
+
+# derive periodic increment (period diameter or basal area increment)
+gd.BAInc<-
+  X %>% 
+  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
+  filter(TreeID %in% gd.treeList) %>% 
+  mutate(BA=d2ba(DBH)) %>% # convert DBH to basal area
+  dplyr::select(Plot,TreeID,Species,Year,BA) %>% 
+  pivot_wider(names_from=Year,values_from=BA) %>% 
+  mutate(PBAI_2019=`2019`-`2009`,
+         PBAI_2009=`2009`-`1997`,
+         PBAI_1997=`1997`-`1994`,
+         PBAI_1994=`1994`-`1992`) %>% 
+  dplyr::select(Plot,TreeID,Species,starts_with("PBAI_"))
+
+gd.Inc<-
+  X %>% 
+  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
+  filter(TreeID %in% gd.treeList) %>% 
+  left_join(filter(volCoef,VolType=="total"),by="Species") %>% 
+  mutate(vol=exp(1)^b0*DBH^b1*Height.Pred^b2) %>% 
+  dplyr::select(Plot,TreeID,Species,Year,Vol=vol) %>% 
+  pivot_wider(names_from=Year,values_from=Vol) %>%
+   mutate(PVI_2019=`2019`-`2009`,
+         PVI_2009=`2009`-`1997`,
+         PVI_1997=`1997`-`1994`,
+         PVI_1994=`1994`-`1992`) %>% 
+  dplyr::select(Plot,TreeID,Species,starts_with("PVI_")) %>% 
+  inner_join(gd.BAInc,by=c("Plot","TreeID","Species")) %>% 
+  pivot_longer(
+    cols = -c("Plot","Species","TreeID"),
+    names_to = c("Var", "Year"),
+    names_sep = "_",
+    values_to = "Inc")
+
+gd.prelim<-
+  X %>% 
+  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
+  filter(TreeID %in% gd.treeList) %>% 
+  dplyr::select(Plot,TreeID,Species,Year,DBH.Init=DBH) %>% 
+  mutate(Year=fct_recode(Year,tmp_1994="1992",
+                         tmp_1997="1994",
+                         tmp_2009="1997",
+                         tmp_2019="2009")) %>% 
+  mutate(Year=str_remove(as.character(Year),pattern="tmp_")) %>% 
+  mutate(Year=factor(Year)) %>% 
+  mutate(BA.Init=d2ba(DBH.Init)) %>% 
+  dplyr::select(Plot:Year,BA.Init)
+
+gd.Dat <-
+  X %>% 
+  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
+  filter(TreeID %in% gd.treeList) %>% 
+  dplyr::select(Plot,TreeID,Species,Year,Status,DBH,Height) %>% 
+  left_join(gd.Inc,by=c("Plot","TreeID","Species","Year")) %>% 
+  left_join(gd.prelim,by=c("Plot","TreeID","Species","Year")) %>% 
+  filter(Status=="Live") %>% 
+  drop_na(Inc)
+
+# clean up some objects
+  rm(gd.prelim,gd.BAInc,gd.Inc)
+
+# Plots of GD
+
+  ## Basal area
+  gd.cumu.Inc<-  
+    gd.Dat %>% 
+    filter(Var=="PBAI") %>% 
+    arrange(Plot,Year,BA.Init) %>% 
+    group_by(Plot,Year) %>%
+
+    # CUmulative increment
+    mutate(Cumu.Inc=cumsum(Inc)) %>% 
+    mutate(Cumu.Inc2=max(Cumu.Inc)) %>% 
+    mutate(Cumu.Prop.Inc=Cumu.Inc/Cumu.Inc2) %>% 
+    dplyr::select(-Cumu.Inc2) %>%
+
+    # Cumulative stand basal area
+    mutate(Cumu.BA=cumsum(BA.Init)) %>% 
+    mutate(Cumu.BA2=max(Cumu.BA)) %>% 
+    mutate(Cumu.Prop.BA=Cumu.BA/Cumu.BA2) %>% 
+    dplyr::select(-Cumu.BA2)
+
+    # Calculate Growth Dominance statistic
+    gd.stats<-
+      gd.cumu.Inc %>% 
+      mutate(GDS=(Cumu.Prop.Inc+lag(Cumu.Prop.Inc,default=0))*(Cumu.Prop.BA-lag(Cumu.Prop.BA,default=0))) %>% 
+      ungroup() %>% 
+      group_by(Plot,Year) %>% 
+      summarise(GDS=1-sum(GDS))
+
+    gd.stats %>% 
+      ggplot()+
+      aes(x=Year,y=GDS)+
+      geom_point()+
+      facet_wrap(~Plot)
+
+    ggplot()+
+    aes(x=Cumu.Prop.BA,y=Cumu.Prop.Inc)+
+    geom_point()+
+    geom_line()+
+    geom_abline(slope=1,intercept=0,linetype = "dashed")+
+    theme(aspect.ratio = 1)+
+    facet_wrap(~Year,scales="free_y")
+
+```
 

### 2021-03-06 manuscript_draft.Rmd
growth dominance stats

@@
       group_by(Plot,Year) %>% 
-      summarise(GDS=1-sum(GDS))
+      summarise(GDS=1-sum(GDS)) %>% 
+      left_join(plotDat,by="Plot")
 
     gd.stats %>% 
+      filter(Year=="1994") %>% 
       ggplot()+
-      aes(x=Year,y=GDS)+
-      geom_point()+
-      facet_wrap(~Plot)
+      aes(x=Pre.Harvest,y=GDS)+
+      geom_point()
 

### 2021-03-16 manuscript_draft.Rmd
remove 1992 data from analysis

--- 2021-03-06 manuscript_draft.Rmd
@@
   # Create TU lookup
-  mutate(BA.Target=paste0("TU:",BA.Target))
+  mutate(BA.Target=paste0("TU:",BA.Target)) %>%
+
+  # need to remove plots 16-19 because they were only measured once
+  filter(!Plot %in% c(16,17,18,19))
 
@@
 
-To test the hypothesis that tree-level radial growth response over the 1992-2019 period varied with interactions between (i) tree diameter at the time of treatment (1992); (ii) treatment; and (iii) species, we fit linear mixed effects models with log-transformed 1992-2019 diameter increment as the response variable, and treatment, species and log-transformed 1992 diameter as interacting fixed effects.
+To test the hypothesis that tree-level radial growth response over the 1994-2019 period varied with interactions between (i) tree diameter at the time of final treatment (1994); (ii) treatment; and (iii) species, we fit linear mixed effects models with log-transformed 1994-2019 diameter increment as the response variable, and treatment, species and log-transformed 1992 diameter as interacting fixed effects.
 To account for plot-level productivity differences, we included plot as a random intercept effect.
 Models were fit using the 'lme4' package for R [@].
-We also fit a similar model with the 1992-2019 diameter increment expressed as a percentage of 1992 diameter.
+We also fit a similar model with the 1994-2019 diameter increment expressed as a percentage of 1994 diameter.
 Plots of residuals and fit lines with orginal data were examined to evaluate model goodness of fit.
@@
 
-### Basal area increment from 1992-2019
+### Basal area increment from 1994-2019
 
@@
   filter(Status=="Live") %>% # filter for live trees
-  filter(Year %in% c(1992,2019)) %>% #filter for start and end dates
+  filter(Year %in% c(1994,2019)) %>% #filter for start and end dates
   group_by(TreeID) %>% 
@@
   pivot_wider(names_from=Year,values_from=BA) %>% 
-  mutate(PI=`2019`-`1992`) %>% # 
+  mutate(PI=`2019`-`1994`) %>% # 
   filter(PI>0) %>% # one tree has a PI of 0, has to be removed
-  mutate(PIpct=PI/`1992`*100) %>% # express periodic increment a a percentage
-  dplyr::select(Plot,TreeID,Species,PBAI=PI,PBAIpct=PIpct,Init=`1992`) %>% 
+  mutate(PIpct=PI/`1994`*100) %>% # express periodic increment a a percentage
+  dplyr::select(Plot,TreeID,Species,PBAI=PI,PBAIpct=PIpct,Init=`1994`) %>% 
   left_join(plotDat,by="Plot")
@@
   arrange(BA.Target,Init) %>% 
-  rename(DBH.1992="Init")
+  rename(DBH.1994="Init")
 
@@
   arrange(Species,contrast,Init) %>% 
-  rename(DBH.1992="Init")
-
-```
-
-The tree-level basal area increment model indicated that interactions between log-transformed tree size, species and treatment explained `r round(r.squaredGLMM(fitBA)[2]*100,1)`% of the variation in log-transformed 1992-2019 tree-level BAI, and `r round(r.squaredGLMM(fitBApct)[2]*100,1)`% of the variation in basal area increment as a percentage of tree size in 1992 (p<0.0001 for both, `r tableNums("fitBA.table",display="cite")`).
+  rename(DBH.1994="Init")
+
+```
+
+The tree-level basal area increment model indicated that interactions between log-transformed tree size, species and treatment explained `r round(r.squaredGLMM(fitBA)[2]*100,1)`% of the variation in log-transformed 1994-2019 tree-level BAI, and `r round(r.squaredGLMM(fitBApct)[2]*100,1)`% of the variation in basal area increment as a percentage of tree size in 1994 (p<0.0001 for both, `r tableNums("fitBA.table",display="cite")`).
 Estimated marginal means from the model showed that between species and among treatments, basal area increment increased with tree size, meaning that larger trees increased basal area more than smaller trees (`r figNums("treeGrowth.plot",display="cite")`), although smaller trees increased their size proportionally more than large trees (`r figNums("treeGrowthPct.plot",display="cite")`).
@@
   X %>% 
-   filter(Year %in% c(1992,2019)) %>% 
+   filter(Year %in% c(1994,2019)) %>% 
    filter(Status=="Live") %>%
@@
   # Filter for years of interest
-  filter(Year %in% c(1992,2019)) %>% 
+  filter(Year %in% c(1994,2019)) %>% 
   mutate(Year=fct_drop(Year)) %>% 
@@
 plotData.X<-
-# Determine which plots have measurements in 1992 and 2019
+
   X %>%
@@
   pivot_wider(names_from = Year,values_from = val) %>% 
-  mutate(Period1=(`1994`-`1992`)/3,
-         Period2=(`1997`-`1994`)/3,
-         Period3=(`2009`-`1997`)/11,
-         Period4=(`2019`-`2009`)/11) %>% 
-  dplyr::select(BA.Target:var,Period1:Period4) %>% 
-  pivot_longer(cols=Period1:Period4,names_to="Period",values_to="Values") %>% 
+  mutate(Period1=(`1997`-`1994`)/3,
+         Period2=(`2009`-`1997`)/11,
+         Period3=(`2019`-`2009`)/11) %>% 
+  dplyr::select(BA.Target:var,Period1:Period3) %>% 
+  pivot_longer(cols=Period1:Period3,names_to="Period",values_to="Values") %>% 
   pivot_wider(names_from=var,values_from=Values) %>%
@@
   mutate(Period=dplyr::recode(Period,
-                       Period1="1992-1994",
-                       Period2="1994-1997",
-                       Period3="1997-2009",
-                       Period4="2009-2019"))
+                       Period1="1994-1997",
+                       Period2="1997-2009",
+                       Period3="2009-2019"))
 
@@
   rbind(data.frame(Variable="Stand density",sph.contrast)) %>% 
-  filter(Year_revpairwise %in% c("2019 - 1992")) %>% 
+  filter(Year_revpairwise %in% c("2019 - 1994")) %>% 
   dplyr::select(Variable,`Year contrast`="Year_revpairwise",`Treatment contrast`="BA.Target_pairwise",everything()) %>% 
@@
 
-# Create species mean contrasts table
-
-```
-
-### Rate of change from 1992 to 2019
-All treatment units, including the control, increased volume, basal area and stand density over the 1992-2019 period (`r figNums("standPlots",display="cite")` and `r appNums("standPlot.table",display="cite")`).
+```
+
+### Rate of change from 1994 to 2019
+All treatment units, including the control, increased volume, basal area and stand density over the 1994-2019 period (`r figNums("standPlots",display="cite")` and `r appNums("standPlot.table",display="cite")`).
 Mean annual basal area and volume increment peaked during the 1997-2009 period in the two harvested treatments, and peaked in the 1994-1997 period in the control treatment unit (`r figNums("perInc.plot",display="cite")`).
-From 1992 to 2019, estimated marginal means of basal area increment in the low RBA unit exceeded marginal means in the high RBA and control unit by an estimated
+From 1994 to 2019, estimated marginal means of basal area increment in the low RBA unit exceeded marginal means in the high RBA and control unit by an estimated
 `r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Basal area") %>%   pull(estimate) %>% round(.,1) %>% .[1]`
@@
 
+# # sgr.Dat
+#   sgr.Dat<-
+#     gd.Dat %>%
+#     filter(Var=="PBAI") %>% 
+#     ungroup() %>% 
+#     group_by(Plot) %>% 
+#     mutate(Stand.Inc=sum(Inc),
+#            Stand.BA=sum(BA.Init)) %>% 
+#     mutate(Prop.growth=Inc/Stand.Inc,
+#            Prop.size=BA.Init/Stand.BA) %>% 
+#     left_join(plotDat,by="Plot")
+
 # Plots of GD
@@
     mutate(Cumu.Prop.BA=Cumu.BA/Cumu.BA2) %>% 
-    dplyr::select(-Cumu.BA2)
+    dplyr::select(-Cumu.BA2) %>%
+
+    # Bring in plot data
+     left_join(plotDat,by="Plot")
 
@@
       group_by(Plot,Year) %>% 
-      summarise(GDS=1-sum(GDS))
-
-    gd.stats %>% 
+      summarise(GDS=1-sum(GDS)) %>% 
+      left_join(plotDat,by="Plot") %>% 
+      ungroup() %>% 
+      group_by(BA.Target,Year) %>% 
+      summarise(GDS.mean=round(mean(GDS),3),
+                GDS.sd=round(sd(GDS),2)) %>% 
+      mutate(GDS.label=paste0("GDS: ",GDS.mean," (",GDS.sd,") "))
+
+  #   gd.stats.plot<-
+  #     gd.stats %>% 
+  #   filter(Year %in% c(2009,2019)) %>% 
+  # ggplot()+
+  # aes(x=BA.Target,y=GDS,fill=Year)+
+  # geom_boxplot()
+  #
+
+  # Plotting
+  # ## Growth Dominance statistic
+  gd.plot<-
+    gd.cumu.Inc %>%
+    filter(Year %in% c(2009,2019)) %>%
     ggplot()+
-      aes(x=Year,y=GDS)+
-      geom_point()+
-      facet_wrap(~Plot)
-
-    ggplot()+
-    aes(x=Cumu.Prop.BA,y=Cumu.Prop.Inc)+
-    geom_point()+
-    geom_line()+
-    geom_abline(slope=1,intercept=0,linetype = "dashed")+
-    theme(aspect.ratio = 1)+
-    facet_wrap(~Year,scales="free_y")
-
-```
+    aes(x=Cumu.Prop.BA,y=Cumu.Prop.Inc,group=BA.Target)+
+    geom_point(colour="grey")+
+    geom_smooth(se=F,size=1.5)+
+    geom_abline(slope=1,intercept=0,linetype = "dashed",size=1.5)+
+    theme(aspect.ratio = 1,
+          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
+    facet_grid(Year~BA.Target,scales="free_y")+
+    xlab("Cumulative proportion of stand size (basal area)")+
+    ylab("Cumulative proportion of stand growth (basal area)")+
+    facet_grid(Year~BA.Target)+
+    geom_text(data=filter(gd.stats %>% filter(Year %in% c(2009,2019))),
+            size=4,
+            mapping = aes(x = Inf, y = 0.1, label = GDS.label),
+            hjust="right")
+
+  ## Plot measurement year GSD
+      # gd.stats %>% 
+      #   ggplot()+
+      #   aes(x=Year,y=GDS,group=Plot,color=Plot)+
+      #   geom_line()+
+      #   facet_wrap(~BA.Target)
+      #
+
+  ## GSize-Growth relationship statistic
+      # sgr.ind.plot<-
+      #   sgr.Dat %>% 
+      #  ggplot()+
+      #   aes(x=Prop.size,y=Prop.growth,group=Plot,color=BA.Target)+
+      #   geom_point()+
+      #   geom_smooth(aes(group=Plot),method="lm",se=F,color="black")+
+      #   facet_wrap(~Year)+
+      #       theme(aspect.ratio = 1)+
+      #     facet_grid(BA.Target~Year)
+
+```
+
+Across all plots and time periods, smaller trees contributed proportionally more to stand basal area growth than larger trees (`r figNums("gds.plot",display="cite")`).
+Over the last two measurement periods, the growth dominance statistic decreased in all plots except one (not shown), indicating that individual tree growth contributions to stand growth became more proportional to tree size over the last 11 growing seasons.
 

### 2021-03-16 manuscript_draft.Rmd
moved growth dominance statistic seciton around

@@
 bibliography: ep1162.bib
-
+editor_options: 
+  markdown: 
+    wrap: 72
 ---
@@
 
-we can put text below each comment to describe how it was addressed, and then delete comments for any issues once they are resolved.
+we can put text below each comment to describe how it was addressed, and
+then delete comments for any issues once they are resolved.
 
@@
 
-- The term "units" is used in several places, where it seems that you mean "treatments" (or treatment types, treatment unit).
-
-- The use of the term "plot" or "plots" is used appropriately as far as I can see.
-
-- The term "fir" is used to describe subalpine fir or 'balsam' in several parts of the text, including possibly figure and table captions.
-I know what species you are referring to, of course, but others, especially cursory readers, may misconstrue it.
-The colloquial use of 'fir" is spread across both Abies and Douglas-fir.
-"Balsam" as a term has well-documented ambiguities, and while Canadians tend to use balsam for Abies, Americans often call Abies "firs", as in the various species of true fir.
-I would recommend using Abies as the shorthand descriptor in the text and captions for subalpine fir.
-It is a short word, and we have only 1 Abies species at this site.
-Also, my experience in the on-line-search literature world and with things like Google Scholar is that Abies is a much better and more diagnostic search word for this subject matter than fir or balsam.
-"Spruce-Abies" is very unambiguous as a descriptor for this forest type.
-
-- Rate of change (increase, decrease) in vol, BA, and sph: In both in your stats tables and figure captions, and in the text narrative, ensure that what you are describing is clearly and accurately described as either:  (a) the incremental change in values between two measurement periods, or (b)  the difference in absolute value in certain parameters (e.g.
-standing volume) between two measurement periods, or between two treatments.
-Also, the annualized growth rate of something averaged over a measurement period (e.g.
-mean PBAI, like BA growth in m2/yr over a period) is different than the rate of change in a growth rate over time (e.g.
-is a growth rate increasing or decreasing).
-I got confused a few times on this issue.
+-   The term "units" is used in several places, where it seems that you
+    mean "treatments" (or treatment types, treatment unit).
+
+-   The use of the term "plot" or "plots" is used appropriately as far
+    as I can see.
+
+-   The term "fir" is used to describe subalpine fir or 'balsam' in
+    several parts of the text, including possibly figure and table
+    captions. I know what species you are referring to, of course, but
+    others, especially cursory readers, may misconstrue it. The
+    colloquial use of 'fir" is spread across both Abies and Douglas-fir.
+    "Balsam" as a term has well-documented ambiguities, and while
+    Canadians tend to use balsam for Abies, Americans often call Abies
+    "firs", as in the various species of true fir. I would recommend
+    using Abies as the shorthand descriptor in the text and captions for
+    subalpine fir. It is a short word, and we have only 1 Abies species
+    at this site. Also, my experience in the on-line-search literature
+    world and with things like Google Scholar is that Abies is a much
+    better and more diagnostic search word for this subject matter than
+    fir or balsam. "Spruce-Abies" is very unambiguous as a descriptor
+    for this forest type.
+
+-   Rate of change (increase, decrease) in vol, BA, and sph: In both in
+    your stats tables and figure captions, and in the text narrative,
+    ensure that what you are describing is clearly and accurately
+    described as either: (a) the incremental change in values between
+    two measurement periods, or (b) the difference in absolute value in
+    certain parameters (e.g. standing volume) between two measurement
+    periods, or between two treatments. Also, the annualized growth rate
+    of something averaged over a measurement period (e.g. mean PBAI,
+    like BA growth in m2/yr over a period) is different than the rate of
+    change in a growth rate over time (e.g. is a growth rate increasing
+    or decreasing). I got confused a few times on this issue.
 
@@
 
-- Great: Very clear and straightforward. i.e. - log-transformed treatment (initial BA) was a very statistically significant predictor of tree response across dbh classes for both Sx and Bl.
-- Don't know what all the random effects outputs are. I am sure they are meaningful, but I just dont know what they mean.
-- Number of observations of 510: Should ensure it is clear that only trees > 7.5 cm dbh are counted.
-- Would be good to define the dbh class limits somewhere. In Appendix 6 table, we describe the class mid-point.
-- For Appendix 6, I assume that the null hypothesis for the ratio is = 1.
+-   Great: Very clear and straightforward. i.e. - log-transformed
+    treatment (initial BA) was a very statistically significant
+    predictor of tree response across dbh classes for both Sx and Bl.
+-   Don't know what all the random effects outputs are. I am sure they
+    are meaningful, but I just dont know what they mean.
+-   Number of observations of 510: Should ensure it is clear that only
+    trees \> 7.5 cm dbh are counted.
+-   Would be good to define the dbh class limits somewhere. In Appendix
+    6 table, we describe the class mid-point.
+-   For Appendix 6, I assume that the null hypothesis for the ratio is
+    = 1.
 
@@
 
-- I am thinking that mortality should / could be analysed as a frequency response - i.e.
-- expected vs actual distribution of mortality by size class and/or species.
-e.g.
-a Chi-squared or allied type of analysis.
-Null hypothesis is that mortality of stems is randomly distributed across different sizes and species of trees.
-H1 is that it is non-random.
-- I'd suggest that we don't compare total mortality (e.g # of stems) across treatment units, but rather frequency relative to initial number of stems (i.e.
-- at beginning of observation period).
-That way we normalize mortality rates relative to initial stem densities.
-Otherwise, results could be mis-leading - e.g.
-Control treatments could have more mortality in overall sph, simply because they have more initial stems.
-- ACTION ITEM FOR ME:  I need to check my original datasets and data sorting iterations to see how I handled mortality observations for the 1992 and 1994 plot remeasurements.
-- Note for you:  All of the mortality you've analysed and identified so far is mortality that is incremental following 92/94, so it is still solid analysis data, just with that qualification.
-I will clarify the 92/94 mortalilty observations for the Methods section; we did observe dead trees at these times, but these mortality pre-dated our treatments and/or first measurements.
-Anyway, probably not a big deal, but this needs examination and clarification to explain this.
-If necessary, I will quantify observations of dead trees observed on these years, because they are actually > 0.
-
-Testing for density-dependence of mortality responses... some ideas or hypotheses..  Note that f (___) means "a function of"...
-
-  *   MORT RATE of a size class = f (Sum of BA in that plot > than that size class).
-Would assume asymmetric competition where only bigger trees exert competition on smaller trees.
-  *   MORT RATE of a size class = f ( Sum of BA in that plot that is either > or < that size class).
-Assumes symmetric competition where both big and small trees exert competition on their neighbors.
-
-Caveats:   (a) our plots and tree distributions might be too spatially variable to get a reliable test of these hypotheses, and (b) BA would be the BA at the beginning of a measurement period.
-And (c) a tree that died of suppression would probably show reduced growth rates prior to death; but trees that had abiotic damage and mortality - e.g.
-- snow or wind breakage - might not.
-Complicated stuff !!
+-   I am thinking that mortality should / could be analysed as a
+    frequency response - i.e. - expected vs actual distribution of
+    mortality by size class and/or species. e.g. a Chi-squared or allied
+    type of analysis. Null hypothesis is that mortality of stems is
+    randomly distributed across different sizes and species of trees. H1
+    is that it is non-random.
+-   I'd suggest that we don't compare total mortality (e.g \# of stems)
+    across treatment units, but rather frequency relative to initial
+    number of stems (i.e. - at beginning of observation period). That
+    way we normalize mortality rates relative to initial stem densities.
+    Otherwise, results could be mis-leading - e.g. Control treatments
+    could have more mortality in overall sph, simply because they have
+    more initial stems.
+-   ACTION ITEM FOR ME: I need to check my original datasets and data
+    sorting iterations to see how I handled mortality observations for
+    the 1992 and 1994 plot remeasurements.
+-   Note for you: All of the mortality you've analysed and identified so
+    far is mortality that is incremental following 92/94, so it is still
+    solid analysis data, just with that qualification. I will clarify
+    the 92/94 mortalilty observations for the Methods section; we did
+    observe dead trees at these times, but these mortality pre-dated our
+    treatments and/or first measurements. Anyway, probably not a big
+    deal, but this needs examination and clarification to explain this.
+    If necessary, I will quantify observations of dead trees observed on
+    these years, because they are actually \> 0.
+
+Testing for density-dependence of mortality responses... some ideas or
+hypotheses.. Note that f (\_\_\_) means "a function of"...
+
+-   MORT RATE of a size class = f (Sum of BA in that plot \> than that
+    size class). Would assume asymmetric competition where only bigger
+    trees exert competition on smaller trees.
+-   MORT RATE of a size class = f ( Sum of BA in that plot that is
+    either \> or \< that size class). Assumes symmetric competition
+    where both big and small trees exert competition on their neighbors.
+
+Caveats: (a) our plots and tree distributions might be too spatially
+variable to get a reliable test of these hypotheses, and (b) BA would be
+the BA at the beginning of a measurement period. And (c) a tree that
+died of suppression would probably show reduced growth rates prior to
+death; but trees that had abiotic damage and mortality - e.g. - snow or
+wind breakage - might not. Complicated stuff !!
 
@@
 -   Really solid stats and rationale for a single ht:dbh model. Great !
-- What is really interesting and useful about this is that the ht:dbh relationship is seemingly statistically independent of treatment or, that is, initial density,  A very useful simplifying assumption proven for future modellers of these stands.
-We should highlight this in either our results or discussion.
+-   What is really interesting and useful about this is that the ht:dbh
+    relationship is seemingly statistically independent of treatment or,
+    that is, initial density, A very useful simplifying assumption
+    proven for future modellers of these stands. We should highlight
+    this in either our results or discussion.
 
@@
 -   See notes above about use of term "units" vs treatment (or similar).
-- Clarify effects of treatment and/or species on growth rate vs standing vol / BA at each measurement period.
-- A Figure 7b or equivalent could visually present changes in growth rate over time. (e.g. previous graphs we've developed)
-- Under Observations, # 2 - the 2nd sentence needs work to edit and clarify.
+-   Clarify effects of treatment and/or species on growth rate vs
+    standing vol / BA at each measurement period.
+-   A Figure 7b or equivalent could visually present changes in growth
+    rate over time. (e.g. previous graphs we've developed)
+-   Under Observations, \# 2 - the 2nd sentence needs work to edit and
+    clarify.
 
@@
 
-- We should define what we mean by the term inverse J or inverse J-shaped dbh distribution. Just for clarity.
-- Should we quantitatively describe these distributions? Don't have to, just asking the question.
-- Is the merch limit for min merch tree size 18.5 cm dbh, or 17.5 cm dbh? I remember the latter from various cruises and cruise compilations.
+-   We should define what we mean by the term inverse J or inverse
+    J-shaped dbh distribution. Just for clarity.
+-   Should we quantitatively describe these distributions? Don't have
+    to, just asking the question.
+-   Is the merch limit for min merch tree size 18.5 cm dbh, or 17.5 cm
+    dbh? I remember the latter from various cruises and cruise
+    compilations.
 -   There are two Figure 2's, one is for vol and one is for sph.
-- Clarify in figures and text that sph is based on trees 7.5 cm dbh and greater. Increases in sph can be from ingress of trees growing from size classes below this limit.
-- A suggestion:  Have a table comparing total volume to merch volume by treatment, etc?  Would be useful for some readers, and communicate MAI and volume growth rates too.
+-   Clarify in figures and text that sph is based on trees 7.5 cm dbh
+    and greater. Increases in sph can be from ingress of trees growing
+    from size classes below this limit.
+-   A suggestion: Have a table comparing total volume to merch volume by
+    treatment, etc? Would be useful for some readers, and communicate
+    MAI and volume growth rates too.
 
@@
 
-- First sentence:  Clarify what you mean by "organized similarly". Does this refer to frequency distribution ? What constitutes similarity?
+-   First sentence: Clarify what you mean by "organized similarly". Does
+    this refer to frequency distribution ? What constitutes similarity?
 
@@
 
-Last winter, as well as keeping reference copies of the initial “raw” dataset following data entry and related stuff, I kept copies each stage of data cleaning and prep (about 10 iterations).
-
-We did have “dead tree” observations in the first ~ 2 year of monitoring. The key thing to see is : if I can determine is when the trees were dead, and/or when death was initiated, and why.  
-Timing of tree death and observations of causal factors of tree death in these first two years will help inform our interpretation and data handling of these observations.
-
-If trees were dead or moribund at the first “initial” observation or two (as soon after the logging as possible), this is – arguably -  different than mortality that occurs over time in response to the basal area treatment and/or could be considered to be density- or treatment-dependent in some way.
-
-The 1992 data was in mid-March really early after logging, and before snowmelt (in retrospect not a great idea but we had resources available before the end of the fiscal year at the time).
-The 1994 data was from when the plots were re-visited under snow-free conditions.
-We did a thinning and sanitation culling of trees in the treatment units and PSP’s in 1994 to remove logging-damaged trees and most deciduous, to bring the units and PSP’s down to their final treatment BA.
-1994 remeasurements were completed after that.
+Last winter, as well as keeping reference copies of the initial "raw"
+dataset following data entry and related stuff, I kept copies each stage
+of data cleaning and prep (about 10 iterations).
+
+We did have "dead tree" observations in the first \~ 2 year of
+monitoring. The key thing to see is : if I can determine is when the
+trees were dead, and/or when death was initiated, and why.\
+Timing of tree death and observations of causal factors of tree death in
+these first two years will help inform our interpretation and data
+handling of these observations.
+
+If trees were dead or moribund at the first "initial" observation or two
+(as soon after the logging as possible), this is -- arguably - different
+than mortality that occurs over time in response to the basal area
+treatment and/or could be considered to be density- or
+treatment-dependent in some way.
+
+The 1992 data was in mid-March really early after logging, and before
+snowmelt (in retrospect not a great idea but we had resources available
+before the end of the fiscal year at the time). The 1994 data was from
+when the plots were re-visited under snow-free conditions. We did a
+thinning and sanitation culling of trees in the treatment units and
+PSP's in 1994 to remove logging-damaged trees and most deciduous, to
+bring the units and PSP's down to their final treatment BA. 1994
+remeasurements were completed after that.
 
@@
 
-1.	The very-early observed mortality may be a result of pre-existing stand conditions not the created, treatment-induced stand conditions. “stochastic” vs potentially treatment induced. 
-2.	It can be argued that trees dead at time of treatment or very soon  thereafter were never taking up growing space in the post-harvest stand, and therefore would not contribute to the initial “green” basal area going forward.
-3.	We do have 1992 data in most plots, but it was 1994 when the BA in each trmt unit and embedded PSP was finalized. 1994 is the most homogeneous start point for all PSP’s.
-
-Regarding tree mortality in the period from 1992 to 1994 at Summit Lake EP 1162:
-
-I reviewed our data regarding this, any relevant project records, and my own recollection of our monitoring of tree mortality during this period.
+1.  The very-early observed mortality may be a result of pre-existing
+    stand conditions not the created, treatment-induced stand
+    conditions. "stochastic" vs potentially treatment induced.
+2.  It can be argued that trees dead at time of treatment or very soon
+    thereafter were never taking up growing space in the post-harvest
+    stand, and therefore would not contribute to the initial "green"
+    basal area going forward.
+3.  We do have 1992 data in most plots, but it was 1994 when the BA in
+    each trmt unit and embedded PSP was finalized. 1994 is the most
+    homogeneous start point for all PSP's.
+
+Regarding tree mortality in the period from 1992 to 1994 at Summit Lake
+EP 1162:
+
+I reviewed our data regarding this, any relevant project records, and my
+own recollection of our monitoring of tree mortality during this period.
 Here is my train of thought, and what I have concluded on this topic:
@@
 
-Trees that were dead at the time of the 1992 PSP establishment were not tagged or recorded.
-The rationale for this was that these trees were dead and therefore not taking up up growing space in the stands.
-Dead trees in 1992 would have been either snags / dead trees pre-existing the harvest, or trees fatally damaged or killed by logging activities (e.g.
-– trees with stems snapped and crowns broken off, no live limbs) or trees smashed to the ground by felled trees).
+Trees that were dead at the time of the 1992 PSP establishment were not
+tagged or recorded. The rationale for this was that these trees were
+dead and therefore not taking up up growing space in the stands. Dead
+trees in 1992 would have been either snags / dead trees pre-existing the
+harvest, or trees fatally damaged or killed by logging activities (e.g.
+-- trees with stems snapped and crowns broken off, no live limbs) or
+trees smashed to the ground by felled trees).
 
@@
 
-EP 1162 treatment units, including the PSP area contained therein, were assessed in Spring 1994 to review stand density / basal area, and stand condition, in order to:  
-a)	identify and mark-to-cut residual trees that were considered too badly scarred or damaged to constitute good growing stock for trial purposes, 
-b)	where needed, to lower the initial BA closer to the target level, 
-c)	to either fell or girdle birch trees, especially in the PSP’s themselves, but usually also within the TU as a whole, and 
-d)	to fell any dead, dangerous, and/or moribund trees within the treatment units and PSP’s. 
-Generally most if not all of these trees felled in the sanitation felling were understory or subdominant trees, and occasionally a lower codominant tree.
-Tagged trees felled by the sanitation felling were documented on field sheets and our Excel datasets, so we can retrieve the species and size of these trees as needed.
-However, sanitation-felled trees included both live, moribund, and dead trees – generally all poorer quality or vigor though.
-
-After felling, the felled trees were bucked to a 2 to 3 m length so as to lay these felled trees to as close to the ground as possibly, maximize decay, and minimize physical obstructions within the TU’s and PSP’s.
-
-As a result of the sanitation felling, potentially some trees which died or were damaged between 1992 and Summer 1994 were felled without recording these mortality events and precise tree vigor beforehand.
-Some minor mortality occurred between 1992 and 94, but was obscured by the post-harvest sanitation treatment.
-
-The 1994 Remeasurement occurred in Fall 1994, with a few plot remeasurements occurring in Winter 1994/95 (i.e. no intervening growing season).
-
-For scientific rigour, we should not state that there was zero mortality between 1992 and 1994.
-Rather, I suggest, it is more accurate to state that data on tree mortalilty between 1992 and 1994 was not collected, and that the Summer 1994 sanitation felling treatment did not allow accurate documentation of tree mortality in this initial period on a retrospective basis.
-
-We don’t, therefore, have data on mortality between 1992 and 1994, though my recollection is that it was minor in nature.
-
-The difference in basal area between 1992 and 1994 is predominantly a product of the sanitation felling of some basal area, with minor natural mortality.
-
-Mortality occurring between Fall 1994 and Fall 1997 remeasurements was captured accurately by these two sequential remeasurements.
-
-“The difference in basal area between 1992 and 1994 is a result of basal area reductions through the sanitation felling of some stems and some minor natural mortality, combined with basal area increases through the growth of remaining live stems.”
-
-A review of the various plots indicates that between 1992 and 1994, some PSP’s decreased in BA (as sanitation removals and minor mortality exceeded growth of live stems), while others increased in BA as tree growth exceeded BA removals.
-
-  *   Is there any change in which sizes or size classes of trees die as stand volume increases?
-- e.g.
-is competition asymmetrical or symmetrical?
-Currently the y axis is total number of trees, I believe.
-  *   Not sure if we have enough sample size of trees to look at species dynamics (Sx vs Bl).
-  *   I am wondering if the same relationship that appears between volume and tree mortality, is mirrored in the relationship between BA and tree mortality. A bit more on this below...
-
-Foresters (like me) do like to use BA per ha as a density measure for field planning and stand assessment because it seems to be broadly useful and is easily assessed via stand walkthroughs or plots.
-Sapwood BA is pretty well correlated with foliar biomass of the stand - a biological linkage.
-Tree BA is often correlated well with things like crown width.
-
-I hadn't thought alot previously about vol per ha as a density measurement, but it does have the characteristic of integrating not just BA, but also tree height, so it integrates tree growing space occupancy in both in horizontal dimensions (crown width and probably root spread) but also vertical.
-Vol ha theoretically provides more of a 3D measure of how the trees are occupying a given land area.
-The fact that volume is of commercial interest is a bonus.
-Key assumptions here: That trees are healthy and most importantly, crowns are health and growing relatively optimally relative to their growing space opportunities.
-Once crown vigor etc starts to decline with age, pests or pathogens, then the relationship could start to break down.
+EP 1162 treatment units, including the PSP area contained therein, were
+assessed in Spring 1994 to review stand density / basal area, and stand
+condition, in order to:\
+a) identify and mark-to-cut residual trees that were considered too
+badly scarred or damaged to constitute good growing stock for trial
+purposes, b) where needed, to lower the initial BA closer to the target
+level, c) to either fell or girdle birch trees, especially in the PSP's
+themselves, but usually also within the TU as a whole, and d) to fell
+any dead, dangerous, and/or moribund trees within the treatment units
+and PSP's. Generally most if not all of these trees felled in the
+sanitation felling were understory or subdominant trees, and
+occasionally a lower codominant tree. Tagged trees felled by the
+sanitation felling were documented on field sheets and our Excel
+datasets, so we can retrieve the species and size of these trees as
+needed. However, sanitation-felled trees included both live, moribund,
+and dead trees -- generally all poorer quality or vigor though.
+
+After felling, the felled trees were bucked to a 2 to 3 m length so as
+to lay these felled trees to as close to the ground as possibly,
+maximize decay, and minimize physical obstructions within the TU's and
+PSP's.
+
+As a result of the sanitation felling, potentially some trees which died
+or were damaged between 1992 and Summer 1994 were felled without
+recording these mortality events and precise tree vigor beforehand. Some
+minor mortality occurred between 1992 and 94, but was obscured by the
+post-harvest sanitation treatment.
+
+The 1994 Remeasurement occurred in Fall 1994, with a few plot
+remeasurements occurring in Winter 1994/95 (i.e. no intervening growing
+season).
+
+For scientific rigour, we should not state that there was zero mortality
+between 1992 and 1994. Rather, I suggest, it is more accurate to state
+that data on tree mortalilty between 1992 and 1994 was not collected,
+and that the Summer 1994 sanitation felling treatment did not allow
+accurate documentation of tree mortality in this initial period on a
+retrospective basis.
+
+We don't, therefore, have data on mortality between 1992 and 1994,
+though my recollection is that it was minor in nature.
+
+The difference in basal area between 1992 and 1994 is predominantly a
+product of the sanitation felling of some basal area, with minor natural
+mortality.
+
+Mortality occurring between Fall 1994 and Fall 1997 remeasurements was
+captured accurately by these two sequential remeasurements.
+
+"The difference in basal area between 1992 and 1994 is a result of basal
+area reductions through the sanitation felling of some stems and some
+minor natural mortality, combined with basal area increases through the
+growth of remaining live stems."
+
+A review of the various plots indicates that between 1992 and 1994, some
+PSP's decreased in BA (as sanitation removals and minor mortality
+exceeded growth of live stems), while others increased in BA as tree
+growth exceeded BA removals.
+
+-   Is there any change in which sizes or size classes of trees die as
+    stand volume increases? - e.g. is competition asymmetrical or
+    symmetrical? Currently the y axis is total number of trees, I
+    believe.
+-   Not sure if we have enough sample size of trees to look at species
+    dynamics (Sx vs Bl).
+-   I am wondering if the same relationship that appears between volume
+    and tree mortality, is mirrored in the relationship between BA and
+    tree mortality. A bit more on this below...
+
+Foresters (like me) do like to use BA per ha as a density measure for
+field planning and stand assessment because it seems to be broadly
+useful and is easily assessed via stand walkthroughs or plots. Sapwood
+BA is pretty well correlated with foliar biomass of the stand - a
+biological linkage. Tree BA is often correlated well with things like
+crown width.
+
+I hadn't thought alot previously about vol per ha as a density
+measurement, but it does have the characteristic of integrating not just
+BA, but also tree height, so it integrates tree growing space occupancy
+in both in horizontal dimensions (crown width and probably root spread)
+but also vertical. Vol ha theoretically provides more of a 3D measure of
+how the trees are occupying a given land area. The fact that volume is
+of commercial interest is a bonus. Key assumptions here: That trees are
+healthy and most importantly, crowns are health and growing relatively
+optimally relative to their growing space opportunities. Once crown
+vigor etc starts to decline with age, pests or pathogens, then the
+relationship could start to break down.
 
@@
 
-This section will describe the methods we used for data analyses, not field data collection.
+This section will describe the methods we used for data analyses, not
+field data collection.
 
@@
 
-Our data analyses and summaries examined the effects of the different levels of stand-level basal area density treatments on three main attributes of stand development following treatment.
-These include, in order:
-
-1. Post-treatment stand basal area dynamics, including the rate and pattern of basal area re-growth and recovery over time following different levels of initial basal area density reductions;
-2. Stand structural outcomes, including treatment influences on the overall abundance of trees (in sph) by diameter class, within the different treatment types, and;
-3.
-Tree species composition outcomes, including treatment influences on the relative abundance and size class distributions of subalpine fir and hybrid white spruce within the different treatment types.
-
-Analyses are grouped into three main categories (below).  These are treated as separate categories in the Results section.
+Our data analyses and summaries examined the effects of the different
+levels of stand-level basal area density treatments on three main
+attributes of stand development following treatment. These include, in
+order:
+
+1.  Post-treatment stand basal area dynamics, including the rate and
+    pattern of basal area re-growth and recovery over time following
+    different levels of initial basal area density reductions;
+2.  Stand structural outcomes, including treatment influences on the
+    overall abundance of trees (in sph) by diameter class, within the
+    different treatment types, and;
+3.  Tree species composition outcomes, including treatment influences on
+    the relative abundance and size class distributions of subalpine fir
+    and hybrid white spruce within the different treatment types.
+
+Analyses are grouped into three main categories (below). These are
+treated as separate categories in the Results section.
 
@@
 
-To test the hypothesis that tree-level radial growth response over the 1994-2019 period varied with interactions between (i) tree diameter at the time of final treatment (1994); (ii) treatment; and (iii) species, we fit linear mixed effects models with log-transformed 1994-2019 diameter increment as the response variable, and treatment, species and log-transformed 1992 diameter as interacting fixed effects.
-To account for plot-level productivity differences, we included plot as a random intercept effect.
-Models were fit using the 'lme4' package for R [@].
-We also fit a similar model with the 1994-2019 diameter increment expressed as a percentage of 1994 diameter.
-Plots of residuals and fit lines with orginal data were examined to evaluate model goodness of fit.
-Variance explained for each model are reported using the approach described in Nakagawa et al.
-[-@Nakagawa2013], as implemented in the MuMIn package for R [@barton2009mumin].
-Pairwise contrasts were performed using the 'emmeans' package for R [@] and Tukey HSD method of *p* value adjustment was used to control for multiplicity.
-We report all significant results at p<0.05.
+To test the hypothesis that tree-level radial growth response over the
+1994-2019 period varied with interactions between (i) tree diameter at
+the time of final treatment (1994); (ii) treatment; and (iii) species,
+we fit linear mixed effects models with log-transformed 1994-2019
+diameter increment as the response variable, and treatment, species and
+log-transformed 1992 diameter as interacting fixed effects. To account
+for plot-level productivity differences, we included plot as a random
+intercept effect. Models were fit using the 'lme4' package for R [\@].
+We also fit a similar model with the 1994-2019 diameter increment
+expressed as a percentage of 1994 diameter. Plots of residuals and fit
+lines with orginal data were examined to evaluate model goodness of fit.
+Variance explained for each model are reported using the approach
+described in Nakagawa et al. [-@Nakagawa2013], as implemented in the
+MuMIn package for R [@barton2009mumin]. Pairwise contrasts were
+performed using the 'emmeans' package for R [\@] and Tukey HSD method of
+*p* value adjustment was used to control for multiplicity. We report all
+significant results at p\<0.05.
+
+For each plot and measurement period, we computed the growth dominance statistic (GDS) to evaluate the relationship between individual tree size and its growth as a proportion of the stand's overall growth [@Forrester; @West; @Binkey]
 
@@
 #### Stand-level volume and basal area estimates
-Total and merchantable volumes were estimated for trees using species- and region-specific equations [@nigh2016] based on tree height and DBH.
-Consistent with forest management practices in British Columbia, we estimated volumes for trees meeting a minimum diameter utilization limit of 18.5cm DBH.
-For trees lacking a tree height measurement, we estimated height by fitting a single diameter-height model developed with a linear mixed effects model, with log-transformed DBH as a fixed effect, and tree nested within plot as random intercepts.
-Trees with broken tops were omitted from diameter-height models.
-We summed tree-level estimates of volume and basal area to generate stand-level estimates for each plot.
-To test the hypothesis that changes in stand attributes over 27 years varied by treatment, we fit separate linear models with stand total volume, basal area, and density as response variables, and the interaction between treatment and year as the independent variable.
+
+Total and merchantable volumes were estimated for trees using species-
+and region-specific equations [@nigh2016] based on tree height and DBH.
+Consistent with forest management practices in British Columbia, we
+estimated volumes for trees meeting a minimum diameter utilization limit
+of 18.5cm DBH. For trees lacking a tree height measurement, we estimated
+height by fitting a single diameter-height model developed with a linear
+mixed effects model, with log-transformed DBH as a fixed effect, and
+tree nested within plot as random intercepts. Trees with broken tops
+were omitted from diameter-height models. We summed tree-level estimates
+of volume and basal area to generate stand-level estimates for each
+plot. To test the hypothesis that changes in stand attributes over 27
+years varied by treatment, we fit separate linear models with stand
+total volume, basal area, and density as response variables, and the
+interaction between treatment and year as the independent variable.
 
@@
 
-The tree-level basal area increment model indicated that interactions between log-transformed tree size, species and treatment explained `r round(r.squaredGLMM(fitBA)[2]*100,1)`% of the variation in log-transformed 1994-2019 tree-level BAI, and `r round(r.squaredGLMM(fitBApct)[2]*100,1)`% of the variation in basal area increment as a percentage of tree size in 1994 (p<0.0001 for both, `r tableNums("fitBA.table",display="cite")`).
-Estimated marginal means from the model showed that between species and among treatments, basal area increment increased with tree size, meaning that larger trees increased basal area more than smaller trees (`r figNums("treeGrowth.plot",display="cite")`), although smaller trees increased their size proportionally more than large trees (`r figNums("treeGrowthPct.plot",display="cite")`).
-Species contrasts showed that spruce trees increased  basal area significantly more than than fir in the high-removal and control treatment units, whereas basal growth between the two species was close to equal in the low-removal unit (`r appNums("sppContrast.table",display="cite")`).
-Within-species contrasts showed that spruce basal area increment in the high-removal unit exceeded spruce growth in both other units (`r appNums("trtContrast.table",display="cite")`), whereas fir basal area increment in the high- and low-removal units was higher than the control unit, and there were no significant differences between the two harvested units (`r appNums("trtContrast.table",display="cite")`).
+The tree-level basal area increment model indicated that interactions
+between log-transformed tree size, species and treatment explained
+`r round(r.squaredGLMM(fitBA)[2]*100,1)`% of the variation in
+log-transformed 1994-2019 tree-level BAI, and
+`r round(r.squaredGLMM(fitBApct)[2]*100,1)`% of the variation in basal
+area increment as a percentage of tree size in 1994 (p\<0.0001 for both,
+`r tableNums("fitBA.table",display="cite")`). Estimated marginal means
+from the model showed that between species and among treatments, basal
+area increment increased with tree size, meaning that larger trees
+increased basal area more than smaller trees
+(`r figNums("treeGrowth.plot",display="cite")`), although smaller trees
+increased their size proportionally more than large trees
+(`r figNums("treeGrowthPct.plot",display="cite")`). Species contrasts
+showed that spruce trees increased basal area significantly more than
+than fir in the high-removal and control treatment units, whereas basal
+growth between the two species was close to equal in the low-removal
+unit (`r appNums("sppContrast.table",display="cite")`). Within-species
+contrasts showed that spruce basal area increment in the high-removal
+unit exceeded spruce growth in both other units
+(`r appNums("trtContrast.table",display="cite")`), whereas fir basal
+area increment in the high- and low-removal units was higher than the
+control unit, and there were no significant differences between the two
+harvested units (`r appNums("trtContrast.table",display="cite")`).
 
@@
 ### Tree height modeling
-Tree heights had a positive nonlinear relationship with diameter, for both species and across all time periods (`r appNums("heightDiam.plot",display="cite")`).
-We originally developed separate models for each species and time period, however, a comparison of height-diameter models showed that intercepts and slopes did not vary significantly (i.e., p>0.05) between species or time periods (not shown).
-Therefore, we developed a single height-diameter linear mixed-effects model using `r X %>% filter(Species %in% c("Fir","Spruce")) %>% filter(Status=="Live") %>% filter(is.na(DBT)) %>% drop_na(Height) %>% nrow()` tree height-diameter observations in the dataset.
-This model explained `r round(MuMIn::r.squaredGLMM(htDiamFit)[1]*100,1)`% of variation in log-transformed heights, with an intercept of `r round(fixef(htDiamFit)[1],3)` and slope coefficient of  `r round(fixef(htDiamFit)[2],3)` * log-transformed DBH (not shown).
-Residual and predicted vs actual height plots were assessed to ensure goodness of fit (`r appNums("htModelFit.plot",display="cite")`).
-Using this model, we generated height estimates for trees without height measurements, and then estimated tree total volume as per Nigh [-@nigh2016].
+
+Tree heights had a positive nonlinear relationship with diameter, for
+both species and across all time periods
+(`r appNums("heightDiam.plot",display="cite")`). We originally developed
+separate models for each species and time period, however, a comparison
+of height-diameter models showed that intercepts and slopes did not vary
+significantly (i.e., p\>0.05) between species or time periods (not
+shown). Therefore, we developed a single height-diameter linear
+mixed-effects model using
+`r X %>% filter(Species %in% c("Fir","Spruce")) %>% filter(Status=="Live") %>% filter(is.na(DBT)) %>% drop_na(Height) %>% nrow()`
+tree height-diameter observations in the dataset. This model explained
+`r round(MuMIn::r.squaredGLMM(htDiamFit)[1]*100,1)`% of variation in
+log-transformed heights, with an intercept of
+`r round(fixef(htDiamFit)[1],3)` and slope coefficient of
+`r round(fixef(htDiamFit)[2],3)` \* log-transformed DBH (not shown).
+Residual and predicted vs actual height plots were assessed to ensure
+goodness of fit (`r appNums("htModelFit.plot",display="cite")`). Using
+this model, we generated height estimates for trees without height
+measurements, and then estimated tree total volume as per Nigh
+[-@nigh2016].
 
@@
 ### Rate of change from 1994 to 2019
-All treatment units, including the control, increased volume, basal area and stand density over the 1994-2019 period (`r figNums("standPlots",display="cite")` and `r appNums("standPlot.table",display="cite")`).
-Mean annual basal area and volume increment peaked during the 1997-2009 period in the two harvested treatments, and peaked in the 1994-1997 period in the control treatment unit (`r figNums("perInc.plot",display="cite")`).
-From 1994 to 2019, estimated marginal means of basal area increment in the low RBA unit exceeded marginal means in the high RBA and control unit by an estimated
+
+All treatment units, including the control, increased volume, basal area
+and stand density over the 1994-2019 period
+(`r figNums("standPlots",display="cite")` and
+`r appNums("standPlot.table",display="cite")`). Mean annual basal area
+and volume increment peaked during the 1997-2009 period in the two
+harvested treatments, and peaked in the 1994-1997 period in the control
+treatment unit (`r figNums("perInc.plot",display="cite")`). From 1994 to
+2019, estimated marginal means of basal area increment in the low RBA
+unit exceeded marginal means in the high RBA and control unit by an
+estimated
 `r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Basal area") %>%   pull(estimate) %>% round(.,1) %>% .[1]`
@@
 `r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Basal area") %>%   pull(estimate) %>% round(.,1) %>% .[2]`
-m² per hectare, respectively (p<0.01 for both, `r appNums("standFit.contrasts",display="cite")`).
-
-The low RBA unit also significantly increased its stand density by an estimated 
+m² per hectare, respectively (p\<0.01 for both,
+`r appNums("standFit.contrasts",display="cite")`).
+
+The low RBA unit also significantly increased its stand density by an
+estimated
 `r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Stand density") %>%   pull(estimate) %>% round(.,0)  %>%   .[2]`
-stems per hectare more than the control unit (p<0.001).
-The estimated rate of change in tree density did not differ significantly between the two harvested units.
-The rate of change in volume in trees over 18.5cm DBH was barely significant (p=0.049) between the two harvested units, and the difference was not significant between the control and harvested units.
+stems per hectare more than the control unit (p\<0.001). The estimated
+rate of change in tree density did not differ significantly between the
+two harvested units. The rate of change in volume in trees over 18.5cm
+DBH was barely significant (p=0.049) between the two harvested units,
+and the difference was not significant between the control and harvested
+units.
 
 ### Stand attributes by year
-By year, mean plot volume, basal area and stand density did not differ significantly between both harvested units (i.e., p>0.05, `r appNums("standFitM.contrasts",display="cite")` and `r figNums("standPlots",display="cite")`).
-Contrasts between the high-removal and control treatment units showed that volume, basal area and stand density differences were highest and significant (i.e., p<0.05) in the 1992 measurement year, and these differences diminshed over each subsequent measurement year.
-For example, mean plot density in the high-removal treatment unit was approximately 600 trees per hectare less than the control in 1994 (p<0.05); by 2019, that difference had decreased to 219 trees per hectare (p>0.05).
-Contrasts between the low-removal and the control treatment units showed a similar trend with basal area and stand density; the control unit had significanly higher basal area and stand density means in 1992, but the difference between the units diminished over each subsequent measurement, becoming non-significant (i.e., p>0.05) in 2019.
-Plot means of volume did not differ significantly between the treatment units in any measurement year.
+
+By year, mean plot volume, basal area and stand density did not differ
+significantly between both harvested units (i.e., p\>0.05,
+`r appNums("standFitM.contrasts",display="cite")` and
+`r figNums("standPlots",display="cite")`). Contrasts between the
+high-removal and control treatment units showed that volume, basal area
+and stand density differences were highest and significant (i.e.,
+p\<0.05) in the 1994 measurement year, and these differences diminshed
+over each subsequent measurement year. For example, mean plot density in
+the high-removal treatment unit was approximately
+
+`r standFitM.contrasts %>% filter(Contrast=="TU:10 - TU:Control") %>% filter(Variable=="Stand density") %>%  filter(Year==1994) %>% mutate(estimate=abs(round(estimate,0))) %>%  pull(estimate)`
+
+trees per hectare less than the control in 1994 (p\<0.05); by 2019, that
+difference had decreased to
+
+`r standFitM.contrasts %>% filter(Contrast=="TU:10 - TU:Control") %>% filter(Variable=="Stand density") %>%  filter(Year==2019) %>% mutate(estimate=abs(round(estimate,0))) %>%  pull(estimate)`
+
+trees per hectare (p\>0.05). Contrasts between the low-removal and the
+control treatment units showed a similar trend with basal area and stand
+density; the control unit had significantly higher basal area and stand
+density means in 1994, but the difference between the units diminished
+over each subsequent measurement, becoming non-significant (i.e.,
+p\>0.05) in 2019. Plot means of volume did not differ significantly
+between the treatment units in any measurement year.
 
@@
 
-Across all plots and time periods, smaller trees contributed proportionally more to stand basal area growth than larger trees (`r figNums("gds.plot",display="cite")`).
-Over the last two measurement periods, the growth dominance statistic decreased in all plots except one (not shown), indicating that individual tree growth contributions to stand growth became more proportional to tree size over the last 11 growing seasons.
+Across all plots and time periods, smaller trees contributed
+proportionally more to stand basal area growth than larger trees
+(`r figNums("gds.plot",display="cite")`). Over the last two measurement
+periods, the growth dominance statistic decreased in all plots except
+one (not shown), indicating that individual tree growth contributions to
+stand growth became more proportional to tree size over the last 11
+growing seasons.
 
@@
 
-Among treatments and throughout the stand development after treatment, the stands maintained a negative exponential (also referred to as 'reverse-J') diameter class distribution (`r figNums("diamK.plot",display="cite")`).
-Immediately after the partial harvest, the diameter class distribution of volume differed among treatments.
-In the 27 years since partial harvest, volume in all three treatment units was mostly concentrated in the 22.5-27.5cm DBH class, and mostly comprised of fir (`r figNums("diamVol.plot",display="cite")`).
-Over both time periods shown in the figure and all three treatments, most of the spruce volume was concentrated in DBH classes spanning 37.5 to 47.5cm.
-Basal area trends were very similar to volume (not shown).
+Among treatments and throughout the stand development after treatment,
+the stands maintained a negative exponential (also referred to as
+'reverse-J') diameter class distribution
+(`r figNums("diamK.plot",display="cite")`).\
+Immediately after the partial harvest, the diameter class distribution
+of volume differed among treatments. In the 27 years since partial
+harvest, volume in all three treatment units was mostly concentrated in
+the 22.5-27.5cm DBH class, and mostly comprised of fir
+(`r figNums("diamVol.plot",display="cite")`). Over both time periods
+shown in the figure and all three treatments, most of the spruce volume
+was concentrated in DBH classes spanning 37.5 to 47.5cm. Basal area
+trends were very similar to volume (not shown).
 
@@
 
-A comparison of proportional species composition by diameter class, treatment and year showed that fir and spruce volume proportion by diameter class varied similarly between the high RBA and control units (`r figNums("sppCompDiam.plot",display="cite")`).
-In 1992, the largest diameter classes were dominated by spruce, whereas fir was the dominant species in the smallest diameter classes.
-This pattern was also found for basal area and tree density (not shown).
-By 2019, the fir-spruce composition by distribution class was maintained in these two units, with small increases in spruce composition in the smallest diameter class.
-
-In the low RBA, the species composition distribution by diameter class was different.
-Here, both fir and spruce maintained similar proportions across diameter classes smaller than 37.5cm DBH.
-Spruce formed the leading species in diameter classes above 37.5cm DBH, with the exception of the largest class.
+A comparison of proportional species composition by diameter class,
+treatment and year showed that fir and spruce volume proportion by
+diameter class varied similarly between the high RBA and control units
+(`r figNums("sppCompDiam.plot",display="cite")`). In 1992, the largest
+diameter classes were dominated by spruce, whereas fir was the dominant
+species in the smallest diameter classes. This pattern was also found
+for basal area and tree density (not shown). By 2019, the fir-spruce
+composition by distribution class was maintained in these two units,
+with small increases in spruce composition in the smallest diameter
+class.
+
+In the low RBA, the species composition distribution by diameter class
+was different. Here, both fir and spruce maintained similar proportions
+across diameter classes smaller than 37.5cm DBH. Spruce formed the
+leading species in diameter classes above 37.5cm DBH, with the exception
+of the largest class.
 
@@
 
-A total of `r deadList %>% nrow()` trees (`r deadList %>% filter(Species=="Fir") %>% nrow()` fir, `r deadList %>% filter(Species=="Spruce") %>% nrow()` spruce, and `r deadList %>% filter(Species=="Fd") %>% nrow()` Douglas-fir) died in `r length(unique(deadList$Plot))` plots from 1994 to 2019.
-The number of trees that died per plot from 1994 to 2019 increased significantly with plot basal area in 1994 (`r figNums("mort.plot",display="cite")`, p < 0.0001).
-Mortality was generally not attributable to a particular forest health factor, and in the majority ofcases, we could not ascertain a casual agent.
-Similar linear models fitted separately for fir and spruce showed that fir mortality had a similar relationship to plot basal area, whereas spruce mortality did not (not shown).
-The diameter distribution of fir mortality followed a negative exponential distribution, with mortality concentrated in smaller diameter classes (`r figNums("mortSppDiam.plot",display="cite")`).
-Spruce mortality was concentrated in larger diameter classes.
-By treatment unit, the control unit had the highest mortality, whereas the high-removal treatment unit had the lowest mortality (`r figNums("mortDiam.plot",display="cite")`).
-The majority of mortality occured during the most recent measurement period of 2009-2019.
+A total of `r deadList %>% nrow()` trees
+(`r deadList %>% filter(Species=="Fir") %>% nrow()` fir,
+`r deadList %>% filter(Species=="Spruce") %>% nrow()` spruce, and
+`r deadList %>% filter(Species=="Fd") %>% nrow()` Douglas-fir) died in
+`r length(unique(deadList$Plot))` plots from 1994 to 2019. The number of
+trees that died per plot from 1994 to 2019 increased significantly with
+plot basal area in 1994 (`r figNums("mort.plot",display="cite")`, p \<
+0.0001). Mortality was generally not attributable to a particular forest
+health factor, and in the majority ofcases, we could not ascertain a
+casual agent. Similar linear models fitted separately for fir and spruce
+showed that fir mortality had a similar relationship to plot basal area,
+whereas spruce mortality did not (not shown). The diameter distribution
+of fir mortality followed a negative exponential distribution, with
+mortality concentrated in smaller diameter classes
+(`r figNums("mortSppDiam.plot",display="cite")`). Spruce mortality was
+concentrated in larger diameter classes. By treatment unit, the control
+unit had the highest mortality, whereas the high-removal treatment unit
+had the lowest mortality (`r figNums("mortDiam.plot",display="cite")`).
+The majority of mortality occured during the most recent measurement
+period of 2009-2019.
 
@@
 ## Uneven-aged stand management
-This study describes tree- and stand-level responses to partial harvest treatments in an uneven-aged stand.  
-- THere are numerous studies documenting responses of even-aged stands to partial harvest treatments such as thinning.
-In even-aged stands, a common objective of thinning treatments is to re-allocate growing space and resources to the most valulable trees in the stand that have capacity to increase production.
-The objectives of partial harvesting in uneven-aged stands is more often to provide a sustained production of wood over multiple generations of trees.
-In this context, partial harvesting in uneven-aged stands seeks to provide adequate growing space and resources to trees across a range of sizes and densities.
-By quantifying growth responses at the tree- and stand-level, this study addresses questions regarding the manipulation of structure in an uneven-aged stand to achieve ...
+
+This study describes tree- and stand-level responses to partial harvest
+treatments in an uneven-aged stand.\
+- THere are numerous studies documenting responses of even-aged stands
+to partial harvest treatments such as thinning. In even-aged stands, a
+common objective of thinning treatments is to re-allocate growing space
+and resources to the most valulable trees in the stand that have
+capacity to increase production. The objectives of partial harvesting in
+uneven-aged stands is more often to provide a sustained production of
+wood over multiple generations of trees. In this context, partial
+harvesting in uneven-aged stands seeks to provide adequate growing space
+and resources to trees across a range of sizes and densities. By
+quantifying growth responses at the tree- and stand-level, this study
+addresses questions regarding the manipulation of structure in an
+uneven-aged stand to achieve ...
 
@@
 
-At the tree-level, our results suggest that while both hybrid white spruce and subalpine fir respond to thinning with faster radial growth, heavy thinning can elicit a stronger radial growth response in spruce, compared to fir.
-Spruce trees in the low-removal harvest treatment did not respond with a significant growth release.
-THus, our results suggest that a relatively high removal of the stand will result in increased spruce growth, and this release will exceed that of co-occurring fir trees.
-Other studies have shown that residual spruce growth responds favourably to heavy thinning.
-A 10-year analysis of white spruce growth underneath an aspen canopy found that a complete or high overstory removal resulted in immediate and significant spruce radial and height growth responses, compared to control and partial cutting treatments (Smith et al.
-2016).
-In a 35-year old white spruce stand, diameter increment was maximized with the heaviest thinning intensity, however, the treatment thinning from below (Stiell 1970).
-Frank (1973) also showed that 70-year old white spruce basal increment increased with the amount of thinning.
-Competition for light is a key factor that limits white spruce growth, and opening up the stand to allow for increased light penetration may be a primary mechanism that enhances white spruce growth responses [@Lieffers1999].
-High basal area removal may also increase air and soil temperatures, which can preferentially favour spruce growth responses compared to fir [@Lajzerowicz2004].
-The relatively low spruce growth response in the low-removal treatment may be explained in that the gaps created may have been too small to elicit as strong a growth response as observed in the high-removal thinning treatment.
-Light levels may differ little between forest canopies that are closed or contain small gaps [@Bartemucci2002], and strong vertical light gradients may result in relatively little light regime changes for understorey trees.
-
-In comparison, both partial harvest treatments stimulated a growth response in fir, compared to the control.
-Although both species are similar in shade tolerance, firs are considered to be better adapted to low light conditions in the understory, allowing them to persist longer in undisturbed forest  (Antos 2000) and potentially able to respond faster to even modest changes in understory light (Dumais et al 2014, Messier et al.
-1999).
-Although fir growth may respond at lower levels of overstorey removal, the growth response potential of fir may be lower than spruce.
-Comparative studies of spruce and fir growth rates present differing evidence regarding species differences in release potential to thinning.
-In mature Engelmann spruce (*Picea engelmannii*) – subalpine fir forests in western North America, understory spruce growth rates exceeded those of fir, allowing spruce to achieve canopy position faster, including after canopy gaps were created through disturbance (Antos et al.
-2000, Andrus et al.
-2018).
-In mature forest in central BC, spruce growth exceeded that of subalpine fir after age 80 (Eis and Craigdallie 1983).
-Because spruce tends to live longer than subalpine fir, higher growth rates at later ages may allow this species to become larger than fir.
-In contrast, thinning studies in spruce-fir forests in eastern North America report that balsam fir (*Abies balsamifera*) responds to release with faster height and diameter growth than co-occurring red spruce (*Picea rubens*) (Dumais and Prevost, 2007; Dumais et al.
-2014, Messier et al.
-1999).
-Other studies have found no difference in growth response between spruce and fir species after treatment (MCaughey and Smith1982).
-Our study also suggests that at lower levels of basal area removal, spruce and fir growth responses are similar.
-
-Post-treatment growth rates increased with tree size at the time of treatment, for both species.
-This has been found in other studies as well (references) and may be related to size-asymmetric competition whereby larger trees preempt resources, especially light, from smaller individuals [@Bose2017].
-Thinning treatments may exacerbate size-assymetric competition [@Cordonnier2015], and growth increases from treatment will be predominantly concentrated in the larger tree size classes [@Bose2018; @Schwinning1998].
-Smaller trees will likely continue to experience competition and will not benefit as much from the thinning treatment.
-Thinning treatments may not promote growth responses in the largest dominant trees in a stand either, as these trees may not experience significant changes in resource availability or may have achieved maximal growth rates [@Ryan1997, Bose2017].
-Forest managers should account for potential size differential when planning thinning treatments.
-For example, a silviculture treatment that seeks to release a suppressed understorey may have to ensure that sufficient overstorey trees are removed to reduce assymetric competition after treatment.
+At the tree-level, our results suggest that while both hybrid white
+spruce and subalpine fir respond to thinning with faster radial growth,
+heavy thinning can elicit a stronger radial growth response in spruce,
+compared to fir. Spruce trees in the low-removal harvest treatment did
+not respond with a significant growth release. THus, our results suggest
+that a relatively high removal of the stand will result in increased
+spruce growth, and this release will exceed that of co-occurring fir
+trees. Other studies have shown that residual spruce growth responds
+favourably to heavy thinning. A 10-year analysis of white spruce growth
+underneath an aspen canopy found that a complete or high overstory
+removal resulted in immediate and significant spruce radial and height
+growth responses, compared to control and partial cutting treatments
+(Smith et al. 2016). In a 35-year old white spruce stand, diameter
+increment was maximized with the heaviest thinning intensity, however,
+the treatment thinning from below (Stiell 1970). Frank (1973) also
+showed that 70-year old white spruce basal increment increased with the
+amount of thinning. Competition for light is a key factor that limits
+white spruce growth, and opening up the stand to allow for increased
+light penetration may be a primary mechanism that enhances white spruce
+growth responses [@Lieffers1999]. High basal area removal may also
+increase air and soil temperatures, which can preferentially favour
+spruce growth responses compared to fir [@Lajzerowicz2004]. The
+relatively low spruce growth response in the low-removal treatment may
+be explained in that the gaps created may have been too small to elicit
+as strong a growth response as observed in the high-removal thinning
+treatment. Light levels may differ little between forest canopies that
+are closed or contain small gaps [@Bartemucci2002], and strong vertical
+light gradients may result in relatively little light regime changes for
+understorey trees.
+
+In comparison, both partial harvest treatments stimulated a growth
+response in fir, compared to the control. Although both species are
+similar in shade tolerance, firs are considered to be better adapted to
+low light conditions in the understory, allowing them to persist longer
+in undisturbed forest (Antos 2000) and potentially able to respond
+faster to even modest changes in understory light (Dumais et al 2014,
+Messier et al. 1999). Although fir growth may respond at lower levels of
+overstorey removal, the growth response potential of fir may be lower
+than spruce. Comparative studies of spruce and fir growth rates present
+differing evidence regarding species differences in release potential to
+thinning. In mature Engelmann spruce (*Picea engelmannii*) -- subalpine
+fir forests in western North America, understory spruce growth rates
+exceeded those of fir, allowing spruce to achieve canopy position
+faster, including after canopy gaps were created through disturbance
+(Antos et al. 2000, Andrus et al. 2018). In mature forest in central BC,
+spruce growth exceeded that of subalpine fir after age 80 (Eis and
+Craigdallie 1983). Because spruce tends to live longer than subalpine
+fir, higher growth rates at later ages may allow this species to become
+larger than fir. In contrast, thinning studies in spruce-fir forests in
+eastern North America report that balsam fir (*Abies balsamifera*)
+responds to release with faster height and diameter growth than
+co-occurring red spruce (*Picea rubens*) (Dumais and Prevost, 2007;
+Dumais et al. 2014, Messier et al. 1999). Other studies have found no
+difference in growth response between spruce and fir species after
+treatment (MCaughey and Smith1982). Our study also suggests that at
+lower levels of basal area removal, spruce and fir growth responses are
+similar.
+
+Post-treatment growth rates increased with tree size at the time of
+treatment, for both species. This has been found in other studies as
+well (references) and may be related to size-asymmetric competition
+whereby larger trees preempt resources, especially light, from smaller
+individuals [@Bose2017]. Thinning treatments may exacerbate
+size-assymetric competition [@Cordonnier2015], and growth increases from
+treatment will be predominantly concentrated in the larger tree size
+classes [@Bose2018; @Schwinning1998]. Smaller trees will likely continue
+to experience competition and will not benefit as much from the thinning
+treatment. Thinning treatments may not promote growth responses in the
+largest dominant trees in a stand either, as these trees may not
+experience significant changes in resource availability or may have
+achieved maximal growth rates [@Ryan1997, Bose2017]. Forest managers
+should account for potential size differential when planning thinning
+treatments. For example, a silviculture treatment that seeks to release
+a suppressed understorey may have to ensure that sufficient overstorey
+trees are removed to reduce assymetric competition after treatment.
 
@@
 
-Stand density 
-The thinned treatments increased tree density faster than the control, and within three decades, there were no significant density differences between the two harvested treatments and control.
-This likely resulted from two processes; growth and mortality.
-Increasing stand density reflects the growth of smaller trees that achieved the minimum diameter class (i.e., 7.5cm DBH) in this study.
-The faster increases in stand density of the thinned treatments compared to the control likely reflect that smaller trees in the thinned stands responded with faster diameter increment after treatment (as previously discussed).
-We note that our study did not account for new seedlings that may have become established over the period of the study.
-Based on increasing densities in the smaller diameter classes, these stands will likely maintain a negative-exponential diameter distribution, which is considered typical of older, uneven-aged stands in this region.
-
-  Stand densities between all units likely also converged because of different mortality rates.
-Tree mortality increased with stand-level basal area, and the thinned stands had the lowest mortality over the time of the study.
-Most of the mortality was concentrated in fir trees among the smaller size classes.
-Density-dependent tree mortality is common in younger, even-aged stands, reflecting intertree competition for light and other resources, and resulting in increasing uniformity of tree distribution [@Larson2015].
-In the uneven-aged stands of our study, density-dependent mortality likely reflects indicate the effects of assymetric competition and suppression of .
-In the unthinned stands, smaller trees died more frequently as a result of suppression from the larger trees, whereas thinning reduced this competition.
-Density-dependent mortality in uneven-aged, mixed-species forests can also vary with drought and site-level conditions such as aspect and soil moisture regime, which needs to be considered when planning stand thinning treatments [@VanGunst2016].
+Stand density The thinned treatments increased tree density faster than
+the control, and within three decades, there were no significant density
+differences between the two harvested treatments and control. This
+likely resulted from two processes; growth and mortality. Increasing
+stand density reflects the growth of smaller trees that achieved the
+minimum diameter class (i.e., 7.5cm DBH) in this study. The faster
+increases in stand density of the thinned treatments compared to the
+control likely reflect that smaller trees in the thinned stands
+responded with faster diameter increment after treatment (as previously
+discussed). We note that our study did not account for new seedlings
+that may have become established over the period of the study. Based on
+increasing densities in the smaller diameter classes, these stands will
+likely maintain a negative-exponential diameter distribution, which is
+considered typical of older, uneven-aged stands in this region.
+
+Stand densities between all units likely also converged because of
+different mortality rates. Tree mortality increased with stand-level
+basal area, and the thinned stands had the lowest mortality over the
+time of the study. Most of the mortality was concentrated in fir trees
+among the smaller size classes. Density-dependent tree mortality is
+common in younger, even-aged stands, reflecting intertree competition
+for light and other resources, and resulting in increasing uniformity of
+tree distribution [@Larson2015]. In the uneven-aged stands of our study,
+density-dependent mortality likely reflects indicate the effects of
+assymetric competition and suppression of . In the unthinned stands,
+smaller trees died more frequently as a result of suppression from the
+larger trees, whereas thinning reduced this competition.
+Density-dependent mortality in uneven-aged, mixed-species forests can
+also vary with drought and site-level conditions such as aspect and soil
+moisture regime, which needs to be considered when planning stand
+thinning treatments [@VanGunst2016].
 
@@
 
-The low-removal thinning treatment also maintained a species composition by diameter class similar to the unthinned stands.
-In these units, subalpine fir was the leading species by frequency in the smaller diameter classes, and spruce was the leading species in the larger classes.
-This is a stand structure and composition common at low- to mid-elevations in the interior of British Columbia.
-In closed forests with undisturbed forest floors, subalpine fir can have higher seedling establishment success than spruce, leading to higher densities of fir in the understorey.
-Compared to spruce, subalpine fir can be relatively short-lived, and can die after 250 years in the canopy.
-Spruce can live for over 400 years and thus can achieve larger sizes and dominant canopy positions.
-
-The high-removal unit had a somewhat different species composition by diameter class.
-Here, fir was the leading species in all diameter classes.
-This means that the high-removal thinning treatment converted the stand composition to fir-leading in all tree sizes.
-This result may be concern for forest managers, as spruce is often preferred as a crop species over subalpine fir [].
-However, the stands in this treatment had more spruce in the smaller diameter classes than the other treatment units and over the 27-year period of this study, spruce monotonically increased its composition (by volume and basal area) compared to fir.
-Over time, we expect that spruce will continue to increase its proportion relative to fir and eventually achieve a similar species composition - diameter distribution as the other units.
+The low-removal thinning treatment also maintained a species composition
+by diameter class similar to the unthinned stands. In these units,
+subalpine fir was the leading species by frequency in the smaller
+diameter classes, and spruce was the leading species in the larger
+classes. This is a stand structure and composition common at low- to
+mid-elevations in the interior of British Columbia. In closed forests
+with undisturbed forest floors, subalpine fir can have higher seedling
+establishment success than spruce, leading to higher densities of fir in
+the understorey. Compared to spruce, subalpine fir can be relatively
+short-lived, and can die after 250 years in the canopy. Spruce can live
+for over 400 years and thus can achieve larger sizes and dominant canopy
+positions.
+
+The high-removal unit had a somewhat different species composition by
+diameter class. Here, fir was the leading species in all diameter
+classes. This means that the high-removal thinning treatment converted
+the stand composition to fir-leading in all tree sizes. This result may
+be concern for forest managers, as spruce is often preferred as a crop
+species over subalpine fir []. However, the stands in this treatment had
+more spruce in the smaller diameter classes than the other treatment
+units and over the 27-year period of this study, spruce monotonically
+increased its composition (by volume and basal area) compared to fir.
+Over time, we expect that spruce will continue to increase its
+proportion relative to fir and eventually achieve a similar species
+composition - diameter distribution as the other units.
 
@@
 
-, reflecting intertree competiton for light and other resources.
-In older stands with more complex structures, tree mortality tends to be primarily caused by senescence and spatially autocorrelated forest agents.
-Given that the forests in this study are older and have a stand structure typial of an uneven-aged stand,.
-, which tends to result in spatially aggregated tree distributions but by Thinning is considered a silviculture treatment that improves growth efficiency of the stand by redistibuting growing space and capturing mortality that would otherwise reduce net stand growth.
-Our findings support suggest that this occured, and that the thinning treament captured a portion of the stand that would have been lost eventually to natural mortality.
-The thinning treatments may have captured some volume through trees that would have died naturally, potentially increasing the net recoverable volume from these stands.
-[@Valkonen2020] found similar mortality in uneven-aged norway spruce stands.
+, reflecting intertree competiton for light and other resources. In
+older stands with more complex structures, tree mortality tends to be
+primarily caused by senescence and spatially autocorrelated forest
+agents. Given that the forests in this study are older and have a stand
+structure typial of an uneven-aged stand,. , which tends to result in
+spatially aggregated tree distributions but by Thinning is considered a
+silviculture treatment that improves growth efficiency of the stand by
+redistibuting growing space and capturing mortality that would otherwise
+reduce net stand growth. Our findings support suggest that this occured,
+and that the thinning treament captured a portion of the stand that
+would have been lost eventually to natural mortality. The thinning
+treatments may have captured some volume through trees that would have
+died naturally, potentially increasing the net recoverable volume from
+these stands. [@Valkonen2020] found similar mortality in uneven-aged
+norway spruce stands.
 
 ### Stand-level growth responses
-Over the 27-year period of this study, plots with high-removal harvest treatments accrued basal area significantly faster than the control and low-removal harvest treatment, likely due to increases in tree-level growth and stand density discussed previously.
-As a result, differences in basal area between the high-removal harvest unit and the control reduced from ca.
-17m2/hectare in 1992 to ca.9m2/hectare in 2019.
-
-In contrast, basal area increment between the low-removal unit and the control did not differ significantly over the 1992-2019 period, and the difference between these two units remained relatively constant over all measurement periods at ca.
-10m2/hectare.
-The similar rates of basal area increment between the low-removal thinning treatment and the control are difficult to explain.
-Tree-level results showed that fir grew faster in the low-removal unit than the control, but spruce did not.
-As well, by 2019 there were no significant differences in stand density between the two units.
-Thus, it appears that while the fir trees may have increased growth as a response of low-removal, the growth increases were insufficient to cause a significant change in basal area increment.
-Further monitoring in this study will help determine the future trajectory of the low-removal thinning treatment.
-
-While our study found that high-removal partial harvest resulted in faster basal area growth, we did not find evidence that thinning changed volume increment in trees greater than 17.5cm DBH.
-A possible explanation is that tree growth responses to thinning might be primarily concentrated in radial growth, as opposed to height growth; this would be important because the volume formula used in this study is based on both tree diameter and height [@Nigh2016].
-We note that it is often difficult to precisely capture height growth changes in mature trees using conventional field-based methods [@Vepakommaetal2011], which makes capturing height changes problematic in these sorts of studies.
-We also utilized a 17.5cm DBH limit for volume calculations; as such, our volume measures only reflect the contribution of the larger trees in the stand, and do not include volume increment in smaller trees.
-
-The partial harvest treatment appeared to change the timing of peak stand growth.
-In the control unit, Volume and basal area increment peaked in the 1994-1997 measurement period, and declined thereafter, with the strongest decline in annual growth over the last measurement period of 2009-2019.
-Age-related declines in stand productivity may be associated with decreasing water-use efficiency and reduced photosynthesis, with the age of maximal growth rate varying with factors such as species and site conditions [@Ryan1997].
-In both thinned treatments, maximal growth rates peaked in the 1997-2009 period.
-Thus, it appears the thinning treatment may have delayed the period of maximal growth rates by approximately 10 years.
-The timing of thinning treatments can be important.
-Stands that are past their period of age-related maximal growth rate may not have as much capacity to respond to treatments as a younger stand, although this is liekly a more important factor for shade-intolerant species rather than the spruce and fir trees in this study [@Bose2017].
-At the time of this study, the high-removal thinning treatment had maintianed an annual basal area and volume increment that was similar to previous measurement periods; if this high level of increment continues, differences in basal area and volume will continue to diminish compared to the control, and this stand may exceed the low-removal thinning treatment.
+
+Over the 27-year period of this study, plots with high-removal harvest
+treatments accrued basal area significantly faster than the control and
+low-removal harvest treatment, likely due to increases in tree-level
+growth and stand density discussed previously. As a result, differences
+in basal area between the high-removal harvest unit and the control
+reduced from ca. 17m2/hectare in 1992 to ca.9m2/hectare in 2019.
+
+In contrast, basal area increment between the low-removal unit and the
+control did not differ significantly over the 1992-2019 period, and the
+difference between these two units remained relatively constant over all
+measurement periods at ca. 10m2/hectare. The similar rates of basal area
+increment between the low-removal thinning treatment and the control are
+difficult to explain. Tree-level results showed that fir grew faster in
+the low-removal unit than the control, but spruce did not. As well, by
+2019 there were no significant differences in stand density between the
+two units. Thus, it appears that while the fir trees may have increased
+growth as a response of low-removal, the growth increases were
+insufficient to cause a significant change in basal area increment.
+Further monitoring in this study will help determine the future
+trajectory of the low-removal thinning treatment.
+
+While our study found that high-removal partial harvest resulted in
+faster basal area growth, we did not find evidence that thinning changed
+volume increment in trees greater than 17.5cm DBH. A possible
+explanation is that tree growth responses to thinning might be primarily
+concentrated in radial growth, as opposed to height growth; this would
+be important because the volume formula used in this study is based on
+both tree diameter and height [@Nigh2016]. We note that it is often
+difficult to precisely capture height growth changes in mature trees
+using conventional field-based methods [@Vepakommaetal2011], which makes
+capturing height changes problematic in these sorts of studies. We also
+utilized a 17.5cm DBH limit for volume calculations; as such, our volume
+measures only reflect the contribution of the larger trees in the stand,
+and do not include volume increment in smaller trees.
+
+The partial harvest treatment appeared to change the timing of peak
+stand growth. In the control unit, Volume and basal area increment
+peaked in the 1994-1997 measurement period, and declined thereafter,
+with the strongest decline in annual growth over the last measurement
+period of 2009-2019. Age-related declines in stand productivity may be
+associated with decreasing water-use efficiency and reduced
+photosynthesis, with the age of maximal growth rate varying with factors
+such as species and site conditions [@Ryan1997]. In both thinned
+treatments, maximal growth rates peaked in the 1997-2009 period. Thus,
+it appears the thinning treatment may have delayed the period of maximal
+growth rates by approximately 10 years. The timing of thinning
+treatments can be important. Stands that are past their period of
+age-related maximal growth rate may not have as much capacity to respond
+to treatments as a younger stand, although this is liekly a more
+important factor for shade-intolerant species rather than the spruce and
+fir trees in this study [@Bose2017]. At the time of this study, the
+high-removal thinning treatment had maintianed an annual basal area and
+volume increment that was similar to previous measurement periods; if
+this high level of increment continues, differences in basal area and
+volume will continue to diminish compared to the control, and this stand
+may exceed the low-removal thinning treatment.
 
@@
 ## Management considerations
-1.
-Both species responded with increased growth after partial harvest.
-High removal of basal area in these forests stimulated stronger spruce growth.
-This information can help silviculturists manipulate stand structure to preferentially ...
-2.
-Partial harvesting at both intensities delayed peak volume and basal area increment by approximately 10 years.
-Annual increment in the control unit declined faster over the last measurement period compared to previous periods, perhaps suggesting that the stand is becoming mature and losses due to mortality may substantially offset growth increment.
-3. GIven that larger trees responded with faster growth than smaller trees, silviculturists need to consider the implications of this on residual stand structure and response after treatment.
+
+1.  Both species responded with increased growth after partial harvest.
+    High removal of basal area in these forests stimulated stronger
+    spruce growth. This information can help silviculturists manipulate
+    stand structure to preferentially ...
+2.  Partial harvesting at both intensities delayed peak volume and basal
+    area increment by approximately 10 years. Annual increment in the
+    control unit declined faster over the last measurement period
+    compared to previous periods, perhaps suggesting that the stand is
+    becoming mature and losses due to mortality may substantially offset
+    growth increment.
+3.  GIven that larger trees responded with faster growth than smaller
+    trees, silviculturists need to consider the implications of this on
+    residual stand structure and response after treatment.
 
@@
 
-*The dependent variable in the first model (BAI) is tree basal area increment from 1992-2019.
-The dependent variable in the second model (BAI (%)) is basal area increment from 1992-2019 as a percentage of the tree's basal area in 1992.*
+*The dependent variable in the first model (BAI) is tree basal area
+increment from 1992-2019. The dependent variable in the second model
+(BAI (%)) is basal area increment from 1992-2019 as a percentage of the
+tree's basal area in 1992.*
 
@@
 ```
-*Trt, treatment; Sq., squares, Num., numerator; Den., denominator; DF, degrees of freedom*
+
+*Trt, treatment; Sq., squares, Num., numerator; Den., denominator; DF,
+degrees of freedom*
 
@@
 
-*Note that in this figure, we show volume for all diameter classes.
-This is different than the stand-level estimate of volume, where we only estimate volume for trees that would be considered merchantable under conventional harvest in British Columbia (i.e., minimum 18.5cm DBH).
-*
+*Note that in this figure, we show volume for all diameter classes. This
+is different than the stand-level estimate of volume, where we only
+estimate volume for trees that would be considered merchantable under
+conventional harvest in British Columbia (i.e., minimum 18.5cm DBH).*
 
@@
 
-*Note this figure is almost identical for compairsons of basal area, volume or tree density. For brevity, we only show basal area.*
+*Note this figure is almost identical for compairsons of basal area,
+volume or tree density. For brevity, we only show basal area.*
 
@@
 
-*Volume is m3/hectare, BA = basal area (m2/ha), SPH = stems per hectare, QMD = quadratic mean diameter (cm) at 1.3m height.  Values in brackets are the standard error of the mean.*
+*Volume is m3/hectare, BA = basal area (m2/ha), SPH = stems per hectare,
+QMD = quadratic mean diameter (cm) at 1.3m height. Values in brackets
+are the standard error of the mean.*
 
@@
 *Volume is m3/hectare, BA = basal area (m2/ha), SPH = stems per hectare.
-Values in brackets in Period column are number of growing seasons in that period.
-Values in brackets in Volume, BA and SPH columns are the standard error of the mean.*
+Values in brackets in Period column are number of growing seasons in
+that period. Values in brackets in Volume, BA and SPH columns are the
+standard error of the mean.*
 

### 2021-03-16 manuscript_draft.Rmd
include growth dominance coefficient information into discussion

@@
 csl: cjfr.csl
-bibliography: ep1162.bib
+bibliography: library.bib
 editor_options: 
@@
 diameter increment as the response variable, and treatment, species and
-log-transformed 1992 diameter as interacting fixed effects. To account
+log-transformed 1994 diameter as interacting fixed effects. To account
 for plot-level productivity differences, we included plot as a random
@@
 
-# Filter only for trees that have measurements in 1992 and 2019 
+# Filter only for trees that have measurements in 1994 and 2019 
 treeList<- # assign variable
@@
 
-# We will produce diameter class stand attributes for two time periods 1992 and 2019.
+# We will produce diameter class stand attributes for two time periods 1994 and 2019.
 
@@
 
-  # Determine which plots have measurements in 1992 and 2019
+  # Determine which plots have measurements in 1994 and 2019
   X %>% 
@@
 the high-removal treatment unit was approximately
-
 `r standFitM.contrasts %>% filter(Contrast=="TU:10 - TU:Control") %>% filter(Variable=="Stand density") %>%  filter(Year==1994) %>% mutate(estimate=abs(round(estimate,0))) %>%  pull(estimate)`
-
 trees per hectare less than the control in 1994 (p\<0.05); by 2019, that
 difference had decreased to 
-
 `r standFitM.contrasts %>% filter(Contrast=="TU:10 - TU:Control") %>% filter(Variable=="Stand density") %>%  filter(Year==2019) %>% mutate(estimate=abs(round(estimate,0))) %>%  pull(estimate)`
-
 trees per hectare (p\>0.05). Contrasts between the low-removal and the
@@
 
-### Growth dominance statistics
+### Growth dominance coefficients
 
@@
 
-# Filter only for trees that have measurements in 1992 and 2019 
+# Filter only for trees that have measurements in 1994 and 2019 
 gd.treeList<- # assign variable
@@
 
-Across all plots and time periods, smaller trees contributed
+The growth dominance coefficient was negative for all plots and time periods, indicating that smaller trees contributed
 proportionally more to stand basal area growth than larger trees
 (`r figNums("gds.plot",display="cite")`). Over the last two measurement
-periods, the growth dominance statistic decreased in all plots except
+periods, the growth dominance coefficient increased in all plots except
 one (not shown), indicating that individual tree growth contributions to
@@
 diameter class varied similarly between the high RBA and control units
-(`r figNums("sppCompDiam.plot",display="cite")`). In 1992, the largest
+(`r figNums("sppCompDiam.plot",display="cite")`). In 1994, the largest
 diameter classes were dominated by spruce, whereas fir was the dominant
@@
 A total of `r deadList %>% nrow()` trees
-(`r deadList %>% filter(Species=="Fir") %>% nrow()` fir,
-`r deadList %>% filter(Species=="Spruce") %>% nrow()` spruce, and
-`r deadList %>% filter(Species=="Fd") %>% nrow()` Douglas-fir) died in
+(`r deadList %>% filter(Species=="Fir") %>% nrow()` fir and
+`r deadList %>% filter(Species=="Spruce") %>% nrow()` spruce) died in
 `r length(unique(deadList$Plot))` plots from 1994 to 2019. The number of
@@
 plot basal area in 1994 (`r figNums("mort.plot",display="cite")`, p \<
-0.0001). Mortality was generally not attributable to a particular forest
-health factor, and in the majority ofcases, we could not ascertain a
-casual agent. Similar linear models fitted separately for fir and spruce
+0.0001). In the majority of cases, we could not identify a casual agent, and overall, mortality was generally not attributable to a particular forest
+health factor. Linear models fitted separately for fir and spruce
 showed that fir mortality had a similar relationship to plot basal area,
@@
 compared to fir. Spruce trees in the low-removal harvest treatment did
-not respond with a significant growth release. THus, our results suggest
+not respond with a significant growth release. Thus, our results suggest
 that a relatively high removal of the stand will result in increased
@@
 the treatment thinning from below (Stiell 1970). Frank (1973) also
-showed that 70-year old white spruce basal increment increased with the
+showed that basal area increment in a 70-year old white spruce stand increased with the
 amount of thinning. Competition for light is a key factor that limits
@@
 In comparison, both partial harvest treatments stimulated a growth
-response in fir, compared to the control. Although both species are
-similar in shade tolerance, firs are considered to be better adapted to
+response in fir, compared to the control. Firs are considered to be better adapted than spruce to
 low light conditions in the understory, allowing them to persist longer
@@
 than spruce. Comparative studies of spruce and fir growth rates present
-differing evidence regarding species differences in release potential to
-thinning. In mature Engelmann spruce (*Picea engelmannii*) -- subalpine
+differing evidence regarding genus-level differences in release potential to
+thinning. In mature Engelmann spruce (*Picea engelmannii*) - subalpine
 fir forests in western North America, understory spruce growth rates
@@
 faster, including after canopy gaps were created through disturbance
-(Antos et al. 2000, Andrus et al. 2018). In mature forest in central BC,
+(Antos et al. 2000, Andrus et al. 2018). In a mature forest in central BC,
 spruce growth exceeded that of subalpine fir after age 80 (Eis and
@@
 experience significant changes in resource availability or may have
-achieved maximal growth rates [@Ryan1997, Bose2017]. Forest managers
-should account for potential size differential when planning thinning
+achieved maximal growth rates [@Ryan1997, Bose2017].
+
+Although growth responses were concentrated in the larger trees, our analysis of growth dominance indicated that smaller trees contributed proportionally more to stand growth than larger trees.
+Binkley et al.
+(2006) propose that GDC in developing forests increases from 0 to positive during early development, reflecting the growth dominance of larger robust trees as canopy closure is achieved.
+As the forests age, the GDC may switch to negative, reflecting several co-occuring processes.
+Older trees may start to senesce, with growth reductions and increased mortality.
+Younger trees may increase their resource use efficiency, and/or increase their allocation to leaf biomass (Baret et al.
+2017) partially in response to increasing resource availiabilty as older trees senesce.
+Data from old Englemann spruce – subalpine fir forests in Colorado (Binkley et al.
+2006) as well as old-growth boreal mixedwood stands (Baret et al.
+2017) support this proposed model of growth dominance development.
+While we found a similar negative growth dominance in the stands of this study, a comparison between the two recent measurement periods showed that growth dominance actually increased over time (i.e., tree growth became increasingly proportional to tree size).
+This finding is unique from other studies, and suggests that either larger trees are increasing their growth rates relative to smaller trees, or smaller trees are beginning to experience increased competition and may be experiencing lower growth rates.
+Further data from these stands will help evaluate the long-term trend in GDC.
+
+Our results suggest that thinning did not substantially alter the growth dominance of the stand within close to 30 years after treatment.
+Similar results were reported after thinning in uneven- and even-aged hardwood stands in eastern Canada [@Lemire2020].
+While thinning treatments stimulated a growth response in all trees, the increase in growth was more or less proportional to tree size.
+A significant decrease in GDS could be potentially achieved in older stands where the treatment seeks to release suppressed larger trees [@Lemire2020].
+Conversely, thinning even-aged red pine (*Pinus resinosa*)  stands reduced growth dominance compared to unthinned stands, regardless of thinning type, presumably by allocating more growing space to previously suppressed trees [@Bradford2010].
+The use of GDC to evaluate the effects of partial cutting may be insufficient on its own, and should be used in conjunction with tree-level analyses [@Lemire2020].
+Forest managers should account for potential size differential when planning thinning
 treatments. For example, a silviculture treatment that seeks to release
 a suppressed understorey may have to ensure that sufficient overstorey
-trees are removed to reduce assymetric competition after treatment.
+trees are removed to reduce assymetric competition and growth dominance after treatment.
 
@@
 
-Stand density The thinned treatments increased tree density faster than
+The thinned treatments increased tree density faster than
 the control, and within three decades, there were no significant density
@@
 in basal area between the high-removal harvest unit and the control
-reduced from ca. 17m2/hectare in 1992 to ca.9m2/hectare in 2019.
+reduced from ca. 17m2/hectare in 1994 to ca.9m2/hectare in 2019.
 
 In contrast, basal area increment between the low-removal unit and the
-control did not differ significantly over the 1992-2019 period, and the
+control did not differ significantly over the 1994-2019 period, and the
 difference between these two units remained relatively constant over all
@@
 *The dependent variable in the first model (BAI) is tree basal area
-increment from 1992-2019. The dependent variable in the second model
-(BAI (%)) is basal area increment from 1992-2019 as a percentage of the
-tree's basal area in 1992.*
+increment from 1994-2019. The dependent variable in the second model
+(BAI (%)) is basal area increment from 1994-2019 as a percentage of the
+tree's basal area in 1994.*
 
@@
                  position=position_dodge(0.05))+
-  xlab("1992 DBH (cm)")+
-  ylab("Tree BAI from 1992-2019 (m²)")+
+  xlab("1994 DBH (cm)")+
+  ylab("Tree BAI from 1994-2019 (m²)")+
   facet_wrap(~BA.Target)+
@@
                  position=position_dodge(0.05))+
-  xlab("1992 DBH (cm)")+
-  ylab("Tree BAI from 1992-2019 (% increase)")+
+  xlab("1994 DBH (cm)")+
+  ylab("Tree BAI from 1994-2019 (% increase)")+
   facet_wrap(~BA.Target)+
@@
 
-`r figNums("treeGrowthPct.plot",caption="Estimated marginal means of spruce and fir basal area increment (expressed as a percentage increase relative to 1992 DBH) across five inital DBH classes and three treatments.
+`r figNums("treeGrowthPct.plot",caption="Estimated marginal means of spruce and fir basal area increment (expressed as a percentage increase relative to 1994 DBH) across five inital DBH classes and three treatments.
 Whiskers are standard error of the mean.
@@
                    # position=position_dodge(.9))+
-   scale_x_continuous(limits=c(1992,2019),breaks=c(1992,1994,1997,2009,2019),
-                     minor_breaks = c(1992,1994,1997,2009,2019),
+   scale_x_continuous(limits=c(1994,2019),breaks=c(1994,1997,2009,2019),
+                     minor_breaks = c(1994,1997,2009,2019),
                      name="Measurement year")+ # stretch out x-axis
@@
                    position=position_dodge(0))+
- scale_x_continuous(limits=c(1992,2019),breaks=c(1992,1994,1997,2009,2019),
-                     minor_breaks = c(1992,1994,1997,2009,2019),
+ scale_x_continuous(limits=c(1994,2019),breaks=c(1994,1997,2009,2019),
+                     minor_breaks = c(1994,1997,2009,2019),
                      name="Measurement year")+ # stretch out x-axis
@@
 
-`r appNums("trtContrast.table",caption="Contrasts of predicted 1992-2019 tree growth among treatments within each species.  Ratio of predicted growth between treatments is in 'ratio' column.")`
+`r appNums("trtContrast.table",caption="Contrasts of predicted 1994-2019 tree growth among treatments within each species.  Ratio of predicted growth between treatments is in 'ratio' column.")`
 
@@
     # Insert number of growing seasons column
-  add_column(Num.Seasons=c(rep(3,6),rep(11,6)),.after="Period") %>% 
+  add_column(Num.Seasons=c(rep(3,3),rep(11,6)),.after="Period") %>% 
   mutate(Period=paste0(Period," (",Num.Seasons,")")) %>% 
@@
   valign(valign="top",part="body") %>% 
-  hline(i=seq(from=3,to=12,by=3),border=fp_border(color="black")) %>% 
+  hline(i=seq(from=3,to=9,by=3),border=fp_border(color="black")) %>% 
   fix_border_issues() %>% 

### 2021-03-24 manuscript_draft.Rmd
manny work

@@
 csl: cjfr.csl
-bibliography: library.bib
+bibliography: C:\\Users\\hgriesba\\Google Drive\\Papers\\library.bib
 editor_options: 
@@
 library(officer)
-library(sjPlot)
 library(emmeans)
@@
 for plot-level productivity differences, we included plot as a random
-intercept effect. Models were fit using the 'lme4' package for R [\@].
+intercept effect. Models were fit using the 'lme4' package [@Bates2015] for the R statistical environment [@R-base].
 We also fit a similar model with the 1994-2019 diameter increment
@@
 MuMIn package for R [@barton2009mumin]. Pairwise contrasts were
-performed using the 'emmeans' package for R [\@] and Tukey HSD method of
+performed using the 'emmeans' package for R [@Lenth2019] and Tukey Honestly Significant Difference test of
 *p* value adjustment was used to control for multiplicity. We report all
-significant results at p\<0.05.
-
-For each plot and measurement period, we computed the growth dominance statistic (GDS) to evaluate the relationship between individual tree size and its growth as a proportion of the stand's overall growth [@Forrester; @West; @Binkey]
+significant results at p<0.05.
+
+Growth dominance in a stand considers the contribution of individual trees to total stand growth as a function of their size [@Binkley2006].
+The growth dominance statistic [GDS; @Binkley2006; @West2014] represents growth dominance as a single metric, whereby a GDS value of 0 indicates a stand where the growth of individual trees, expressed as a proportion of stand total growth, is equal to their size, expressed as a proportion of stand total size (e.g., basal area).
+A negative (positive) GDS indicates a stand where smaller (larger) trees contribute more to stand growth relative to their proportional size.
+For formulae used to compute GDS, we refer the reader to West [-@West2014].
 
@@
 (`r figNums("gds.plot",display="cite")`). Over the last two measurement
-periods, the growth dominance coefficient increased in all plots except
+periods of 1997-2009 and 2009-2019, the growth dominance coefficient increased in all plots except
 one (not shown), indicating that individual tree growth contributions to
@@
 ## Uneven-aged stand management
-
-This study describes tree- and stand-level responses to partial harvest
-treatments in an uneven-aged stand.\
-- THere are numerous studies documenting responses of even-aged stands
-to partial harvest treatments such as thinning. In even-aged stands, a
-common objective of thinning treatments is to re-allocate growing space
-and resources to the most valulable trees in the stand that have
-capacity to increase production. The objectives of partial harvesting in
-uneven-aged stands is more often to provide a sustained production of
-wood over multiple generations of trees. In this context, partial
-harvesting in uneven-aged stands seeks to provide adequate growing space
-and resources to trees across a range of sizes and densities. By
-quantifying growth responses at the tree- and stand-level, this study
-addresses questions regarding the manipulation of structure in an
-uneven-aged stand to achieve ...
+The successful application of partial harvesting in uneven-aged, multi-species stands can be complicated [@smith1997practice], especially where the objectives include the sustained production of wood over multiple generations of trees.
+In this context, partial harvesting in uneven-aged stands seeks to re-allocate growing space and resources to trees across a range of sizes and densities.
+Selection thinning is a typical thinning treatment in uneven-aged stands, whereby individual or groups of dominant trees are removed to recover their economic value as well as create growing space for understory trees.
+While this treatment
 
@@
 growth responses, compared to control and partial cutting treatments
-(Smith et al. 2016). In a 35-year old white spruce stand, diameter
+[@smith2016sprucing]. In a 35-year old white spruce stand, diameter
 increment was maximized with the heaviest thinning intensity, however,
-the treatment thinning from below (Stiell 1970). Frank (1973) also
+the treatment thinning from below [@Stiell1980]. Frank [-@frank1973course] also
 showed that basal area increment in a 70-year old white spruce stand increased with the
@@
 low light conditions in the understory, allowing them to persist longer
-in undisturbed forest (Antos 2000) and potentially able to respond
-faster to even modest changes in understory light (Dumais et al 2014,
-Messier et al. 1999). Although fir growth may respond at lower levels of
+in undisturbed forest [@Antos2000] and potentially able to respond
+faster to even modest changes in understory light [@Dumais2014; @Messier1999]. Although fir growth may respond at lower levels of
 overstorey removal, the growth response potential of fir may be lower
@@
 faster, including after canopy gaps were created through disturbance
-(Antos et al. 2000, Andrus et al. 2018). In a mature forest in central BC,
-spruce growth exceeded that of subalpine fir after age 80 (Eis and
-Craigdallie 1983). Because spruce tends to live longer than subalpine
+[@Antos2000; @Andrus2018]. In a mature forest in central BC,
+spruce growth exceeded that of subalpine fir after age 80 [@eis1983regeneration]. Because spruce tends to live longer than subalpine
 fir, higher growth rates at later ages may allow this species to become
@@
 responds to release with faster height and diameter growth than
-co-occurring red spruce (*Picea rubens*) (Dumais and Prevost, 2007;
-Dumais et al. 2014, Messier et al. 1999). Other studies have found no
+co-occurring red spruce (*Picea rubens*) [@Dumais2008; @Dumais2008a; @Dumais2014; @Messier1999]. Other studies have found no
 difference in growth response between spruce and fir species after
-treatment (MCaughey and Smith1982). Our study also suggests that at
+treatment [@mccaughey1982understory]. Our study also suggests that at
 lower levels of basal area removal, spruce and fir growth responses are
@@
 whereby larger trees preempt resources, especially light, from smaller
-individuals [@Bose2017]. Thinning treatments may exacerbate
-size-assymetric competition [@Cordonnier2015], and growth increases from
+individuals [@Bose]. Thinning treatments may exacerbate
+size-assymetric competition [@cordonnier2015gini], and growth increases from
 treatment will be predominantly concentrated in the larger tree size
-classes [@Bose2018; @Schwinning1998]. Smaller trees will likely continue
+classes [@Bose; @schwinning1998mechanisms]. Smaller trees will likely continue
 to experience competition and will not benefit as much from the thinning
@@
 experience significant changes in resource availability or may have
-achieved maximal growth rates [@Ryan1997, Bose2017].
+achieved maximal growth rates [@ryan1997age, @Bose].
 
@@
 Binkley et al.
-(2006) propose that GDC in developing forests increases from 0 to positive during early development, reflecting the growth dominance of larger robust trees as canopy closure is achieved.
+[-@Binkley2006] propose that GDC in developing forests increases from 0 to positive during early development, reflecting the growth dominance of larger robust trees as canopy closure is achieved.
 As the forests age, the GDC may switch to negative, reflecting several co-occuring processes.
 Older trees may start to senesce, with growth reductions and increased mortality.
-Younger trees may increase their resource use efficiency, and/or increase their allocation to leaf biomass (Baret et al.
-2017) partially in response to increasing resource availiabilty as older trees senesce.
-Data from old Englemann spruce – subalpine fir forests in Colorado (Binkley et al.
-2006) as well as old-growth boreal mixedwood stands (Baret et al.
-2017) support this proposed model of growth dominance development.
+Younger trees may increase their resource use efficiency, and/or increase their allocation to leaf biomass [@baret2017long] partially in response to increasing resource availiabilty as older trees senesce.
+Data from old Englemann spruce – subalpine fir forests in Colorado [@Binkley2006] as well as old-growth boreal mixedwood stands [@baret2017long] support this proposed model of growth dominance development.
 While we found a similar negative growth dominance in the stands of this study, a comparison between the two recent measurement periods showed that growth dominance actually increased over time (i.e., tree growth became increasingly proportional to tree size).
@@
 be important because the volume formula used in this study is based on
-both tree diameter and height [@Nigh2016]. We note that it is often
+both tree diameter and height [@nigh2016]. We note that it is often
 difficult to precisely capture height growth changes in mature trees
-using conventional field-based methods [@Vepakommaetal2011], which makes
+using conventional field-based methods [@vepakomma2011response], which makes
 capturing height changes problematic in these sorts of studies. We also
@@
 photosynthesis, with the age of maximal growth rate varying with factors
-such as species and site conditions [@Ryan1997]. In both thinned
+such as species and site conditions [@ryan1997age]. In both thinned
 treatments, maximal growth rates peaked in the 1997-2009 period. Thus,
@@
 important factor for shade-intolerant species rather than the spruce and
-fir trees in this study [@Bose2017]. At the time of this study, the
+fir trees in this study [@Bose]. At the time of this study, the
 high-removal thinning treatment had maintianed an annual basal area and

### 2021-03-25 manuscript_draft.Rmd
manny work

@@
 
-Growth dominance in a stand considers the contribution of individual trees to total stand growth as a function of their size [@Binkley2006].
-The growth dominance statistic [GDS; @Binkley2006; @West2014] represents growth dominance as a single metric, whereby a GDS value of 0 indicates a stand where the growth of individual trees, expressed as a proportion of stand total growth, is equal to their size, expressed as a proportion of stand total size (e.g., basal area).
+Growth dominance in a stand considers the contribution of individual trees to total stand growth as a function of their size [@Binkley2006; @Forrester2019; @West2014].
+The growth dominance statistic (GDS) expresses growth dominance as a single metric, where a GDS value of 0 indicates a stand where the growth of individual trees, expressed as a proportion of stand total growth, is equal to their size, expressed as a proportion of stand total size (e.g., basal area).
 A negative (positive) GDS indicates a stand where smaller (larger) trees contribute more to stand growth relative to their proportional size.
-For formulae used to compute GDS, we refer the reader to West [-@West2014].
+For more information on the formula used to compute GDS, we refer the reader to West [-@West2014].
 
@@
 interaction between treatment and year as the independent variable.
-
-#### Mortality
 
@@
 
-All treatment units, including the control, increased volume, basal area
-and stand density over the 1994-2019 period
-(`r figNums("standPlots",display="cite")` and
-`r appNums("standPlot.table",display="cite")`). Mean annual basal area
-and volume increment peaked during the 1997-2009 period in the two
-harvested treatments, and peaked in the 1994-1997 period in the control
-treatment unit (`r figNums("perInc.plot",display="cite")`). From 1994 to
-2019, estimated marginal means of basal area increment in the low RBA
-unit exceeded marginal means in the high RBA and control unit by an
-estimated
-`r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Basal area") %>%   pull(estimate) %>% round(.,1) %>% .[1]`
+All treatment units, including the control, increased volume, basal area and stand density over the 1994-2019 period (`r figNums("standPlots",display="cite")`).
+Over this period, the low RBA treatment unit accrued basal area and density significantly faster (p<0.05) than the high RBA unit and control, and the high RBA unit accrued stand density faster than the control (p<0.05).
+The rate of volume increment did not differ significantly between treatments and control.
+Contrasts of estimated marginal means showed that basal area increment in the low RBA unit exceeded the high RBA and control unit by an estimated `r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Basal area") %>%   pull(estimate) %>% round(.,1) %>% .[1]`
 and
@@
 
-The low RBA unit also significantly increased its stand density by an
-estimated
-`r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Stand density") %>%   pull(estimate) %>% round(.,0)  %>%   .[2]`
-stems per hectare more than the control unit (p\<0.001). The estimated
-rate of change in tree density did not differ significantly between the
-two harvested units. The rate of change in volume in trees over 18.5cm
-DBH was barely significant (p=0.049) between the two harvested units,
-and the difference was not significant between the control and harvested
-units.
-
-### Stand attributes by year
-
-By year, mean plot volume, basal area and stand density did not differ
-significantly between both harvested units (i.e., p\>0.05,
-`r appNums("standFitM.contrasts",display="cite")` and
-`r figNums("standPlots",display="cite")`). Contrasts between the
-high-removal and control treatment units showed that volume, basal area
-and stand density differences were highest and significant (i.e.,
-p\<0.05) in the 1994 measurement year, and these differences diminshed
-over each subsequent measurement year. For example, mean plot density in
-the high-removal treatment unit was approximately
-`r standFitM.contrasts %>% filter(Contrast=="TU:10 - TU:Control") %>% filter(Variable=="Stand density") %>%  filter(Year==1994) %>% mutate(estimate=abs(round(estimate,0))) %>%  pull(estimate)`
-trees per hectare less than the control in 1994 (p\<0.05); by 2019, that
-difference had decreased to 
-`r standFitM.contrasts %>% filter(Contrast=="TU:10 - TU:Control") %>% filter(Variable=="Stand density") %>%  filter(Year==2019) %>% mutate(estimate=abs(round(estimate,0))) %>%  pull(estimate)`
-trees per hectare (p\>0.05). Contrasts between the low-removal and the
-control treatment units showed a similar trend with basal area and stand
-density; the control unit had significantly higher basal area and stand
-density means in 1994, but the difference between the units diminished
-over each subsequent measurement, becoming non-significant (i.e.,
-p\>0.05) in 2019. Plot means of volume did not differ significantly
-between the treatment units in any measurement year.
+In 1994, the low RBA unit had 
+`r standFitM.contrasts %>%   filter(Contrast=="TU:10 - TU:20") %>%  filter(Variable=="Basal area") %>%  filter(Year==1994) %>%  pull(estimate) %>%   abs() %>%   round(.,1)`
+and
+ `r standFitM.contrasts %>%   filter(Contrast=="TU:10 - TU:Control") %>% filter(Variable=="Basal area") %>%   filter(Year==1994) %>%   pull(estimate) %>%   abs() %>%   round(.,1)`
+ less basal area per hectare than the high RBA unit and control, respectively, and by 2019, those differences had decreased to 
+`r standFitM.contrasts %>%   filter(Contrast=="TU:10 - TU:20") %>%  filter(Variable=="Basal area") %>%  filter(Year==2019) %>%  pull(estimate) %>%   abs() %>%   round(.,1)`
+and
+`r standFitM.contrasts %>%   filter(Contrast=="TU:10 - TU:Control") %>% filter(Variable=="Basal area") %>%   filter(Year==2019) %>%   pull(estimate) %>%   abs() %>%   round(.,1)`, respectively.
+In terms of stand density, the low RBA unit accrued `r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Stand density") %>%   pull(estimate) %>% round(.,0) %>% .[1]`
+and
+`r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Stand density") %>%   pull(estimate) %>% round(.,0) %>% .[2]` more trees per hectare from 1994-2019, than the high RBA and control, respectively (p<0.05for both), and the high RBA accrued  `r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Stand density") %>%   pull(estimate) %>% round(.,0) %>% .[3]` more trees than the control (p<0.05) over the same time period.
+By 2019, there were no significant stand density differences between the treatments and control.
+
+The rate of change varied by measurement period.  In the low RBA unit, basal area and volume increment peaked at 
+`r perInc.X %>% filter(BA.Target=="TU:10") %>% filter(Period=="1997-2009") %>% pull("BA") %>%   round(.,2)` m2/ha and
+ `r perInc.X %>% filter(BA.Target=="TU:10") %>% filter(Period=="1997-2009") %>% pull("Vol") %>%   round(.,1)` m3/ha, respectively, in the 1997-2009 period; increments declined to `r perInc.X %>% filter(BA.Target=="TU:10") %>% filter(Period=="2009-2019") %>% pull("BA") %>%   round(.,2)` m2/ha and
+ `r perInc.X %>% filter(BA.Target=="TU:10") %>% filter(Period=="2009-2019") %>% pull("Vol") %>%   round(.,1)` m3/ha respectively in the subsequent measurement period of 2009-2019.
+In the high RBA unit, periodic annual basal area and volume increment increased with each measurement period and peaked in the 2009-2019 measurement period at `r perInc.X %>% filter(BA.Target=="TU:20") %>% filter(Period=="2009-2019") %>% pull("BA") %>%   round(.,2)` m2/ha and
+ `r perInc.X %>% filter(BA.Target=="TU:20") %>% filter(Period=="2009-2019") %>% pull("Vol") %>%   round(.,1)` m3/ha respectively.
+Conversely, basal area and volume increment in the control unit was highest in the 1994-1997 period, and declined with each subsequent measurement period to `r perInc.X %>% filter(BA.Target=="TU:Control") %>% filter(Period=="2009-2019") %>% pull("BA") %>%   round(.,2)` m2/ha and
+ `r perInc.X %>% filter(BA.Target=="TU:Control") %>% filter(Period=="2009-2019") %>% pull("Vol") %>%   round(.,1)` m3/ha respectively..
 
@@
          Num.dead=replace_na(Num.dead,0)) %>% 
-  mutate(prop.K=Num.dead/SPH,
-         prop.BA=BA.dead/BA)
+  mutate(prop.K=Num.dead/SPH*100,
+         prop.BA=BA.dead/BA*100)
 
@@
 plot basal area in 1994 (`r figNums("mort.plot",display="cite")`, p \<
-0.0001). In the majority of cases, we could not identify a casual agent, and overall, mortality was generally not attributable to a particular forest
+0.0001). In the majority of cases, we could not identify a causal agent, and overall, mortality was generally not attributable to a particular forest
 health factor. Linear models fitted separately for fir and spruce
@@
 The successful application of partial harvesting in uneven-aged, multi-species stands can be complicated [@smith1997practice], especially where the objectives include the sustained production of wood over multiple generations of trees.
-In this context, partial harvesting in uneven-aged stands seeks to re-allocate growing space and resources to trees across a range of sizes and densities.
-Selection thinning is a typical thinning treatment in uneven-aged stands, whereby individual or groups of dominant trees are removed to recover their economic value as well as create growing space for understory trees.
-While this treatment
+In this context, partial harvesting seeks to re-allocate growing space and resources to trees across a range of species and size classes to ensure their survival and productivity.
+Knowledge gleaned from thinning treatments in even-aged stands can help inform silviculture practices in more complex stands, however, complexities associated with competition between size classes and species requires additional consideration.
 
@@
 treatment. Light levels may differ little between forest canopies that
-are closed or contain small gaps [@Bartemucci2002], and strong vertical
+are closed or contain small gaps, and strong vertical
 light gradients may result in relatively little light regime changes for
-understorey trees.
+understorey trees [@Bartemucci2002].
 
@@
 difference in growth response between spruce and fir species after
-treatment [@mccaughey1982understory]. Our study also suggests that at
-lower levels of basal area removal, spruce and fir growth responses are
+treatment [@mccaughey1982understory]. Our study also suggests that spruce and fir growth responses to lower levels of basal area removal are
 similar.
@@
 to experience competition and will not benefit as much from the thinning
-treatment. Thinning treatments may not promote growth responses in the
+treatment. However, thinning treatments may not promote growth responses in the
 largest dominant trees in a stand either, as these trees may not
 experience significant changes in resource availability or may have
-achieved maximal growth rates [@ryan1997age, @Bose].
+already achieved maximal growth rates [@ryan1997age, @Bose].
 
@@
 Older trees may start to senesce, with growth reductions and increased mortality.
-Younger trees may increase their resource use efficiency, and/or increase their allocation to leaf biomass [@baret2017long] partially in response to increasing resource availiabilty as older trees senesce.
-Data from old Englemann spruce – subalpine fir forests in Colorado [@Binkley2006] as well as old-growth boreal mixedwood stands [@baret2017long] support this proposed model of growth dominance development.
+Concurrently, younger trees may increase their resource use efficiency, and/or increase their allocation to leaf biomass [@baret2017long], partially in response to increasing resource availability resulting from the senescence of older trees.
+Data from old-growth Engelmann spruce – subalpine fir forests in Colorado [@Binkley2006] as well as old-growth boreal mixedwood stands [@baret2017long] support this proposed model of growth dominance development.
 While we found a similar negative growth dominance in the stands of this study, a comparison between the two recent measurement periods showed that growth dominance actually increased over time (i.e., tree growth became increasingly proportional to tree size).
@@
 
-Our results suggest that thinning did not substantially alter the growth dominance of the stand within close to 30 years after treatment.
+Our results suggest that thinning did not substantially alter the growth dominance of the stand at any point 27 years after treatment.
 Similar results were reported after thinning in uneven- and even-aged hardwood stands in eastern Canada [@Lemire2020].
@@
 be concern for forest managers, as spruce is often preferred as a crop
-species over subalpine fir []. However, the stands in this treatment had
+species in British Columbia and elsewhere [@Deal]. However, the stands in this treatment had
 more spruce in the smaller diameter classes than the other treatment
@@
 ### Tree mortality
-
-, reflecting intertree competiton for light and other resources. In
-older stands with more complex structures, tree mortality tends to be
-primarily caused by senescence and spatially autocorrelated forest
-agents. Given that the forests in this study are older and have a stand
-structure typial of an uneven-aged stand,. , which tends to result in
-spatially aggregated tree distributions but by Thinning is considered a
-silviculture treatment that improves growth efficiency of the stand by
-redistibuting growing space and capturing mortality that would otherwise
-reduce net stand growth. Our findings support suggest that this occured,
-and that the thinning treament captured a portion of the stand that
-would have been lost eventually to natural mortality. The thinning
-treatments may have captured some volume through trees that would have
-died naturally, potentially increasing the net recoverable volume from
-these stands. [@Valkonen2020] found similar mortality in uneven-aged
-norway spruce stands.
+Our results suggest that mortality increased with stand density.
+The majority of trees that died were fir, and these were predominantly concentrated in the smaller size classes.
+These findings concur with other studies that found that risk or tree mortality increases with stand density, competition and tree-level factors such as live crown and growth [@Valkonen2020; @VanGunst2016].
+Competition from overstory trees is likely a contributing cause to most of the mortality recorded in our study.
+Because we did not record trees below 7.5cm DBH, we cannot quantify total tree mortality in the stand.
+Given the relatively low mortality in the thinning treatments, it does appear that partial harvesting may effectively capture some volume that would otherwise be lost to mortality, potentially affecting the total net volume recovered from a stand.
+Density-dependent mortality may vary with other factors such as drought [@VanGunst2016], and climate change may alter competitive interactions between species and size classes.
+We also note that as of 2019, there was a large spruce bark beetle (*Dendroctonus rufipennis*) outbreak in the study region; presumably, the larger spruce trees in this study may be removed by this forest health agent in the short-term.
+This may provide an opportunity for further study into stand responses after mortality of larger stems.
 
@@
 The partial harvest treatment appeared to change the timing of peak
-stand growth. In the control unit, Volume and basal area increment
+stand growth. In the control unit, volume and basal area increment
 peaked in the 1994-1997 measurement period, and declined thereafter,
@@
 age-related maximal growth rate may not have as much capacity to respond
-to treatments as a younger stand, although this is liekly a more
+to treatments as a younger stand, although this is likely a more
 important factor for shade-intolerant species rather than the spruce and
 fir trees in this study [@Bose]. At the time of this study, the
-high-removal thinning treatment had maintianed an annual basal area and
+high-removal thinning treatment had maintained an annual basal area and
 volume increment that was similar to previous measurement periods; if
@@
 
-`r appNums("standFit.contrasts",caption="Contrasts of 1992-2019 stand structure development over 1992-2019 between treatments.
-Contrasts are shown for three variables: volume, basal area and stand density.")`
+`r appNums("standFit.contrasts",caption="Contrasts of 1992-2019 stand structure development between treatments.  Contrasts are shown for three variables: volume, basal area and stand density.")`
 

### 2021-03-25 manuscript_draft.Rmd
Work on manny

@@
 
+We calculated growth dominance statistic (GDS) for each plot and time period.
 Growth dominance in a stand considers the contribution of individual trees to total stand growth as a function of their size [@Binkley2006; @Forrester2019; @West2014].
-The growth dominance statistic (GDS) expresses growth dominance as a single metric, where a GDS value of 0 indicates a stand where the growth of individual trees, expressed as a proportion of stand total growth, is equal to their size, expressed as a proportion of stand total size (e.g., basal area).
+The GDS expresses growth dominance as a single metric, where a value of 0 indicates a stand where the growth of individual trees, expressed as a proportion of stand total growth (in this case, basal area increment), is equal to their size, expressed as a proportion of stand total size (in this case, basal area).
 A negative (positive) GDS indicates a stand where smaller (larger) trees contribute more to stand growth relative to their proportional size.
@@
  `r standFitM.contrasts %>%   filter(Contrast=="TU:10 - TU:Control") %>% filter(Variable=="Basal area") %>%   filter(Year==1994) %>%   pull(estimate) %>%   abs() %>%   round(.,1)`
- less basal area per hectare than the high RBA unit and control, respectively, and by 2019, those differences had decreased to 
+ less basal area (m² per hectare) than the high RBA unit and control, respectively, and by 2019, those differences had decreased to 
 `r standFitM.contrasts %>%   filter(Contrast=="TU:10 - TU:20") %>%  filter(Variable=="Basal area") %>%  filter(Year==2019) %>%  pull(estimate) %>%   abs() %>%   round(.,1)`
 and
-`r standFitM.contrasts %>%   filter(Contrast=="TU:10 - TU:Control") %>% filter(Variable=="Basal area") %>%   filter(Year==2019) %>%   pull(estimate) %>%   abs() %>%   round(.,1)`, respectively.
+`r standFitM.contrasts %>%   filter(Contrast=="TU:10 - TU:Control") %>% filter(Variable=="Basal area") %>%   filter(Year==2019) %>%   pull(estimate) %>%   abs() %>%   round(.,1)` m² per hectare, respectively.
 In terms of stand density, the low RBA unit accrued `r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Stand density") %>%   pull(estimate) %>% round(.,0) %>% .[1]`
 and
-`r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Stand density") %>%   pull(estimate) %>% round(.,0) %>% .[2]` more trees per hectare from 1994-2019, than the high RBA and control, respectively (p<0.05for both), and the high RBA accrued  `r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Stand density") %>%   pull(estimate) %>% round(.,0) %>% .[3]` more trees than the control (p<0.05) over the same time period.
+`r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Stand density") %>%   pull(estimate) %>% round(.,0) %>% .[2]` more trees per hectare from 1994-2019, than the high RBA and control, respectively (p<0.05 for both), and the high RBA accrued  `r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Stand density") %>%   pull(estimate) %>% round(.,0) %>% .[3]` more trees than the control (p<0.05) over the same time period.
 By 2019, there were no significant stand density differences between the treatments and control.
@@
 The rate of change varied by measurement period.  In the low RBA unit, basal area and volume increment peaked at 
-`r perInc.X %>% filter(BA.Target=="TU:10") %>% filter(Period=="1997-2009") %>% pull("BA") %>%   round(.,2)` m2/ha and
- `r perInc.X %>% filter(BA.Target=="TU:10") %>% filter(Period=="1997-2009") %>% pull("Vol") %>%   round(.,1)` m3/ha, respectively, in the 1997-2009 period; increments declined to `r perInc.X %>% filter(BA.Target=="TU:10") %>% filter(Period=="2009-2019") %>% pull("BA") %>%   round(.,2)` m2/ha and
- `r perInc.X %>% filter(BA.Target=="TU:10") %>% filter(Period=="2009-2019") %>% pull("Vol") %>%   round(.,1)` m3/ha respectively in the subsequent measurement period of 2009-2019.
-In the high RBA unit, periodic annual basal area and volume increment increased with each measurement period and peaked in the 2009-2019 measurement period at `r perInc.X %>% filter(BA.Target=="TU:20") %>% filter(Period=="2009-2019") %>% pull("BA") %>%   round(.,2)` m2/ha and
- `r perInc.X %>% filter(BA.Target=="TU:20") %>% filter(Period=="2009-2019") %>% pull("Vol") %>%   round(.,1)` m3/ha respectively.
-Conversely, basal area and volume increment in the control unit was highest in the 1994-1997 period, and declined with each subsequent measurement period to `r perInc.X %>% filter(BA.Target=="TU:Control") %>% filter(Period=="2009-2019") %>% pull("BA") %>%   round(.,2)` m2/ha and
- `r perInc.X %>% filter(BA.Target=="TU:Control") %>% filter(Period=="2009-2019") %>% pull("Vol") %>%   round(.,1)` m3/ha respectively..
+`r perInc.X %>% filter(BA.Target=="TU:10") %>% filter(Period=="1997-2009") %>% pull("BA") %>%   round(.,2)` m²/ha and
+ `r perInc.X %>% filter(BA.Target=="TU:10") %>% filter(Period=="1997-2009") %>% pull("Vol") %>%   round(.,1)` m³/ha, respectively, in the 1997-2009 period; increments declined to `r perInc.X %>% filter(BA.Target=="TU:10") %>% filter(Period=="2009-2019") %>% pull("BA") %>%   round(.,2)` m²/ha and
+ `r perInc.X %>% filter(BA.Target=="TU:10") %>% filter(Period=="2009-2019") %>% pull("Vol") %>%   round(.,1)` m³/ha respectively in the subsequent measurement period of 2009-2019.
+In the high RBA unit, periodic annual basal area and volume increment increased with each measurement period and peaked in the 2009-2019 measurement period at `r perInc.X %>% filter(BA.Target=="TU:20") %>% filter(Period=="2009-2019") %>% pull("BA") %>%   round(.,2)` m²/ha and
+ `r perInc.X %>% filter(BA.Target=="TU:20") %>% filter(Period=="2009-2019") %>% pull("Vol") %>%   round(.,1)` m³/ha respectively.
+Conversely, basal area and volume increment in the control unit was highest in the 1994-1997 period, and declined with each subsequent measurement period to `r perInc.X %>% filter(BA.Target=="TU:Control") %>% filter(Period=="2009-2019") %>% pull("BA") %>%   round(.,2)` m²/ha and
+ `r perInc.X %>% filter(BA.Target=="TU:Control") %>% filter(Period=="2009-2019") %>% pull("Vol") %>%   round(.,1)` m³/ha respectively..
 

### 2021-05-06 manuscript_draft.Rmd
updtae negExp

@@
 # To-do list
+
+1. COmplete Q factor coding
 
@@
 
+#### Growth dominance coefficients
+*Need to describe GDC methods here*
+
+#### Diameter distributions
+We quantified the stand structure in each plot and time period by computing Q factor, which is the ratio between adjacent diameter classes.  Diameter classes were 5cm wide.
+
 # Results
@@
 'reverse-J') diameter class distribution
-(`r figNums("diamK.plot",display="cite")`).\
+(`r figNums("diamK.plot",display="cite")`).
 Immediately after the partial harvest, the diameter class distribution
@@
 trends were very similar to volume (not shown).
+
+Q factor
+```{r compute Q factor}
+
+# Fit negative exponential distribution to plots/time combinations, and then combine
+source(here::here("manuscript","negExp.R"))
+
+# compute for each plot and time period
+x.qFact<-
+  X %>% 
+  mutate(DBH.class=as.numeric(as.character(DBH.class))) %>% 
+  group_by(BA.Target,Plot,Year,DBH.class) %>% 
+  summarise(k=n()) %>% 
+  complete(BA.Target,Plot,Year,DBH.class,fill=list(k=0)) %>% 
+  group_by(BA.Target,Year,DBH.class) %>% 
+  summarise(k=mean(k)) %>% 
+  mutate(k=k*20) %>% 
+  filter(k>0)
+
+# function to apply negExp to each year
+nexExp.year<-function(year){
+
+  x<-
+    x.qFact %>% 
+    filter(BA.Target=="TU:10") %>% 
+    filter(Year==year)
+
+    negExp(x$DBH.class,y=x$DBH.class)
+
+  }
+
+}
+
+```
 

### 2021-05-06 data-raw/growthDominance_TRIAL.R

@@ -1,0 +1,16 @@
+# Growth dominance statistic trial
+
+gsd<-
+  readr::read_csv("data-raw/growthDominance_trialData.csv", 
+                                           col_names = FALSE) %>% 
+  as.numeric() %>% 
+  matrix(nrow=15,byrow = T) %>% 
+  as.data.frame()
+
+
+names(gsd)=c("S1","D1","SD1","S2","D2","SD2")                
+                
+                
+gsd %>% 
+  mutate(SD1a=(S1-lag(S1,default=0))*(D1+lag(D1,default=0))/10000) %>% 
+

### 2021-05-06 manuscript/negExp.R

@@ -1,0 +1,41 @@
+# Script to calculate negative exponential diameter distribution of tree stems
+
+negExp<-function(xVar,yVar) {
+  
+  require(broom)
+  
+   # Estimate the rest parameters using a linear model
+  model.0 <- lm(log(yVar) ~ xVar)  
+  alpha.0 <- exp(coef(model.0)[1])
+  beta.0 <- coef(model.0)[2]
+  
+  # Starting parameters
+  start <- list(alpha = alpha.0, beta = beta.0)
+  
+  # fit model
+  model <- nls(yVar ~ alpha * exp(beta * xVar), start = start)
+  
+  # calculate Q factor
+  q<-
+    tidy(model) %>% 
+    slice(2) %>% 
+    pull(estimate) %>% 
+    abs() %>% 
+    exp()
+  
+  # return coef
+  tidy(model) %>% 
+    mutate(q.factor=c(NA,q)) %>% 
+    return()
+    
+}
+
+
+
+
+
+
+
+
+
+
